---
title: "00-manuscript"
format: docx
bibliography: resources/Relief-bibliography.bib
csl: resources/acta-physiologica.csl
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

```
 

## Introduction
<!-- Would it be natural to mention the functional paper findings here, as a "we observed this, hence we want to explore the molecular data"?-->
Skeletal muscle mass and strength decline with age, best prevented or treated with regular resistance training (RT). The decline in muscle mass and -strength advances upon reaching approximately 50 years of age and accelerates further after reaching 70, and, if left unchecked, has a high probability of leading to sarcopenia [@fragala_resistance_2019; @timmons_molecular_2016]. Muscular weakness and low muscle mass *per se* is further linked to diabetes, disability, cognitive decline, osteoporosis and early all-cause mortality, prevented/treated by either attaining or maintaining a healthy skeletal muscle mass [@fragala_resistance_2019; @li_associations_2018; @scheerman_sarcopenia_2021]. Older adults are expected to gain similar benefits from RT as the young, i.e., increased muscle mass, -strength and overall physical performance [@chen_is_2023; @marzuca-nassr_muscle_2024]. To ensure maximal gains from a period of RT, the anabolic stimulus provided from exercising must be sufficient [@chen_is_2023; @marzuca-nassr_muscle_2024; @paulussen_anabolic_2021], of which training volume is a principal determinant [@schoenfeld_dose-response_2017].  

Whether the anabolic stimulus is sufficient may in part depend on how responsive aging skeletal muscle is to it. Earlier findings suggest that older adults have a less responsive skeletal muscle mass compared to young, with a lower muscle protein synthesis (MPS) response to a session of resistance exercise training (RT) or protein ingestion [@atherton_muscle_2012; @timmons_molecular_2016; @wall_aging_2015]. However, baseline levels of MPS do not differ by age, and the activity in MPS-influencing proteins (e.g. mTOR) is elevated in fasted muscle from older adults compared to the young [@hodson_molecular_2019; @timmons_molecular_2016; @markofski_effect_2015]. The increased baseline mTOR activity could represent a compensation to increased relative muscle loading due to age-related loss of muscle mass, or dysregulation of the mTOR complex [@timmons_molecular_2016]. Further, likely as a consequence of altered mTOR regulation, ribosomal gene transcription and ribosome biogenesis is reduced in older adults [@chaillou_does_2023; @hodson_molecular_2019; @timmons_molecular_2016]. The increase in ribosomal content following onset of RT is believed to provide support for skeletal muscle fibre hypertrophy, through increased translational capacity yielding a higher baseline MPS [@figueiredo_regulation_2019; @figueiredo_revisiting_2019]. Indeed, regardless of age, individuals who respond to RT exhibit an accumulation in markers of ribosome biogenesis (e.g., total RNA and/or rRNA), compared to those who do not experience benefits of RT [@hammarstrom_benefits_2020; @stec_ribosome_2016; @stec_effects_2015]. 

Given the dose-response relationship between training volume and gains in muscle mass and strength observed in younger adults [@schoenfeld_dose-response_2017], and recent findings that older "non-responders" can achieve improved hypertrophy with higher training volume [@lixandrao_higher_2024], increasing training volume may represent a strategy to enhance the anabolic stimulus and thereby support the ribosome biogenesis response necessary for adaptation. Therefore, by comparing low and moderate training-volume protocols in healthy young and older adults, we aim to; determine if there are differences in response to RT between young and older adults, and if these differences are reflected in markers of ribosome biogenesis.


## Methods

```{r}
#| label: fig-study-design
#| fig-cap: "Study overview and weekly training volume. (A) Participant recruitment and randomization flow, and testing modalities used throughout the study: dual-energy x-ray absorptiometry / ultrasound (DXA/US), muscle biopsy, oral glucose tolerance test (OGGT), magnetic resonance imaging (MRI), strength testing (STR), and resistance training (RT). LV = low volume, MV = moderate volume, CON = non-training control. (B) Weekly training volume load (kg × repetitions) for the leg press exercise across the intervention, shown separately by age group (young, old) and training volume condition (low, moderate; solid vs. dashed lines respectively). Shaded ribbons represent the 95% confidence interval around the weekly mean. Vertical dashed lines and accompanying icons mark key study timepoints: baseline testing (week 1) and the start of training (week 2), a muscle biopsy at week 4, mid-intervention testing (week 6.5), and post-intervention testing (week 12); icons above each timepoint indicate which assessments were performed."
#| fig-width: 6.85
#| fig-height: 6.3

readRDS("figures/fig-study-design.RDS")

```
<!--
Question: Should I use the weekly volume for all exercises, or is it sufficient to use leg press only?
-->

### Study Overview

This study was conducted as an age-stratified within-participant trial employing a unilateral design. All participants performed a whole-body unilateral training program, with training protocols (low and moderate) randomly assigned to either leg/arm. The intervention lasted for 12 weeks, with 10 weeks of exercising and two weeks of testing (Figure 1). The participants exercised for a total of 24 training sessions, two sessions per week for the first five weeks, increased to three sessions per week for the five last weeks, including a taper down to two sessions for the 10th and final week. The effects of the intervention on skeletal muscle mass was measured as thigh muscle volume (MRI: baseline, post), thigh muscle thickness (ultrasound: baseline, half-way, post) and lean body mass (DXA: baseline, half-way, post). Skeletal muscle strength was assessed as peak torque with the humac dynamometer, 1RM leg-press with the Keiser Airpress and maximal voluntary bicep curl contraction with the MuscleLab system (all baseline, half-way, post). Glucose tolerance was measured by the oral glucose tolerance test at baseline and post-intervention. Molecular data was gathered via micro biopsy at baseline, after three weeks of RT (six training session) and post-intervention. All tests were performed at the same time of day (± 2hrs), and measurement of lean body mass, muscle thickness, glucose tolerance, and sampling of muscle tissue was conducted in an overnight fasted state. Participants were asked to maintain their habitual physical activity level throughout the intervention, and to not perform any additional type of heavy exercise.

All participants gave their written informed consent before study enrollment. The study was approved by the Regional Committee for Medical and Health Research Ethics - South-East Norway (ID nr. 255224), pre-registered at clinicaltrials.gov (Identifier: NCT05063279), and conducted according to the Helsinki Declaration.


### Participants

Thirty-two young (24 ± 3.1 years) and 49 old (75.5 ± 5.6 years) adults were recruited locally; 10 of the old participants were randomly allocated to a control group. Inclusion criteria were age 20–30 or above 70 years (Figure 1). Exclusion criteria were: insufficient Norwegian proficiency; RT more than once weekly and/or endurance training three or more times weekly within the last six months; unstable cardiovascular disease; illness or injury contraindicating RT; serious mental illness; or allergy to local anaesthesia.


### Resistance training protocol

Resistance training was performed as a whole-body training program. Each RT-session contained three lower-body exercises (unilateral leg press, unilateral knee extension and -flexion) and three upper-body exercises (unilateral chest-press, unilateral bent-over rows, and unilateral bicep curls), using individualized settings in all exercise apparatuses. The low training volume condition was performed as a single set of 10RM per exercise, and the moderate training volume condition was performed as three sets of 10RM per exercise. As a general warm-up, the participants cycled on an indoor exercicse bicycle for 5-10 minutes. Before lower-body exercises the participants performed 1 set of 10 repetitions in bilateral leg press, and before upper-body exercises they performed a warm-up set of 10 push-ups, modified to their individual fitness level (e.g., standard, from the knees, or against a wall). To ensure progressive loading and an adequate stimulation of the skeletal muscle, the exercise load was increased in the following set if the participants lifted more than 12 repetitions. Resting time between each set was two minutes. For safety and standardization purposes, all sessions were monitored by trained personnel. To ensure sufficient recovery, at least one day separated training days throughout the intervention.


### Skeletal muscle sampling and analyses

Muscle biopsies were collected from m. vastus lateralis. The biopsies were sampled under antiseptic and anaesthetised conditions (Lidokain 10 mg ml-1, Mylan Hospital AS, Oslo Norway), using the micro-biopsy technique with a 12-14 gauge needle (Universal Plus, Mermaid medical A/S, Stenløse, Denmark) operated with a spring loaded biopsy gun (Bard Magnum, Bard Norway A/S, Oslo, Norway). Samples were divided into four aliquots for RNA, protein, immunohistochemistry and DNA analyses. At week 3 and 12 measurements, there were 48 hours between the previous training session and biopsy sampling. 


### Protein extraction and western blotting

The wet muscle tissue was freeze-dried for 24h and dissected, then homogenised with a plastic rod in an ice cold lysis buffer. Briefly, 150ul of RIPA buffer was added per mg muscle tissue dry weight, with 1ul protease/phosphatase inhibitors per 100ul lysis buffer. The samples were kept on ice at all times except for when manually disrupting to homogenise the tissue, and then incubated for 30 min on a rotator while on ice. Following the incubation, the samples were spun in the centrifuge for 40 min, at 15 000g at 4°C. Total protein was then diluted to 2ug/ul stock and aliquoted to run the samples in triplicates for protein concentration determination and for western blotting. Protein concentration was determined in a 1:10 dilution (Pierce Detergent Compatible Bradford Assay Reagent, Thermo Fisher Scientific, Oslo, Norway).

Western blots were done using the Jess Automated Western Blot System (Bio-Techne, Minneapolis, USA), according to the manufacturer's protocol, optimized for our lab. The stock solution (2ug/ul) was prepared with fluorescent master mix from the EZ standard pack (Bio-Techne) and 0.1X sample buffer (Bio-Techne) and thus diluted to 0.75 ug/ul for the assay. The assay samples were then briefly vortexed, denatured at 95°C, vortexed and spun down prior to loading the assay plate. To control for within-plate variation, we used a total protein normalization reagent (Bio-Techne) for all assays. A specific between-plate normalization was not used, however, all samples from one participant were analyzed on the same plate. Primary antibodies were purchased from Santa Cruz Biotechnology (Texas, USA): UBF 1:50, UBF F-9, sc-13125; and Cell Signaling Technology (Massachusetts, USA): c-Myc 1:15, c-Myc D84C12. Ready-to-use secondary antibodies (anti-mouse and anti-rabbit) were purchased from Bio-Techne. After loading, the assay plate was centrifuged at 2500 rpm for 5 min at room temperature. The assay was then run on the Jess system, with the chemiluminescence detection module.

### Total RNA extraction and RNA sequencing

Total RNA was extracted using TRIzol from muscle tissue homogenised using 0.5mm RNase-free Zirconium beads (\~50 ul; Next Advanced, Averill Park, NY, USA) and mechanical agitation (Bullet Blender, Next Advanced, Averill Park, NY, USA). Chloroform (Sigma-Aldrich, Oslo, Norway) was used for phase separation, and the RNA pellet was precipitated with isopropanol (VWR International, Oslo, Norway). For assessment of RNA content and purity, RNA was eluted in TE buffer (1:2) and assessed via spectrophotometry. Samples with a 260/280 ratio below 1.8 were excluded. The RNA stock was stored at -80°C until further analyses.


```{r}
#| code-fold: true

library(dplyr)
library(english)
library(reliefdata)
library(patchwork)

# Participants / samples -- Metadata
metadata <- reliefdata::relief_seqsamples |> 
  mutate(participant = as.character(participant)) |> 
  dplyr::rename(sample_weight = weight) |> 
  inner_join(relief_participants ) |> 
  inner_join(relief_volume |> 
               dplyr::select(-arm) |> 
              mutate(participant = as.character(participant)) ) |> 
  dplyr::select(participant, time, age, leg,sex, condition, allocation, sample_weight, seq_sample_id) |> 
  mutate(age = if_else(age < 40, "young", "old")) |> 
  arrange(seq_sample_id)

metadata <- metadata |>
  mutate(time = if_else(
    participant == "3003" & time == "t2" & leg == "L" & condition == "mod",
    "t1",
    time
  )) # t1 L sample wrongfully labelled as t2 for this participant, this fixes it short-term, but it should be looked into closer. 



# Read gene-level data 
dat <- readRDS("data-raw-rnaseq/relief_gene.RDS"); dat <- dat$expected_count


# Number of samples to RNA-seq
nsamp <- stringr::str_to_sentence(as.english(nrow(metadata)))
nsamp_included <- (ncol(dat) - 1)



```

`r nsamp` total RNA samples were selected for RNA-sequencing analysis of which `r nsamp_included` samples passed initial quality assessment (RIN > 8) and were used in downstream analyses. The samples were selected from the pool of intervention participants that had both baseline and post samples collected. Purified mRNA (poly-T oligo-attached magnetic beads) was used for library preparation and subsequent 150-bp paired-end sequencing (Illumina NovaSeq, Novogene Europe, Cambridge, United Kingdom). Raw reads were filtered (SOFTWARE, settings) and aligned to the Human genome (Release) followed by gene and transcript level quantification using RSEM (version).

Gene counts (filtered for low-abundance targets) were modeled using a regularized negative binomial generalized linear mixed model [@glmmmtb] including model parameters for sex, time (baseline and post-intervention), age-group (young and old) and condition (low- and moderate-volume) and the interaction between time age and condition. A varying intercept (random effect) was used for participant, and the dispersion parameter was estimated using the `nbinom2` parameterization available in `glmmTMB`[@glmmmtb]. The effective library size [@robinson2010;@edger] was included as an offset to account for sequencing depth. Regularization through an empirical Bayes shrinkage approach[@efron2010] was achieved using a two-step approach. First, naive models were fitted for all target counts (gene/transcript) using our in-house pipeline [@seqwrap] and estimates for all model parameters (except the intercept) were used to create priors for subsequent informed models. Fixed effect priors were constructed using a normal distribution with mean and standard deviations from each parameter value from the first model pass, the prior for the varying intercepts were constructed using a gamma distribution with the shape parameter informed by the mean from the naive model estimates and the  scale parameter fixed to 2. Finally, the dispersion parameter priors were constructed by fitting a trend model (Local Polynomial Regression Fit) to the association between raw mean log counts and log dispersion across all targets. The predicted dispersion at target-specific raw mean log count was used as the mean and the model residual standard error was used as the standard deviation to define a normally distributed target-specific prior.


### Statistical analyses
<!-- Rework to match molecular data only -->

All statistical analyses were performed in R (version 4.5.1; R Core Team, 2025), and except for RNA sequencing, all outcomes were analysed with Bayesian multilevel models in brms [version 2.23.0;@brms], and contrasts were derived from posterior draws with marginaleffects [version 0.32.0;@marginaleffects]. Muscle volume, muscle thickness, upper-body lean mass, 1RM leg press, bicep curl MVC, UBF and c-Myc were log-transformed and standardized. For these outcomes, each combination of allocation, age group, volume condition and sex was given its own baseline and change over time. Varying effects accounted for participants (intercept and change over time) and for limbs within participants, with additional terms for repeated tests where these were performed. Peak torque was modelled as a quadratic function of angular velocity. Total RNA was modelled at the level of technical replicates, with tissue weight as a covariate. OGTT glucose was modelled as a B-spline function of sampling time. Weakly informative priors were used throughout and checked by prior predictive simulation. Convergence was assessed by R-hat (< 1.01) and effective sample size (> 400). Group changes were averaged across sexes with equal weight. They were back-transformed and reported as percent change, with between-group contrasts reported as ratios expressed in percent. Estimates are given as posterior means with 95% equal-tailed credible intervals (CI) and the posterior probability of an effect in the direction of the estimate (P(>0) or P(<0))<!-- Evidence for all other effects was described from the posterior probability in the direction of the estimate: P(>0), the probability that the effect is positive, for positive estimates, and P(<0), the probability that it is negative, for negative estimates.-->. Effects were termed credible when the 95% CI excluded zero. Otherwise they were described as likely (P ≥ 90%), weak evidence (P ≥ 75%) or no evidence (P < 75%). RNA-seq counts were modelled with negative binomial mixed models (glmmTMB [@glmmmtb]) regularized by empirical Bayes priors [@seqwrap; @efron2010]. Values are given to one decimal place, or two where rounding would give 0.0%. P-values were adjusted with the Benjamini–Hochberg procedure within each contrast (FDR < 5%), followed by rank-based gene-set enrichment analysis [@xu2024a]. Full model specifications are given in the Supplementary Material. 


## Results
<!--
Missing: 
- RNA seq analysis is under revision, results text/fig may need updating after. 
- Should we extend the volume load analysis? This is currently only a summary of the total load, and how much more mod lifted than low.
-->
### Exercise Volume Load

Moderate volume exercise was estimated to ~2.8× higher than low in both age groups (young mod: 137438.7 ± 54491.7 kg, young low: 50693.5 ± 25707.7 kg, old mod: 118124.8 ± 38224.6 kg, old low: 42134.1 ± 14352.5 kg). 

### Protein

```{r}
#| label: fig-protein-rna
#| fig-cap: "UBF and c-Myc (high peak, ~60 kDa) protein content, and total RNA yield, before and 3 weeks into (and after, where applicable) the training intervention. Panels A, C, and E: predicted values at each measured timepoint, corrected area (arbitrary units) for UBF and c-Myc, and total RNA (ng, adjusted for tissue weight). Error bars represent the 95% credible interval around the posterior mean (point and line). Panels B, D, and F: posterior distributions for five key between-group and between-condition contrasts — Old: Intervention vs. Control, Young vs. Old (intervention), Young: Mod vs. Low, Old: Mod vs. Low, and Volume effect: Young vs. Old, expressed as the ratio of change between the two groups or conditions being compared. For UBF and c-Myc, only the pre-to-week 3 change is shown, reflecting the timepoints measured. For total RNA, both the pre-to-week 3 change (grey) and the pre-to-post change (blue) are shown, per the legend. Points show the posterior mean ratio with 95% credible intervals; the dashed vertical line indicates no difference between groups (ratio = 1). Percentages indicate the posterior probability that the ratio exceeds 1 — i.e., the probability that the first-named group or condition in each comparison showed the larger increase."
#| fig-width: 13
#| fig-height: 13

readRDS(here("figures/fig-protein-rna.rds"))

```

__UBF Protein__
Older adults in the intervention group (averaged across volume conditions) tended to increase UBF protein content by week 3 (19.1%, 95% CI [−7.0%, 49.0%], P(>0) = 90.1%), as did the control group (24.9%, 95% CI [−15.7%, 64.3%], P(>0) = 90.8%), with no evidence of a difference between them (−2.6%, 95% CI [−33.7%, 38.1%], P(<0) = 58.6%). There was weak evidence of an increase in young adults (averaged across volume conditions; 15.7%, 95% CI [−8.0%, 44.2%], P(>0) = 88.5%), with no evidence of a difference from older adults (−1.5%, 95% CI [−29.0%, 33.8%], P(<0) = 56.8%). Within young adults, there was weak evidence of an increase with both low-volume (15.2%, 95% CI [−18.7%, 52.5%], P(>0) = 80.8%) and moderate-volume training (17.6%, 95% CI [−15.2%, 60.7%], P(>0) = 83.2%), with no evidence of a difference between them (4.4%, 95% CI [−34.0%, 54.8%], P(>0) = 53.9%). In older adults, there was weak evidence of an increase with low-volume training (27.1%, 95% CI [−13.3%, 74.7%], P(>0) = 88.9%) and no evidence with moderate-volume training (13.1%, 95% CI [−18.8%, 55.0%], P(>0) = 74.9%), again with no evidence of a difference between them (−8.2%, 95% CI [−44.1%, 41.9%], P(<0) = 67.8%). There was no evidence that the volume effect differed between age groups (20.1%, 95% CI [−39.8%, 112.3%], P(>0) = 67.2%).

__c-Myc Protein__
The difference in c-Myc (high peak) protein content between the old intervention group (averaged across volume conditions) and the control group was large by week 3 (99.8%, 95% CI [17.5%, 216.3%], P(>0) = 99.3%). This difference was driven mainly by a decrease in the control group (−34.5%, 95% CI [−56.0%, −5.5%], P(<0) = 98.6%), while the intervention group tended to increase (25.9%, 95% CI [−8.9%, 68.1%], P(>0) = 92.0%). There was weak evidence of an increase in young adults (averaged across volume conditions; 18.6%, 95% CI [−12.5%, 57.5%], P(>0) = 85.9%), with no evidence of a difference from older adults (−3.5%, 95% CI [−37.8%, 45.0%], P(<0) = 61.5%). Within young adults, there was no evidence of a change with low-volume training (4.4%, 95% CI [−30.1%, 49.1%], P(>0) = 55.2%), while moderate-volume training tended to increase c-Myc (37.3%, 95% CI [−12.2%, 106.2%], P(>0) = 91.5%). There was weak evidence of a larger increase under moderate volume (36.2%, 95% CI [−23.8%, 127.4%], P(>0) = 83.5%). The same pattern held in older adults: there was no evidence of a change with low-volume training (9.6%, 95% CI [−29.9%, 64.4%], P(>0) = 62.0%), moderate-volume training tended to increase c-Myc (47.9%, 95% CI [−4.7%, 118.4%], P(>0) = 95.5%), and there was weak evidence of a larger increase under moderate volume (41.0%, 95% CI [−24.3%, 138.0%], P(>0) = 84.5%). There was no evidence that the volume effect differed between age groups (5.2%, 95% CI [−55.8%, 114.3%], P(>0) = 46.7%).

### Total RNA
<!-- total amount of RNA, accounting for tissue mass, but not normalizing to RNA/mg -->

Older adults in the intervention group (averaged across volume conditions) increased RNA yield by 14.1% from baseline to post (95% CI [5.3%, 23.5%], P(>0) = 99.9%). The control group tended to increase by a similar amount (13.5%, 95% CI [−4.1%, 34.2%], P(>0) = 93.2%), and there was no evidence of a difference between them (0.5%, 95% CI [−16.7%, 20.3%], P(>0) = 52.4%). The two groups differed in timing: the intervention group had already increased by week 3 (9.2%, 95% CI [1.2%, 18.2%], P(>0) = 98.6%), whereas the control group had not changed at that point (−0.9%, 95% CI [−15.1%, 15.2%], P(<0) = 55.0%). The difference between them at week 3 was still uncertain (10.6%, 95% CI [−7.0%, 30.8%], P(>0) = 87.6%). Young adults (averaged across volume conditions) increased by 11.6% by post (95% CI [2.6%, 21.4%], P(>0) = 99.6%), with no evidence of a difference from older adults (−2.4%, 95% CI [−12.6%, 9.5%], P(<0) = 65.3%). At week 3, however, young adults had increased by 17.4% (95% CI [7.8%, 28.0%], P(>0) > 99.9%), with weak evidence of a larger increase than older adults (7.4%, 95% CI [−4.5%, 20.8%], P(>0) = 88.9%). Between week 3 and post, young adults' RNA yield likely fell back relative to older adults' (−9.0%, 95% CI [−20.4%, 4.5%], P(<0) = 91.9%). Within young adults, both low-volume (15.7%, 95% CI [3.5%, 29.5%], P(>0) = 99.5%) and moderate-volume training (19.2%, 95% CI [6.4%, 33.5%], P(>0) = 99.9%) increased RNA yield by week 3. At post, the increase was 12.2% under moderate volume (95% CI [0.4%, 25.9%], P(>0) = 97.8%), while low volume tended to increase it by 10.8% (95% CI [−0.7%, 24.1%], P(>0) = 96.6%). There was no evidence of a difference between volume conditions at either time point (week 3: 2.9%, 95% CI [−10.8%, 19.9%], P(>0) = 65.7%; post: 1.1%, 95% CI [−12.6%, 17.1%], P(>0) = 56.0%). In older adults, low-volume (12.3%, 95% CI [1.0%, 24.6%], P(>0) = 98.3%) and moderate-volume training (15.8%, 95% CI [4.4%, 28.3%], P(>0) = 99.8%) both increased RNA yield by post, having tended to increase already by week 3 (low: 10.3%, 95% CI [−0.3%, 22.0%], P(>0) = 97.1%; moderate: 8.3%, 95% CI [−2.3%, 19.9%], P(>0) = 93.3%). There was no evidence of a difference between volume conditions at either time point (week 3: −1.9%, 95% CI [−14.0%, 12.3%], P(<0) = 60.4%; post: 3.1%, 95% CI [−9.9%, 17.8%], P(>0) = 67.2%). There was no evidence that the volume effect differed between age groups, either at week 3 (5.0%, 95% CI [−14.6%, 27.5%], P(>0) = 68.4%) or at post (−1.7%, 95% CI [−20.0%, 20.0%], P(<0) = 57.2%).

### RNA-sequencing

```{r}
#| label: fig-rnaseq-combined
#| fig-cap: "Baseline and training-related transcriptional differences between age groups. (A) Baseline age differences (Young − Old): volcano plot of gene-level estimates (log₂ fold-change) against FDR (log₂ scale), with the dashed line marking FDR = 5%. Genes with the strongest evidence of differential expression are labeled. (B) Rank-based gene-set enrichment analysis (GSEA) of the baseline age contrast, showing the running enrichment score (top) and per-gene contrast effect size (z-score, bottom) across gene rank position, for three representative gene sets (oxidative phosphorylation, heterochromatin formation, complement activation, classical pathway). (C) Age × training-response difference (Young − Old): volcano plot of gene-level estimates for the interaction contrast, in the same format as (A). (D) GSEA of the age × training-response interaction contrast, in the same format as (B), for three representative gene sets (RNA splicing, angiogenesis, aerobic respiration)."
#| fig-width: 10
#| fig-height: 10


library(seqwrap)
m2r <- readRDS("data-out/m2-rnaseq.RDS")
m2s <- seqwrap_summarise(m2r, verbose = FALSE)

n_diff_expressed <- m2s$summaries |> 
  dplyr::filter(term == "ageyoung") |> 
  dplyr::mutate(p.adj = p.adjust(p.value, method = "BH")) |> 
  dplyr::filter(p.adj < 0.05) |> 
  nrow()



n_sig_young_t3 <- m2s$summaries |> 
  dplyr::filter(term == "young_t3") |> 
  dplyr::mutate(p.adj = p.adjust(p.value, method = "BH")) |> 
  dplyr::filter(p.adj < 0.05) |> 
  nrow()

n_sig_old_t3 <- m2s$summaries |> 
  dplyr::filter(term == "old_t3") |> 
  dplyr::mutate(p.adj = p.adjust(p.value, method = "BH")) |> 
  dplyr::filter(p.adj < 0.05) |> 
  nrow()

n_sig_old_volume_t3 <- m2s$summaries |> 
  dplyr::filter(term == "old_volume_t3") |> 
  dplyr::mutate(p.adj = p.adjust(p.value, method = "BH")) |> 
  dplyr::filter(p.adj < 0.05) |> 
  nrow()


readRDS(here("figures/fig-rnaseq.RDS"))

```

_Baseline_
Comparing gene-expression profiles between age groups at baseline (Young − Old) revealed marked differences, with `r n_diff_expressed` genes showing robust differences between the groups (FDR < 5%, Fig 6A/B). Rank-based gene-set enrichment analysis [@xu2024a] indicated that young participants had relatively higher expression of genes related to aerobic energy metabolism (oxidative phosphorylation) and relatively lower expression of genes related to immune system activation (complement activation, classical pathway) and epigenetic regulation of gene expression (heterochromatin formation). Training induced a large, broad transcriptional response in both age groups: `r n_sig_young_t3` genes changed from baseline to post in young participants, and `r n_sig_old_t3` in old participants (FDR < 5%, Fig 6C/D). Individual genes with the strongest training-induced changes were predominantly collagen and basement-membrane genes in both groups (e.g., COL4A1, COL4A2, LAMB1, PXDN in young; COL4A1, LAMA4, COL3A1, COL1A2, SPARC in old).

_Training-effect_
Rank-based gene-set enrichment analysis [@xu2024a] showed that the dominant biology of the training response was similar between age groups: 6 of the top 10 enriched categories were shared, with matching direction and comparable magnitude in both young and old (RNA splicing, ribonucleoprotein complex biogenesis, integrin-mediated signaling pathway, cell adhesion mediated by integrin, collagen metabolic process, and mRNA transport). Training up-regulated genes related to ECM/connective tissue remodeling (e.g., integrin-mediated signaling and adhesion, collagen metabolic process) and down-regulated genes related to RNA-processing and ribosome biogenesis (e.g., RNA splicing, ribonucleoprotein complex biogenesis, mRNA transport) in both groups. While essentially no genes (1 gene at 5% and 0 at 1%) showed a statistically significant age-dependent training response after FDR correction, gene-set enrichment analysis of the age × time interaction contrast revealed that the magnitude of the shared training response differed somewhat between age groups in specific processes: the down-regulation of RNA splicing and related mRNA-processing categories was more pronounced in young than old, while the up-regulation of angiogenesis and aerobic respiration/cellular respiration categories was more pronounced in young than old.

A smaller set of categories differed by age: young participants showed additional enrichment for cytoplasmic translation, mesenchymal cell migration, and angiogenesis, while old participants showed additional enrichment for coagulation- and vascular-related categories. In contrast to the broadly age-independent training effect, the effect of training volume (moderate vs. low) was age-dependent: a substantial number of genes differed by volume in old participants (`r n_sig_old_volume_t3` genes, FDR < 5%), while only one gene showed a volume effect in young participants. Among old participants, higher training volume was associated with relatively higher expression of genes related to mitochondrial biogenesis and oxidative metabolism (e.g., inner mitochondrial membrane organization, respiratory electron transport chain, purine-containing compound biosynthesis) and relatively lower expression of genes related to muscle contractile function and translation (e.g., skeletal muscle contraction, cytoplasmic translation, muscle organ development, striated muscle cell differentiation, ribosomal small subunit biogenesis). This volume effect was considerably more modest than the main training effect, both in the number of genes involved and in statistical strength. 


## Discussion

<!-- P1: Opening paragraph -->
Old individuals in the intervention increased molecular markers of ribosome biogenesis (total RNA, UBF, c-Myc) in a similar fashion to young, though on a delayed timescale, and the RNA-sequencing data showed a marked difference between age groups at baseline that became more similar after training. 

<!-- P2: Ribosome biogenesis - RNA, UBF, cMyc, rnaseq? -->
Briefly, both UBF and c-Myc protein increased from baseline to week 3, however UBF content increased similarly in the old control, suggesting this was a general temporal effect rather than a training-specific response. On the contrary, c-Myc content was markedly lower in the old control group than in old intervention. Total RNA also increased in most groups including control, and differed by timing and not magnitude. The young increased primarily from baseline to week 3, and the old caught up by post, with no clear difference between age groups or intervention and control, and no volume condition effect. Lastly, the RNA-seq data revealed a largly age-independent training response dominated by up-regulation of ECM/collagen genes and down-regulation of RNA-processing/ribosome-biogenesis genes from baseline to post. The Age×time interaction is essentially null at the single-gene level, but GSEA shows old adults have a smaller-magnitude version of the same shared response (less RNA-splicing down-regulation, less angiogenesis/aerobic-respiration up-regulation). Training volume was impactful in the old (mitochondrial biogenesis up, contractile/translation genes down with more volume) but not at all in young. 

Findings from the present study suggests that ribosome biogenesis may be delayed rather than blunted in the elderly, as elderly reach close to the same levels of change as young, only at a later time point. The young group had the expected early rise following both moderate and low-volume exercise (19 and 16% respectively), which eased only modestly by post (12 and 11%), remaining clearly elevated above baseline rather than returning toward it — a pattern consistent with previous findings from our lab [@lian_glucose_2024; @hammarstrom_ribosome_2022]. While the elderly intervention group did not match this change after 3 weeks (mod = 8%, low = 10%), they achieved near similar increases at post (mod = 16%, low = 13%). These changes were clear across both volume conditions in the young at week 3, and for moderate-volume at post; low-volume in the young showed only a weaker, less certain trend toward an increase at post. In the elderly, a clear increase emerged only at post. Control also showed a similar baseline-to-post increase as old (13%), though this was a weaker, less certain trend than the changes seen in the trained groups. From baseline to week 3, the change in control was considered flat, at -1%. A possible explanation here could be timing of the strength tests, as post-strength testing was conducted three-four days prior to the biopsy - whereas at baseline biopsies were sampled before strength testing, and there were no strength testing at all around the week 3 biopsy sampling. Given that total RNA is quite sensitive to such stimuli [SOURCE ACUTE STUDIES], the strength testing alone may have caused the observed spike in total RNA in the control group, and does indeed present a methodological weakness of a poorly timed test. 

Changes in UBF and c-Myc protein matches the total RNA pattern quite well from baseline to week 3, with UBF tending to increase across all four intervention subgroups, albeit weakly, while c-Myc did not increase following low-volume exercise in the old intervention group. In the control group, c-Myc content decreased while content of UBF increased in a similar fashion to the old intervention group. Observed changes in protein content in the intervention group matches previous short-term/acute study findings [@figueiredo_impact_2016; @lian_glucose_2024], and the fact that low-volume training did not yield meaningful change in the old may further corroborate the narrative that increasing RT volume is more important for the elderly [@lixandrao_higher_2024]. What does not quite add up in this story, is the increase of UBF content in the control group, paired with a decrease in c-Myc, since UBF is expected to be a downstream and training-specific target, triggered by c-Myc's transcriptional control [@figueiredo_impact_2016].

<!-- RNA-seq -->
The RNA-seq pattern observed in the present study quite closely matches previous findings. An earlier metformin trial in older adults found progressive resistance training to significantly increase expression of genes involved in ECM remodelling and down-regulate genes in RNA-processing pathways [@kulkarni_metformin_2020], the same two dominant, opposite-direction gene sets observed in the present study. Recently, an aged-muscle reanalysis across modalities found that combined training and resistance training were mainly linked to ECM, collagen, adhesion, basement membrane, also similar to our results [@yang_transcriptomic_2026]. Impaired transcriptional responses in mitochondrial, ECM and immune system genes have also been described as defining features of age-related anabolic resistance to resistance training [@deane_proteomic_2022], among these, baseline expression of genes related to oxidative phosphorylation was relatively lower in old compared to young, while genes related to activation of the immune system were relatively higher in the old, in the present study. The down-regulation of RNA-processing/ribosome-biogenesis gene categories at post compared to baseline may at first glance seem inconsitent with the increased total RNA from baseline to post. On the contrary, this fits well within the interpretation that once translational capacity has been built up, during the early-phase, further biogenesis-gene transcription is dialed back via feedback [@figueiredo_ribosome_2015; @hammarstrom_ribosome_2022]. 
 

## References




















