---
title: "Protein sensitivity runs"
format: html
editor: source
editor_options:
  chunk_output_type: console
---

```{r}
#| label: packages-and-settings
#| eval: true
#| message: false
#| warning: false
#| echo: true

library(brms)
library(marginaleffects)
library(tidyverse)
library(reliefdata)

# Disables a safety warning in marginaleffects. We confirm the order of effects
# for the contrasts ourselves (check_grid() below).
options(marginaleffects_safe = FALSE)

```

### Protein sensitivity runs (UBF, c-Myc high)

**Runs.** Each is fitted for UBF and c-Myc high. The outcome is log-transformed and
standardised per protein and per run (see `make_prot_dat()`).

| Run | Outcome | Data | Model |
|-----|---------|------|-------|
| A (primary) | log(corr.area) | all samples | base |
| B | log(area) | all samples | base (no Compass normalisation) |
| C | log(corr.area) | excludes normalisation factor < 0.5 and Compass-flagged capillaries | base |
| C2 (UBF only) | log(corr.area) | as C, with the stricter cutoff 0.4 | base |
| D | log(corr.area) | all samples | base plus a plate intercept |

**Why corr.area is primary.** `corr.area = area / norm_factor`, where the factor is one
value per capillary, scaled so that reference capillary 2 on each plate is exactly 1. The
plate-specific scale therefore cancels in within-participant changes. The factor is mostly
capillary-specific noise: it correlates only weakly across proteins for the same sample
(r ~ 0.22) and did not reduce within-participant variance. Run B shows whether the result
depends on the normalisation at all; Runs C/C2 show whether it depends on heavily corrected
or flagged capillaries.

**Design.** Each participant has three different samples: one baseline (randomised to the
left or right leg, assumed equal for both legs) and one week-3 sample per leg. There are no
technical replicates. The sample suffix (e.g. 26.3) is an identifier added at import into
`reliefdata` and says nothing about time or leg; always use the `time` and `leg` columns.
A baseline can therefore be among the samples removed in Run C.

```{r}
#| label: protein-data
#| message: false

west.dat <- reliefdata::relief_western |>
  mutate(participant = as.character(participant)) |>
  select(participant, leg, sampleid = sample, time, capillary, peak,
         protein, position, mw_kda, height, area, area_p, corr.area,
         width, plate) |>
  filter(!sampleid %in% c("Biot. Ladder", "DDH2O"),
         !participant %in% c("3084", "3090"))   # no measurement

condition <- reliefdata::relief_volume |>
  mutate(participant = as.character(participant))

participants <- reliefdata::relief_participants |>
  mutate(age_group = if_else(age < 40, "yng", "old")) |>
  select(participant, age, sex, allocation, age_group)

# cMyc_low (~48 kDa) is not analysed: within-participant SD is ~0.91 on the log scale
# (c-Myc high: 0.48) and no model fitted it reliably. Reported in the supplement.
analysed_proteins <- c("UBF", "cMyc_high")

# Capillaries with a Compass warning, excluded in Runs C/C2 (matched on protein + plate + sample).
# TODO: add the UBF and peak-fit warnings after checking them in Compass.
flags <- tribble(
  ~protein,    ~plate, ~sampleid, ~warning,
  "cMyc_high", "P2",   "16.3",    "standards: low confidence",
  "cMyc_high", "P5",   "26.3",    "standards: low confidence"
)

make_prot_dat <- function(prot, col = "corr.area", min_factor = 0, drop_flagged = FALSE) {
  stopifnot(prot %in% analysed_proteins)

  west.dat |>
    full_join(condition) |>
    select(-arm) |>
    full_join(participants) |>
    # Removes NAs introduced by the joins (a baseline exists for one leg only) and blank capillaries
    filter(!is.na(sampleid), protein == prot) |>
    mutate(across(c(area, corr.area), as.numeric),
           norm_factor = area / corr.area) |>
    left_join(flags |> mutate(flagged = TRUE) |> select(protein, plate, sampleid, flagged),
              by = c("protein", "plate", "sampleid")) |>
    mutate(flagged = coalesce(flagged, FALSE)) |>
    filter(norm_factor >= min_factor,           # min_factor = 0 keeps every row with a valid ratio
           !(drop_flagged & flagged)) |>
    # Mean and SD come from the rows actually fitted in this run, so each fit is
    # back-transformed with its own sd_y. A raw-scale SD here gives Inf after exp().
    mutate(y    = log(.data[[col]]),
           ystd = (y - mean(y, na.rm = TRUE)) / sd(y, na.rm = TRUE),
           condition = if_else(allocation == "con", "con", condition),
           tx = paste0(allocation, "_", age_group, "_", condition),
           tx = factor(tx, levels = c("int_old_low", "int_old_mod", "int_yng_low", "int_yng_mod", "con_old_con")),
           tx_sex = paste0(tx, "_", sex),
           # Level order matters: the contrast indices (b1...b20) below depend on it
           tx_sex = factor(tx_sex, levels = c(
             "int_old_low_female", "int_old_low_male", "int_old_mod_female", "int_old_mod_male",
             "int_yng_low_female", "int_yng_low_male", "int_yng_mod_female", "int_yng_mod_male",
             "con_old_con_female", "con_old_con_male")),
           time = factor(time, levels = c("pre", "w3")))
}

# C2 exists for UBF only: for c-Myc high every sample below 0.4 is already removed by
# C (factor cutoff or flag), so C2 would duplicate C.
prot_runs <- expand_grid(protein = analysed_proteins, run = c("A", "B", "C", "C2", "D")) |>
  filter(!(run == "C2" & protein == "cMyc_high")) |>
  mutate(data = pmap(list(protein, run), \(p, r)
           make_prot_dat(p,
                         col          = if (r == "B") "area" else "corr.area",
                         min_factor   = case_when(r == "C" ~ 0.5, r == "C2" ~ 0.4, .default = 0),
                         drop_flagged = r %in% c("C", "C2"))),
         n      = map_int(data, nrow),
         mean_y = map_dbl(data, \(d) mean(d$y)),   # for back-transforming predictions (not needed for contrasts)
         sd_y   = map_dbl(data, \(d) sd(d$y)))

prot_runs |> select(protein, run, n, mean_y, sd_y)

# Accessors, so nothing below depends on row position
run_data <- function(p, r) prot_runs |> filter(protein == p, run == r) |> pull(data) |> pluck(1)
run_fit  <- function(p, r) prot_runs |> filter(protein == p, run == r) |> pull(fit)  |> pluck(1)
run_sd   <- function(p, r) prot_runs |> filter(protein == p, run == r) |> pull(sd_y)
run_mean <- function(p, r) prot_runs |> filter(protein == p, run == r) |> pull(mean_y)

```

#### Data checks

```{r}
#| label: protein-data-checks

# 1. One row per participant x leg x time (no technical replicates): every count should be 1
prot_runs |>
  filter(run == "A") |>
  mutate(rows_per_cell = map(data, \(d) d |> count(participant, leg, time) |> count(n, name = "cells"))) |>
  select(protein, rows_per_cell) |>
  unnest(rows_per_cell)

# 2. Which samples does Run C remove? Time and leg are shown because the sample order is
#    randomised, so a removed sample can be a baseline.
prot_runs |>
  filter(run %in% c("A", "C")) |>
  select(run, data) |>                 # protein is already a column inside data
  unnest(data) |>
  distinct(protein, run, participant, leg, time, sampleid, plate, capillary, mw_kda,
           area, corr.area, norm_factor, flagged) |>
  group_by(protein, participant, leg, time, sampleid) |>
  filter(n_distinct(run) == 1, run == "A") |>    # present in A, absent in C
  ungroup()

# 3. Participants left without a baseline or without both week-3 samples, per run.
#    Report this next to the contrasts: a difference between A and C can come from
#    the exclusion criterion or from losing a participant's within-person change.
prot_runs |>
  mutate(cells = map(data, \(d) d |>
           summarise(n_pre = sum(time == "pre"), n_w3 = sum(time == "w3"), .by = participant) |>
           summarise(no_baseline = sum(n_pre == 0),
                     one_w3_only = sum(n_w3 == 1),
                     no_w3       = sum(n_w3 == 0)))) |>
  select(protein, run, n, cells) |>
  unnest(cells)

```

#### Models

The same specification is used for every run and both proteins, so the runs differ only in
the data (or, for D, in the plate term).

- **No random time slope.** Each participant contributes one baseline and two week-3
  samples (fewer in the sensitivity runs), which cannot support participant-specific change.
  The model with `(1 + time | participant)` sampled poorly for UBF (up to 85 divergences, ESS
  down to ~74) while giving the same contrasts as this model (see the supplement chunk
  at the end).
- **Priors.** `normal(0, 1)` on coefficients and `exponential(1)` on all sds and sigma, on
  the standardised scale. These are the priors of the initial protein analyses (mu2/mhi2), so
  no new prior predictive checks were run.
- **Sampler settings.** 10,000 iterations, `adapt_delta = 0.99`, `max_treedepth = 15`.
  `init = 0.1` avoids chains getting stuck early in warmup, and `seed` makes the fits reproducible.

```{r}
#| label: protein-fits

form_A <- bf(ystd ~ 0 + tx_sex + tx_sex:time + (1 | participant) + (1 | participant:leg))
form_D <- bf(ystd ~ 0 + tx_sex + tx_sex:time + (1 | participant) + (1 | participant:leg) + (1 | plate))

pri_A <- c(
  prior(normal(0, 1), class = "b"),
  prior(exponential(1), class = sd, group = "participant"),
  prior(exponential(1), class = sd, group = "participant:leg"),
  prior(exponential(1), class = sigma)
)
pri_D <- c(pri_A, prior(exponential(1), class = "sd", group = "plate"))

prot_runs <- prot_runs |>
  mutate(fit = pmap(list(protein, run, data), \(p, r, d)
    brm(if (r == "D") form_D else form_A,
        prior = if (r == "D") pri_D else pri_A,
        data  = d,
        cores = 4, iter = 10000,
        control = list(adapt_delta = 0.99, max_treedepth = 15),
        init = 0.1, seed = 1,
        file = file.path("fits/04-protein", paste0(p, "_", r)),
        file_refit = "on_change")))

# Sampler diagnostics for one fit: divergences, max Rhat, min effective sample size
sampler_diag <- function(m) {
  np <- nuts_params(m)
  tibble(divergences = sum(np$Value[np$Parameter == "divergent__"]),
         max_rhat    = max(rhat(m), na.rm = TRUE),
         min_ess     = min(neff_ratio(m) * ndraws(m), na.rm = TRUE))
}

# Goes in the supplement. Every fit should have 0 divergences, Rhat < 1.01 and ESS > 400.
prot_diag <- prot_runs |>
  mutate(diag = map(fit, sampler_diag)) |>
  select(protein, run, diag) |>
  unnest(diag) |>
  print()

if (any(prot_diag$divergences > 0 | prot_diag$max_rhat > 1.01 | prot_diag$min_ess < 400)) {
  warning("At least one fit fails the diagnostic thresholds: do not interpret its contrasts yet.")
}

```

#### Contrasts

```{r}
#| label: protein-contrasts

# Same hypotheses as mu2_pred_contr / mhi2_pred_contr. Each delta averages the two sexes
# 50/50 on the log (standardised) scale, so results do not depend on the sex balance of a group.
# The indices rely on the datagrid order: b(2k-1) = pre and b(2k) = w3 for the k-th
# tx_sex level. check_grid() verifies that order for every fit.
hyp_prot <- c(
  "int_old_low_delta"    = "((b2-b1)+(b4-b3))/2 = 0",
  "int_old_mod_delta"    = "((b6-b5)+(b8-b7))/2 = 0",
  "int_yng_low_delta"    = "((b10-b9)+(b12-b11))/2 = 0",
  "int_yng_mod_delta"    = "((b14-b13)+(b16-b15))/2 = 0",
  "con_old_con_delta"    = "((b18-b17)+(b20-b19))/2 = 0",
  "int_old_pooled_delta" = "(((b2-b1)+(b4-b3))/2 + ((b6-b5)+(b8-b7))/2)/2 = 0",
  "int_yng_pooled_delta" = "(((b10-b9)+(b12-b11))/2 + ((b14-b13)+(b16-b15))/2)/2 = 0",
  "yng_vs_old_delta"     = "(((b14-b13)+(b16-b15))/2 + ((b10-b9)+(b12-b11))/2)/2 - (((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2)/2 = 0",
  "old_vs_con_delta"     = "((((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2)/2) - (((b18-b17)+(b20-b19))/2) = 0",
  "yng_mod_vs_low"       = "((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2 = 0",
  "old_mod_vs_low"       = "((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2 = 0",
  "yng_vs_old_vol"       = "(((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2) - (((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2) = 0"
)

check_grid <- function(m) {
  g <- datagrid(model = m, tx_sex = unique, time = unique)
  stopifnot(identical(as.character(g$time[1:2]), c("pre", "w3")),
            g$tx_sex[1] == g$tx_sex[2],
            identical(as.character(g$tx_sex[1]), "int_old_low_female"))
  invisible(TRUE)
}

# Draw-level contrasts on the standardised log scale, back-transformed to ratios with the
# fit's OWN sd_y (standardisation is per protein and per run). Population-level effects only
# (re_formula = NA). p_dir is the posterior probability in the direction of the estimate;
# per the reporting rules it is omitted when the 95% interval excludes zero.
prot_contr <- function(fit, sd_y, ndraws = 4000) {
  check_grid(fit)
  predictions(fit,
              newdata = datagrid(tx_sex = unique, time = unique),
              re_formula = NA, ndraws = ndraws,
              hypothesis = hyp_prot) |>
    posterior_draws() |>
    mutate(ratio = exp(draw * sd_y)) |>
    summarise(.by = hypothesis,
              mean    = mean(ratio),
              lower95 = quantile(ratio, 0.025),
              upper95 = quantile(ratio, 0.975),
              p_pos   = mean(ratio > 1)) |>
    mutate(pct      = (mean - 1) * 100,
           pct_lo   = (lower95 - 1) * 100,
           pct_hi   = (upper95 - 1) * 100,
           credible = lower95 > 1 | upper95 < 1,
           p_dir    = if_else(mean >= 1, p_pos, 1 - p_pos))
}

prot_runs <- prot_runs |>
  mutate(contr = map2(fit, sd_y, \(m, s) prot_contr(m, s)))

contr_all <- prot_runs |>
  select(protein, run, n, contr) |>
  unnest(contr) |>
  mutate(hypothesis = factor(hypothesis, levels = names(hyp_prot)),
         run        = factor(run, levels = c("A", "B", "C", "C2", "D")))

saveRDS(contr_all, "data/data-gen/contrast_protein_runs.rds")

# Side-by-side comparison of runs (A at the top of each panel)
contr_all |>
  ggplot(aes(pct, fct_rev(run))) +
  geom_pointrange(aes(xmin = pct_lo, xmax = pct_hi)) +
  geom_vline(xintercept = 0, lty = 2) +
  facet_grid(protein ~ hypothesis, scales = "free_x", labeller = label_wrap_gen(12)) +
  labs(x = "Change / difference (%)", y = "Run")

# Compact table: percent change [95% CI] by run (C2 is UBF only)
contr_all |>
  mutate(est = sprintf("%.1f [%.1f, %.1f]", pct, pct_lo, pct_hi)) |>
  select(protein, hypothesis, run, est) |>
  pivot_wider(names_from = run, values_from = est) |>
  arrange(protein, hypothesis)

```

#### Results, models check

Sensitivity analyses (protein). The primary analysis used log-transformed Compass-corrected area (corr.area; Run A). To assess dependence on the normalisation and on technical quality, models were refitted using uncorrected area (B), excluding capillaries with a normalisation factor < 0.5 or a Compass warning (C; for UBF also with the stricter cutoff 0.4, C2), and with a random plate intercept (D). All models used the same specification and priors as the primary analysis, and all fits showed no divergent transitions (max R̂ ≤ 1.01, min ESS = 1,588). For c-Myc (high), estimates were nearly identical across runs (all within a few percentage points), and the credible findings (a decrease in the control group and a difference between older intervention and control adults) were unchanged. For UBF, no intervention-group change or group comparison reached credibility in any run; the control-group increase had an interval including zero in the primary analysis and excluding zero in runs B, C and C2.

Directive from this testing:
The case for A is a design argument, not a result. corr.area corrects each capillary for how much total protein it contained, which is the reason to normalise at all. The earlier analysis was less supportive: the factor looked mostly like capillary-specific noise (r ≈ 0.22 across proteins) and didn’t reduce within-participant variance. That is an argument for B, since normalising a noisy factor can add noise. Run B shows it doesn’t matter much for these data. For c-Myc high, A and B agree within a few points, and for UBF they agree except for the control group’s band. So the choice between A and B is not decisive, and I’d keep A because it was your pre-specified analysis and the standard approach with this instrument. If a reviewer pushes, B is the answer, and the table shows the conclusions hold.

So the primary analysis is A, with B, C/C2 and D as sensitivity.

#### Primary results (Run A)

Everything the paper reports for the proteins comes from Run A. The output files below keep
the names and columns of the earlier UBF / c-Myc files, so the existing table and figure
scripts read them unchanged. NOTE: this overwrites the old `pred_*.rds` and `contrast_*.rds`
in `data/data-gen/` (the old mu2/mhi2 fits in `fits/02-protein/` can regenerate them).

```{r}
#| label: protein-primary-results

# Reporting bands (same as the RNA results): credible = 95% CI excludes zero;
# otherwise likely (>= 90%), weak (75-90%), no evidence (< 75%) by direction probability.
band <- \(credible, p_dir) case_when(credible ~ "credible", p_dir >= 0.90 ~ "likely",
                                     p_dir >= 0.75 ~ "weak", .default = "none")

# Contrasts: percent change [95% CI]; P is given in the direction of the estimate and
# omitted when the interval excludes zero.
res_A <- contr_all |> 
  filter(run == "A") |> 
  mutate(band = band(credible, p_dir),
         dir  = if_else(mean >= 1, "P(>0)", "P(<0)"),
         text = sprintf("%.1f%%, 95%% CI [%.1f%%, %.1f%%]%s", pct, pct_lo, pct_hi,
                        if_else(credible, "", sprintf(", %s = %.1f%%", dir, p_dir * 100))),
         text = str_replace_all(text, "-", "\u2212")) |>      # typographic minus
  select(protein, hypothesis, pct, pct_lo, pct_hi, p_dir, band, text)

res_A |> select(protein, hypothesis, band, text) |> print(n = Inf)

# Same columns as the earlier contrast_ubf / contrast_cmyc_high files
save_contrast <- function(p, file) {
  contr_all |> 
    filter(run == "A", protein == p) |> 
    transmute(hypothesis = as.character(hypothesis), mean, lower95, upper95,
              pos = p_pos, pct_change = pct, pct_lower = pct_lo, pct_upper = pct_hi,
              credible = if_else(credible, "Yes", "No")) |> 
    saveRDS(file)
}
save_contrast("UBF",       "data/data-gen/contrast_ubf.rds")
save_contrast("cMyc_high", "data/data-gen/contrast_cmyc_high.rds")

# Predictions for the figures: population-level (re_formula = NA) per treatment group, sex
# and time, back-transformed to corrected-area units with this fit's own mean_y and sd_y.
# 4000 draws (the earlier plotting code used 50, too few for 95% intervals).
tx_levels <- c("int_old_low", "int_old_mod", "int_yng_low", "int_yng_mod", "con_old_con")

make_pred <- function(p, ndraws = 4000) {
  predictions(run_fit(p, "A"),
              newdata = datagrid(tx_sex = unique, time = unique),
              re_formula = NA, ndraws = ndraws) |> 
    posterior_draws() |> 
    mutate(sex   = if_else(grepl("_female", tx_sex), "female", "male"),
           tx    = factor(gsub("_(female|male)$", "", tx_sex), levels = tx_levels),
           value = exp(draw * run_sd(p, "A") + run_mean(p, "A"))) |> 
    summarise(.by = c(tx, sex, time),
              m     = mean(value),
              lwr50 = quantile(value, 0.25),  upr50 = quantile(value, 0.75),
              lwr90 = quantile(value, 0.05),  upr90 = quantile(value, 0.95),
              lwr95 = quantile(value, 0.025), upr95 = quantile(value, 0.975))
}

pred_ubf        <- make_pred("UBF")
pred_cmyc_high  <- make_pred("cMyc_high")

# Sanity check: values should be on the scale of the raw corrected areas
bind_rows(UBF = pred_ubf, cMyc_high = pred_cmyc_high, .id = "protein") |> 
  summarise(.by = protein, min = min(m), max = max(m))
prot_runs |> filter(run == "A") |> 
  mutate(raw_median = map_dbl(data, \(d) median(d$corr.area))) |> select(protein, raw_median)

saveRDS(pred_ubf,       "data/data-gen/pred_ubf.rds")
saveRDS(pred_cmyc_high, "data/data-gen/pred_cmyc_high.rds")
```

#### Main results

_UBF_
Older adults in the intervention group (averaged across volume conditions) tended to increase UBF protein content by week 3 (20.4%, 95% CI [−5.6%, 50.4%], P(>0) = 93.3%), as did the control group (27.6%, 95% CI [−4.2%, 64.8%], P(>0) = 95.2%), with no evidence of a difference between them (−3.9%, 95% CI [−33.2%, 35.9%], P(<0) = 63.1%). Young adults (averaged across volume conditions) also tended to increase UBF (16.9%, 95% CI [−7.1%, 45.2%], P(>0) = 91.0%), with no evidence of a difference from older adults (−1.4%, 95% CI [−29.7%, 33.9%], P(<0) = 57.0%). Within young adults, there was weak evidence of an increase with both low-volume (17.1%, 95% CI [−14.4%, 55.1%], P(>0) = 83.2%) and moderate-volume training (18.2%, 95% CI [−15.8%, 60.2%], P(>0) = 82.6%), with no evidence of a difference between them (3.4%, 95% CI [−34.7%, 56.9%], P(>0) = 51.1%). In older adults, low-volume training tended to increase UBF (29.8%, 95% CI [−8.4%, 77.3%], P(>0) = 92.0%), whereas there was no evidence of a change with moderate-volume training (13.2%, 95% CI [−19.0%, 57.1%], P(>0) = 74.6%), and no evidence of a difference between them (−10.1%, 95% CI [−44.9%, 41.5%], P(<0) = 71.6%). There was no evidence that the volume effect differed between age groups (21.9%, 95% CI [−41.3%, 117.3%], P(>0) = 67.9%).

_c-Myc High Peak_
c-Myc (high peak) decreased in the control group by week 3 (−35.8%, 95% CI [−55.4%, −11.2%]). Older adults in the intervention group (averaged across volume conditions) tended to increase c-Myc (26.0%, 95% CI [−6.7%, 65.7%], P(>0) = 93.5%), and their change was greater than that of the control group (102.9%, 95% CI [22.4%, 211.1%]), a difference driven by the decrease in the control group. Young adults (averaged across volume conditions) showed weak evidence of an increase (19.8%, 95% CI [−9.1%, 54.3%], P(>0) = 89.5%), with no evidence of a difference from older adults (−2.9%, 95% CI [−34.8%, 41.5%], P(<0) = 60.1%). Within young adults, there was no evidence of a change with low-volume training (4.4%, 95% CI [−28.5%, 48.4%], P(>0) = 54.9%), whereas moderate-volume training tended to increase c-Myc (40.4%, 95% CI [−9.1%, 106.1%], P(>0) = 93.5%), with weak evidence of a difference between them (39.6%, 95% CI [−24.4%, 134.6%], P(>0) = 85.2%). In older adults, there was no evidence of a change with low-volume training (10.8%, 95% CI [−28.8%, 62.4%], P(>0) = 65.5%), whereas moderate-volume training tended to increase c-Myc (46.6%, 95% CI [−5.8%, 116.2%], P(>0) = 95.6%), again with weak evidence of a difference between them (38.3%, 95% CI [−26.1%, 138.5%], P(>0) = 82.6%). There was no evidence that the volume effect differed between age groups (10.9%, 95% CI [−54.8%, 140.8%], P(>0) = 50.0%).


#### Supplement: choice of model structure

Record of why the random time slope was dropped. Not evaluated when rendering; the fits are
stored in `fits/04-protein/supplement/` and reload instantly if the chunk is run by hand.

Results at the time of writing (UBF run A): the full model gave 40 divergences, the same
model at `adapt_delta = 0.999` gave 6 (and a low-BFMI chain), tighter sd priors
(`exponential(2)`) gave 0, and the model without slope gave 0. All four gave
contrasts within ~3 percentage points of each other with similar interval widths. In the sparse
UBF run C2 the full model with tighter priors still gave 26 divergences (ESS ~420), while the
model without slope gave 0 (ESS ~1,600). The slope sd, the intercept-slope correlation and the
leg sd all had posteriors reaching zero: with three samples per participant the data cannot
split the variance among these terms.

```{r}
#| label: supplement-model-structure
#| eval: false

form_full <- bf(ystd ~ 0 + tx_sex + tx_sex:time + (1 + time | participant) + (1 | participant:leg))
pri_full  <- c(
  prior(normal(0, 1), class = "b"),
  prior(lkj(2), class = "cor", group = "participant"),
  prior(exponential(1), class = sd, group = "participant"),
  prior(exponential(1), class = sd, group = "participant:leg"),
  prior(exponential(1), class = sigma))
pri_full_tight <- c(
  prior(normal(0, 1), class = "b"),
  prior(lkj(2), class = "cor", group = "participant"),
  prior(exponential(2), class = sd, group = "participant"),
  prior(exponential(2), class = sd, group = "participant:leg"),
  prior(exponential(1), class = sigma))

fit_spec <- function(form, pri, data, file, ad = 0.99) {
  brm(form, prior = pri, data = data, cores = 4, iter = 10000,
      control = list(adapt_delta = ad, max_treedepth = 15),
      init = 0.1, seed = 1, file = file, file_refit = "on_change")
}

# --- UBF run A: full model vs variants vs the model used in the analysis -------------
d_A <- run_data("UBF", "A")
spec_A <- list(
  full       = fit_spec(form_full, pri_full,       d_A, "fits/04-protein/supplement/UBF_A_full"),
  full_ad999 = fit_spec(form_full, pri_full,       d_A, "fits/04-protein/supplement/UBF_A_full_ad999", ad = 0.999),
  full_tight = fit_spec(form_full, pri_full_tight, d_A, "fits/04-protein/supplement/UBF_A_full_tight"),
  noslope    = run_fit("UBF", "A"))

map_dfr(spec_A, sampler_diag, .id = "spec")

map_dfr(spec_A, \(f) prot_contr(f, run_sd("UBF", "A")), .id = "spec") |>
  mutate(est = sprintf("%.1f [%.1f, %.1f]", pct, pct_lo, pct_hi)) |>
  select(spec, hypothesis, est) |>
  pivot_wider(names_from = spec, values_from = est)

# Which variance terms are weakly identified (lower limits near 0, wide intervals)?
summary(spec_A$full)$random

# --- UBF run C2: the sparsest case ------------------------------------------------
d_c2 <- run_data("UBF", "C2")
spec_C2 <- list(
  full_tight = fit_spec(form_full, pri_full_tight, d_c2, "fits/04-protein/supplement/UBF_C2_full_tight"),
  noslope    = run_fit("UBF", "C2"))

map_dfr(spec_C2, sampler_diag, .id = "spec")
```
