---
title: "Molecular"
format: html
editor: source
editor_options: 
  chunk_output_type: console
---


```{r}
#| label: packages-and-settings
#| eval: true
#| message: false
#| warning: false
#| echo: true

# Loading packages
library(brms)
library(marginaleffects)
library(tidyverse)
library(reliefdata)


# This setting disables a safety warning message for the marginaleffects package
# we should still confirm the order of effects for the contrasts
options(marginaleffects_safe = FALSE)


```

### Total RNA

Total RNA was extracted via the trizol protocol, and measured with the NanoDrop spectrophotometry. Biopsies were sampled before, 3 weeks/6 sessions into the intervention, and after the intervetion. Samples were run with 4 replicates on the spectrophotometer, to estimate the rna concentration. 

With this data, we are interested in estimating:

1.  Within condition (low- and moderate-volume, and non-training control conditions) change from baseline.
2.  Differences between age-groups at baseline.
3.  Average between age-group differences in change from baseline (young vs. old in the intervention group)
4.  Average between control- vs. intervention-group changes from baseline.
5.  Average effect of condition in the intervention groups (low- vs. moderate volume)
6.  Differences between age-groups in effect of condition.


#### Data -The relationship between total RNA and mg tissue

In practice, weighing small tissue pieces carries real measurement error, which is amplified when dividing by small mg values — potentially contributing to the sampling instability seen
in the UBF and c-Myc models. Rather than pre-computing a fixed ratio, we model the relationship between measured RNA and mg weight directly (on the log-log scale, consistent with the project's existing log + standardization convention). The aim is to let the mg-RNA
relationship be estimated from the data rather than assumed, which should:

- allow uncertainty to scale appropriately for extreme mg observations,
  rather than being artificially inflated or deflated by the ratio, and
- yield more precise estimates in the well-populated middle of the mg
  distribution, where the fixed-ratio approach was not actually using the
  available information efficiently.

```{r}

# Data

conc.dat <- reliefdata::relief_rnaconc |> 
  mutate(participant = as.character(participant)) |> 
  mutate(participant = ifelse(is.na(participant), "blank", participant),
         leg = ifelse(is.na(leg), "blank", leg),
         time = ifelse(is.na(time), "blank", time),
         weight = ifelse(is.na(weight), "0", weight)) |> 
  mutate(total.rna = conc * 4 * 25) |> 
  dplyr::select(participant, leg, time, weight, well, plate, sample, total.rna) |> 
  filter(participant != "blank") |> 
  mutate(weight = as.numeric(weight)) |> 
  print()

condition <- reliefdata::relief_volume |> 
  mutate(participant = as.character(participant))

participants <- reliefdata::relief_participants |> 
  mutate(age_group = if_else(age < 40, "yng", "old")) |> 
  dplyr::select(participant, age, sex, height, bodyweight = weight, allocation, age_group)


rna.dat <- conc.dat |> 
  full_join(condition) |> 
  select(-arm) |> 
  full_join(participants) |> 
  group_by(participant, leg, time, age_group, allocation, condition, sex) |> 
  summarise(
    mean_rna = mean(total.rna, na.rm = TRUE), # average technical replicates (conc side)
    weight = first(weight)                    # one weight measurement per sample
  ) |> 
  ungroup() |> 
  mutate(
    y     = log(mean_rna),
    ystd  = (y - mean(y, na.rm = TRUE)) / sd(y, na.rm = TRUE),
    mg    = log(weight),
    mgstd = (mg - mean(mg, na.rm = TRUE)) / sd(mg, na.rm = TRUE),
    condition = if_else(allocation == "con", "con", condition),
    tx = paste0(allocation, "_", age_group, "_", condition),
    tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con")),
    tx_sex = paste0(tx, "_", sex),
    tx_sex = factor(tx_sex, levels = c(
      "int_old_low_female", "int_old_low_male",
      "int_old_mod_female", "int_old_mod_male", 
      "int_yng_low_female", "int_yng_low_male",
      "int_yng_mod_female", "int_yng_mod_male", 
      "con_old_con_female", "con_old_con_male"
    )),
    time = factor(time, levels = c("pre", "w3", "post")),
    sex_centered = if_else(sex == "female", -0.5, 0.5)
  ) |> 
  filter(!is.na(y)) |> 
  filter(!is.na(mg)) |> 
  filter(!is.na(time)) |> 
  print()
    
# Checking for mg = 0, which would produce inf
#sum(is.infinite(rna.dat$mg))
# zero
  

## Save sd and mean for back transformation
mean_rna <- mean(rna.dat$y, na.rm = TRUE)
sd_rna <- sd(rna.dat$y, na.rm = TRUE)
mean_mg <- mean(rna.dat$mg, na.rm = TRUE)
sd_mg <- sd(rna.dat$mg, na.rm = TRUE)

# A preliminary figure of observed data
# RNA plotted against weight - the relationship to be estimated in the moodel

rna_mg_plot <- rna.dat %>% 
  ggplot(aes(weight, mean_rna, color = time)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Tissue weight (mg, log scale)", 
       y = "Total RNA (ng, log scale)")


rna.dat %>% 
  ggplot(aes(mgstd, ystd, color = time)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +

  labs(x = "Tissue weight (mg, log scale, standardized)", 
       y = "Total RNA (ng, log scale, standardized)")



## Data for model 4
# Model 4 will have a random intercept per biopsy replicate, so Im making a separate data frame here without summarising to mean_rna per biopsy.


rna_rep.dat <- conc.dat |> 
  full_join(condition) |> 
  select(-arm) |> 
  full_join(participants) |> 
  mutate(
    y     = log(total.rna),
    ystd  = (y - mean(y, na.rm = TRUE)) / sd(y, na.rm = TRUE),
    mg    = log(weight),
    mgstd = (mg - mean(mg, na.rm = TRUE)) / sd(mg, na.rm = TRUE),
    condition = if_else(allocation == "con", "con", condition),
    tx = paste0(allocation, "_", age_group, "_", condition),
    tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con")),
    tx_sex = paste0(tx, "_", sex),
    tx_sex = factor(tx_sex, levels = c(
      "int_old_low_female", "int_old_low_male",
      "int_old_mod_female", "int_old_mod_male", 
      "int_yng_low_female", "int_yng_low_male",
      "int_yng_mod_female", "int_yng_mod_male", 
      "con_old_con_female", "con_old_con_male"
    )),
    time = factor(time, levels = c("pre", "w3", "post")),
    sex_centered = if_else(sex == "female", -0.5, 0.5)
  ) |> 
  filter(!is.na(y)) |> 
  filter(!is.na(mg)) |> 
  filter(!is.na(time)) |> 
  print()

mean_rna_rep <- mean(rna_rep.dat$y, na.rm = TRUE)
sd_rna_rep   <- sd(rna_rep.dat$y, na.rm = TRUE)
mean_mg_rep  <- mean(rna_rep.dat$mg, na.rm = TRUE)
sd_mg_rep    <- sd(rna_rep.dat$mg, na.rm = TRUE)

```

#### Models

```{r}
# Setting up formulas
form0 <- bf(ystd ~ mgstd) # No grouping, a clean look at the relationship

form1 <- bf(ystd ~ mgstd + (1|participant)) # Now adding a random intercept for participant, to account for repeated measures
# this also means that the model now separates between-participant variance (sd(Intercept)) from residual variance(sigma), i.e., sigma should drop compared to m0.

form2 <- bf(ystd ~ mgstd + tx + time + (1|participant))
# this model has tx and time added, to also explore age, allocation and time effects on the relationship between RNA yield and mg. The initial idea was to only check age and time, but using age_group here would collapse int_old and con_old, which should stay separate. "tx" fixes that, but also adds the allocation and condition. An alternative would be to create a new factor that includes only age group and allocation (int_young, int_old, con_old). Written as is, this model is additive, and tests "does group shift RNA level" and "does time shift RNA level".

# What model 2 cannot tell us, is whether the time trajectory itself differs by group - does the control group's RNA stay flat across time while interventions group rise? We can test this by adding tx:time. Probably also time to expand the random effects, with time and tx in the model. Adding time lets each participant's time trajectory vary, and one for leg lets the model capture if a person's low-volume leg runs differently from the moderate-volume leg.

form3 <- bf(ystd ~ mgstd + tx * time + (1 + time | participant) + (1 | participant:leg))

form4 <- bf(ystd ~ mgstd + tx * time + (1 + time | participant) + (1 | participant:leg) + 
  (1|participant:sample))

# This model expands on m3, adding (1|participant:sample). Using participant:sample here instead of "just" sample, for the nesting structure to be explicit and self-documented (i.e., "sample nested within participant").

# Priors
get_prior(form0, data = rna.dat)
get_prior(form1, data = rna.dat)
get_prior(form2, data = rna.dat)
get_prior(form3, data = rna.dat)
get_prior(form4, data = rna_rep.dat)


pri0 <- c(
  prior(normal(0, 1), class = "Intercept"),
  prior(normal(1, 1), class = "b", coef = "mgstd"),
  prior(exponential(1), class = "sigma")
) 
# since both ystd and mgstd are standardized (mean 0, SD 1), a normal(0, 1) prior on the slope is a reasonable weakly-informative choice - meaning "we don't strongly expect the standardized slope to exceed roughly ±2"

# Intercept normal(0, 1) - since ystd is standardized (mean 0), the intercept represents predicted ystd when mgstd = 0 (i.e., at average log-mg). We'd expect this to be close to 0 if mg fully explains the centering, so a weakly-informative prior centered at 0 is sensible, consistent with how we've set priors in previous brms

# b for mgstd, normal(0, 1) - same logic as before: standardized slope, so ±2 SD as a generous plausible range

# sigma, exponential(1) - matches our convention in prior brms work


pri1 <- c(
  prior(normal(0, 1), class = "Intercept"),
  prior(normal(0, 1), class = "b", coef = "mgstd"),
  prior(exponential(1), class = "sd", group = "participant"),
  prior(exponential(1), class = "sigma")
)

# sd gets added as a variance component for participant



pri2 <- c(
  prior(normal(0, 1), class = "Intercept"),
  prior(normal(0, 1), class = "b", coef = "mgstd"),
  prior(normal(0, 1), class = "b", coef = "timew3"),
  prior(normal(0, 1), class = "b", coef = "timepost"),
  prior(normal(0, 1), class = "b", coef = "txcon_old_con"),
  prior(normal(0, 1), class = "b", coef = "txint_old_mod"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_low"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod"),
  prior(exponential(1), class = "sd", group = "participant"),
  prior(exponential(1), class = "sigma")
)

# the reference level here is int_old_low (absorbed into the intercept), since its the first level in the tx factor order. So the intercept now represents predicted ystd for an int_old_low participant at time = pre (pre being the reference level of time) and average mg. Each tx coefficient is then a shift relative to int_old_low, and con_olc_con specifically gives the intervention-vs-control directly as a single coefficient.

# currently no pred/contr functions, since none of the models has any interactions, they're all single coefficients with e.g. int_old_low/pre (m2) as the reference. 



pri3 <- c(
  prior(normal(0, 1), class = "Intercept"),
  prior(normal(1, 1), class = "b", coef = "mgstd"),
  prior(normal(0, 1), class = "b", coef = "timew3"),
  prior(normal(0, 1), class = "b", coef = "timepost"),
  prior(normal(0, 1), class = "b", coef = "txcon_old_con"),
  prior(normal(0, 1), class = "b", coef = "txint_old_mod"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_low"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod"),
  prior(normal(0, 1), class = "b", coef = "txcon_old_con:timew3"),
  prior(normal(0, 1), class = "b", coef = "txcon_old_con:timepost"),
  prior(normal(0, 1), class = "b", coef = "txint_old_mod:timew3"),
  prior(normal(0, 1), class = "b", coef = "txint_old_mod:timepost"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_low:timew3"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_low:timepost"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod:timew3"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod:timepost"),
  prior(lkj(2), class = "cor", group = "participant"),
  prior(exponential(1), class = "sd", group = "participant"),
  prior(exponential(1), class = "sd", group = "participant:leg"),
  prior(exponential(1), class = "sigma")
)


pri4 <- c(
  prior(normal(0, 1), class = "Intercept"),
  prior(normal(1, 1), class = "b", coef = "mgstd"),  
  prior(normal(0, 1), class = "b", coef = "timew3"),
  prior(normal(0, 1), class = "b", coef = "timepost"),
  prior(normal(0, 1), class = "b", coef = "txcon_old_con"),
  prior(normal(0, 1), class = "b", coef = "txint_old_mod"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_low"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod"),
  prior(normal(0, 1), class = "b", coef = "txcon_old_con:timew3"),
  prior(normal(0, 1), class = "b", coef = "txcon_old_con:timepost"),
  prior(normal(0, 1), class = "b", coef = "txint_old_mod:timew3"),
  prior(normal(0, 1), class = "b", coef = "txint_old_mod:timepost"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_low:timew3"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_low:timepost"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod:timew3"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod:timepost"),
  prior(lkj(2), class = "cor", group = "participant"),
  prior(exponential(1), class = "sd", group = "participant"),
  prior(exponential(1), class = "sd", group = "participant:leg"),
  prior(exponential(1), class = "sd", group = "participant:sample"),
  prior(exponential(1), class = "sigma")
)



# adding lkj(2) for the correlation matrix between participant-level random intercept/slopes


m0.pp <- brm(form0,
             prior = pri0,
             data = rna.dat,
             file = "fits/XX/m0.pp",
             file_refit = "on_change",
             sample_prior = "only")
#pp_check(m0.pp, ndraws = 100)

# Prior predictive simulation 
as_draws_df(m0.pp) |> 
  filter(.draw %in% 1:100) |> 
  expand_grid(mgstd = seq(from = -3, to = 3, by = 0.2)) |> 
  # predict
  mutate(y = b_Intercept + b_mgstd * mgstd) |> 
  ggplot(aes(mgstd, y, group = .draw)) + 
  geom_line(alpha = 0.2) +
  
  geom_point(data = rna.dat, 
             aes(mgstd, ystd, group = NULL))
  
 
# Update model to the posterior

m0.post <- update(m0.pp,
                  sample_prior = "no",
                  iter = 4000,
                  file = "fits/XX/m0.post",
                  file_refit = "on_change")

# Posterior predictive simulation 
as_draws_df(m0.post) |> 
  filter(.draw %in% 1:100) |> 
  expand_grid(mgstd = seq(from = -4, to = 3, by = 0.5)) |> 
  # predict
  mutate(y = b_Intercept + b_mgstd * mgstd) |> 
  ggplot(aes(mgstd, y, group = .draw)) + 
  geom_line(alpha = 0.2) +
  
  geom_point(data = rna.dat, 
             aes(mgstd, ystd, group = NULL))



  



m1.pp <- brm(form1,
             prior = pri1,
             data = rna.dat,
             file = "fits/XX/m1.pp",
             file_refit = "on_change",
             sample_prior = "only")
#pp_check(m1.pp, ndraws = 100)

as_draws_df(m1.pp) |> 
  filter(.draw %in% 1:100) |> 
  expand_grid(mgstd = seq(from = -3, to = 3, by = 0.2)) |> 
  # predict
  mutate(y = b_Intercept + b_mgstd * mgstd) |> 
  ggplot(aes(mgstd, y, group = .draw)) + 
  geom_line(alpha = 0.2) +
  
  geom_point(data = rna.dat, 
             aes(mgstd, ystd, group = NULL))





m2.pp <- brm(form2,
             prior = pri2,
             data = rna.dat,
             file = "fits/XX/m2.pp",
             file_refit = "on_change",
             sample_prior = "only")
#pp_check(m2.pp, ndraws = 100)


m3.pp <- brm(form3,
             prior = pri3,
             data = rna.dat,
             file = "fits/XX/m3.pp",
             file_refit = "on_change",
             sample_prior = "only")
#pp_check(m3.pp, ndraws = 100)



m4.pp <- brm(form4,
             prior = pri4,
             data = rna_rep.dat,
             file = "fits/XX/m4.pp",
             file_refit = "on_change",
             sample_prior = "only")


```

```{r}
### Posteriors

## Model 3

pred <- predictions(
    m3.pp,
    type = "prediction",
    newdata = datagrid(tx = unique,
                        time = unique),
    ndraws = 1000) 



prior <- get_draws(pred) |> 
  dplyr::select(drawid, draw, participant, tx, time) |> 
  mutate(fit = "prior")




m3.post <- update(m3.pp,
                   sample_prior = "no",
                   cores = 4,
                   control = list(adapt_delta = 0.95,
                                  max_treedepth = 15),
                   iter = 4000,
                   file = "fits/XX/m3.post",
                   file_refit = "on_change")

pp_check(m3.post, ndraws = 100)

pred.post <- predictions(
    m3.post,
    type = "prediction",
    newdata = datagrid(tx = unique,
                        time = unique),
    ndraws = 4000) 


post <- get_draws(pred.post) |> 
  dplyr::select(drawid, draw, participant, tx, time) |> 
  mutate(fit = "posterior")

bind_rows(prior, 
          post) |> 
  ggplot(aes(draw, color = fit)) + 
  geom_line(stat = "density") +
  facet_wrap(tx ~ time)



# Generate predictions (swapping from type = "predictions" to "response")
pred.post <- predictions(
    m3.post,
    type = "response",   
    newdata = datagrid(tx = unique,
                        time = unique),
    ndraws = 4000)

# Setting up contrasts for plotting
all_contrasts <- get_draws(pred.post) |> 
  dplyr::select(drawid, draw, participant, tx, time) |> 
  mutate(grp = paste0(tx, time)) |> 
  dplyr::select(-tx, -time) |> 
  pivot_wider(names_from = grp, values_from = draw) |> 
  mutate(
    # Within-group time deltas
    int_old_low_w3      = int_old_loww3  - int_old_lowpre,
    int_old_low_post     = int_old_lowpost - int_old_lowpre,
    int_old_low_w3post   = int_old_lowpost - int_old_loww3,
    
    int_old_mod_w3       = int_old_modw3  - int_old_modpre,
    int_old_mod_post      = int_old_modpost - int_old_modpre,
    int_old_mod_w3post    = int_old_modpost - int_old_modw3,
    
    int_yng_low_w3        = int_yng_loww3  - int_yng_lowpre,
    int_yng_low_post       = int_yng_lowpost - int_yng_lowpre,
    int_yng_low_w3post     = int_yng_lowpost - int_yng_loww3,
    
    int_yng_mod_w3         = int_yng_modw3  - int_yng_modpre,
    int_yng_mod_post        = int_yng_modpost - int_yng_modpre,
    int_yng_mod_w3post      = int_yng_modpost - int_yng_modw3,
    
    con_old_con_w3           = con_old_conw3  - con_old_conpre,
    con_old_con_post          = con_old_conpost - con_old_conpre,
    con_old_con_w3post        = con_old_conpost - con_old_conw3,
    
    # Pooled intervention deltas (NOW that the within-group deltas above exist)
    int_old_pooled_w3   = (int_old_low_w3 + int_old_mod_w3) / 2,
    int_old_pooled_post = (int_old_low_post + int_old_mod_post) / 2,
    int_yng_pooled_w3   = (int_yng_low_w3 + int_yng_mod_w3) / 2,
    int_yng_pooled_post = (int_yng_low_post + int_yng_mod_post) / 2,
    
    # Volume effect diffs (mod vs low), within age group
    old_mod_vs_low_w3      = int_old_mod_w3 - int_old_low_w3,
    old_mod_vs_low_post     = int_old_mod_post - int_old_low_post,
    old_mod_vs_low_w3post   = int_old_mod_w3post - int_old_low_w3post,
    
    yng_mod_vs_low_w3       = int_yng_mod_w3 - int_yng_low_w3,
    yng_mod_vs_low_post      = int_yng_mod_post - int_yng_low_post,
    yng_mod_vs_low_w3post    = int_yng_mod_w3post - int_yng_low_w3post,
    
    # Average group diffs: young vs old (averaging low/mod within age)
    yng_vs_old_w3         = ((int_yng_low_w3 + int_yng_mod_w3) / 2) - ((int_old_low_w3 + int_old_mod_w3) / 2),
    yng_vs_old_post         = ((int_yng_low_post + int_yng_mod_post) / 2) - ((int_old_low_post + int_old_mod_post) / 2),
    yng_vs_old_w3post        = ((int_yng_low_w3post + int_yng_mod_w3post) / 2) - ((int_old_low_w3post + int_old_mod_w3post) / 2),
    
    # Old intervention vs control
    old_vs_con_w3          = ((int_old_low_w3 + int_old_mod_w3) / 2) - con_old_con_w3,
    old_vs_con_post          = ((int_old_low_post + int_old_mod_post) / 2) - con_old_con_post,
    old_vs_con_w3post         = ((int_old_low_w3post + int_old_mod_w3post) / 2) - con_old_con_w3post,
    
    # Volume effect diffs between age groups
    yng_vs_old_vol_w3      = yng_mod_vs_low_w3 - old_mod_vs_low_w3,
    yng_vs_old_vol_post      = yng_mod_vs_low_post - old_mod_vs_low_post,
    yng_vs_old_vol_w3post     = yng_mod_vs_low_w3post - old_mod_vs_low_w3post
  ) |> 
  dplyr::select(drawid, 
                int_old_low_w3:int_old_low_w3post, int_old_mod_w3:int_old_mod_w3post,
                int_yng_low_w3:int_yng_low_w3post, int_yng_mod_w3:int_yng_mod_w3post,
                con_old_con_w3:con_old_con_w3post,
                int_old_pooled_w3, int_old_pooled_post,
                int_yng_pooled_w3, int_yng_pooled_post,
                old_mod_vs_low_w3:old_mod_vs_low_w3post, yng_mod_vs_low_w3:yng_mod_vs_low_w3post,
                yng_vs_old_w3:yng_vs_old_w3post, old_vs_con_w3:old_vs_con_w3post,
                yng_vs_old_vol_w3:yng_vs_old_vol_w3post) |> 
  pivot_longer(names_to = "contrast", values_to = "value", cols = -drawid) |> 
  mutate(
    delta = case_when(
      grepl("w3post$", contrast) ~ "w3 to post",
      grepl("w3$", contrast) ~ "pre to w3",
      grepl("post$", contrast) ~ "pre to post"
    ),
    contrast_group = case_when(
      grepl("^int_|^con_old_con", contrast) ~ "within-group",
      grepl("_mod_vs_low", contrast) ~ "volume effect",
      grepl("^yng_vs_old_vol", contrast) ~ "volume x age",
      grepl("^yng_vs_old_w3|^yng_vs_old_post", contrast) ~ "age effect",
      grepl("^old_vs_con", contrast) ~ "intervention vs control"
    )
  )

## Back-transformation for reporting
# Prior to this back-transformation, the values of ystd and mgstd are log-transformed, mean-centered and sd-standardized - this could indeed be reported as a diff in SD's, but its not very intuitive. Therefore, the below code back-transforms ystd to fold-change. 
# For within-group deltas (e.g. int_old_low_w3) this is a straightforward "RNA at w3 is fc_est x RNA at pre", e.g. 1.15 = 15% higher
# For between group/comparison contrasts (e.g., old_vs_con_w3) the value being exponentiated is already a difference of differences (a diff in log-fold-changes), so the resulting fold_change means "the intervention group's fold-change in is fc_est x the control group's fold-change" - a ratio of ratios

all_contrasts_fc <- all_contrasts |> 
  mutate(fold_change = exp(value * sd_rna))

saveRDS(all_contrasts_fc, "./data/data-gen/rna_mg_contrasts_fc.rds")

contrast_summary <- all_contrasts_fc |> 
  group_by(contrast, contrast_group, delta) |> 
  summarise(
    ystd_est   = median(value),
    ystd_lower = quantile(value, 0.025),
    ystd_upper = quantile(value, 0.975),
    fc_est     = median(fold_change),
    fc_lower   = quantile(fold_change, 0.025),
    fc_upper   = quantile(fold_change, 0.975),
    pos        = sum(fold_change > 1) / n(),   
    .groups = "drop"
  ) |> 
  arrange(contrast_group, contrast, delta)

saveRDS(contrast_summary, "data/data-gen/rna_mg_contrast_summary.rds")

print(contrast_summary, n = 30)




## Model 4

pred_rep <- predictions(
  m4.pp,
  type = "prediction",
  newdata = datagrid(tx = unique,
                     time = unique),
  ndraws = 1000)

prior_rep <- get_draws(pred_rep) |> 
  select(drawid, draw, participant, tx, time) |> 
  mutate(fit = "prior")

# Fit the posterior

m4.post <- update(m4.pp,
                  sample_prior = "no",
                  cores = 4,
                  control = list(adapt_delta = 0.95,
                                 max_treedepth = 15),
                  iter = 4000,
                  file = "fits/XX/m4.post",
                  file_refit = "on_change")
pp_check(m4.post, ndraws = 100)


pred.post_rep <- predictions(
    m4.post,
    type = "response",   
    newdata = datagrid(tx = unique,
                        time = unique),
    re_formula = NA,
    ndraws = 4000)


post_rep <- get_draws(pred.post_rep) |> 
  select(drawid, draw, participant, tx, time) |> 
  mutate(fit = "posterior")


bind_rows(prior_rep, post_rep) |> 
  ggplot(aes(draw, color = fit)) +
  geom_line(stat = "density") +
  facet_wrap(tx ~ time)


# Setting up contrasts for plotting

all_contrasts_rep <- get_draws(pred.post_rep) |> 
  dplyr::select(drawid, draw, participant, tx, time) |> 
  mutate(grp = paste0(tx, time)) |> 
  dplyr::select(-tx, -time) |> 
  pivot_wider(names_from = grp, values_from = draw) |> 
  mutate(
    # Within-group time deltas
    int_old_low_w3      = int_old_loww3  - int_old_lowpre,
    int_old_low_post     = int_old_lowpost - int_old_lowpre,
    int_old_low_w3post   = int_old_lowpost - int_old_loww3,
    
    int_old_mod_w3       = int_old_modw3  - int_old_modpre,
    int_old_mod_post      = int_old_modpost - int_old_modpre,
    int_old_mod_w3post    = int_old_modpost - int_old_modw3,
    
    int_yng_low_w3        = int_yng_loww3  - int_yng_lowpre,
    int_yng_low_post       = int_yng_lowpost - int_yng_lowpre,
    int_yng_low_w3post     = int_yng_lowpost - int_yng_loww3,
    
    int_yng_mod_w3         = int_yng_modw3  - int_yng_modpre,
    int_yng_mod_post        = int_yng_modpost - int_yng_modpre,
    int_yng_mod_w3post      = int_yng_modpost - int_yng_modw3,
    
    con_old_con_w3           = con_old_conw3  - con_old_conpre,
    con_old_con_post          = con_old_conpost - con_old_conpre,
    con_old_con_w3post        = con_old_conpost - con_old_conw3,
    
    # Pooled intervention deltas — goes here, after within-group deltas exist
    int_old_pooled_w3   = (int_old_low_w3 + int_old_mod_w3) / 2,
    int_old_pooled_post = (int_old_low_post + int_old_mod_post) / 2,
    int_yng_pooled_w3   = (int_yng_low_w3 + int_yng_mod_w3) / 2,
    int_yng_pooled_post = (int_yng_low_post + int_yng_mod_post) / 2,
    
    # Volume effect diffs (mod vs low), within age group
    old_mod_vs_low_w3      = int_old_mod_w3 - int_old_low_w3,
    old_mod_vs_low_post     = int_old_mod_post - int_old_low_post,
    old_mod_vs_low_w3post   = int_old_mod_w3post - int_old_low_w3post,
    
    yng_mod_vs_low_w3       = int_yng_mod_w3 - int_yng_low_w3,
    yng_mod_vs_low_post      = int_yng_mod_post - int_yng_low_post,
    yng_mod_vs_low_w3post    = int_yng_mod_w3post - int_yng_low_w3post,
    
    # Average group diffs: young vs old (averaging low/mod within age)
    yng_vs_old_w3         = ((int_yng_low_w3 + int_yng_mod_w3) / 2) - ((int_old_low_w3 + int_old_mod_w3) / 2),
    yng_vs_old_post         = ((int_yng_low_post + int_yng_mod_post) / 2) - ((int_old_low_post + int_old_mod_post) / 2),
    yng_vs_old_w3post        = ((int_yng_low_w3post + int_yng_mod_w3post) / 2) - ((int_old_low_w3post + int_old_mod_w3post) / 2),
    
    # Old intervention vs control
    old_vs_con_w3          = ((int_old_low_w3 + int_old_mod_w3) / 2) - con_old_con_w3,
    old_vs_con_post          = ((int_old_low_post + int_old_mod_post) / 2) - con_old_con_post,
    old_vs_con_w3post         = ((int_old_low_w3post + int_old_mod_w3post) / 2) - con_old_con_w3post,
    
    # Volume effect diffs between age groups
    yng_vs_old_vol_w3      = yng_mod_vs_low_w3 - old_mod_vs_low_w3,
    yng_vs_old_vol_post      = yng_mod_vs_low_post - old_mod_vs_low_post,
    yng_vs_old_vol_w3post     = yng_mod_vs_low_w3post - old_mod_vs_low_w3post
  ) |> 
  dplyr::select(drawid, 
                int_old_low_w3:int_old_low_w3post, int_old_mod_w3:int_old_mod_w3post,
                int_yng_low_w3:int_yng_low_w3post, int_yng_mod_w3:int_yng_mod_w3post,
                con_old_con_w3:con_old_con_w3post,
                int_old_pooled_w3, int_old_pooled_post,
                int_yng_pooled_w3, int_yng_pooled_post,
                old_mod_vs_low_w3:old_mod_vs_low_w3post, yng_mod_vs_low_w3:yng_mod_vs_low_w3post,
                yng_vs_old_w3:yng_vs_old_w3post, old_vs_con_w3:old_vs_con_w3post,
                yng_vs_old_vol_w3:yng_vs_old_vol_w3post) |> 
  pivot_longer(names_to = "contrast", values_to = "value", cols = -drawid) |> 
  mutate(
    delta = case_when(
      grepl("w3post$", contrast) ~ "w3 to post",
      grepl("w3$", contrast) ~ "pre to w3",
      grepl("post$", contrast) ~ "pre to post"
    ),
    contrast_group = case_when(
      grepl("^int_|^con_old_con", contrast) ~ "within-group",
      grepl("_mod_vs_low", contrast) ~ "volume effect",
      grepl("^yng_vs_old_vol", contrast) ~ "volume x age",
      grepl("^yng_vs_old_w3|^yng_vs_old_post", contrast) ~ "age effect",
      grepl("^old_vs_con", contrast) ~ "intervention vs control"
    )
  )


# Back-transformation

all_contrasts_fc_rep <- all_contrasts_rep |> 
  mutate(fold_change = exp(value * sd_rna_rep))


saveRDS(all_contrasts_fc_rep, "./data/data-gen/rna_mg_contrasts_fc_m4.rds")


contrast_summary_rep <- all_contrasts_fc_rep |> 
  group_by(contrast, contrast_group, delta) |> 
  summarise(
    ystd_est   = median(value),
    ystd_lower = quantile(value, 0.025),
    ystd_upper = quantile(value, 0.975),
    fc_est     = median(fold_change),
    fc_lower   = quantile(fold_change, 0.025),
    fc_upper   = quantile(fold_change, 0.975),
    pos        = sum(fold_change > 1) / n(),   
    .groups = "drop"
  ) |>
  arrange(contrast_group, contrast, delta)


saveRDS(contrast_summary_rep, "./data/data-gen/rna_mg_contrast_summary_m4.rds")

print(contrast_summary_rep, n = 34)


# The replicate-level random effect tightened the estimates without changing the substantive conclusions, which is what it was supposed to do. 

summary(m4.post)

# For plotting figures

pred_rna <- get_draws(pred.post_rep) |>
  dplyr::select(drawid, draw, tx, time) |>
  mutate(draw_bt = exp(draw * sd_rna_rep + mean_rna_rep)) |>
  summarise(.by = c(tx, time),
            m     = mean(draw_bt),
            lwr50 = quantile(draw_bt, 0.25), upr50 = quantile(draw_bt, 0.75),
            lwr90 = quantile(draw_bt, 0.05), upr90 = quantile(draw_bt, 0.95),
            lwr95 = quantile(draw_bt, 0.025), upr95 = quantile(draw_bt, 0.975))

saveRDS(pred_rna, "data/data-gen/pred_rna.rds")

contrast_rna <- contrast_summary_rep |>
  filter(delta %in% c("pre to post", "pre to w3")) |>
  mutate(hypothesis = case_when(
    str_remove(contrast, "_(w3|post)$") == "yng_vs_old"       ~ "yng_vs_old_delta",
    str_remove(contrast, "_(w3|post)$") == "old_vs_con"       ~ "old_vs_con_delta",
    str_remove(contrast, "_(w3|post)$") == "con_old_con" ~ "con_old_con_delta",
    str_remove(contrast, "_(w3|post)$") == "yng_mod_vs_low"   ~ "yng_mod_vs_low",
    str_remove(contrast, "_(w3|post)$") == "old_mod_vs_low"   ~ "old_mod_vs_low",
    str_remove(contrast, "_(w3|post)$") == "yng_vs_old_vol"   ~ "yng_vs_old_vol",
    str_remove(contrast, "_(w3|post)$") == "int_old_pooled"   ~ "int_old_pooled_delta",
    str_remove(contrast, "_(w3|post)$") == "int_yng_pooled"   ~ "int_yng_pooled_delta"
  )) |>
  filter(!is.na(hypothesis)) |>
  transmute(hypothesis, delta, mean = fc_est, lower95 = fc_lower, upper95 = fc_upper, pos)

saveRDS(contrast_rna, "data/data-gen/contrast_rna.rds")



```


#### Results

Older adults in the intervention group (averaged across volume conditions) increased RNA yield by 14.1% from baseline to post (95% CI [5.3%, 23.5%], P(>0) = 99.9%). The control group tended to increase by a similar amount (13.5%, 95% CI [−4.1%, 34.2%], P(>0) = 93.2%), and there was no evidence of a difference between them (0.5%, 95% CI [−16.7%, 20.3%], P(>0) = 52.4%). The two groups differed in timing: the intervention group had already increased by week 3 (9.2%, 95% CI [1.2%, 18.2%], P(>0) = 98.6%), whereas the control group had not changed at that point (−0.9%, 95% CI [−15.1%, 15.2%], P(<0) = 55.0%). The difference between them at week 3 was still uncertain (10.6%, 95% CI [−7.0%, 30.8%], P(>0) = 87.6%). Young adults (averaged across volume conditions) increased by 11.6% by post (95% CI [2.6%, 21.4%], P(>0) = 99.6%), with no evidence of a difference from older adults (−2.4%, 95% CI [−12.6%, 9.5%], P(<0) = 65.3%). At week 3, however, young adults had increased by 17.4% (95% CI [7.8%, 28.0%], P(>0) > 99.9%), with weak evidence of a larger increase than older adults (7.4%, 95% CI [−4.5%, 20.8%], P(>0) = 88.9%). Between week 3 and post, young adults' RNA yield likely fell back relative to older adults' (−9.0%, 95% CI [−20.4%, 4.5%], P(<0) = 91.9%). Within young adults, both low-volume (15.7%, 95% CI [3.5%, 29.5%], P(>0) = 99.5%) and moderate-volume training (19.2%, 95% CI [6.4%, 33.5%], P(>0) = 99.9%) increased RNA yield by week 3. At post, the increase was 12.2% under moderate volume (95% CI [0.4%, 25.9%], P(>0) = 97.8%), while low volume tended to increase it by 10.8% (95% CI [−0.7%, 24.1%], P(>0) = 96.6%). There was no evidence of a difference between volume conditions at either time point (week 3: 2.9%, 95% CI [−10.8%, 19.9%], P(>0) = 65.7%; post: 1.1%, 95% CI [−12.6%, 17.1%], P(>0) = 56.0%). In older adults, low-volume (12.3%, 95% CI [1.0%, 24.6%], P(>0) = 98.3%) and moderate-volume training (15.8%, 95% CI [4.4%, 28.3%], P(>0) = 99.8%) both increased RNA yield by post, having tended to increase already by week 3 (low: 10.3%, 95% CI [−0.3%, 22.0%], P(>0) = 97.1%; moderate: 8.3%, 95% CI [−2.3%, 19.9%], P(>0) = 93.3%). There was no evidence of a difference between volume conditions at either time point (week 3: −1.9%, 95% CI [−14.0%, 12.3%], P(<0) = 60.4%; post: 3.1%, 95% CI [−9.9%, 17.8%], P(>0) = 67.2%). There was no evidence that the volume effect differed between age groups, either at week 3 (5.0%, 95% CI [−14.6%, 27.5%], P(>0) = 68.4%) or at post (−1.7%, 95% CI [−20.0%, 20.0%], P(<0) = 57.2%).

<!--
Consider for discussion/methods: Accounting for technical replicates. To separate genuine biological/sample-level variability from technical noise arising from repeated concentration readings of the same tissue extract, a random effect for replicate (nested within participant) was added to the model, fit on the full replicate-level dataset (1552 observations) rather than pre-averaged values. This partitioning showed that approximately 78% of the variance previously absorbed into the residual term reflected genuine sample-to-sample biological variability, with the remaining ~22% attributable to technical replicate noise. Accounting for this structure sharpened estimates throughout without altering the underlying pattern of results — several contrasts that were moderate-to-strong in the summarised-replicate model became strong once technical noise was properly partitioned out.
-->

<!--
Summary. Once tissue mass was properly accounted for via direct modeling, and technical replicate variance was separated from genuine sample-level biological variability, RNA yield showed a robust, near-universal increase across the training period in essentially all groups. Age and intervention status did not alter the total magnitude of this increase, but did modulate its timing — younger and intervention participants tended to rise earlier, with older and control participants showing comparatively more of their increase later in the protocol. Training volume showed no credible influence at any stage.
-->



### Western - UBF and c-Myc


1.  Within condition (low- and moderate-volume, and non-training control conditions) change from baseline.
2.  Differences between age-groups at baseline.
3.  Average between age-group differences in change from baseline (young vs. old in the intervention group)
4.  Average between control- vs. intervention-group changes from baseline.
5.  Average effect of condition in the intervention groups (low- vs. moderate volume)
6.  Differences between age-groups in effect of condition.

The first contrast will be built for descriptive purposes and serves as the building block for comparisons in contrast 3-4.

### UBF

#### Data

In the code chunk below we plot the data grouped by participant leg.

```{r}

# Data

west.dat <- reliefdata::relief_western |> 
  mutate(participant = as.character(participant)) |> 
  select(participant, leg, sampleid = sample, time, capillary, peak,
         protein, position, mw_kda, height, area, area_p, corr.area,
         width, plate) |> 
  filter(!sampleid %in% c("Biot. Ladder", "DDH2O"),
         !participant %in% c("3084", "3090")) |>  #no measurement
  print()

condition <- reliefdata::relief_volume |> 
  mutate(participant = as.character(participant))

participants <- reliefdata::relief_participants |> 
  mutate(age_group = if_else(age < 40, "yng", "old")) |> 
  select(participant, age, sex, allocation, age_group)


ubf.dat <- west.dat |> 
  full_join(condition) |> 
  select(-arm) |> 
  full_join(participants) |> 
  filter(!is.na(sampleid), #filters out NA's introduced by joining, since not both legs are represented at pre
         protein == "UBF") |> 
  mutate(corr.area = as.numeric(corr.area)) |> 
  mutate(y = log(corr.area),
         ystd = (y - mean(y, na.rm = TRUE)) / sd(y, na.rm = TRUE),
         condition = if_else(allocation == "con", "con", condition),
         tx = paste0(allocation, "_", age_group, "_", condition),
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con")),
         tx_sex = paste0(tx, "_", sex),
         tx_sex = factor(tx_sex, levels = c(
           "int_old_low_female", "int_old_low_male",
           "int_old_mod_female", "int_old_mod_male", 
           "int_yng_low_female", "int_yng_low_male",
           "int_yng_mod_female", "int_yng_mod_male", 
           "con_old_con_female", "con_old_con_male"
         )),
         time = factor(time, levels = c("pre", "w3")),
         sex_centered = if_else(sex == "female", -0.5, 0.5)) |> 
  print()

 

## Save sd and mean for back transform
mean_ubf <- mean(ubf.dat$y, na.rm = TRUE)
sd_ubf   <- sd(ubf.dat$y, na.rm = TRUE)


# A preliminary figure of observed data

ubf.dat |> 
  # Change factor names for plotting
  mutate(time = factor(time, labels = c("Pre", "Week 3")), 
         age_group = factor(age_group, labels = c("Old", "Young")), 
         allocation = factor(allocation, labels = c("Control", "Intervention") ), 
         sex = factor(sex, labels = c("Female", "Male"))) |> 
  
  ggplot(aes(time, corr.area,
             color = sex,
             group = paste(participant, leg))) + 
  geom_line() + 
  facet_wrap(~ paste(age_group, allocation))  + 
  labs(x = "", y = "Corrected Area UBF") + 
  theme(axis.title.y = ggtext::element_markdown())




```

#### Models

```{r}
form0 <- bf(ystd ~   0 + tx + tx:time + sex + (1|participant))
form1 <- bf(ystd ~  time + allocation + allocation:time +  (1|participant) +  (1|participant:leg) + (time | tx_sex))
form2 <- bf(ystd ~   0 + tx_sex + tx_sex:time  + (1 + time|participant) +  (1|participant:leg))
form3 <- bf(ystd ~   time + allocation + allocation:time +  (1 + time|participant) +  (1|participant:leg) + (time | tx_sex))


# Priors
get_prior(form0, data = ubf.dat)
get_prior(form1, data = ubf.dat)
get_prior(form2, data = ubf.dat)
get_prior(form3, data = ubf.dat)

pri0 <- c(
  # Sex diffs
  prior(normal(0, 1), class = "b", coef = "sexmale"), 
  # Group-level intercepts - all parameters specifically referenced
  prior(normal(0, 1), class = "b", coef = "txcon_old_con"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_low"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_mod"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_low"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod"),
  
  # time-Post effects
  prior(normal(0, 1), class = "b", coef = "txcon_old_con:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_low:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_mod:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_low:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod:timew3"),
  
  # Random effect (participant)
  prior(exponential(1), class = sd), 
  
  prior(exponential(1), class = sigma)
)


pri1 <- c(
  
  # Intercept + global time slope
  prior(normal(0, 1), class = "Intercept"), 
  prior(normal(0, 1), class = "b"), 
  
  # Correlations of varying effects within groups
  #prior(lkj(2), class = "cor", group = "tx_sex"),
  
  # Random effect (participant)
  prior(exponential(1), class = sd), 
  
  prior(exponential(1), class = sigma)
  
)


pri2 <- c(
  
  # All fixed effects
  prior(normal(0, 1), class = "b"), 

  prior(lkj(2), class = "cor", group = "participant"),
  
  # Random effect (participant and tx_sex) specifically referenced 
  prior(exponential(1), class = sd, group = "participant"),
  prior(exponential(1), class = sd, group = "participant:leg"),
  # Sigma
  prior(exponential(1), class = sigma)
  
)


pri3 <- c(
  
  # Intercept + global time slope
  prior(normal(0, 1), class = "Intercept"), 
  prior(normal(0, 1), class = "b"), 
  
  # Correlations of varying effects within groups
 # prior(lkj(2), class = "cor", group = "tx_sex"),
  prior(lkj(2), class = "cor", group = "participant"),
  
  # Random effect (participant)
  prior(exponential(1), class = sd), 
  
  prior(exponential(1), class = sigma)
  
)




mu0.pp <- brm(form0, 
             prior = pri0, 
             data = ubf.dat, 
             file = "fits/05-molecular/mu0",
             file_refit = "on_change",
             sample_prior = "only")

mu1.pp <- brm(form1, 
             prior = pri1, 
             data = ubf.dat, 
             file = "fits/05-molecular/mu1",
             file_refit = "on_change",
             sample_prior = "only")

mu2.pp <- brm(form2, 
             prior = pri2, 
             data = ubf.dat, 
             file = "fits/05-molecular/mu2",
             file_refit = "on_change",
             sample_prior = "only")

mu3.pp <- brm(form3, 
             prior = pri3, 
             data = ubf.dat, 
             file = "fits/05-molecular/mu3",
             file_refit = "on_change",
             sample_prior = "only")

# Set up predictions/contrasts for model 0

#predictions(mu0.pp, 
#            newdata = datagrid(tx = unique, 
#                               time = unique),
#            )

mu0_pred_contr <- function(m, nd1, nd2) {
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified with the datagrid
    newdata = datagrid(tx = unique, 
                       time = unique), 
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
    hypothesis = c(
              "int_old_low_delta" = "b2 - b1 = 0", 
               "int_old_mod_delta" = "b4 - b3 = 0", 
               "int_yng_low_delta" = "b6 - b5 = 0", 
               "int_yng_mod_delta" = "b8 - b7 = 0", 
               "con_old_con_delta" = "b10 - b9 = 0", 
              # Average group diffs
              "yng_vs_old_delta" = "(((b8 - b7) + (b6 - b5)) / 2)  - (((b2 - b1) + (b4 - b3))/2) = 0", 
              "old_vs_con_delta" = "((b2 - b1) + (b4 - b3))/2  - (b10 - b9) = 0",
              # Volume effects diffs
              "yng_mod_vs_low" = "(b8 - b7) - (b6 - b5) = 0",
              "old_mod_vs_low" = "(b4 - b3) - (b2 - b1) = 0", 
              # volume effect diffs between age groups
              "yng_vs_old_vol" = "((b8 - b7) - (b6 - b5)) - ((b4 - b3) - (b2 - b1)) = 0"
                        ))
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = datagrid(tx = unique, 
                               time = unique, 
                               sex = unique), 
      re_formula = NA)
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )

}
  




mu1_pred_contr <- function(m, nd1, nd2) {
  
  # We need a custom grid to avoid counterfactual tx_sex/allocation 
  # groupings
  grid <- ubf.dat |> 
    distinct(tx_sex, allocation) |> 
    expand_grid(time = unique(ubf.dat$time))
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified from the above grid
    newdata = grid, 
    
    re_formula = ~ (time | tx_sex),
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
    by = c("tx_sex", "time"), 
    hypothesis = c(
       "int_old_low_delta" = "((b2-b1)+(b4-b3))/2 = 0", 
       "int_old_mod_delta" = "((b6-b5)+(b8-b7))/2 = 0", 
       "int_yng_low_delta" = "((b10-b9)+(b12-b11))/2 = 0", 
       "int_yng_mod_delta" = "((b14-b13)+(b16-b15))/2  = 0", 
       "con_old_con_delta" = "((b18-b17)+(b20-b19))/2 = 0", 
      # Average group diffs
      "yng_vs_old_delta" = "(((b14-b13)+(b16-b15))/2 + ((b10-b9)+(b12-b11))/2)/2 - (((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2)/2 = 0 ",
      "old_vs_con_delta" = "((((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2) / 2)  - (((b18-b17)+(b20-b19))/2) = 0",
      # Volume effects diffs
      "yng_mod_vs_low" = "((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2 = 0",
      "old_mod_vs_low" = "((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2 = 0", 
      # volume effect diffs between age groups
      "yng_vs_old_vol" = "(((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2) - (((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2) = 0"
      )
    )
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = grid, 
      re_formula = ~ (time | tx_sex))
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )

}

mu2_pred_contr <- function(m, nd1, nd2) {
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified with the datagrid
    newdata = datagrid(tx_sex = unique, 
                       time = unique), 
    re_formula = NA,
    
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
      hypothesis = c(
              "int_old_low_delta" = "((b2-b1)+(b4-b3))/2 = 0", 
              "int_old_mod_delta" = "((b6-b5)+(b8-b7))/2 = 0", 
              "int_yng_low_delta" = "((b10-b9)+(b12-b11))/2 = 0", 
              "int_yng_mod_delta" = "((b14-b13)+(b16-b15))/2  = 0", 
              "con_old_con_delta" = "((b18-b17)+(b20-b19))/2 = 0", 
              "int_old_pooled_delta" = "(((b2-b1)+(b4-b3))/2 + ((b6-b5)+(b8-b7))/2)/2 = 0",
              "int_yng_pooled_delta" = "(((b10-b9)+(b12-b11))/2 + ((b14-b13)+(b16-b15))/2)/2 = 0",
              
               # Average group diffs
              "yng_vs_old_delta" = "(((b14-b13)+(b16-b15))/2 + ((b10-b9)+(b12-b11))/2)/2 - (((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2)/2 = 0 ",
              "old_vs_con_delta" = "((((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2) / 2)  - (((b18-b17)+(b20-b19))/2) = 0",
              # Volume effects diffs
              "yng_mod_vs_low" = "((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2 = 0",
              "old_mod_vs_low" = "((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2 = 0", 
              # volume effect diffs between age groups
              "yng_vs_old_vol" = "(((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2) - (((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2) = 0"
              
            ))
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = datagrid(tx_sex = unique, 
                               time = unique), 
        re_formula = NA)
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )

}

mu3_pred_contr <- function(m, nd1, nd2) {
  
  # We need a custom grid to avoid counterfactual tx_sex/allocation 
  # groupings
  grid <- ubf.dat |> 
    distinct(tx_sex, allocation) |> 
    expand_grid(time = unique(ubf.dat$time))
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified from the above grid
    newdata = grid, 
    
    re_formula = ~ (time | tx_sex),
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
    by = c("tx_sex", "time"), 
    hypothesis = c(
       "int_old_low_delta" = "((b2-b1)+(b4-b3))/2 = 0", 
       "int_old_mod_delta" = "((b6-b5)+(b8-b7))/2 = 0", 
       "int_yng_low_delta" = "((b10-b9)+(b12-b11))/2 = 0", 
       "int_yng_mod_delta" = "((b14-b13)+(b16-b15))/2  = 0", 
       "con_old_con_delta" = "((b18-b17)+(b20-b19))/2 = 0", 
      # Average group diffs
      "yng_vs_old_delta" = "(((b14-b13)+(b16-b15))/2 + ((b10-b9)+(b12-b11))/2)/2 - (((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2)/2 = 0 ",
      "old_vs_con_delta" = "((((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2) / 2)  - (((b18-b17)+(b20-b19))/2) = 0",
      # Volume effects diffs
      "yng_mod_vs_low" = "((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2 = 0",
      "old_mod_vs_low" = "((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2 = 0", 
      # volume effect diffs between age groups
      "yng_vs_old_vol" = "(((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2) - (((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2) = 0"
      )
    )
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = grid, 
      re_formula = ~ (time | tx_sex))
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )
}



  
mu0_preds <- mu0_pred_contr(mu0.pp, nd1 = 1000, nd2 = 50)
mu1_preds <- mu1_pred_contr(mu1.pp, nd1 = 1000, nd2 = 50)
mu2_preds <- mu2_pred_contr(mu2.pp, nd1 = 1000, nd2 = 50)
mu3_preds <- mu3_pred_contr(mu3.pp, nd1 = 1000, nd2 = 50)


# Prior predictive simulations

p1 <- bind_rows(
  posterior_draws(mu0_preds$pred) |> 
  data.frame() |> 
    mutate(model = "mu0"), 
  
  posterior_draws(mu1_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |>
    mutate(model = "mu1"), 
  
  posterior_draws(mu2_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
    mutate(model = "mu2"), 
  
  posterior_draws(mu3_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |>
    mutate(model = "mu3")
    ) |> 
 
  ggplot(aes(time, draw, group = paste(drawid, tx, sex), 
             color = sex)) + geom_line() + 
  facet_grid(model ~ tx)






p2 <- bind_rows(
  posterior_draws(mu0_preds$contr) |> 
  mutate(model = "mu0"), 
    posterior_draws(mu1_preds$contr) |> 
  mutate(model = "mu1"), 
           posterior_draws(mu2_preds$contr) |> 
  mutate(model = "mu2"), 
   posterior_draws(mu3_preds$contr) |> 
  mutate(model = "mu3")) |>  
  
  ggplot(aes(draw, color = model)) + geom_line(stat = "density") + 
  facet_wrap(~ hypothesis) +
  geom_vline(xintercept = 0, lty = 2) + 
  coord_cartesian(xlim = c(-5, 5))



```

```{r}
### Get the posterior distribution

mu0.post <- update(mu0.pp, 
                  sample_prior = "no", 
                  iter = 4000,
                  file = "fits/05-molecular/mu0.post",
                  file_refit = "on_change")

# rhat checks out, 0 above 1.01, highest at 1.003 though

# neff ratio checks out, min ESS at 1347, with zero below 400


mu1.post <- update(mu1.pp, 
                  sample_prior = "no", 
                  cores = 4,
                  control = list(adapt_delta = 0.95, 
                                 max_treedepth = 15),
                  iter = 4000,
                  file = "fits/05-molecular/mu1.post",
                  file_refit = "on_change")

# rhat checks out, 0 above 1.01, highest at 1.003

# checks out barely, min ESS at 1083, with zero below 400


mu2.post <- update(mu2.pp, 
                  sample_prior = "no", 
                  cores = 4,
                   control = list(adapt_delta = 0.99, 
                                 max_treedepth = 15),
                  iter = 10000,
                  file = "fits/05-molecular/mu2.post",
                  file_refit = "on_change")

# Warning messages:
# 1: There were 6 divergent transitions after warmup. See
# https://mc-stan.org/misc/warnings.html#divergent-transitions-after-warmup
# to find out why this is a problem and how to eliminate them. 
# 2: Examine the pairs() plot to diagnose sampling problems

# Final settings: adapt_delta = 0.99, iter = 10000
# rhat checks out, 0 above 1.01, highest at 1.006
# checks out barely, min ESS at 597, with zero below 400



mu3.post <- update(mu3.pp, 
                  sample_prior = "no", 
                  cores = 4,
                   control = list(adapt_delta = 0.99, 
                                 max_treedepth = 15),
                  iter = 10000,
                  file = "fits/05-molecular/mu3.post",
                  file_refit = "on_change")

# Bumping up to 10k, changing adapt_delta to 0.99 due to low ESS and increased divergences

# Warning messages:
# 1: There were 3 divergent transitions after warmup. See
# https://mc-stan.org/misc/warnings.html#divergent-transitions-after-warmup
# to find out why this is a problem and how to eliminate them. 
# 2: There were 1 chains where the estimated Bayesian Fraction of Missing Information was low. See
# https://mc-stan.org/misc/warnings.html#bfmi-low 
# 3: Examine the pairs() plot to diagnose sampling problems

#rhat(mu3.post) 
#rhat_vals <- brms::rhat(mu3.post)
#max(rhat_vals)
#rhat_vals[rhat_vals > 1.01] 
# rhat checks out, 0 above 1.01, highest at 1.004

#neff_ratio(mu3.post)

#total_draws <- brms::ndraws(mu3.post)
#total_draws

#neff_vals <- brms::neff_ratio(mu3.post)
#ess_vals <- neff_vals * total_draws

#min(ess_vals)                  
#ess_vals[ess_vals < 400]       
# checks out barely, min ESS at 583, with zero below 400

## Posterior predictive checks



pp_check(mu3.post, type = "dens_overlay_grouped", 
         group = "tx_sex")

pp_check(mu2.post)

# These looks like the models might need some work/more data




```

```{r}
# Group comparisons

mu0_preds <- mu0_pred_contr(mu0.post, nd1 = 4000, nd2 = 50)
mu1_preds <- mu1_pred_contr(mu1.post, nd1 = 4000, nd2 = 50)
mu2_preds <- mu2_pred_contr(mu2.post, nd1 = 4000, nd2 = 50)
mu3_preds <- mu3_pred_contr(mu3.post, nd1 = 4000, nd2 = 50)


p1 <- bind_rows(
  posterior_draws(mu0_preds$pred) |> 
  data.frame() |> 
    mutate(model = "mu0"), 
  
  posterior_draws(mu1_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |>
    mutate(model = "mu1"), 
  
  posterior_draws(mu2_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
    mutate(model = "mu2"), 
  
  posterior_draws(mu3_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
    mutate(model = "mu3")
  ) |> 
 
  ggplot(aes(time, draw, group = paste(drawid, tx, sex), 
             color = sex)) + geom_line() + 
  facet_grid(model ~ tx)





p2 <- bind_rows(
  posterior_draws(mu0_preds$contr) |> 
  mutate(model = "mu0"), 
    posterior_draws(mu1_preds$contr) |> 
  mutate(model = "mu1"), 
           posterior_draws(mu2_preds$contr) |> 
  mutate(model = "mu2"), 
  posterior_draws(mu3_preds$contr) |> 
  mutate(model = "mu3")) |>  
  
  ggplot(aes(draw, color = model)) + geom_line(stat = "density") + 
  facet_wrap(~ hypothesis) +
  geom_vline(xintercept = 0, lty = 2)



############### Predictions for plotting #####################



p1 <- posterior_draws(mu2_preds$pred) |> 
    mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
  
  #mutate(draw = draw * sd_corr + mean_corr, 
         #draw = exp(draw)) |> 
  
  
  summarise(.by = c(tx, sex, time), 
            m = mean(draw), 
            lwr50 = quantile(draw, 0.25), 
            upr50 = quantile(draw, 0.75),  
            lwr90 = quantile(draw, 0.05), 
            upr90 = quantile(draw, 0.95),
            lwr95 = quantile(draw, 0.025), 
            upr95 = quantile(draw, 0.975)) |> 
  
  
  ggplot(aes(time, m, color = sex, group = paste(tx, sex))) + 
  

  
  geom_segment(aes(y = c(lwr50), yend = c(upr50), 
                   x = time, xend = time), 
               
              linewidth = 2) +
    geom_segment(aes(y = c(lwr90), yend = c(upr90), 
                   x = time, xend = time), 
             
              linewidth = 1.5) +
    geom_segment(aes(y = c(lwr95), yend = c(upr95), 
                   x = time, xend = time), 
          
              linewidth = 1) +

  geom_line() + 
  
    geom_point(size = 3, shape = 21, 
             aes(color = NULL, fill = sex)) +
  
  facet_wrap(~ tx, ncol = 5)



p2 <- posterior_draws(mu3_preds$contr) |> 
 
  
  #mutate(draw = draw * sd_corr , 
  #       draw = exp(draw)) |> 
  
  ggplot(aes(draw)) + geom_line(stat = "density") + 
  facet_wrap(~ hypothesis) +
  geom_vline(xintercept = 1, lty = 2) 





## Posterior probability
# The code below produces a table with hypothesis and posterior probability of direction, the proportion of posterior draws where the effect exceeds 1. A pos value of thus 1 means 100% of posterior draws showed the ratio exceeding 1 for that contrast, very strong evidence of a positive effect. Values at 0.5 on the other hand, are essentially a coin flip, no credible directional effect. Useful as a supplementary statistic, perhaps?


## Here I added mean, lower and upper to the summarise to get the mean ratio, as well as 95% ETI in addition to pos. 
 
posterior_draws(mu2_preds$contr) |>
  mutate(draw = draw * sd_ubf,
         draw = exp(draw)) |>
  summarise(.by = hypothesis,
            mean = mean(draw),
            lower95 = quantile(draw, 0.025),
            upper95 = quantile(draw, 0.975),
            pos = sum(draw > 1) / n())


# Since these are ratios (from the exp() transform), the null value is 1, not 0 - i.e., its about whether the 95% ETI excludes 1, not zero. These mean/CI tables should be reported and intepreted as ratios (fold-changes), e.g. "mean ratio = 1.15, 95% ETI [1.08, 1.23] - interpretable as "a 15% increase", not as an absolute cm3 diff. 



# For figure plotting


pred_ubf <- posterior_draws(mu2_preds$pred) |>
  mutate(
    sex = if_else(grepl("_female", tx_sex), "female", "male"),
    tx = gsub("_female", "", tx_sex),
    tx = gsub("_male", "", tx),
    tx = factor(tx, levels = c("int_old_low", "int_old_mod", "int_yng_low", "int_yng_mod", "con_old_con"))
  ) |>
  mutate(draw = draw * sd_ubf + mean_ubf,
         draw = exp(draw)) |>
  summarise(.by = c(tx, sex, time),
            m = mean(draw),
            lwr50 = quantile(draw, 0.25),
            upr50 = quantile(draw, 0.75),
            lwr90 = quantile(draw, 0.05),
            upr90 = quantile(draw, 0.95),
            lwr95 = quantile(draw, 0.025),
            upr95 = quantile(draw, 0.975))

saveRDS(pred_ubf, "data/data-gen/pred_ubf.rds")


contrast_ubf <- posterior_draws(mu2_preds$contr) |>
  mutate(draw = draw * sd_ubf, draw = exp(draw)) |>
  summarise(.by = hypothesis,
            mean = mean(draw),
            lower95 = quantile(draw, 0.025),
            upper95 = quantile(draw, 0.975),
            pos = sum(draw > 1) / n()) |>
  mutate(pct_change = (mean - 1) * 100,
         pct_lower  = (lower95 - 1) * 100,
         pct_upper  = (upper95 - 1) * 100,
         credible   = ifelse(lower95 > 1 | upper95 < 1, "Yes", "No"))

saveRDS(contrast_ubf, "data/data-gen/contrast_ubf.rds")




```


#### Results UBF

Older adults in the intervention group (averaged across volume conditions) tended to increase UBF protein content by week 3 (19.1%, 95% CI [−7.0%, 49.0%], P(>0) = 90.1%), as did the control group (24.9%, 95% CI [−15.7%, 64.3%], P(>0) = 90.8%), with no evidence of a difference between them (−2.6%, 95% CI [−33.7%, 38.1%], P(<0) = 58.6%). There was weak evidence of an increase in young adults (averaged across volume conditions; 15.7%, 95% CI [−8.0%, 44.2%], P(>0) = 88.5%), with no evidence of a difference from older adults (−1.5%, 95% CI [−29.0%, 33.8%], P(<0) = 56.8%). Within young adults, there was weak evidence of an increase with both low-volume (15.2%, 95% CI [−18.7%, 52.5%], P(>0) = 80.8%) and moderate-volume training (17.6%, 95% CI [−15.2%, 60.7%], P(>0) = 83.2%), with no evidence of a difference between them (4.4%, 95% CI [−34.0%, 54.8%], P(>0) = 53.9%). In older adults, there was weak evidence of an increase with low-volume training (27.1%, 95% CI [−13.3%, 74.7%], P(>0) = 88.9%) and no evidence with moderate-volume training (13.1%, 95% CI [−18.8%, 55.0%], P(>0) = 74.9%), again with no evidence of a difference between them (−8.2%, 95% CI [−44.1%, 41.9%], P(<0) = 67.8%). There was no evidence that the volume effect differed between age groups (20.1%, 95% CI [−39.8%, 112.3%], P(>0) = 67.2%).



### c-Myc

#### Data

In the code chunk below we plot the data grouped by participant leg.

```{r}

# Data

west.dat <- reliefdata::relief_western |> 
  mutate(participant = as.character(participant)) |> 
  select(participant, leg, sampleid = sample, time, capillary, peak,
         protein, position, mw_kda, height, area, area_p, corr.area,
         width, plate) |> 
  filter(!sampleid %in% c("Biot. Ladder", "DDH2O"),
         !participant %in% c("3084", "3090")) |>  #no measurement
  print()

condition <- reliefdata::relief_volume |> 
  mutate(participant = as.character(participant))

participants <- reliefdata::relief_participants |> 
  mutate(age_group = if_else(age < 40, "yng", "old")) |> 
  select(participant, age, sex, allocation, age_group)

# Both cMyc peaks are noted in the runs: cMyc_low = ~49 kDa, cMyc_high = ~60 kDa.
# Decision: both peaks should be analysedm but as separate outcomes (not combined/summed),
# since they likely represent distinct isoforms/phosphorylation states rather than
# duplicate measurements of the same signal.

# Shared cleaning/join logic as a function, applied separately per peak below
build_myc_dat <- function(protein_filter) {
  west.dat |> 
    full_join(condition) |> 
    select(-arm) |> 
    full_join(participants) |> 
    filter(!is.na(sampleid), #filters out NA's introduced by joining, since not both legs are represented at pre
           protein == protein_filter) |> 
    mutate(corr.area = as.numeric(corr.area)) |> 
    mutate(y = log(corr.area),
           ystd = (y - mean(y, na.rm = TRUE)) / sd(y, na.rm = TRUE),
           condition = if_else(allocation == "con", "con", condition),
           tx = paste0(allocation, "_", age_group, "_", condition),
           tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con")),
           tx_sex = paste0(tx, "_", sex),
           tx_sex = factor(tx_sex, levels = c(
             "int_old_low_female", "int_old_low_male",
             "int_old_mod_female", "int_old_mod_male", 
             "int_yng_low_female", "int_yng_low_male",
             "int_yng_mod_female", "int_yng_mod_male", 
             "con_old_con_female", "con_old_con_male"
           )),
           time = factor(time, levels = c("pre", "w3")),
           sex_centered = if_else(sex == "female", -0.5, 0.5)) |> 
    print()
}

myc_high.dat <- build_myc_dat("cMyc_high")
myc_low.dat  <- build_myc_dat("cMyc_low")

## Save sd and mean for back transform, separately per peak
mean_cmyc_high <- mean(myc_high.dat$y, na.rm = TRUE)
sd_cmyc_high   <- sd(myc_high.dat$y, na.rm = TRUE)

mean_cmyc_low <- mean(myc_low.dat$y, na.rm = TRUE)
sd_cmyc_low   <- sd(myc_low.dat$y, na.rm = TRUE)


# A preliminary figure of observed data, per peak

myc_high.dat |> 
  mutate(time = factor(time, labels = c("Pre", "Week 3")), 
         age_group = factor(age_group, labels = c("Old", "Young")), 
         allocation = factor(allocation, labels = c("Control", "Intervention")), 
         sex = factor(sex, labels = c("Female", "Male"))) |> 
  ggplot(aes(time, corr.area, color = sex, group = paste(participant, leg))) + 
  geom_line() + 
  facet_wrap(~ paste(age_group, allocation)) + 
  labs(x = "", y = "Corrected Area c-Myc (high, ~60 kDa)") + 
  theme(axis.title.y = ggtext::element_markdown())

myc_low.dat |> 
  mutate(time = factor(time, labels = c("Pre", "Week 3")), 
         age_group = factor(age_group, labels = c("Old", "Young")), 
         allocation = factor(allocation, labels = c("Control", "Intervention")), 
         sex = factor(sex, labels = c("Female", "Male"))) |> 
  ggplot(aes(time, corr.area, color = sex, group = paste(participant, leg))) + 
  geom_line() + 
  facet_wrap(~ paste(age_group, allocation)) + 
  labs(x = "", y = "Corrected Area c-Myc (low, ~49 kDa)") + 
  theme(axis.title.y = ggtext::element_markdown())



```

### c-Myc High Peak

#### Models

```{r}
form0 <- bf(ystd ~   0 + tx + tx:time + sex + (1|participant))
form1 <- bf(ystd ~  time + allocation + allocation:time +  (1|participant) +  (1|participant:leg) + (time | tx_sex))
form2 <- bf(ystd ~   0 + tx_sex + tx_sex:time  + (1 + time|participant) +  (1|participant:leg))
form3 <- bf(ystd ~   time + allocation + allocation:time +  (1 + time|participant) +  (1|participant:leg) + (time | tx_sex))


# Priors cmyc high
get_prior(form0, data = myc_high.dat)
get_prior(form1, data = myc_high.dat)
get_prior(form2, data = myc_high.dat)
get_prior(form3, data = myc_high.dat)

pri0 <- c(
  # Sex diffs
  prior(normal(0, 1), class = "b", coef = "sexmale"), 
  # Group-level intercepts - all parameters specifically referenced
  prior(normal(0, 1), class = "b", coef = "txcon_old_con"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_low"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_mod"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_low"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod"),
  
  # time-Post effects
  prior(normal(0, 1), class = "b", coef = "txcon_old_con:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_low:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_mod:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_low:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod:timew3"),
  
  # Random effect (participant)
  prior(exponential(1), class = sd), 
  
  prior(exponential(1), class = sigma)
)


pri1 <- c(
  
  # Intercept + global time slope
  prior(normal(0, 1), class = "Intercept"), 
  prior(normal(0, 1), class = "b"), 
  
  # Correlations of varying effects within groups
  #prior(lkj(2), class = "cor", group = "tx_sex"),
  
  # Random effect (participant)
  prior(exponential(1), class = sd), 
  
  prior(exponential(1), class = sigma)
  
)


pri2 <- c(
  
  # All fixed effects
  prior(normal(0, 1), class = "b"), 

  prior(lkj(2), class = "cor", group = "participant"),
  
  # Random effect (participant and tx_sex) specifically referenced 
  prior(exponential(1), class = sd, group = "participant"),
  prior(exponential(1), class = sd, group = "participant:leg"),
  # Sigma
  prior(exponential(1), class = sigma)
  
)


pri3 <- c(
  
  # Intercept + global time slope
  prior(normal(0, 1), class = "Intercept"), 
  prior(normal(0, 1), class = "b"), 
  
  # Correlations of varying effects within groups
 # prior(lkj(2), class = "cor", group = "tx_sex"),
  prior(lkj(2), class = "cor", group = "participant"),
  
  # Random effect (participant)
  prior(exponential(1), class = sd), 
  
  prior(exponential(1), class = sigma)
  
)




mhi0.pp <- brm(form0, 
             prior = pri0, 
             data = myc_high.dat, 
             file = "fits/05-molecular/mhi0",
             file_refit = "on_change",
             sample_prior = "only")

mhi1.pp <- brm(form1, 
             prior = pri1, 
             data = myc_high.dat, 
             file = "fits/05-molecular/mhi1",
             file_refit = "on_change",
             sample_prior = "only")

mhi2.pp <- brm(form2, 
             prior = pri2, 
             data = myc_high.dat, 
             file = "fits/05-molecular/mhi2",
             file_refit = "on_change",
             sample_prior = "only")

mhi3.pp <- brm(form3, 
             prior = pri3, 
             data = myc_high.dat, 
             file = "fits/05-molecular/mhi3",
             file_refit = "on_change",
             sample_prior = "only")

# Set up predictions/contrasts for model 0

#predictions(mu0.pp, 
#            newdata = datagrid(tx = unique, 
#                               time = unique),
#            )

mhi0_pred_contr <- function(m, nd1, nd2) {
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified with the datagrid
    newdata = datagrid(tx = unique, 
                       time = unique), 
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
    hypothesis = c(
              "int_old_low_delta" = "b2 - b1 = 0", 
               "int_old_mod_delta" = "b4 - b3 = 0", 
               "int_yng_low_delta" = "b6 - b5 = 0", 
               "int_yng_mod_delta" = "b8 - b7 = 0", 
               "con_old_con_delta" = "b10 - b9 = 0", 
              # Average group diffs
              "yng_vs_old_delta" = "(((b8 - b7) + (b6 - b5)) / 2)  - (((b2 - b1) + (b4 - b3))/2) = 0", 
              "old_vs_con_delta" = "((b2 - b1) + (b4 - b3))/2  - (b10 - b9) = 0",
              # Volume effects diffs
              "yng_mod_vs_low" = "(b8 - b7) - (b6 - b5) = 0",
              "old_mod_vs_low" = "(b4 - b3) - (b2 - b1) = 0", 
              # volume effect diffs between age groups
              "yng_vs_old_vol" = "((b8 - b7) - (b6 - b5)) - ((b4 - b3) - (b2 - b1)) = 0"
                        ))
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = datagrid(tx = unique, 
                               time = unique, 
                               sex = unique), 
      re_formula = NA)
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )

}
  




mhi1_pred_contr <- function(m, nd1, nd2) {
  
  # We need a custom grid to avoid counterfactual tx_sex/allocation 
  # groupings
  grid <- myc_high.dat |> 
    distinct(tx_sex, allocation) |> 
    expand_grid(time = unique(myc_high.dat$time))
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified from the above grid
    newdata = grid, 
    
    re_formula = ~ (time | tx_sex),
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
    by = c("tx_sex", "time"), 
    hypothesis = c(
       "int_old_low_delta" = "((b2-b1)+(b4-b3))/2 = 0", 
       "int_old_mod_delta" = "((b6-b5)+(b8-b7))/2 = 0", 
       "int_yng_low_delta" = "((b10-b9)+(b12-b11))/2 = 0", 
       "int_yng_mod_delta" = "((b14-b13)+(b16-b15))/2  = 0", 
       "con_old_con_delta" = "((b18-b17)+(b20-b19))/2 = 0", 
      # Average group diffs
      "yng_vs_old_delta" = "(((b14-b13)+(b16-b15))/2 + ((b10-b9)+(b12-b11))/2)/2 - (((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2)/2 = 0 ",
      "old_vs_con_delta" = "((((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2) / 2)  - (((b18-b17)+(b20-b19))/2) = 0",
      # Volume effects diffs
      "yng_mod_vs_low" = "((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2 = 0",
      "old_mod_vs_low" = "((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2 = 0", 
      # volume effect diffs between age groups
      "yng_vs_old_vol" = "(((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2) - (((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2) = 0"
      )
    )
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = grid, 
      re_formula = ~ (time | tx_sex))
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )

}

mhi2_pred_contr <- function(m, nd1, nd2) {
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified with the datagrid
    newdata = datagrid(tx_sex = unique, 
                       time = unique), 
    re_formula = NA,
    
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
      hypothesis = c(
              "int_old_low_delta" = "((b2-b1)+(b4-b3))/2 = 0", 
              "int_old_mod_delta" = "((b6-b5)+(b8-b7))/2 = 0", 
              "int_yng_low_delta" = "((b10-b9)+(b12-b11))/2 = 0", 
              "int_yng_mod_delta" = "((b14-b13)+(b16-b15))/2  = 0", 
              "con_old_con_delta" = "((b18-b17)+(b20-b19))/2 = 0", 
              "int_old_pooled_delta" = "(((b2-b1)+(b4-b3))/2 + ((b6-b5)+(b8-b7))/2)/2 = 0",
              "int_yng_pooled_delta" = "(((b10-b9)+(b12-b11))/2 + ((b14-b13)+(b16-b15))/2)/2 = 0",
              
               # Average group diffs
              "yng_vs_old_delta" = "(((b14-b13)+(b16-b15))/2 + ((b10-b9)+(b12-b11))/2)/2 - (((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2)/2 = 0 ",
              "old_vs_con_delta" = "((((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2) / 2)  - (((b18-b17)+(b20-b19))/2) = 0",
              # Volume effects diffs
              "yng_mod_vs_low" = "((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2 = 0",
              "old_mod_vs_low" = "((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2 = 0", 
              # volume effect diffs between age groups
              "yng_vs_old_vol" = "(((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2) - (((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2) = 0"
              
            ))
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = datagrid(tx_sex = unique, 
                               time = unique), 
        re_formula = NA)
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )

}

mhi3_pred_contr <- function(m, nd1, nd2) {
  
  # We need a custom grid to avoid counterfactual tx_sex/allocation 
  # groupings
  grid <- myc_high.dat |> 
    distinct(tx_sex, allocation) |> 
    expand_grid(time = unique(myc_high.dat$time))
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified from the above grid
    newdata = grid, 
    
    re_formula = ~ (time | tx_sex),
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
    by = c("tx_sex", "time"), 
    hypothesis = c(
       "int_old_low_delta" = "((b2-b1)+(b4-b3))/2 = 0", 
       "int_old_mod_delta" = "((b6-b5)+(b8-b7))/2 = 0", 
       "int_yng_low_delta" = "((b10-b9)+(b12-b11))/2 = 0", 
       "int_yng_mod_delta" = "((b14-b13)+(b16-b15))/2  = 0", 
       "con_old_con_delta" = "((b18-b17)+(b20-b19))/2 = 0", 
      # Average group diffs
      "yng_vs_old_delta" = "(((b14-b13)+(b16-b15))/2 + ((b10-b9)+(b12-b11))/2)/2 - (((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2)/2 = 0 ",
      "old_vs_con_delta" = "((((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2) / 2)  - (((b18-b17)+(b20-b19))/2) = 0",
      # Volume effects diffs
      "yng_mod_vs_low" = "((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2 = 0",
      "old_mod_vs_low" = "((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2 = 0", 
      # volume effect diffs between age groups
      "yng_vs_old_vol" = "(((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2) - (((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2) = 0"
      )
    )
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = grid, 
      re_formula = ~ (time | tx_sex))
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )
}



  
mhi0_preds <- mhi0_pred_contr(mhi0.pp, nd1 = 1000, nd2 = 50)
mhi1_preds <- mhi1_pred_contr(mhi1.pp, nd1 = 1000, nd2 = 50)
mhi2_preds <- mhi2_pred_contr(mhi2.pp, nd1 = 1000, nd2 = 50)
mhi3_preds <- mhi3_pred_contr(mhi3.pp, nd1 = 1000, nd2 = 50)


# Prior predictive simulations

p1 <- bind_rows(
  posterior_draws(mhi0_preds$pred) |> 
  data.frame() |> 
    mutate(model = "mhi0"), 
  
  posterior_draws(mhi1_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |>
    mutate(model = "mhi1"), 
  
  posterior_draws(mhi2_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
    mutate(model = "mhi2"), 
  
  posterior_draws(mhi3_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |>
    mutate(model = "mhi3")
    ) |> 
 
  ggplot(aes(time, draw, group = paste(drawid, tx, sex), 
             color = sex)) + geom_line() + 
  facet_grid(model ~ tx)






p2 <- bind_rows(
  posterior_draws(mhi0_preds$contr) |> 
  mutate(model = "mhi0"), 
    posterior_draws(mhi1_preds$contr) |> 
  mutate(model = "mhi1"), 
           posterior_draws(mhi2_preds$contr) |> 
  mutate(model = "mhi2"), 
   posterior_draws(mhi3_preds$contr) |> 
  mutate(model = "mhi3")) |>  
  
  ggplot(aes(draw, color = model)) + geom_line(stat = "density") + 
  facet_wrap(~ hypothesis) +
  geom_vline(xintercept = 0, lty = 2) + 
  coord_cartesian(xlim = c(-5, 5))



```

```{r}
### Get the posterior distribution

mhi0.post <- update(mhi0.pp, 
                  sample_prior = "no", 
                  iter = 4000,
                  file = "fits/05-molecular/mhi0.post",
                  file_refit = "on_change")



mhi1.post <- update(mhi1.pp, 
                  sample_prior = "no", 
                  cores = 4,
                  control = list(adapt_delta = 0.95, 
                                 max_treedepth = 15),
                  iter = 4000,
                  file = "fits/05-molecular/mhi1.post",
                  file_refit = "on_change")



mhi2.post <- update(mhi2.pp, 
                  sample_prior = "no", 
                  cores = 4,
                   control = list(adapt_delta = 0.99, 
                                 max_treedepth = 15),
                  iter = 10000,
                  file = "fits/05-molecular/mhi2.post",
                  file_refit = "on_change")

# Starting this at 10k iterations and adapt_delta 0.99, since it was required for the UBF data

# rhat checks out, 0 above 1.01, highest at 1.001
# ESS cheks out, min ESS at 3305, with zero below 400



mhi3.post <- update(mhi3.pp, 
                  sample_prior = "no", 
                  cores = 4,
                   control = list(adapt_delta = 0.99, 
                                 max_treedepth = 15),
                  iter = 10000,
                  file = "fits/05-molecular/mhi3.post",
                  file_refit = "on_change")

# Starting this at 10k iterations and adapt_delta 0.99, since it was required for the UBF data
# rhat checks out, 0 above 1.01, highest at 1.001
# ESS checks out, min ESS at 3633, with zero below 400

#rhat(mhi3.post) 
#rhat_vals <- brms::rhat(mhi3.post)
#max(rhat_vals)
#rhat_vals[rhat_vals > 1.01] 

#neff_ratio(mhi3.post)

#total_draws <- brms::ndraws(mhi3.post)
#total_draws

#neff_vals <- brms::neff_ratio(mhi3.post)
#ess_vals <- neff_vals * total_draws

#min(ess_vals)                  
#ess_vals[ess_vals < 400]       



## Posterior predictive checks

pp_check(mhi3.post, type = "dens_overlay_grouped", 
         group = "tx_sex")

pp_check(mhi2.post)

# These looks like the models might need some work




```

```{r}
# Group comparisons

mhi0_preds <- mhi0_pred_contr(mhi0.post, nd1 = 4000, nd2 = 50)
mhi1_preds <- mhi1_pred_contr(mhi1.post, nd1 = 4000, nd2 = 50)
mhi2_preds <- mhi2_pred_contr(mhi2.post, nd1 = 4000, nd2 = 50)
mhi3_preds <- mhi3_pred_contr(mhi3.post, nd1 = 4000, nd2 = 50)


p1 <- bind_rows(
  posterior_draws(mhi0_preds$pred) |> 
  data.frame() |> 
    mutate(model = "mhi0"), 
  
  posterior_draws(mhi1_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |>
    mutate(model = "mhi1"), 
  
  posterior_draws(mhi2_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
    mutate(model = "mhi2"), 
  
  posterior_draws(mhi3_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
    mutate(model = "mhi3")
  ) |> 
 
  ggplot(aes(time, draw, group = paste(drawid, tx, sex), 
             color = sex)) + geom_line() + 
  facet_grid(model ~ tx)





p2 <- bind_rows(
  posterior_draws(mhi0_preds$contr) |> 
  mutate(model = "mhi0"), 
    posterior_draws(mhi1_preds$contr) |> 
  mutate(model = "mhi1"), 
           posterior_draws(mhi2_preds$contr) |> 
  mutate(model = "mhi2"), 
  posterior_draws(mhi3_preds$contr) |> 
  mutate(model = "mhi3")) |>  
  
  ggplot(aes(draw, color = model)) + geom_line(stat = "density") + 
  facet_wrap(~ hypothesis) +
  geom_vline(xintercept = 0, lty = 2)



############### Predictions for plotting #####################





# These are looking a bit nonsensical, so I guess something here needs more work, but I cant see what at 23:27 in the evening.

p1 <- posterior_draws(mhi2_preds$pred) |> 
    mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
  
  #mutate(draw = draw * sd_corr + mean_corr, 
         #draw = exp(draw)) |> 
  
  
  summarise(.by = c(tx, sex, time), 
            m = mean(draw), 
            lwr50 = quantile(draw, 0.25), 
            upr50 = quantile(draw, 0.75),  
            lwr90 = quantile(draw, 0.05), 
            upr90 = quantile(draw, 0.95),
            lwr95 = quantile(draw, 0.025), 
            upr95 = quantile(draw, 0.975)) |> 
  
  
  ggplot(aes(time, m, color = sex, group = paste(tx, sex))) + 
  

  
  geom_segment(aes(y = c(lwr50), yend = c(upr50), 
                   x = time, xend = time), 
               
              linewidth = 2) +
    geom_segment(aes(y = c(lwr90), yend = c(upr90), 
                   x = time, xend = time), 
             
              linewidth = 1.5) +
    geom_segment(aes(y = c(lwr95), yend = c(upr95), 
                   x = time, xend = time), 
          
              linewidth = 1) +

  geom_line() + 
  
    geom_point(size = 3, shape = 21, 
             aes(color = NULL, fill = sex)) +
  
  facet_wrap(~ tx, ncol = 5)



p2 <- posterior_draws(mhi3_preds$contr) |> 
 
  
  #mutate(draw = draw * sd_corr , 
  #       draw = exp(draw)) |> 
  
  ggplot(aes(draw)) + geom_line(stat = "density") + 
  facet_wrap(~ hypothesis) +
  geom_vline(xintercept = 1, lty = 2) 





## Posterior probability
# The code below produces a table with hypothesis and posterior probability of direction, the proportion of posterior draws where the effect exceeds 1. A pos value of thus 1 means 100% of posterior draws showed the ratio exceeding 1 for that contrast, very strong evidence of a positive effect. Values at 0.5 on the other hand, are essentially a coin flip, no credible directional effect. Useful as a supplementary statistic, perhaps?


## Here I added mean, lower and upper to the summarise to get the mean ratio, as well as 95% ETI in addition to pos. 
 
posterior_draws(mhi2_preds$contr) |>
  mutate(draw = draw * sd_cmyc_high,
         draw = exp(draw)) |>
  summarise(.by = hypothesis,
            mean = mean(draw),
            lower95 = quantile(draw, 0.025),
            upper95 = quantile(draw, 0.975),
            pos = sum(draw > 1) / n())


# Since these are ratios (from the exp() transform), the null value is 1, not 0 - i.e., its about whether the 95% ETI excludes 1, not zero. These mean/CI tables should be reported and intepreted as ratios (fold-changes), e.g. "mean ratio = 1.15, 95% ETI [1.08, 1.23] - interpretable as "a 15% increase", not as an absolute cm3 diff. 


# For figure plotting


pred_cmyc_high <- posterior_draws(mhi2_preds$pred) |>
  mutate(
    sex = if_else(grepl("_female", tx_sex), "female", "male"),
    tx = gsub("_female", "", tx_sex),
    tx = gsub("_male", "", tx),
    tx = factor(tx, levels = c("int_old_low", "int_old_mod", "int_yng_low", "int_yng_mod", "con_old_con"))
  ) |>
  mutate(draw = draw * sd_cmyc_high + mean_cmyc_high,
         draw = exp(draw)) |>
  summarise(.by = c(tx, sex, time),
            m = mean(draw),
            lwr50 = quantile(draw, 0.25),
            upr50 = quantile(draw, 0.75),
            lwr90 = quantile(draw, 0.05),
            upr90 = quantile(draw, 0.95),
            lwr95 = quantile(draw, 0.025),
            upr95 = quantile(draw, 0.975))

saveRDS(pred_cmyc_high, "data/data-gen/pred_cmyc_high.rds")


contrast_cmyc_high <- posterior_draws(mhi2_preds$contr) |>
  mutate(draw = draw * sd_cmyc_high, draw = exp(draw)) |>
  summarise(.by = hypothesis,
            mean = mean(draw),
            lower95 = quantile(draw, 0.025),
            upper95 = quantile(draw, 0.975),
            pos = sum(draw > 1) / n()) |>
  mutate(pct_change = (mean - 1) * 100,
         pct_lower  = (lower95 - 1) * 100,
         pct_upper  = (upper95 - 1) * 100,
         credible   = ifelse(lower95 > 1 | upper95 < 1, "Yes", "No"))

saveRDS(contrast_cmyc_high, "data/data-gen/contrast_cmyc_high.rds")



```

#### Results cMyc High Peak

The difference in c-Myc (high peak) protein content between the old intervention group (averaged across volume conditions) and the control group was large by week 3 (99.8%, 95% CI [17.5%, 216.3%], P(>0) = 99.3%). This difference was driven mainly by a decrease in the control group (−34.5%, 95% CI [−56.0%, −5.5%], P(<0) = 98.6%), while the intervention group tended to increase (25.9%, 95% CI [−8.9%, 68.1%], P(>0) = 92.0%). There was weak evidence of an increase in young adults (averaged across volume conditions; 18.6%, 95% CI [−12.5%, 57.5%], P(>0) = 85.9%), with no evidence of a difference from older adults (−3.5%, 95% CI [−37.8%, 45.0%], P(<0) = 61.5%). Within young adults, there was no evidence of a change with low-volume training (4.4%, 95% CI [−30.1%, 49.1%], P(>0) = 55.2%), while moderate-volume training tended to increase c-Myc (37.3%, 95% CI [−12.2%, 106.2%], P(>0) = 91.5%). There was weak evidence of a larger increase under moderate volume (36.2%, 95% CI [−23.8%, 127.4%], P(>0) = 83.5%). The same pattern held in older adults: there was no evidence of a change with low-volume training (9.6%, 95% CI [−29.9%, 64.4%], P(>0) = 62.0%), moderate-volume training tended to increase c-Myc (47.9%, 95% CI [−4.7%, 118.4%], P(>0) = 95.5%), and there was weak evidence of a larger increase under moderate volume (41.0%, 95% CI [−24.3%, 138.0%], P(>0) = 84.5%). There was no evidence that the volume effect differed between age groups (5.2%, 95% CI [−55.8%, 114.3%], P(>0) = 46.7%).











<!--

### c-Myc Low Peaks

Experiencing model-complexity issues here, specifically from mlo2.post and mlo3.post. It's possibly a reflection of this peaks signal-to-noise ratio, but Im not sure. 

#### Models

```{r}
form0 <- bf(ystd ~   0 + tx + tx:time + sex + (1|participant))
form1 <- bf(ystd ~  time + allocation + allocation:time +  (1|participant) +  (1|participant:leg) + (time | tx_sex))
form2 <- bf(ystd ~   0 + tx_sex + tx_sex:time  + (1 + time|participant) +  (1|participant:leg))
form3 <- bf(ystd ~   time + allocation + allocation:time +  (1 + time|participant) +  (1|participant:leg) + (time | tx_sex))


# Priors cmyc high
get_prior(form0, data = myc_low.dat)
get_prior(form1, data = myc_low.dat)
get_prior(form2, data = myc_low.dat)
get_prior(form3, data = myc_low.dat)

pri0 <- c(
  # Sex diffs
  prior(normal(0, 1), class = "b", coef = "sexmale"), 
  # Group-level intercepts - all parameters specifically referenced
  prior(normal(0, 1), class = "b", coef = "txcon_old_con"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_low"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_mod"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_low"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod"),
  
  # time-Post effects
  prior(normal(0, 1), class = "b", coef = "txcon_old_con:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_low:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_mod:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_low:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod:timew3"),
  
  # Random effect (participant)
  prior(exponential(1), class = sd), 
  
  prior(exponential(1), class = sigma)
)


pri1 <- c(
  
  # Intercept + global time slope
  prior(normal(0, 1), class = "Intercept"), 
  prior(normal(0, 1), class = "b"), 
  
  # Correlations of varying effects within groups
  #prior(lkj(2), class = "cor", group = "tx_sex"),
  
  # Random effect (participant)
  prior(exponential(1), class = sd), 
  
  prior(exponential(1), class = sigma)
  
)


pri2 <- c(
  
  # All fixed effects
  prior(normal(0, 1), class = "b"), 

  prior(lkj(2), class = "cor", group = "participant"),
  
  # Random effect (participant and tx_sex) specifically referenced 
  prior(exponential(1), class = sd, group = "participant"),
  prior(exponential(1), class = sd, group = "participant:leg"),
  # Sigma
  prior(exponential(1), class = sigma)
  
)


pri3 <- c(
  
  # Intercept + global time slope
  prior(normal(0, 1), class = "Intercept"), 
  prior(normal(0, 1), class = "b"), 
  
  # Correlations of varying effects within groups
 # prior(lkj(2), class = "cor", group = "tx_sex"),
  prior(lkj(2), class = "cor", group = "participant"),
  
  # Random effect (participant)
  prior(exponential(1), class = sd), 
  
  prior(exponential(1), class = sigma)
  
)




mlo0.pp <- brm(form0, 
             prior = pri0, 
             data = myc_low.dat, 
             file = "fits/05-molecular/mlo0",
             file_refit = "on_change",
             sample_prior = "only")

mlo1.pp <- brm(form1, 
             prior = pri1, 
             data = myc_low.dat, 
             file = "fits/05-molecular/mlo1",
             file_refit = "on_change",
             sample_prior = "only")

mlo2.pp <- brm(form2, 
             prior = pri2, 
             data = myc_low.dat, 
             file = "fits/05-molecular/mlo2",
             file_refit = "on_change",
             sample_prior = "only")

mlo3.pp <- brm(form3, 
             prior = pri3, 
             data = myc_low.dat, 
             file = "fits/05-molecular/mlo3",
             file_refit = "on_change",
             sample_prior = "only")

# Set up predictions/contrasts for model 0

#predictions(mu0.pp, 
#            newdata = datagrid(tx = unique, 
#                               time = unique),
#            )

mlo0_pred_contr <- function(m, nd1, nd2) {
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified with the datagrid
    newdata = datagrid(tx = unique, 
                       time = unique), 
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
    hypothesis = c(
              "int_old_low_delta" = "b2 - b1 = 0", 
               "int_old_mod_delta" = "b4 - b3 = 0", 
               "int_yng_low_delta" = "b6 - b5 = 0", 
               "int_yng_mod_delta" = "b8 - b7 = 0", 
               "con_old_con_delta" = "b10 - b9 = 0", 
              # Average group diffs
              "yng_vs_old_delta" = "(((b8 - b7) + (b6 - b5)) / 2)  - (((b2 - b1) + (b4 - b3))/2) = 0", 
              "old_vs_con_delta" = "((b2 - b1) + (b4 - b3))/2  - (b10 - b9) = 0",
              # Volume effects diffs
              "yng_mod_vs_low" = "(b8 - b7) - (b6 - b5) = 0",
              "old_mod_vs_low" = "(b4 - b3) - (b2 - b1) = 0", 
              # volume effect diffs between age groups
              "yng_vs_old_vol" = "((b8 - b7) - (b6 - b5)) - ((b4 - b3) - (b2 - b1)) = 0"
                        ))
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = datagrid(tx = unique, 
                               time = unique, 
                               sex = unique), 
      re_formula = NA)
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )

}
  




mlo1_pred_contr <- function(m, nd1, nd2) {
  
  # We need a custom grid to avoid counterfactual tx_sex/allocation 
  # groupings
  grid <- myc_low.dat |> 
    distinct(tx_sex, allocation) |> 
    expand_grid(time = unique(myc_low.dat$time))
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified from the above grid
    newdata = grid, 
    
    re_formula = ~ (time | tx_sex),
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
    by = c("tx_sex", "time"), 
    hypothesis = c(
       "int_old_low_delta" = "((b2-b1)+(b4-b3))/2 = 0", 
       "int_old_mod_delta" = "((b6-b5)+(b8-b7))/2 = 0", 
       "int_yng_low_delta" = "((b10-b9)+(b12-b11))/2 = 0", 
       "int_yng_mod_delta" = "((b14-b13)+(b16-b15))/2  = 0", 
       "con_old_con_delta" = "((b18-b17)+(b20-b19))/2 = 0", 
      # Average group diffs
      "yng_vs_old_delta" = "(((b14-b13)+(b16-b15))/2 + ((b10-b9)+(b12-b11))/2)/2 - (((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2)/2 = 0 ",
      "old_vs_con_delta" = "((((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2) / 2)  - (((b18-b17)+(b20-b19))/2) = 0",
      # Volume effects diffs
      "yng_mod_vs_low" = "((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2 = 0",
      "old_mod_vs_low" = "((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2 = 0", 
      # volume effect diffs between age groups
      "yng_vs_old_vol" = "(((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2) - (((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2) = 0"
      )
    )
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = grid, 
      re_formula = ~ (time | tx_sex))
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )

}

mlo2_pred_contr <- function(m, nd1, nd2) {
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified with the datagrid
    newdata = datagrid(tx_sex = unique, 
                       time = unique), 
    re_formula = NA,
    
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
      hypothesis = c(
              "int_old_low_delta" = "((b2-b1)+(b4-b3))/2 = 0", 
               "int_old_mod_delta" = "((b6-b5)+(b8-b7))/2 = 0", 
               "int_yng_low_delta" = "((b10-b9)+(b12-b11))/2 = 0", 
               "int_yng_mod_delta" = "((b14-b13)+(b16-b15))/2  = 0", 
               "con_old_con_delta" = "((b18-b17)+(b20-b19))/2 = 0", 
              
               # Average group diffs
              "yng_vs_old_delta" = "(((b14-b13)+(b16-b15))/2 + ((b10-b9)+(b12-b11))/2)/2 - (((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2)/2 = 0 ",
              "old_vs_con_delta" = "((((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2) / 2)  - (((b18-b17)+(b20-b19))/2) = 0",
              # Volume effects diffs
              "yng_mod_vs_low" = "((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2 = 0",
              "old_mod_vs_low" = "((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2 = 0", 
              # volume effect diffs between age groups
              "yng_vs_old_vol" = "(((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2) - (((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2) = 0"
              
            ))
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = datagrid(tx_sex = unique, 
                               time = unique), 
        re_formula = NA)
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )

}

mlo3_pred_contr <- function(m, nd1, nd2) {
  
  # We need a custom grid to avoid counterfactual tx_sex/allocation 
  # groupings
  grid <- myc_low.dat |> 
    distinct(tx_sex, allocation) |> 
    expand_grid(time = unique(myc_low.dat$time))
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified from the above grid
    newdata = grid, 
    
    re_formula = ~ (time | tx_sex),
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
    by = c("tx_sex", "time"), 
    hypothesis = c(
       "int_old_low_delta" = "((b2-b1)+(b4-b3))/2 = 0", 
       "int_old_mod_delta" = "((b6-b5)+(b8-b7))/2 = 0", 
       "int_yng_low_delta" = "((b10-b9)+(b12-b11))/2 = 0", 
       "int_yng_mod_delta" = "((b14-b13)+(b16-b15))/2  = 0", 
       "con_old_con_delta" = "((b18-b17)+(b20-b19))/2 = 0", 
      # Average group diffs
      "yng_vs_old_delta" = "(((b14-b13)+(b16-b15))/2 + ((b10-b9)+(b12-b11))/2)/2 - (((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2)/2 = 0 ",
      "old_vs_con_delta" = "((((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2) / 2)  - (((b18-b17)+(b20-b19))/2) = 0",
      # Volume effects diffs
      "yng_mod_vs_low" = "((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2 = 0",
      "old_mod_vs_low" = "((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2 = 0", 
      # volume effect diffs between age groups
      "yng_vs_old_vol" = "(((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2) - (((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2) = 0"
      )
    )
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = grid, 
      re_formula = ~ (time | tx_sex))
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )
}



  
mlo0_preds <- mlo0_pred_contr(mlo0.pp, nd1 = 1000, nd2 = 50)
mlo1_preds <- mlo1_pred_contr(mlo1.pp, nd1 = 1000, nd2 = 50)
mlo2_preds <- mlo2_pred_contr(mlo2.pp, nd1 = 1000, nd2 = 50)
mlo3_preds <- mlo3_pred_contr(mlo3.pp, nd1 = 1000, nd2 = 50)


# Prior predictive simulations

p1 <- bind_rows(
  posterior_draws(mlo0_preds$pred) |> 
  data.frame() |> 
    mutate(model = "mlo0"), 
  
  posterior_draws(mlo1_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |>
    mutate(model = "mlo1"), 
  
  posterior_draws(mlo2_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
    mutate(model = "mlo2"), 
  
  posterior_draws(mlo3_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |>
    mutate(model = "mlo3")
    ) |> 
 
  ggplot(aes(time, draw, group = paste(drawid, tx, sex), 
             color = sex)) + geom_line() + 
  facet_grid(model ~ tx)






p2 <- bind_rows(
  posterior_draws(mlo0_preds$contr) |> 
  mutate(model = "mlo0"), 
    posterior_draws(mlo1_preds$contr) |> 
  mutate(model = "mlo1"), 
           posterior_draws(mlo2_preds$contr) |> 
  mutate(model = "mlo2"), 
   posterior_draws(mlo3_preds$contr) |> 
  mutate(model = "mlo3")) |>  
  
  ggplot(aes(draw, color = model)) + geom_line(stat = "density") + 
  facet_wrap(~ hypothesis) +
  geom_vline(xintercept = 0, lty = 2) + 
  coord_cartesian(xlim = c(-5, 5))



```

```{r}
### Get the posterior distribution

mlo0.post <- update(mlo0.pp, 
                  sample_prior = "no", 
                  iter = 4000,
                  file = "fits/05-molecular/mlo0.post",
                  file_refit = "on_change")



mlo1.post <- update(mlo1.pp, 
                  sample_prior = "no", 
                  cores = 4,
                  control = list(adapt_delta = 0.95, 
                                 max_treedepth = 15),
                  iter = 4000,
                  file = "fits/05-molecular/mlo1.post",
                  file_refit = "on_change")



mlo2.post <- update(mlo2.pp, 
                  sample_prior = "no", 
                  cores = 4,
                   control = list(adapt_delta = 0.999, 
                                 max_treedepth = 15),
                  iter = 10000,
                  file = "fits/05-molecular/mlo2.post",
                  file_refit = "on_change")


# Upping from 0.99 to 0.999 adapt_delta due to low-BFMI warnings

mlo3.post <- update(mlo3.pp, 
                  sample_prior = "no", 
                  cores = 4,
                  control = list(adapt_delta = 0.99, 
                                max_treedepth = 15),
                  iter = 10000,
                 file = "fits/05-molecular/mlo3.post",
                 file_refit = "on_change")


#rhat(mhi3.post) 
#rhat_vals <- brms::rhat(mhi3.post)
#max(rhat_vals)
#rhat_vals[rhat_vals > 1.01] 
# rhat checks out, 0 above 1.01, highest at 1.004

#neff_ratio(mhi3.post)

#total_draws <- brms::ndraws(mhi3.post)
#total_draws

#neff_vals <- brms::neff_ratio(mhi3.post)
#ess_vals <- neff_vals * total_draws

#min(ess_vals)                  
#ess_vals[ess_vals < 400]       
# checks out barely, min ESS at 1128, with zero below 400


## Posterior predictive checks



pp_check(mlo2.post, type = "dens_overlay_grouped", 
         group = "tx_sex")

pp_check(mlo2.post)

# Not sure about these, model 2 looks a bit more divergent than the others ive seen, model 3 looks fine




```

```{r}
# Group comparisons

mhi0_preds <- mhi0_pred_contr(mhi0.post, nd1 = 4000, nd2 = 50)
mhi1_preds <- mhi1_pred_contr(mhi1.post, nd1 = 4000, nd2 = 50)
mhi2_preds <- mhi2_pred_contr(mhi2.post, nd1 = 4000, nd2 = 50)
mhi3_preds <- mhi3_pred_contr(mhi3.post, nd1 = 4000, nd2 = 50)


p1 <- bind_rows(
  posterior_draws(mhi0_preds$pred) |> 
  data.frame() |> 
    mutate(model = "mhi0"), 
  
  posterior_draws(mhi1_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |>
    mutate(model = "mhi1"), 
  
  posterior_draws(mhi2_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
    mutate(model = "mhi2"), 
  
  posterior_draws(mhi3_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
    mutate(model = "mhi3")
  ) |> 
 
  ggplot(aes(time, draw, group = paste(drawid, tx, sex), 
             color = sex)) + geom_line() + 
  facet_grid(model ~ tx)





p2 <- bind_rows(
  posterior_draws(mhi0_preds$contr) |> 
  mutate(model = "mhi0"), 
    posterior_draws(mhi1_preds$contr) |> 
  mutate(model = "mhi1"), 
           posterior_draws(mhi2_preds$contr) |> 
  mutate(model = "mhi2"), 
  posterior_draws(mhi3_preds$contr) |> 
  mutate(model = "mhi3")) |>  
  
  ggplot(aes(draw, color = model)) + geom_line(stat = "density") + 
  facet_wrap(~ hypothesis) +
  geom_vline(xintercept = 0, lty = 2)



############### Predictions for plotting #####################





# These are looking a bit nonsensical, so I guess something here needs more work, but I cant see what at 23:27 in the evening.

p1 <- posterior_draws(mhi2_preds$pred) |> 
    mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
  
  #mutate(draw = draw * sd_corr + mean_corr, 
         #draw = exp(draw)) |> 
  
  
  summarise(.by = c(tx, sex, time), 
            m = mean(draw), 
            lwr50 = quantile(draw, 0.25), 
            upr50 = quantile(draw, 0.75),  
            lwr90 = quantile(draw, 0.05), 
            upr90 = quantile(draw, 0.95),
            lwr95 = quantile(draw, 0.025), 
            upr95 = quantile(draw, 0.975)) |> 
  
  
  ggplot(aes(time, m, color = sex, group = paste(tx, sex))) + 
  

  
  geom_segment(aes(y = c(lwr50), yend = c(upr50), 
                   x = time, xend = time), 
               
              linewidth = 2) +
    geom_segment(aes(y = c(lwr90), yend = c(upr90), 
                   x = time, xend = time), 
             
              linewidth = 1.5) +
    geom_segment(aes(y = c(lwr95), yend = c(upr95), 
                   x = time, xend = time), 
          
              linewidth = 1) +

  geom_line() + 
  
    geom_point(size = 3, shape = 21, 
             aes(color = NULL, fill = sex)) +
  
  facet_wrap(~ tx, ncol = 5)



p2 <- posterior_draws(mhi3_preds$contr) |> 
 
  
  #mutate(draw = draw * sd_corr , 
  #       draw = exp(draw)) |> 
  
  ggplot(aes(draw)) + geom_line(stat = "density") + 
  facet_wrap(~ hypothesis) +
  geom_vline(xintercept = 1, lty = 2) 





## Posterior probability
# The code below produces a table with hypothesis and posterior probability of direction, the proportion of posterior draws where the effect exceeds 1. A pos value of thus 1 means 100% of posterior draws showed the ratio exceeding 1 for that contrast, very strong evidence of a positive effect. Values at 0.5 on the other hand, are essentially a coin flip, no credible directional effect. Useful as a supplementary statistic, perhaps?


## Here I added mean, lower and upper to the summarise to get the mean ratio, as well as 95% ETI in addition to pos. 
 
posterior_draws(mlo2_preds$contr) |>
  mutate(draw = draw * sd_cmyc_low,
         draw = exp(draw)) |>
  summarise(.by = hypothesis,
            mean = mean(draw),
            lower95 = quantile(draw, 0.025),
            upper95 = quantile(draw, 0.975),
            pos = sum(draw > 1) / n())


# Since these are ratios (from the exp() transform), the null value is 1, not 0 - i.e., its about whether the 95% ETI excludes 1, not zero. These mean/CI tables should be reported and intepreted as ratios (fold-changes), e.g. "mean ratio = 1.15, 95% ETI [1.08, 1.23] - interpretable as "a 15% increase", not as an absolute cm3 diff. 

# For figure plotting


pred_cmyc_high <- posterior_draws(mhi2_preds$pred) |>
  mutate(
    sex = if_else(grepl("_female", tx_sex), "female", "male"),
    tx = gsub("_female", "", tx_sex),
    tx = gsub("_male", "", tx),
    tx = factor(tx, levels = c("int_old_low", "int_old_mod", "int_yng_low", "int_yng_mod", "con_old_con"))
  ) |>
  mutate(draw = draw * sd_cmyc_high + mean_cmyc_high,
         draw = exp(draw)) |>
  summarise(.by = c(tx, sex, time),
            m = mean(draw),
            lwr50 = quantile(draw, 0.25),
            upr50 = quantile(draw, 0.75),
            lwr90 = quantile(draw, 0.05),
            upr90 = quantile(draw, 0.95),
            lwr95 = quantile(draw, 0.025),
            upr95 = quantile(draw, 0.975))

saveRDS(pred_cmyc_high, "data/data-gen/pred_cmyc_high.rds")


contrast_cmyc_high <- posterior_draws(mhi2_preds$contr) |>
  mutate(draw = draw * sd_cmyc_high, draw = exp(draw)) |>
  summarise(.by = hypothesis,
            mean = mean(draw),
            lower95 = quantile(draw, 0.025),
            upper95 = quantile(draw, 0.975),
            pos = sum(draw > 1) / n()) |>
  mutate(pct_change = (mean - 1) * 100,
         pct_lower  = (lower95 - 1) * 100,
         pct_upper  = (upper95 - 1) * 100,
         credible   = ifelse(lower95 > 1 | upper95 < 1, "Yes", "No"))

saveRDS(contrast_cmyc_high, "data/data-gen/contrast_cmyc_high.rds")



```







-->







