---
title: "Protein Analysis"
format: html
editor: source
editor_options: 
  chunk_output_type: console
---


```{r}
#| label: packages-and-settings
#| eval: true
#| message: false
#| warning: false
#| echo: true

# Loading packages
library(brms)
library(marginaleffects)
library(tidyverse)
library(reliefdata)


# This setting disables a safety warning message for the marginaleffects package
# we should still confirm the order of effects for the contrasts
options(marginaleffects_safe = FALSE)


```


### Western - UBF and c-Myc


1.  Within condition (low- and moderate-volume, and non-training control conditions) change from baseline.
2.  Differences between age-groups at baseline.
3.  Average between age-group differences in change from baseline (young vs. old in the intervention group)
4.  Average between control- vs. intervention-group changes from baseline.
5.  Average effect of condition in the intervention groups (low- vs. moderate volume)
6.  Differences between age-groups in effect of condition.

The first contrast will be built for descriptive purposes and serves as the building block for comparisons in contrast 3-4.

### UBF

#### Data

In the code chunk below we plot the data grouped by participant leg.

```{r}

# Data

west.dat <- reliefdata::relief_western |> 
  mutate(participant = as.character(participant)) |> 
  select(participant, leg, sampleid = sample, time, capillary, peak,
         protein, position, mw_kda, height, area, area_p, corr.area,
         width, plate) |> 
  filter(!sampleid %in% c("Biot. Ladder", "DDH2O"),
         !participant %in% c("3084", "3090")) |>  #no measurement
  print()

condition <- reliefdata::relief_volume |> 
  mutate(participant = as.character(participant))

participants <- reliefdata::relief_participants |> 
  mutate(age_group = if_else(age < 40, "yng", "old")) |> 
  select(participant, age, sex, allocation, age_group)


ubf.dat <- west.dat |> 
  full_join(condition) |> 
  select(-arm) |> 
  full_join(participants) |> 
  filter(!is.na(sampleid), #filters out NA's introduced by joining, since not both legs are represented at pre
         protein == "UBF") |> 
  mutate(corr.area = as.numeric(corr.area)) |> 
  mutate(y = log(corr.area),
         ystd = (y - mean(y, na.rm = TRUE)) / sd(y, na.rm = TRUE),
         condition = if_else(allocation == "con", "con", condition),
         tx = paste0(allocation, "_", age_group, "_", condition),
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con")),
         tx_sex = paste0(tx, "_", sex),
         tx_sex = factor(tx_sex, levels = c(
           "int_old_low_female", "int_old_low_male",
           "int_old_mod_female", "int_old_mod_male", 
           "int_yng_low_female", "int_yng_low_male",
           "int_yng_mod_female", "int_yng_mod_male", 
           "con_old_con_female", "con_old_con_male"
         )),
         time = factor(time, levels = c("pre", "w3")),
         sex_centered = if_else(sex == "female", -0.5, 0.5)) |> 
  print()

 

## Save sd and mean for back transform
mean_ubf <- mean(ubf.dat$y, na.rm = TRUE)
sd_ubf   <- sd(ubf.dat$y, na.rm = TRUE)


# A preliminary figure of observed data

ubf.dat |> 
  # Change factor names for plotting
  mutate(time = factor(time, labels = c("Pre", "Week 3")), 
         age_group = factor(age_group, labels = c("Old", "Young")), 
         allocation = factor(allocation, labels = c("Control", "Intervention") ), 
         sex = factor(sex, labels = c("Female", "Male"))) |> 
  
  ggplot(aes(time, corr.area,
             color = sex,
             group = paste(participant, leg))) + 
  geom_line() + 
  facet_wrap(~ paste(age_group, allocation))  + 
  labs(x = "", y = "Corrected Area UBF") + 
  theme(axis.title.y = ggtext::element_markdown())




```

#### Models

```{r}
form0 <- bf(ystd ~   0 + tx + tx:time + sex + (1|participant))
form1 <- bf(ystd ~  time + allocation + allocation:time +  (1|participant) +  (1|participant:leg) + (time | tx_sex))
form2 <- bf(ystd ~   0 + tx_sex + tx_sex:time  + (1 + time|participant) +  (1|participant:leg))
form3 <- bf(ystd ~   time + allocation + allocation:time +  (1 + time|participant) +  (1|participant:leg) + (time | tx_sex))


# Priors
get_prior(form0, data = ubf.dat)
get_prior(form1, data = ubf.dat)
get_prior(form2, data = ubf.dat)
get_prior(form3, data = ubf.dat)

pri0 <- c(
  # Sex diffs
  prior(normal(0, 1), class = "b", coef = "sexmale"), 
  # Group-level intercepts - all parameters specifically referenced
  prior(normal(0, 1), class = "b", coef = "txcon_old_con"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_low"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_mod"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_low"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod"),
  
  # time-Post effects
  prior(normal(0, 1), class = "b", coef = "txcon_old_con:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_low:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_mod:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_low:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod:timew3"),
  
  # Random effect (participant)
  prior(exponential(1), class = sd), 
  
  prior(exponential(1), class = sigma)
)


pri1 <- c(
  
  # Intercept + global time slope
  prior(normal(0, 1), class = "Intercept"), 
  prior(normal(0, 1), class = "b"), 
  
  # Correlations of varying effects within groups
  #prior(lkj(2), class = "cor", group = "tx_sex"),
  
  # Random effect (participant)
  prior(exponential(1), class = sd), 
  
  prior(exponential(1), class = sigma)
  
)


pri2 <- c(
  
  # All fixed effects
  prior(normal(0, 1), class = "b"), 

  prior(lkj(2), class = "cor", group = "participant"),
  
  # Random effect (participant and tx_sex) specifically referenced 
  prior(exponential(1), class = sd, group = "participant"),
  prior(exponential(1), class = sd, group = "participant:leg"),
  # Sigma
  prior(exponential(1), class = sigma)
  
)


pri3 <- c(
  
  # Intercept + global time slope
  prior(normal(0, 1), class = "Intercept"), 
  prior(normal(0, 1), class = "b"), 
  
  # Correlations of varying effects within groups
 # prior(lkj(2), class = "cor", group = "tx_sex"),
  prior(lkj(2), class = "cor", group = "participant"),
  
  # Random effect (participant)
  prior(exponential(1), class = sd), 
  
  prior(exponential(1), class = sigma)
  
)




mu0.pp <- brm(form0, 
             prior = pri0, 
             data = ubf.dat, 
             file = "fits/02-protein/mu0",
             file_refit = "on_change",
             sample_prior = "only")

mu1.pp <- brm(form1, 
             prior = pri1, 
             data = ubf.dat, 
             file = "fits/02-protein/mu1",
             file_refit = "on_change",
             sample_prior = "only")

mu2.pp <- brm(form2, 
             prior = pri2, 
             data = ubf.dat, 
             file = "fits/02-protein/mu2",
             file_refit = "on_change",
             sample_prior = "only")

mu3.pp <- brm(form3, 
             prior = pri3, 
             data = ubf.dat, 
             file = "fits/02-protein/mu3",
             file_refit = "on_change",
             sample_prior = "only")

# Set up predictions/contrasts for model 0

#predictions(mu0.pp, 
#            newdata = datagrid(tx = unique, 
#                               time = unique),
#            )

mu0_pred_contr <- function(m, nd1, nd2) {
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified with the datagrid
    newdata = datagrid(tx = unique, 
                       time = unique), 
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
    hypothesis = c(
              "int_old_low_delta" = "b2 - b1 = 0", 
               "int_old_mod_delta" = "b4 - b3 = 0", 
               "int_yng_low_delta" = "b6 - b5 = 0", 
               "int_yng_mod_delta" = "b8 - b7 = 0", 
               "con_old_con_delta" = "b10 - b9 = 0", 
              # Average group diffs
              "yng_vs_old_delta" = "(((b8 - b7) + (b6 - b5)) / 2)  - (((b2 - b1) + (b4 - b3))/2) = 0", 
              "old_vs_con_delta" = "((b2 - b1) + (b4 - b3))/2  - (b10 - b9) = 0",
              # Volume effects diffs
              "yng_mod_vs_low" = "(b8 - b7) - (b6 - b5) = 0",
              "old_mod_vs_low" = "(b4 - b3) - (b2 - b1) = 0", 
              # volume effect diffs between age groups
              "yng_vs_old_vol" = "((b8 - b7) - (b6 - b5)) - ((b4 - b3) - (b2 - b1)) = 0"
                        ))
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = datagrid(tx = unique, 
                               time = unique, 
                               sex = unique), 
      re_formula = NA)
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )

}
  




mu1_pred_contr <- function(m, nd1, nd2) {
  
  # We need a custom grid to avoid counterfactual tx_sex/allocation 
  # groupings
  grid <- ubf.dat |> 
    distinct(tx_sex, allocation) |> 
    expand_grid(time = unique(ubf.dat$time))
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified from the above grid
    newdata = grid, 
    
    re_formula = ~ (time | tx_sex),
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
    by = c("tx_sex", "time"), 
    hypothesis = c(
       "int_old_low_delta" = "((b2-b1)+(b4-b3))/2 = 0", 
       "int_old_mod_delta" = "((b6-b5)+(b8-b7))/2 = 0", 
       "int_yng_low_delta" = "((b10-b9)+(b12-b11))/2 = 0", 
       "int_yng_mod_delta" = "((b14-b13)+(b16-b15))/2  = 0", 
       "con_old_con_delta" = "((b18-b17)+(b20-b19))/2 = 0", 
      # Average group diffs
      "yng_vs_old_delta" = "(((b14-b13)+(b16-b15))/2 + ((b10-b9)+(b12-b11))/2)/2 - (((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2)/2 = 0 ",
      "old_vs_con_delta" = "((((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2) / 2)  - (((b18-b17)+(b20-b19))/2) = 0",
      # Volume effects diffs
      "yng_mod_vs_low" = "((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2 = 0",
      "old_mod_vs_low" = "((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2 = 0", 
      # volume effect diffs between age groups
      "yng_vs_old_vol" = "(((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2) - (((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2) = 0"
      )
    )
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = grid, 
      re_formula = ~ (time | tx_sex))
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )

}

mu2_pred_contr <- function(m, nd1, nd2) {
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified with the datagrid
    newdata = datagrid(tx_sex = unique, 
                       time = unique), 
    re_formula = NA,
    
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
      hypothesis = c(
              "int_old_low_delta" = "((b2-b1)+(b4-b3))/2 = 0", 
              "int_old_mod_delta" = "((b6-b5)+(b8-b7))/2 = 0", 
              "int_yng_low_delta" = "((b10-b9)+(b12-b11))/2 = 0", 
              "int_yng_mod_delta" = "((b14-b13)+(b16-b15))/2  = 0", 
              "con_old_con_delta" = "((b18-b17)+(b20-b19))/2 = 0", 
              "int_old_pooled_delta" = "(((b2-b1)+(b4-b3))/2 + ((b6-b5)+(b8-b7))/2)/2 = 0",
              "int_yng_pooled_delta" = "(((b10-b9)+(b12-b11))/2 + ((b14-b13)+(b16-b15))/2)/2 = 0",
              
               # Average group diffs
              "yng_vs_old_delta" = "(((b14-b13)+(b16-b15))/2 + ((b10-b9)+(b12-b11))/2)/2 - (((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2)/2 = 0 ",
              "old_vs_con_delta" = "((((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2) / 2)  - (((b18-b17)+(b20-b19))/2) = 0",
              # Volume effects diffs
              "yng_mod_vs_low" = "((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2 = 0",
              "old_mod_vs_low" = "((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2 = 0", 
              # volume effect diffs between age groups
              "yng_vs_old_vol" = "(((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2) - (((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2) = 0"
              
            ))
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = datagrid(tx_sex = unique, 
                               time = unique), 
        re_formula = NA)
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )

}

mu3_pred_contr <- function(m, nd1, nd2) {
  
  # We need a custom grid to avoid counterfactual tx_sex/allocation 
  # groupings
  grid <- ubf.dat |> 
    distinct(tx_sex, allocation) |> 
    expand_grid(time = unique(ubf.dat$time))
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified from the above grid
    newdata = grid, 
    
    re_formula = ~ (time | tx_sex),
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
    by = c("tx_sex", "time"), 
    hypothesis = c(
       "int_old_low_delta" = "((b2-b1)+(b4-b3))/2 = 0", 
       "int_old_mod_delta" = "((b6-b5)+(b8-b7))/2 = 0", 
       "int_yng_low_delta" = "((b10-b9)+(b12-b11))/2 = 0", 
       "int_yng_mod_delta" = "((b14-b13)+(b16-b15))/2  = 0", 
       "con_old_con_delta" = "((b18-b17)+(b20-b19))/2 = 0", 
      # Average group diffs
      "yng_vs_old_delta" = "(((b14-b13)+(b16-b15))/2 + ((b10-b9)+(b12-b11))/2)/2 - (((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2)/2 = 0 ",
      "old_vs_con_delta" = "((((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2) / 2)  - (((b18-b17)+(b20-b19))/2) = 0",
      # Volume effects diffs
      "yng_mod_vs_low" = "((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2 = 0",
      "old_mod_vs_low" = "((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2 = 0", 
      # volume effect diffs between age groups
      "yng_vs_old_vol" = "(((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2) - (((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2) = 0"
      )
    )
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = grid, 
      re_formula = ~ (time | tx_sex))
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )
}



  
mu0_preds <- mu0_pred_contr(mu0.pp, nd1 = 1000, nd2 = 50)
mu1_preds <- mu1_pred_contr(mu1.pp, nd1 = 1000, nd2 = 50)
mu2_preds <- mu2_pred_contr(mu2.pp, nd1 = 1000, nd2 = 50)
mu3_preds <- mu3_pred_contr(mu3.pp, nd1 = 1000, nd2 = 50)


# Prior predictive simulations

p1 <- bind_rows(
  posterior_draws(mu0_preds$pred) |> 
  data.frame() |> 
    mutate(model = "mu0"), 
  
  posterior_draws(mu1_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |>
    mutate(model = "mu1"), 
  
  posterior_draws(mu2_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
    mutate(model = "mu2"), 
  
  posterior_draws(mu3_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |>
    mutate(model = "mu3")
    ) |> 
 
  ggplot(aes(time, draw, group = paste(drawid, tx, sex), 
             color = sex)) + geom_line() + 
  facet_grid(model ~ tx)






p2 <- bind_rows(
  posterior_draws(mu0_preds$contr) |> 
  mutate(model = "mu0"), 
    posterior_draws(mu1_preds$contr) |> 
  mutate(model = "mu1"), 
           posterior_draws(mu2_preds$contr) |> 
  mutate(model = "mu2"), 
   posterior_draws(mu3_preds$contr) |> 
  mutate(model = "mu3")) |>  
  
  ggplot(aes(draw, color = model)) + geom_line(stat = "density") + 
  facet_wrap(~ hypothesis) +
  geom_vline(xintercept = 0, lty = 2) + 
  coord_cartesian(xlim = c(-5, 5))



```

```{r}
### Get the posterior distribution

mu0.post <- update(mu0.pp, 
                  sample_prior = "no", 
                  iter = 4000,
                  file = "fits/02-protein/mu0.post",
                  file_refit = "on_change")

# rhat checks out, 0 above 1.01, highest at 1.003 though

# neff ratio checks out, min ESS at 1347, with zero below 400


mu1.post <- update(mu1.pp, 
                  sample_prior = "no", 
                  cores = 4,
                  control = list(adapt_delta = 0.95, 
                                 max_treedepth = 15),
                  iter = 4000,
                  file = "fits/02-protein/mu1.post",
                  file_refit = "on_change")

# rhat checks out, 0 above 1.01, highest at 1.003

# checks out barely, min ESS at 1083, with zero below 400


mu2.post <- update(mu2.pp, 
                  sample_prior = "no", 
                  cores = 4,
                   control = list(adapt_delta = 0.99, 
                                 max_treedepth = 15),
                  iter = 10000,
                  file = "fits/02-protein/mu2.post",
                  file_refit = "on_change")

# Warning messages:
# 1: There were 6 divergent transitions after warmup. See
# https://mc-stan.org/misc/warnings.html#divergent-transitions-after-warmup
# to find out why this is a problem and how to eliminate them. 
# 2: Examine the pairs() plot to diagnose sampling problems

# Final settings: adapt_delta = 0.99, iter = 10000
# rhat checks out, 0 above 1.01, highest at 1.006
# checks out barely, min ESS at 597, with zero below 400



mu3.post <- update(mu3.pp, 
                  sample_prior = "no", 
                  cores = 4,
                   control = list(adapt_delta = 0.99, 
                                 max_treedepth = 15),
                  iter = 10000,
                  file = "fits/02-protein/mu3.post",
                  file_refit = "on_change")

# Bumping up to 10k, changing adapt_delta to 0.99 due to low ESS and increased divergences

# Warning messages:
# 1: There were 3 divergent transitions after warmup. See
# https://mc-stan.org/misc/warnings.html#divergent-transitions-after-warmup
# to find out why this is a problem and how to eliminate them. 
# 2: There were 1 chains where the estimated Bayesian Fraction of Missing Information was low. See
# https://mc-stan.org/misc/warnings.html#bfmi-low 
# 3: Examine the pairs() plot to diagnose sampling problems

#rhat(mu3.post) 
#rhat_vals <- brms::rhat(mu3.post)
#max(rhat_vals)
#rhat_vals[rhat_vals > 1.01] 
# rhat checks out, 0 above 1.01, highest at 1.004

#neff_ratio(mu3.post)

#total_draws <- brms::ndraws(mu3.post)
#total_draws

#neff_vals <- brms::neff_ratio(mu3.post)
#ess_vals <- neff_vals * total_draws

#min(ess_vals)                  
#ess_vals[ess_vals < 400]       
# checks out barely, min ESS at 583, with zero below 400

## Posterior predictive checks



pp_check(mu3.post, type = "dens_overlay_grouped", 
         group = "tx_sex")

pp_check(mu2.post)

# These looks like the models might need some work/more data




```

```{r}
# Group comparisons

mu0_preds <- mu0_pred_contr(mu0.post, nd1 = 4000, nd2 = 50)
mu1_preds <- mu1_pred_contr(mu1.post, nd1 = 4000, nd2 = 50)
mu2_preds <- mu2_pred_contr(mu2.post, nd1 = 4000, nd2 = 50)
mu3_preds <- mu3_pred_contr(mu3.post, nd1 = 4000, nd2 = 50)


p1 <- bind_rows(
  posterior_draws(mu0_preds$pred) |> 
  data.frame() |> 
    mutate(model = "mu0"), 
  
  posterior_draws(mu1_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |>
    mutate(model = "mu1"), 
  
  posterior_draws(mu2_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
    mutate(model = "mu2"), 
  
  posterior_draws(mu3_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
    mutate(model = "mu3")
  ) |> 
 
  ggplot(aes(time, draw, group = paste(drawid, tx, sex), 
             color = sex)) + geom_line() + 
  facet_grid(model ~ tx)





p2 <- bind_rows(
  posterior_draws(mu0_preds$contr) |> 
  mutate(model = "mu0"), 
    posterior_draws(mu1_preds$contr) |> 
  mutate(model = "mu1"), 
           posterior_draws(mu2_preds$contr) |> 
  mutate(model = "mu2"), 
  posterior_draws(mu3_preds$contr) |> 
  mutate(model = "mu3")) |>  
  
  ggplot(aes(draw, color = model)) + geom_line(stat = "density") + 
  facet_wrap(~ hypothesis) +
  geom_vline(xintercept = 0, lty = 2)



############### Predictions for plotting #####################



p1 <- posterior_draws(mu2_preds$pred) |> 
    mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
  
  #mutate(draw = draw * sd_corr + mean_corr, 
         #draw = exp(draw)) |> 
  
  
  summarise(.by = c(tx, sex, time), 
            m = mean(draw), 
            lwr50 = quantile(draw, 0.25), 
            upr50 = quantile(draw, 0.75),  
            lwr90 = quantile(draw, 0.05), 
            upr90 = quantile(draw, 0.95),
            lwr95 = quantile(draw, 0.025), 
            upr95 = quantile(draw, 0.975)) |> 
  
  
  ggplot(aes(time, m, color = sex, group = paste(tx, sex))) + 
  

  
  geom_segment(aes(y = c(lwr50), yend = c(upr50), 
                   x = time, xend = time), 
               
              linewidth = 2) +
    geom_segment(aes(y = c(lwr90), yend = c(upr90), 
                   x = time, xend = time), 
             
              linewidth = 1.5) +
    geom_segment(aes(y = c(lwr95), yend = c(upr95), 
                   x = time, xend = time), 
          
              linewidth = 1) +

  geom_line() + 
  
    geom_point(size = 3, shape = 21, 
             aes(color = NULL, fill = sex)) +
  
  facet_wrap(~ tx, ncol = 5)



p2 <- posterior_draws(mu3_preds$contr) |> 
 
  
  #mutate(draw = draw * sd_corr , 
  #       draw = exp(draw)) |> 
  
  ggplot(aes(draw)) + geom_line(stat = "density") + 
  facet_wrap(~ hypothesis) +
  geom_vline(xintercept = 1, lty = 2) 





## Posterior probability
# The code below produces a table with hypothesis and posterior probability of direction, the proportion of posterior draws where the effect exceeds 1. A pos value of thus 1 means 100% of posterior draws showed the ratio exceeding 1 for that contrast, very strong evidence of a positive effect. Values at 0.5 on the other hand, are essentially a coin flip, no credible directional effect. Useful as a supplementary statistic, perhaps?


## Here I added mean, lower and upper to the summarise to get the mean ratio, as well as 95% ETI in addition to pos. 
 
posterior_draws(mu2_preds$contr) |>
  mutate(draw = draw * sd_ubf,
         draw = exp(draw)) |>
  summarise(.by = hypothesis,
            mean = mean(draw),
            lower95 = quantile(draw, 0.025),
            upper95 = quantile(draw, 0.975),
            pos = sum(draw > 1) / n())


# Since these are ratios (from the exp() transform), the null value is 1, not 0 - i.e., its about whether the 95% ETI excludes 1, not zero. These mean/CI tables should be reported and intepreted as ratios (fold-changes), e.g. "mean ratio = 1.15, 95% ETI [1.08, 1.23] - interpretable as "a 15% increase", not as an absolute cm3 diff. 



# For figure plotting


pred_ubf <- posterior_draws(mu2_preds$pred) |>
  mutate(
    sex = if_else(grepl("_female", tx_sex), "female", "male"),
    tx = gsub("_female", "", tx_sex),
    tx = gsub("_male", "", tx),
    tx = factor(tx, levels = c("int_old_low", "int_old_mod", "int_yng_low", "int_yng_mod", "con_old_con"))
  ) |>
  mutate(draw = draw * sd_ubf + mean_ubf,
         draw = exp(draw)) |>
  summarise(.by = c(tx, sex, time),
            m = mean(draw),
            lwr50 = quantile(draw, 0.25),
            upr50 = quantile(draw, 0.75),
            lwr90 = quantile(draw, 0.05),
            upr90 = quantile(draw, 0.95),
            lwr95 = quantile(draw, 0.025),
            upr95 = quantile(draw, 0.975))

saveRDS(pred_ubf, "data/data-gen/pred_ubf.rds")


contrast_ubf <- posterior_draws(mu2_preds$contr) |>
  mutate(draw = draw * sd_ubf, draw = exp(draw)) |>
  summarise(.by = hypothesis,
            mean = mean(draw),
            lower95 = quantile(draw, 0.025),
            upper95 = quantile(draw, 0.975),
            pos = sum(draw > 1) / n()) |>
  mutate(pct_change = (mean - 1) * 100,
         pct_lower  = (lower95 - 1) * 100,
         pct_upper  = (upper95 - 1) * 100,
         credible   = ifelse(lower95 > 1 | upper95 < 1, "Yes", "No"))

saveRDS(contrast_ubf, "data/data-gen/contrast_ubf.rds")




```


#### Results UBF

Older adults in the intervention group (averaged across volume conditions) tended to increase UBF protein content by week 3 (19.1%, 95% CI [−7.0%, 49.0%], P(>0) = 90.1%), as did the control group (24.9%, 95% CI [−15.7%, 64.3%], P(>0) = 90.8%), with no evidence of a difference between them (−2.6%, 95% CI [−33.7%, 38.1%], P(<0) = 58.6%). There was weak evidence of an increase in young adults (averaged across volume conditions; 15.7%, 95% CI [−8.0%, 44.2%], P(>0) = 88.5%), with no evidence of a difference from older adults (−1.5%, 95% CI [−29.0%, 33.8%], P(<0) = 56.8%). Within young adults, there was weak evidence of an increase with both low-volume (15.2%, 95% CI [−18.7%, 52.5%], P(>0) = 80.8%) and moderate-volume training (17.6%, 95% CI [−15.2%, 60.7%], P(>0) = 83.2%), with no evidence of a difference between them (4.4%, 95% CI [−34.0%, 54.8%], P(>0) = 53.9%). In older adults, there was weak evidence of an increase with low-volume training (27.1%, 95% CI [−13.3%, 74.7%], P(>0) = 88.9%) and no evidence with moderate-volume training (13.1%, 95% CI [−18.8%, 55.0%], P(>0) = 74.9%), again with no evidence of a difference between them (−8.2%, 95% CI [−44.1%, 41.9%], P(<0) = 67.8%). There was no evidence that the volume effect differed between age groups (20.1%, 95% CI [−39.8%, 112.3%], P(>0) = 67.2%).



### c-Myc

#### Data

In the code chunk below we plot the data grouped by participant leg.

```{r}

# Data

west.dat <- reliefdata::relief_western |> 
  mutate(participant = as.character(participant)) |> 
  select(participant, leg, sampleid = sample, time, capillary, peak,
         protein, position, mw_kda, height, area, area_p, corr.area,
         width, plate) |> 
  filter(!sampleid %in% c("Biot. Ladder", "DDH2O"),
         !participant %in% c("3084", "3090")) |>  #no measurement
  print()

condition <- reliefdata::relief_volume |> 
  mutate(participant = as.character(participant))

participants <- reliefdata::relief_participants |> 
  mutate(age_group = if_else(age < 40, "yng", "old")) |> 
  select(participant, age, sex, allocation, age_group)

# Both cMyc peaks are noted in the runs: cMyc_low = ~49 kDa, cMyc_high = ~60 kDa.
# Decision: both peaks should be analysedm but as separate outcomes (not combined/summed),
# since they likely represent distinct isoforms/phosphorylation states rather than
# duplicate measurements of the same signal.

# Shared cleaning/join logic as a function, applied separately per peak below
build_myc_dat <- function(protein_filter) {
  west.dat |> 
    full_join(condition) |> 
    select(-arm) |> 
    full_join(participants) |> 
    filter(!is.na(sampleid), #filters out NA's introduced by joining, since not both legs are represented at pre
           protein == protein_filter) |> 
    mutate(corr.area = as.numeric(corr.area)) |> 
    mutate(y = log(corr.area),
           ystd = (y - mean(y, na.rm = TRUE)) / sd(y, na.rm = TRUE),
           condition = if_else(allocation == "con", "con", condition),
           tx = paste0(allocation, "_", age_group, "_", condition),
           tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con")),
           tx_sex = paste0(tx, "_", sex),
           tx_sex = factor(tx_sex, levels = c(
             "int_old_low_female", "int_old_low_male",
             "int_old_mod_female", "int_old_mod_male", 
             "int_yng_low_female", "int_yng_low_male",
             "int_yng_mod_female", "int_yng_mod_male", 
             "con_old_con_female", "con_old_con_male"
           )),
           time = factor(time, levels = c("pre", "w3")),
           sex_centered = if_else(sex == "female", -0.5, 0.5)) |> 
    print()
}

myc_high.dat <- build_myc_dat("cMyc_high")
myc_low.dat  <- build_myc_dat("cMyc_low")

## Save sd and mean for back transform, separately per peak
mean_cmyc_high <- mean(myc_high.dat$y, na.rm = TRUE)
sd_cmyc_high   <- sd(myc_high.dat$y, na.rm = TRUE)

mean_cmyc_low <- mean(myc_low.dat$y, na.rm = TRUE)
sd_cmyc_low   <- sd(myc_low.dat$y, na.rm = TRUE)


# A preliminary figure of observed data, per peak

myc_high.dat |> 
  mutate(time = factor(time, labels = c("Pre", "Week 3")), 
         age_group = factor(age_group, labels = c("Old", "Young")), 
         allocation = factor(allocation, labels = c("Control", "Intervention")), 
         sex = factor(sex, labels = c("Female", "Male"))) |> 
  ggplot(aes(time, corr.area, color = sex, group = paste(participant, leg))) + 
  geom_line() + 
  facet_wrap(~ paste(age_group, allocation)) + 
  labs(x = "", y = "Corrected Area c-Myc (high, ~60 kDa)") + 
  theme(axis.title.y = ggtext::element_markdown())

myc_low.dat |> 
  mutate(time = factor(time, labels = c("Pre", "Week 3")), 
         age_group = factor(age_group, labels = c("Old", "Young")), 
         allocation = factor(allocation, labels = c("Control", "Intervention")), 
         sex = factor(sex, labels = c("Female", "Male"))) |> 
  ggplot(aes(time, corr.area, color = sex, group = paste(participant, leg))) + 
  geom_line() + 
  facet_wrap(~ paste(age_group, allocation)) + 
  labs(x = "", y = "Corrected Area c-Myc (low, ~49 kDa)") + 
  theme(axis.title.y = ggtext::element_markdown())



```

### c-Myc High Peak

#### Models

```{r}
form0 <- bf(ystd ~   0 + tx + tx:time + sex + (1|participant))
form1 <- bf(ystd ~  time + allocation + allocation:time +  (1|participant) +  (1|participant:leg) + (time | tx_sex))
form2 <- bf(ystd ~   0 + tx_sex + tx_sex:time  + (1 + time|participant) +  (1|participant:leg))
form3 <- bf(ystd ~   time + allocation + allocation:time +  (1 + time|participant) +  (1|participant:leg) + (time | tx_sex))


# Priors cmyc high
get_prior(form0, data = myc_high.dat)
get_prior(form1, data = myc_high.dat)
get_prior(form2, data = myc_high.dat)
get_prior(form3, data = myc_high.dat)

pri0 <- c(
  # Sex diffs
  prior(normal(0, 1), class = "b", coef = "sexmale"), 
  # Group-level intercepts - all parameters specifically referenced
  prior(normal(0, 1), class = "b", coef = "txcon_old_con"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_low"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_mod"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_low"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod"),
  
  # time-Post effects
  prior(normal(0, 1), class = "b", coef = "txcon_old_con:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_low:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_mod:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_low:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod:timew3"),
  
  # Random effect (participant)
  prior(exponential(1), class = sd), 
  
  prior(exponential(1), class = sigma)
)


pri1 <- c(
  
  # Intercept + global time slope
  prior(normal(0, 1), class = "Intercept"), 
  prior(normal(0, 1), class = "b"), 
  
  # Correlations of varying effects within groups
  #prior(lkj(2), class = "cor", group = "tx_sex"),
  
  # Random effect (participant)
  prior(exponential(1), class = sd), 
  
  prior(exponential(1), class = sigma)
  
)


pri2 <- c(
  
  # All fixed effects
  prior(normal(0, 1), class = "b"), 

  prior(lkj(2), class = "cor", group = "participant"),
  
  # Random effect (participant and tx_sex) specifically referenced 
  prior(exponential(1), class = sd, group = "participant"),
  prior(exponential(1), class = sd, group = "participant:leg"),
  # Sigma
  prior(exponential(1), class = sigma)
  
)


pri3 <- c(
  
  # Intercept + global time slope
  prior(normal(0, 1), class = "Intercept"), 
  prior(normal(0, 1), class = "b"), 
  
  # Correlations of varying effects within groups
 # prior(lkj(2), class = "cor", group = "tx_sex"),
  prior(lkj(2), class = "cor", group = "participant"),
  
  # Random effect (participant)
  prior(exponential(1), class = sd), 
  
  prior(exponential(1), class = sigma)
  
)




mhi0.pp <- brm(form0, 
             prior = pri0, 
             data = myc_high.dat, 
             file = "fits/02-protein/mhi0",
             file_refit = "on_change",
             sample_prior = "only")

mhi1.pp <- brm(form1, 
             prior = pri1, 
             data = myc_high.dat, 
             file = "fits/02-protein/mhi1",
             file_refit = "on_change",
             sample_prior = "only")

mhi2.pp <- brm(form2, 
             prior = pri2, 
             data = myc_high.dat, 
             file = "fits/02-protein/mhi2",
             file_refit = "on_change",
             sample_prior = "only")

mhi3.pp <- brm(form3, 
             prior = pri3, 
             data = myc_high.dat, 
             file = "fits/02-protein/mhi3",
             file_refit = "on_change",
             sample_prior = "only")

# Set up predictions/contrasts for model 0

#predictions(mu0.pp, 
#            newdata = datagrid(tx = unique, 
#                               time = unique),
#            )

mhi0_pred_contr <- function(m, nd1, nd2) {
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified with the datagrid
    newdata = datagrid(tx = unique, 
                       time = unique), 
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
    hypothesis = c(
              "int_old_low_delta" = "b2 - b1 = 0", 
               "int_old_mod_delta" = "b4 - b3 = 0", 
               "int_yng_low_delta" = "b6 - b5 = 0", 
               "int_yng_mod_delta" = "b8 - b7 = 0", 
               "con_old_con_delta" = "b10 - b9 = 0", 
              # Average group diffs
              "yng_vs_old_delta" = "(((b8 - b7) + (b6 - b5)) / 2)  - (((b2 - b1) + (b4 - b3))/2) = 0", 
              "old_vs_con_delta" = "((b2 - b1) + (b4 - b3))/2  - (b10 - b9) = 0",
              # Volume effects diffs
              "yng_mod_vs_low" = "(b8 - b7) - (b6 - b5) = 0",
              "old_mod_vs_low" = "(b4 - b3) - (b2 - b1) = 0", 
              # volume effect diffs between age groups
              "yng_vs_old_vol" = "((b8 - b7) - (b6 - b5)) - ((b4 - b3) - (b2 - b1)) = 0"
                        ))
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = datagrid(tx = unique, 
                               time = unique, 
                               sex = unique), 
      re_formula = NA)
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )

}
  




mhi1_pred_contr <- function(m, nd1, nd2) {
  
  # We need a custom grid to avoid counterfactual tx_sex/allocation 
  # groupings
  grid <- myc_high.dat |> 
    distinct(tx_sex, allocation) |> 
    expand_grid(time = unique(myc_high.dat$time))
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified from the above grid
    newdata = grid, 
    
    re_formula = ~ (time | tx_sex),
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
    by = c("tx_sex", "time"), 
    hypothesis = c(
       "int_old_low_delta" = "((b2-b1)+(b4-b3))/2 = 0", 
       "int_old_mod_delta" = "((b6-b5)+(b8-b7))/2 = 0", 
       "int_yng_low_delta" = "((b10-b9)+(b12-b11))/2 = 0", 
       "int_yng_mod_delta" = "((b14-b13)+(b16-b15))/2  = 0", 
       "con_old_con_delta" = "((b18-b17)+(b20-b19))/2 = 0", 
      # Average group diffs
      "yng_vs_old_delta" = "(((b14-b13)+(b16-b15))/2 + ((b10-b9)+(b12-b11))/2)/2 - (((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2)/2 = 0 ",
      "old_vs_con_delta" = "((((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2) / 2)  - (((b18-b17)+(b20-b19))/2) = 0",
      # Volume effects diffs
      "yng_mod_vs_low" = "((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2 = 0",
      "old_mod_vs_low" = "((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2 = 0", 
      # volume effect diffs between age groups
      "yng_vs_old_vol" = "(((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2) - (((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2) = 0"
      )
    )
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = grid, 
      re_formula = ~ (time | tx_sex))
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )

}

mhi2_pred_contr <- function(m, nd1, nd2) {
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified with the datagrid
    newdata = datagrid(tx_sex = unique, 
                       time = unique), 
    re_formula = NA,
    
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
      hypothesis = c(
              "int_old_low_delta" = "((b2-b1)+(b4-b3))/2 = 0", 
              "int_old_mod_delta" = "((b6-b5)+(b8-b7))/2 = 0", 
              "int_yng_low_delta" = "((b10-b9)+(b12-b11))/2 = 0", 
              "int_yng_mod_delta" = "((b14-b13)+(b16-b15))/2  = 0", 
              "con_old_con_delta" = "((b18-b17)+(b20-b19))/2 = 0", 
              "int_old_pooled_delta" = "(((b2-b1)+(b4-b3))/2 + ((b6-b5)+(b8-b7))/2)/2 = 0",
              "int_yng_pooled_delta" = "(((b10-b9)+(b12-b11))/2 + ((b14-b13)+(b16-b15))/2)/2 = 0",
              
               # Average group diffs
              "yng_vs_old_delta" = "(((b14-b13)+(b16-b15))/2 + ((b10-b9)+(b12-b11))/2)/2 - (((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2)/2 = 0 ",
              "old_vs_con_delta" = "((((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2) / 2)  - (((b18-b17)+(b20-b19))/2) = 0",
              # Volume effects diffs
              "yng_mod_vs_low" = "((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2 = 0",
              "old_mod_vs_low" = "((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2 = 0", 
              # volume effect diffs between age groups
              "yng_vs_old_vol" = "(((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2) - (((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2) = 0"
              
            ))
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = datagrid(tx_sex = unique, 
                               time = unique), 
        re_formula = NA)
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )

}

mhi3_pred_contr <- function(m, nd1, nd2) {
  
  # We need a custom grid to avoid counterfactual tx_sex/allocation 
  # groupings
  grid <- myc_high.dat |> 
    distinct(tx_sex, allocation) |> 
    expand_grid(time = unique(myc_high.dat$time))
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified from the above grid
    newdata = grid, 
    
    re_formula = ~ (time | tx_sex),
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
    by = c("tx_sex", "time"), 
    hypothesis = c(
       "int_old_low_delta" = "((b2-b1)+(b4-b3))/2 = 0", 
       "int_old_mod_delta" = "((b6-b5)+(b8-b7))/2 = 0", 
       "int_yng_low_delta" = "((b10-b9)+(b12-b11))/2 = 0", 
       "int_yng_mod_delta" = "((b14-b13)+(b16-b15))/2  = 0", 
       "con_old_con_delta" = "((b18-b17)+(b20-b19))/2 = 0", 
      # Average group diffs
      "yng_vs_old_delta" = "(((b14-b13)+(b16-b15))/2 + ((b10-b9)+(b12-b11))/2)/2 - (((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2)/2 = 0 ",
      "old_vs_con_delta" = "((((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2) / 2)  - (((b18-b17)+(b20-b19))/2) = 0",
      # Volume effects diffs
      "yng_mod_vs_low" = "((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2 = 0",
      "old_mod_vs_low" = "((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2 = 0", 
      # volume effect diffs between age groups
      "yng_vs_old_vol" = "(((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2) - (((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2) = 0"
      )
    )
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = grid, 
      re_formula = ~ (time | tx_sex))
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )
}



  
mhi0_preds <- mhi0_pred_contr(mhi0.pp, nd1 = 1000, nd2 = 50)
mhi1_preds <- mhi1_pred_contr(mhi1.pp, nd1 = 1000, nd2 = 50)
mhi2_preds <- mhi2_pred_contr(mhi2.pp, nd1 = 1000, nd2 = 50)
mhi3_preds <- mhi3_pred_contr(mhi3.pp, nd1 = 1000, nd2 = 50)


# Prior predictive simulations

p1 <- bind_rows(
  posterior_draws(mhi0_preds$pred) |> 
  data.frame() |> 
    mutate(model = "mhi0"), 
  
  posterior_draws(mhi1_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |>
    mutate(model = "mhi1"), 
  
  posterior_draws(mhi2_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
    mutate(model = "mhi2"), 
  
  posterior_draws(mhi3_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |>
    mutate(model = "mhi3")
    ) |> 
 
  ggplot(aes(time, draw, group = paste(drawid, tx, sex), 
             color = sex)) + geom_line() + 
  facet_grid(model ~ tx)






p2 <- bind_rows(
  posterior_draws(mhi0_preds$contr) |> 
  mutate(model = "mhi0"), 
    posterior_draws(mhi1_preds$contr) |> 
  mutate(model = "mhi1"), 
           posterior_draws(mhi2_preds$contr) |> 
  mutate(model = "mhi2"), 
   posterior_draws(mhi3_preds$contr) |> 
  mutate(model = "mhi3")) |>  
  
  ggplot(aes(draw, color = model)) + geom_line(stat = "density") + 
  facet_wrap(~ hypothesis) +
  geom_vline(xintercept = 0, lty = 2) + 
  coord_cartesian(xlim = c(-5, 5))



```

```{r}
### Get the posterior distribution

mhi0.post <- update(mhi0.pp, 
                  sample_prior = "no", 
                  iter = 4000,
                  file = "fits/02-protein/mhi0.post",
                  file_refit = "on_change")



mhi1.post <- update(mhi1.pp, 
                  sample_prior = "no", 
                  cores = 4,
                  control = list(adapt_delta = 0.95, 
                                 max_treedepth = 15),
                  iter = 4000,
                  file = "fits/02-protein/mhi1.post",
                  file_refit = "on_change")



mhi2.post <- update(mhi2.pp, 
                  sample_prior = "no", 
                  cores = 4,
                   control = list(adapt_delta = 0.99, 
                                 max_treedepth = 15),
                  iter = 10000,
                  file = "fits/02-protein/mhi2.post",
                  file_refit = "on_change")

# Starting this at 10k iterations and adapt_delta 0.99, since it was required for the UBF data

# rhat checks out, 0 above 1.01, highest at 1.001
# ESS cheks out, min ESS at 3305, with zero below 400



mhi3.post <- update(mhi3.pp, 
                  sample_prior = "no", 
                  cores = 4,
                   control = list(adapt_delta = 0.99, 
                                 max_treedepth = 15),
                  iter = 10000,
                  file = "fits/02-protein/mhi3.post",
                  file_refit = "on_change")

# Starting this at 10k iterations and adapt_delta 0.99, since it was required for the UBF data
# rhat checks out, 0 above 1.01, highest at 1.001
# ESS checks out, min ESS at 3633, with zero below 400

#rhat(mhi3.post) 
#rhat_vals <- brms::rhat(mhi3.post)
#max(rhat_vals)
#rhat_vals[rhat_vals > 1.01] 

#neff_ratio(mhi3.post)

#total_draws <- brms::ndraws(mhi3.post)
#total_draws

#neff_vals <- brms::neff_ratio(mhi3.post)
#ess_vals <- neff_vals * total_draws

#min(ess_vals)                  
#ess_vals[ess_vals < 400]       



## Posterior predictive checks

pp_check(mhi3.post, type = "dens_overlay_grouped", 
         group = "tx_sex")

pp_check(mhi2.post)

# These looks like the models might need some work




```

```{r}
# Group comparisons

mhi0_preds <- mhi0_pred_contr(mhi0.post, nd1 = 4000, nd2 = 50)
mhi1_preds <- mhi1_pred_contr(mhi1.post, nd1 = 4000, nd2 = 50)
mhi2_preds <- mhi2_pred_contr(mhi2.post, nd1 = 4000, nd2 = 50)
mhi3_preds <- mhi3_pred_contr(mhi3.post, nd1 = 4000, nd2 = 50)


p1 <- bind_rows(
  posterior_draws(mhi0_preds$pred) |> 
  data.frame() |> 
    mutate(model = "mhi0"), 
  
  posterior_draws(mhi1_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |>
    mutate(model = "mhi1"), 
  
  posterior_draws(mhi2_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
    mutate(model = "mhi2"), 
  
  posterior_draws(mhi3_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
    mutate(model = "mhi3")
  ) |> 
 
  ggplot(aes(time, draw, group = paste(drawid, tx, sex), 
             color = sex)) + geom_line() + 
  facet_grid(model ~ tx)





p2 <- bind_rows(
  posterior_draws(mhi0_preds$contr) |> 
  mutate(model = "mhi0"), 
    posterior_draws(mhi1_preds$contr) |> 
  mutate(model = "mhi1"), 
           posterior_draws(mhi2_preds$contr) |> 
  mutate(model = "mhi2"), 
  posterior_draws(mhi3_preds$contr) |> 
  mutate(model = "mhi3")) |>  
  
  ggplot(aes(draw, color = model)) + geom_line(stat = "density") + 
  facet_wrap(~ hypothesis) +
  geom_vline(xintercept = 0, lty = 2)



############### Predictions for plotting #####################





# These are looking a bit nonsensical, so I guess something here needs more work, but I cant see what at 23:27 in the evening.

p1 <- posterior_draws(mhi2_preds$pred) |> 
    mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
  
  #mutate(draw = draw * sd_corr + mean_corr, 
         #draw = exp(draw)) |> 
  
  
  summarise(.by = c(tx, sex, time), 
            m = mean(draw), 
            lwr50 = quantile(draw, 0.25), 
            upr50 = quantile(draw, 0.75),  
            lwr90 = quantile(draw, 0.05), 
            upr90 = quantile(draw, 0.95),
            lwr95 = quantile(draw, 0.025), 
            upr95 = quantile(draw, 0.975)) |> 
  
  
  ggplot(aes(time, m, color = sex, group = paste(tx, sex))) + 
  

  
  geom_segment(aes(y = c(lwr50), yend = c(upr50), 
                   x = time, xend = time), 
               
              linewidth = 2) +
    geom_segment(aes(y = c(lwr90), yend = c(upr90), 
                   x = time, xend = time), 
             
              linewidth = 1.5) +
    geom_segment(aes(y = c(lwr95), yend = c(upr95), 
                   x = time, xend = time), 
          
              linewidth = 1) +

  geom_line() + 
  
    geom_point(size = 3, shape = 21, 
             aes(color = NULL, fill = sex)) +
  
  facet_wrap(~ tx, ncol = 5)



p2 <- posterior_draws(mhi3_preds$contr) |> 
 
  
  #mutate(draw = draw * sd_corr , 
  #       draw = exp(draw)) |> 
  
  ggplot(aes(draw)) + geom_line(stat = "density") + 
  facet_wrap(~ hypothesis) +
  geom_vline(xintercept = 1, lty = 2) 





## Posterior probability
# The code below produces a table with hypothesis and posterior probability of direction, the proportion of posterior draws where the effect exceeds 1. A pos value of thus 1 means 100% of posterior draws showed the ratio exceeding 1 for that contrast, very strong evidence of a positive effect. Values at 0.5 on the other hand, are essentially a coin flip, no credible directional effect. Useful as a supplementary statistic, perhaps?


## Here I added mean, lower and upper to the summarise to get the mean ratio, as well as 95% ETI in addition to pos. 
 
posterior_draws(mhi2_preds$contr) |>
  mutate(draw = draw * sd_cmyc_high,
         draw = exp(draw)) |>
  summarise(.by = hypothesis,
            mean = mean(draw),
            lower95 = quantile(draw, 0.025),
            upper95 = quantile(draw, 0.975),
            pos = sum(draw > 1) / n())


# Since these are ratios (from the exp() transform), the null value is 1, not 0 - i.e., its about whether the 95% ETI excludes 1, not zero. These mean/CI tables should be reported and intepreted as ratios (fold-changes), e.g. "mean ratio = 1.15, 95% ETI [1.08, 1.23] - interpretable as "a 15% increase", not as an absolute cm3 diff. 


# For figure plotting


pred_cmyc_high <- posterior_draws(mhi2_preds$pred) |>
  mutate(
    sex = if_else(grepl("_female", tx_sex), "female", "male"),
    tx = gsub("_female", "", tx_sex),
    tx = gsub("_male", "", tx),
    tx = factor(tx, levels = c("int_old_low", "int_old_mod", "int_yng_low", "int_yng_mod", "con_old_con"))
  ) |>
  mutate(draw = draw * sd_cmyc_high + mean_cmyc_high,
         draw = exp(draw)) |>
  summarise(.by = c(tx, sex, time),
            m = mean(draw),
            lwr50 = quantile(draw, 0.25),
            upr50 = quantile(draw, 0.75),
            lwr90 = quantile(draw, 0.05),
            upr90 = quantile(draw, 0.95),
            lwr95 = quantile(draw, 0.025),
            upr95 = quantile(draw, 0.975))

saveRDS(pred_cmyc_high, "data/data-gen/pred_cmyc_high.rds")


contrast_cmyc_high <- posterior_draws(mhi2_preds$contr) |>
  mutate(draw = draw * sd_cmyc_high, draw = exp(draw)) |>
  summarise(.by = hypothesis,
            mean = mean(draw),
            lower95 = quantile(draw, 0.025),
            upper95 = quantile(draw, 0.975),
            pos = sum(draw > 1) / n()) |>
  mutate(pct_change = (mean - 1) * 100,
         pct_lower  = (lower95 - 1) * 100,
         pct_upper  = (upper95 - 1) * 100,
         credible   = ifelse(lower95 > 1 | upper95 < 1, "Yes", "No"))

saveRDS(contrast_cmyc_high, "data/data-gen/contrast_cmyc_high.rds")



```

#### Results cMyc High Peak

The difference in c-Myc (high peak) protein content between the old intervention group (averaged across volume conditions) and the control group was large by week 3 (99.8%, 95% CI [17.5%, 216.3%], P(>0) = 99.3%). This difference was driven mainly by a decrease in the control group (−34.5%, 95% CI [−56.0%, −5.5%], P(<0) = 98.6%), while the intervention group tended to increase (25.9%, 95% CI [−8.9%, 68.1%], P(>0) = 92.0%). There was weak evidence of an increase in young adults (averaged across volume conditions; 18.6%, 95% CI [−12.5%, 57.5%], P(>0) = 85.9%), with no evidence of a difference from older adults (−3.5%, 95% CI [−37.8%, 45.0%], P(<0) = 61.5%). Within young adults, there was no evidence of a change with low-volume training (4.4%, 95% CI [−30.1%, 49.1%], P(>0) = 55.2%), while moderate-volume training tended to increase c-Myc (37.3%, 95% CI [−12.2%, 106.2%], P(>0) = 91.5%). There was weak evidence of a larger increase under moderate volume (36.2%, 95% CI [−23.8%, 127.4%], P(>0) = 83.5%). The same pattern held in older adults: there was no evidence of a change with low-volume training (9.6%, 95% CI [−29.9%, 64.4%], P(>0) = 62.0%), moderate-volume training tended to increase c-Myc (47.9%, 95% CI [−4.7%, 118.4%], P(>0) = 95.5%), and there was weak evidence of a larger increase under moderate volume (41.0%, 95% CI [−24.3%, 138.0%], P(>0) = 84.5%). There was no evidence that the volume effect differed between age groups (5.2%, 95% CI [−55.8%, 114.3%], P(>0) = 46.7%).







<!--

### c-Myc Low Peaks

Experiencing model-complexity issues here, specifically from mlo2.post and mlo3.post. It's possibly a reflection of this peaks signal-to-noise ratio, but Im not sure. 

#### Models

```{r}
form0 <- bf(ystd ~   0 + tx + tx:time + sex + (1|participant))
form1 <- bf(ystd ~  time + allocation + allocation:time +  (1|participant) +  (1|participant:leg) + (time | tx_sex))
form2 <- bf(ystd ~   0 + tx_sex + tx_sex:time  + (1 + time|participant) +  (1|participant:leg))
form3 <- bf(ystd ~   time + allocation + allocation:time +  (1 + time|participant) +  (1|participant:leg) + (time | tx_sex))


# Priors cmyc high
get_prior(form0, data = myc_low.dat)
get_prior(form1, data = myc_low.dat)
get_prior(form2, data = myc_low.dat)
get_prior(form3, data = myc_low.dat)

pri0 <- c(
  # Sex diffs
  prior(normal(0, 1), class = "b", coef = "sexmale"), 
  # Group-level intercepts - all parameters specifically referenced
  prior(normal(0, 1), class = "b", coef = "txcon_old_con"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_low"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_mod"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_low"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod"),
  
  # time-Post effects
  prior(normal(0, 1), class = "b", coef = "txcon_old_con:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_low:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_old_mod:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_low:timew3"), 
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod:timew3"),
  
  # Random effect (participant)
  prior(exponential(1), class = sd), 
  
  prior(exponential(1), class = sigma)
)


pri1 <- c(
  
  # Intercept + global time slope
  prior(normal(0, 1), class = "Intercept"), 
  prior(normal(0, 1), class = "b"), 
  
  # Correlations of varying effects within groups
  #prior(lkj(2), class = "cor", group = "tx_sex"),
  
  # Random effect (participant)
  prior(exponential(1), class = sd), 
  
  prior(exponential(1), class = sigma)
  
)


pri2 <- c(
  
  # All fixed effects
  prior(normal(0, 1), class = "b"), 

  prior(lkj(2), class = "cor", group = "participant"),
  
  # Random effect (participant and tx_sex) specifically referenced 
  prior(exponential(1), class = sd, group = "participant"),
  prior(exponential(1), class = sd, group = "participant:leg"),
  # Sigma
  prior(exponential(1), class = sigma)
  
)


pri3 <- c(
  
  # Intercept + global time slope
  prior(normal(0, 1), class = "Intercept"), 
  prior(normal(0, 1), class = "b"), 
  
  # Correlations of varying effects within groups
 # prior(lkj(2), class = "cor", group = "tx_sex"),
  prior(lkj(2), class = "cor", group = "participant"),
  
  # Random effect (participant)
  prior(exponential(1), class = sd), 
  
  prior(exponential(1), class = sigma)
  
)




mlo0.pp <- brm(form0, 
             prior = pri0, 
             data = myc_low.dat, 
             file = "fits/05-molecular/mlo0",
             file_refit = "on_change",
             sample_prior = "only")

mlo1.pp <- brm(form1, 
             prior = pri1, 
             data = myc_low.dat, 
             file = "fits/05-molecular/mlo1",
             file_refit = "on_change",
             sample_prior = "only")

mlo2.pp <- brm(form2, 
             prior = pri2, 
             data = myc_low.dat, 
             file = "fits/05-molecular/mlo2",
             file_refit = "on_change",
             sample_prior = "only")

mlo3.pp <- brm(form3, 
             prior = pri3, 
             data = myc_low.dat, 
             file = "fits/05-molecular/mlo3",
             file_refit = "on_change",
             sample_prior = "only")

# Set up predictions/contrasts for model 0

#predictions(mu0.pp, 
#            newdata = datagrid(tx = unique, 
#                               time = unique),
#            )

mlo0_pred_contr <- function(m, nd1, nd2) {
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified with the datagrid
    newdata = datagrid(tx = unique, 
                       time = unique), 
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
    hypothesis = c(
              "int_old_low_delta" = "b2 - b1 = 0", 
               "int_old_mod_delta" = "b4 - b3 = 0", 
               "int_yng_low_delta" = "b6 - b5 = 0", 
               "int_yng_mod_delta" = "b8 - b7 = 0", 
               "con_old_con_delta" = "b10 - b9 = 0", 
              # Average group diffs
              "yng_vs_old_delta" = "(((b8 - b7) + (b6 - b5)) / 2)  - (((b2 - b1) + (b4 - b3))/2) = 0", 
              "old_vs_con_delta" = "((b2 - b1) + (b4 - b3))/2  - (b10 - b9) = 0",
              # Volume effects diffs
              "yng_mod_vs_low" = "(b8 - b7) - (b6 - b5) = 0",
              "old_mod_vs_low" = "(b4 - b3) - (b2 - b1) = 0", 
              # volume effect diffs between age groups
              "yng_vs_old_vol" = "((b8 - b7) - (b6 - b5)) - ((b4 - b3) - (b2 - b1)) = 0"
                        ))
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = datagrid(tx = unique, 
                               time = unique, 
                               sex = unique), 
      re_formula = NA)
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )

}
  




mlo1_pred_contr <- function(m, nd1, nd2) {
  
  # We need a custom grid to avoid counterfactual tx_sex/allocation 
  # groupings
  grid <- myc_low.dat |> 
    distinct(tx_sex, allocation) |> 
    expand_grid(time = unique(myc_low.dat$time))
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified from the above grid
    newdata = grid, 
    
    re_formula = ~ (time | tx_sex),
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
    by = c("tx_sex", "time"), 
    hypothesis = c(
       "int_old_low_delta" = "((b2-b1)+(b4-b3))/2 = 0", 
       "int_old_mod_delta" = "((b6-b5)+(b8-b7))/2 = 0", 
       "int_yng_low_delta" = "((b10-b9)+(b12-b11))/2 = 0", 
       "int_yng_mod_delta" = "((b14-b13)+(b16-b15))/2  = 0", 
       "con_old_con_delta" = "((b18-b17)+(b20-b19))/2 = 0", 
      # Average group diffs
      "yng_vs_old_delta" = "(((b14-b13)+(b16-b15))/2 + ((b10-b9)+(b12-b11))/2)/2 - (((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2)/2 = 0 ",
      "old_vs_con_delta" = "((((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2) / 2)  - (((b18-b17)+(b20-b19))/2) = 0",
      # Volume effects diffs
      "yng_mod_vs_low" = "((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2 = 0",
      "old_mod_vs_low" = "((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2 = 0", 
      # volume effect diffs between age groups
      "yng_vs_old_vol" = "(((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2) - (((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2) = 0"
      )
    )
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = grid, 
      re_formula = ~ (time | tx_sex))
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )

}

mlo2_pred_contr <- function(m, nd1, nd2) {
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified with the datagrid
    newdata = datagrid(tx_sex = unique, 
                       time = unique), 
    re_formula = NA,
    
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
      hypothesis = c(
              "int_old_low_delta" = "((b2-b1)+(b4-b3))/2 = 0", 
               "int_old_mod_delta" = "((b6-b5)+(b8-b7))/2 = 0", 
               "int_yng_low_delta" = "((b10-b9)+(b12-b11))/2 = 0", 
               "int_yng_mod_delta" = "((b14-b13)+(b16-b15))/2  = 0", 
               "con_old_con_delta" = "((b18-b17)+(b20-b19))/2 = 0", 
              
               # Average group diffs
              "yng_vs_old_delta" = "(((b14-b13)+(b16-b15))/2 + ((b10-b9)+(b12-b11))/2)/2 - (((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2)/2 = 0 ",
              "old_vs_con_delta" = "((((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2) / 2)  - (((b18-b17)+(b20-b19))/2) = 0",
              # Volume effects diffs
              "yng_mod_vs_low" = "((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2 = 0",
              "old_mod_vs_low" = "((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2 = 0", 
              # volume effect diffs between age groups
              "yng_vs_old_vol" = "(((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2) - (((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2) = 0"
              
            ))
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = datagrid(tx_sex = unique, 
                               time = unique), 
        re_formula = NA)
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )

}

mlo3_pred_contr <- function(m, nd1, nd2) {
  
  # We need a custom grid to avoid counterfactual tx_sex/allocation 
  # groupings
  grid <- myc_low.dat |> 
    distinct(tx_sex, allocation) |> 
    expand_grid(time = unique(myc_low.dat$time))
  
  contr <-  predictions(
    # The model object
    m, 
    # New data specified from the above grid
    newdata = grid, 
    
    re_formula = ~ (time | tx_sex),
    
    # Number of draws 
    ndraws = nd1,
    # Specify contrasts
    by = c("tx_sex", "time"), 
    hypothesis = c(
       "int_old_low_delta" = "((b2-b1)+(b4-b3))/2 = 0", 
       "int_old_mod_delta" = "((b6-b5)+(b8-b7))/2 = 0", 
       "int_yng_low_delta" = "((b10-b9)+(b12-b11))/2 = 0", 
       "int_yng_mod_delta" = "((b14-b13)+(b16-b15))/2  = 0", 
       "con_old_con_delta" = "((b18-b17)+(b20-b19))/2 = 0", 
      # Average group diffs
      "yng_vs_old_delta" = "(((b14-b13)+(b16-b15))/2 + ((b10-b9)+(b12-b11))/2)/2 - (((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2)/2 = 0 ",
      "old_vs_con_delta" = "((((b6-b5)+(b8-b7))/2 + ((b2-b1)+(b4-b3))/2) / 2)  - (((b18-b17)+(b20-b19))/2) = 0",
      # Volume effects diffs
      "yng_mod_vs_low" = "((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2 = 0",
      "old_mod_vs_low" = "((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2 = 0", 
      # volume effect diffs between age groups
      "yng_vs_old_vol" = "(((b14-b13)+(b16-b15))/2 - ((b10-b9)+(b12-b11))/2) - (((b6-b5)+(b8-b7))/2 - ((b2-b1)+(b4-b3))/2) = 0"
      )
    )
  
    pred <- predictions(
      m, 
      # Number of draws
      ndraws = nd2,
      # Predict only over the data grid                  
            newdata = grid, 
      re_formula = ~ (time | tx_sex))
    
    return(
      list(
      contr = contr, 
      pred = pred
      )
    )
}



  
mlo0_preds <- mlo0_pred_contr(mlo0.pp, nd1 = 1000, nd2 = 50)
mlo1_preds <- mlo1_pred_contr(mlo1.pp, nd1 = 1000, nd2 = 50)
mlo2_preds <- mlo2_pred_contr(mlo2.pp, nd1 = 1000, nd2 = 50)
mlo3_preds <- mlo3_pred_contr(mlo3.pp, nd1 = 1000, nd2 = 50)


# Prior predictive simulations

p1 <- bind_rows(
  posterior_draws(mlo0_preds$pred) |> 
  data.frame() |> 
    mutate(model = "mlo0"), 
  
  posterior_draws(mlo1_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |>
    mutate(model = "mlo1"), 
  
  posterior_draws(mlo2_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
    mutate(model = "mlo2"), 
  
  posterior_draws(mlo3_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |>
    mutate(model = "mlo3")
    ) |> 
 
  ggplot(aes(time, draw, group = paste(drawid, tx, sex), 
             color = sex)) + geom_line() + 
  facet_grid(model ~ tx)






p2 <- bind_rows(
  posterior_draws(mlo0_preds$contr) |> 
  mutate(model = "mlo0"), 
    posterior_draws(mlo1_preds$contr) |> 
  mutate(model = "mlo1"), 
           posterior_draws(mlo2_preds$contr) |> 
  mutate(model = "mlo2"), 
   posterior_draws(mlo3_preds$contr) |> 
  mutate(model = "mlo3")) |>  
  
  ggplot(aes(draw, color = model)) + geom_line(stat = "density") + 
  facet_wrap(~ hypothesis) +
  geom_vline(xintercept = 0, lty = 2) + 
  coord_cartesian(xlim = c(-5, 5))



```

```{r}
### Get the posterior distribution

mlo0.post <- update(mlo0.pp, 
                  sample_prior = "no", 
                  iter = 4000,
                  file = "fits/05-molecular/mlo0.post",
                  file_refit = "on_change")



mlo1.post <- update(mlo1.pp, 
                  sample_prior = "no", 
                  cores = 4,
                  control = list(adapt_delta = 0.95, 
                                 max_treedepth = 15),
                  iter = 4000,
                  file = "fits/05-molecular/mlo1.post",
                  file_refit = "on_change")



mlo2.post <- update(mlo2.pp, 
                  sample_prior = "no", 
                  cores = 4,
                   control = list(adapt_delta = 0.999, 
                                 max_treedepth = 15),
                  iter = 10000,
                  file = "fits/05-molecular/mlo2.post",
                  file_refit = "on_change")


# Upping from 0.99 to 0.999 adapt_delta due to low-BFMI warnings

mlo3.post <- update(mlo3.pp, 
                  sample_prior = "no", 
                  cores = 4,
                  control = list(adapt_delta = 0.99, 
                                max_treedepth = 15),
                  iter = 10000,
                 file = "fits/05-molecular/mlo3.post",
                 file_refit = "on_change")


#rhat(mhi3.post) 
#rhat_vals <- brms::rhat(mhi3.post)
#max(rhat_vals)
#rhat_vals[rhat_vals > 1.01] 
# rhat checks out, 0 above 1.01, highest at 1.004

#neff_ratio(mhi3.post)

#total_draws <- brms::ndraws(mhi3.post)
#total_draws

#neff_vals <- brms::neff_ratio(mhi3.post)
#ess_vals <- neff_vals * total_draws

#min(ess_vals)                  
#ess_vals[ess_vals < 400]       
# checks out barely, min ESS at 1128, with zero below 400


## Posterior predictive checks



pp_check(mlo2.post, type = "dens_overlay_grouped", 
         group = "tx_sex")

pp_check(mlo2.post)

# Not sure about these, model 2 looks a bit more divergent than the others ive seen, model 3 looks fine




```

```{r}
# Group comparisons

mhi0_preds <- mhi0_pred_contr(mhi0.post, nd1 = 4000, nd2 = 50)
mhi1_preds <- mhi1_pred_contr(mhi1.post, nd1 = 4000, nd2 = 50)
mhi2_preds <- mhi2_pred_contr(mhi2.post, nd1 = 4000, nd2 = 50)
mhi3_preds <- mhi3_pred_contr(mhi3.post, nd1 = 4000, nd2 = 50)


p1 <- bind_rows(
  posterior_draws(mhi0_preds$pred) |> 
  data.frame() |> 
    mutate(model = "mhi0"), 
  
  posterior_draws(mhi1_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |>
    mutate(model = "mhi1"), 
  
  posterior_draws(mhi2_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
    mutate(model = "mhi2"), 
  
  posterior_draws(mhi3_preds$pred) |> 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
    mutate(model = "mhi3")
  ) |> 
 
  ggplot(aes(time, draw, group = paste(drawid, tx, sex), 
             color = sex)) + geom_line() + 
  facet_grid(model ~ tx)





p2 <- bind_rows(
  posterior_draws(mhi0_preds$contr) |> 
  mutate(model = "mhi0"), 
    posterior_draws(mhi1_preds$contr) |> 
  mutate(model = "mhi1"), 
           posterior_draws(mhi2_preds$contr) |> 
  mutate(model = "mhi2"), 
  posterior_draws(mhi3_preds$contr) |> 
  mutate(model = "mhi3")) |>  
  
  ggplot(aes(draw, color = model)) + geom_line(stat = "density") + 
  facet_wrap(~ hypothesis) +
  geom_vline(xintercept = 0, lty = 2)



############### Predictions for plotting #####################





# These are looking a bit nonsensical, so I guess something here needs more work, but I cant see what at 23:27 in the evening.

p1 <- posterior_draws(mhi2_preds$pred) |> 
    mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) |> 
  
  #mutate(draw = draw * sd_corr + mean_corr, 
         #draw = exp(draw)) |> 
  
  
  summarise(.by = c(tx, sex, time), 
            m = mean(draw), 
            lwr50 = quantile(draw, 0.25), 
            upr50 = quantile(draw, 0.75),  
            lwr90 = quantile(draw, 0.05), 
            upr90 = quantile(draw, 0.95),
            lwr95 = quantile(draw, 0.025), 
            upr95 = quantile(draw, 0.975)) |> 
  
  
  ggplot(aes(time, m, color = sex, group = paste(tx, sex))) + 
  

  
  geom_segment(aes(y = c(lwr50), yend = c(upr50), 
                   x = time, xend = time), 
               
              linewidth = 2) +
    geom_segment(aes(y = c(lwr90), yend = c(upr90), 
                   x = time, xend = time), 
             
              linewidth = 1.5) +
    geom_segment(aes(y = c(lwr95), yend = c(upr95), 
                   x = time, xend = time), 
          
              linewidth = 1) +

  geom_line() + 
  
    geom_point(size = 3, shape = 21, 
             aes(color = NULL, fill = sex)) +
  
  facet_wrap(~ tx, ncol = 5)



p2 <- posterior_draws(mhi3_preds$contr) |> 
 
  
  #mutate(draw = draw * sd_corr , 
  #       draw = exp(draw)) |> 
  
  ggplot(aes(draw)) + geom_line(stat = "density") + 
  facet_wrap(~ hypothesis) +
  geom_vline(xintercept = 1, lty = 2) 





## Posterior probability
# The code below produces a table with hypothesis and posterior probability of direction, the proportion of posterior draws where the effect exceeds 1. A pos value of thus 1 means 100% of posterior draws showed the ratio exceeding 1 for that contrast, very strong evidence of a positive effect. Values at 0.5 on the other hand, are essentially a coin flip, no credible directional effect. Useful as a supplementary statistic, perhaps?


## Here I added mean, lower and upper to the summarise to get the mean ratio, as well as 95% ETI in addition to pos. 
 
posterior_draws(mlo2_preds$contr) |>
  mutate(draw = draw * sd_cmyc_low,
         draw = exp(draw)) |>
  summarise(.by = hypothesis,
            mean = mean(draw),
            lower95 = quantile(draw, 0.025),
            upper95 = quantile(draw, 0.975),
            pos = sum(draw > 1) / n())


# Since these are ratios (from the exp() transform), the null value is 1, not 0 - i.e., its about whether the 95% ETI excludes 1, not zero. These mean/CI tables should be reported and intepreted as ratios (fold-changes), e.g. "mean ratio = 1.15, 95% ETI [1.08, 1.23] - interpretable as "a 15% increase", not as an absolute cm3 diff. 

# For figure plotting


pred_cmyc_high <- posterior_draws(mhi2_preds$pred) |>
  mutate(
    sex = if_else(grepl("_female", tx_sex), "female", "male"),
    tx = gsub("_female", "", tx_sex),
    tx = gsub("_male", "", tx),
    tx = factor(tx, levels = c("int_old_low", "int_old_mod", "int_yng_low", "int_yng_mod", "con_old_con"))
  ) |>
  mutate(draw = draw * sd_cmyc_high + mean_cmyc_high,
         draw = exp(draw)) |>
  summarise(.by = c(tx, sex, time),
            m = mean(draw),
            lwr50 = quantile(draw, 0.25),
            upr50 = quantile(draw, 0.75),
            lwr90 = quantile(draw, 0.05),
            upr90 = quantile(draw, 0.95),
            lwr95 = quantile(draw, 0.025),
            upr95 = quantile(draw, 0.975))

saveRDS(pred_cmyc_high, "data/data-gen/pred_cmyc_high.rds")


contrast_cmyc_high <- posterior_draws(mhi2_preds$contr) |>
  mutate(draw = draw * sd_cmyc_high, draw = exp(draw)) |>
  summarise(.by = hypothesis,
            mean = mean(draw),
            lower95 = quantile(draw, 0.025),
            upper95 = quantile(draw, 0.975),
            pos = sum(draw > 1) / n()) |>
  mutate(pct_change = (mean - 1) * 100,
         pct_lower  = (lower95 - 1) * 100,
         pct_upper  = (upper95 - 1) * 100,
         credible   = ifelse(lower95 > 1 | upper95 < 1, "Yes", "No"))

saveRDS(contrast_cmyc_high, "data/data-gen/contrast_cmyc_high.rds")



```







-->







