## Methods


```{r}
#| label: fig-study-design
#| fig-width: 6.85
#| fig-height: 6.3


source("figures/fig-study-design.R")   # produces fig1

fig1
```
_Figure 1: Study overview and weekly training volume. (A) Participant recruitment and randomization flow, and testing modalities used throughout the study: dual-energy x-ray absorptiometry / ultrasound (DXA/US), muscle biopsy, oral glucose tolerance test (OGGT), magnetic resonance imaging (MRI), strength testing (STR), and resistance training (RT). LV = low volume, MV = moderate volume, CON = non-training control. (B) Weekly training volume load (kg × repetitions) for the leg press exercise across the intervention, shown separately by age group (young, old) and training volume condition (low, moderate; solid vs. dashed lines respectively). Shaded ribbons represent the 95% confidence interval around the weekly mean. Vertical dashed lines and accompanying icons mark key study timepoints: baseline testing (week 1) and the start of training (week 2), a muscle biopsy at week 4, mid-intervention testing (week 6.5), and post-intervention testing (week 12); icons above each timepoint indicate which assessments were performed._

<!--
Question: Should I use the weekly volume for all exercises, or is it sufficient to use leg press only?
-->

### Study Overview

This study was conducted as an age-stratified within-participant trial employing a unilateral design. All participants performed a whole-body unilateral training program, with training protocols (low and moderate) randomly assigned to either leg/arm. The intervention lasted for 12 weeks, with 10 weeks of exercising and two weeks of testing (Figure 1). The participants exercised for a total of 24 training sessions, two sessions per week for the first five weeks, increased to three sessions per week for the five last weeks, including a taper down to two sessions for the 10th and final week. The effects of the intervention on skeletal muscle mass was measured as thigh muscle volume (MRI: baseline, post), thigh muscle thickness (ultrasound: baseline, half-way, post) and lean body mass (DXA: baseline, half-way, post). Skeletal muscle strength was assessed as peak torque with the humac dynamometer, 1RM leg-press with the Keiser Airpress and maximal voluntary bicep curl contraction with the MuscleLab system (all baseline, half-way, post). Glucose tolerance was measured by the oral glucose tolerance test at baseline and post-intervention. Molecular data was gathered via micro biopsy at baseline, after three weeks of RT (six training session) and post-intervention. All tests were performed at the same time of day (± 2hrs), and measurement of lean body mass, muscle thickness, glucose tolerance, and sampling of muscle tissue was conducted in an overnight fasted state. Participants were asked to maintain their habitual physical activity level throughout the intervention, and to not perform any additional type of heavy exercise.

All participants gave their written informed consent before study enrollment. The study was approved by the Regional Committee for Medical and Health Research Ethics - South-East Norway (ID nr. 255224), pre-registered at clinicaltrials.gov (Identifier: NCT05063279), and conducted according to the Helsinki Declaration.


### Participants

Thirty-two young (24 ± 3.1 years) and 49 old (75.5 ± 5.6 years) adults were recruited locally; 10 of the old participants were randomly allocated to a control group. Inclusion criteria were age 20–30 or above 70 years (Figure 1). Exclusion criteria were: insufficient Norwegian proficiency; RT more than once weekly and/or endurance training three or more times weekly within the last six months; unstable cardiovascular disease; illness or injury contraindicating RT; serious mental illness; or allergy to local anaesthesia.


### Resistance training protocol

Resistance training was performed as a whole-body training program. Each RT-session contained three lower-body exercises (unilateral leg press, unilateral knee extension and -flexion) and three upper-body exercises (unilateral chest-press, unilateral bent-over rows, and unilateral bicep curls), using individualized settings in all exercise apparatuses. The low training volume condition was performed as a single set of 10RM per exercise, and the moderate training volume condition was performed as three sets of 10RM per exercise. As a general warm-up, the participants cycled on an indoor exercicse bicycle for 5-10 minutes. Before lower-body exercises the participants performed 1 set of 10 repetitions in bilateral leg press, and before upper-body exercises they performed a warm-up set of 10 push-ups, modified to their individual fitness level (e.g., standard, from the knees, or against a wall). To ensure progressive loading and an adequate stimulation of the skeletal muscle, the exercise load was increased in the following set if the participants lifted more than 12 repetitions. Resting time between each set was two minutes. For safety and standardization purposes, all sessions were monitored by trained personnel. To ensure sufficient recovery, at least one day separated training days throughout the intervention.


### Measurement of body mass composition

MRI was measured in a non-fasted state, in accordance with the manufacturer’s protocol (S-Scan, Esaote Europe B.V., Maastricht, the Netherlands). The images were analysed using the 3D Slicer software (SOURCE). The layering details within 3D Slicer software were the same for all the images (fat = 300-max, muscle = 1-300, bone = 0-750, and intra-muscular fat = 300-max), as was the calculation of the area to be analyzed (approx. 20% from the trochanter major proximally and 30% from the knee-joint distally). Muscle thickness of m. vastus lateralis and m. rectus femoris were measured using B-mode ultrasonography (SmartUs EXT-1 M, Telemed, Vilnius, Lithuania) with a 39 mm 12 MHz, linear array probe. Transverse images were obtained ~60% distally from the trochanter major towards the femoral lateral epicondyle. Three images were captured for each muscle at each time point, where the probe was relocated to the same position between each image. The position of the probe was marked on the skin and subsequently marked on a soft transparent plastic sheet superimposed on the thigh. Landmarks were marked on the plastic sheet for relocation of the scanned areas during post-training measurements. During analysis, pre and post images from the same participant were analysed consecutively using the Fiji software [@schindelin2012fiji]. The average muscle thickness of the three images captured per muscle per time point was used for further analyses. Dual-energy x-ray absorptiometry (Lunar Prodigy, GE Healthcare, Madison, WI, 320 USA) scans were analyzed with the manufacturer's software, where the region of interest (RoI) were matched on baseline, half-way and post images. 


### Assessment of muscular strength

After a standardized warm-up, strength assessments were performed using a standardized sequence; unilateral knee extension torque (legs), unilateral 1RM leg press, and unilateral maximal voluntary contraction in the bicep curl, using individualized settings. At each time point (baseline, halfway, post), muscle strength was assessed twice, on two separate days. During baseline testing, the first of the two testing days represented familiarisation. The participants had at least one whole day of rest between the previous training session and assessment of muscular strength.

*Knee extension torque*
Maximal isometric and isokinetic knee extension torque were measured with a Humac Norm Dynamometer (CSMi, Stoughton, Massachusetts, USA). Isokinetic peak torque was measured concentrically from 90 to 0 degrees knee angle (extended knee was set to 0 degrees) at angular velocities of 60-, 120- and 240 degrees per second, 2x3 repetitions each, with the first set of each exercise as a sub-maximal warm-up. Isometric knee extensor peak torque was measured at a knee angle of 60 degrees, for a maximum of 10 seconds and two repetitions per test. The isometric tests were concluded when the participants reached a plateau or peak torque development decreased, which on average occurred between 2-4 seconds into the test. Between sets of the same angular velocity the resting time was 30 seconds, and between angular velocities (i.e. from 60- to 120 degrees per second) the resting time was 1 minute.

*One repetition maximum*
The 1RM test was measured with a Keiser AIR300 Leg press (Keiser Corporation, Fresno, USA). As a specific warm-up, the participants performed 4 repetitions at ~50% of 1RM. Maximal one repetition leg press strength was defined as the maximal load lifted in a controlled fashion, starting from a knee angle of 90 degrees, within 5kg. From the ~50% of 1RM warm-up, the load was increased at most by increments between 10-40kg for the initial attempts, and at least between 5-10kg when the participants signified being near their maximal effort. Resting time between 1RM attempts was 30 seconds. 

*Maximal isometric bicep curl*
Maximal isometric force of the bicep muscle was measured with a MuscleLab force sensor (Ergotest Innovation AS, Stathelle, Norway), at a fixed elbow angle of 90 degrees. The participants were given one warm-up attempt per arm, and two attempts to perform their maximal force development in the bicep exercise. The isometric tests were concluded when the participants reached a plateau, peak torque development decreased (which on average occurred between 2-4 seconds into the test), or after 10 seconds.


### Skeletal muscle sampling and analyses

Muscle biopsies were collected from m. vastus lateralis. The biopsies were sampled under antiseptic and anaesthetised conditions (Lidokain 10 mg ml-1, Mylan Hospital AS, Oslo Norway), using the micro-biopsy technique with a 12-14 gauge needle (Universal Plus, Mermaid medical A/S, Stenløse, Denmark) operated with a spring loaded biopsy gun (Bard Magnum, Bard Norway A/S, Oslo, Norway). Samples were divided into four aliquots for RNA, protein, immunohistochemistry and DNA analyses. At week 3 and 12 measurements, there were 48 hours between the previous training session and biopsy sampling. 


#### Total RNA extraction

Total RNA was extracted using TRIzol from muscle tissue homogenised using 0.5mm RNase-free Zirconium beads (\~50 ul; Next Advanced, Averill Park, NY, USA) and mechanical agitation (Bullet Blender, Next Advanced, Averill Park, NY, USA). Chloroform (Sigma-Aldrich, Oslo, Norway) was used for phase separation, and the RNA pellet was precipitated with isopropanol (VWR International, Oslo, Norway). For assessment of RNA content and purity, RNA was eluted in TE buffer (1:2) and assessed via spectrophotometry. Samples with a 260/280 ratio below 1.8 were excluded. The RNA stock was stored at -80°C until further analyses.


#### RNA sequencing

```{r}
#| code-fold: true

library(dplyr)
library(english)
library(reliefdata)
library(patchwork)

# Participants / samples -- Metadata
metadata <- reliefdata::relief_seqsamples |> 
  mutate(participant = as.character(participant)) |> 
  dplyr::rename(sample_weight = weight) |> 
  inner_join(relief_participants ) |> 
  inner_join(relief_volume |> 
               dplyr::select(-arm) |> 
              mutate(participant = as.character(participant)) ) |> 
  dplyr::select(participant, time, age, leg,sex, condition, allocation, sample_weight, seq_sample_id) |> 
  mutate(age = if_else(age < 40, "young", "old")) |> 
  arrange(seq_sample_id)

metadata <- metadata |>
  mutate(time = if_else(
    participant == "3003" & time == "t2" & leg == "L" & condition == "mod",
    "t1",
    time
  )) # t1 L sample wrongfully labelled as t2 for this participant, this fixes it short-term, but it should be looked into closer. 



# Read gene-level data 
dat <- readRDS("data-raw-rnaseq/relief_gene.RDS"); dat <- dat$expected_count


# Number of samples to RNA-seq
nsamp <- stringr::str_to_sentence(as.english(nrow(metadata)))
nsamp_included <- (ncol(dat) - 1)



```


`r nsamp` total RNA samples were selected for RNA-sequencing analysis of which `r nsamp_included` samples passed initial quality assessment (RIN > 8) and were used in downstream analyses. The samples were selected from the pool of intervention participants that had both baseline and post samples collected. Purified mRNA (poly-T oligo-attached magnetic beads) was used for library preparation and subsequent 150-bp paired-end sequencing (Illumina NovaSeq, Novogene Europe, Cambridge, United Kingdom). Raw reads were filtered (SOFTWARE, settings) and aligned to the Human genome (Release) followed by gene and transcript level quantification using RSEM (version).

Gene counts (filtered for low-abundance targets) were modeled using a regularized negative binomial generalized linear mixed model [@glmmmtb] including model parameters for sex, time (baseline and post-intervention), age-group (young and old) and condition (low- and moderate-volume) and the interaction between time age and condition. A varying intercept (random effect) was used for participant, and the dispersion parameter was estimated using the `nbinom2` parameterization available in `glmmTMB`[@glmmmtb]. The effective library size [@robinson2010;@edger] was included as an offset to account for sequencing depth. Regularization through an empirical Bayes shrinkage approach[@efron2010] was achieved using a two-step approach. First, naive models were fitted for all target counts (gene/transcript) using our in-house pipeline [@seqwrap] and estimates for all model parameters (except the intercept) were used to create priors for subsequent informed models. Fixed effect priors were constructed using a normal distribution with mean and standard deviations from each parameter value from the first model pass, the prior for the varying intercepts were constructed using a gamma distribution with the shape parameter informed by the mean from the naive model estimates and the  scale parameter fixed to 2. Finally, the dispersion parameter priors were constructed by fitting a trend model (Local Polynomial Regression Fit) to the association between raw mean log counts and log dispersion across all targets. The predicted dispersion at target-specific raw mean log count was used as the mean and the model residual standard error was used as the standard deviation to define a normally distributed target-specific prior.


### Protein extraction and immunoblotting

The wet muscle tissue was freeze-dried for 24h and dissected, then homogenised with a plastic rod in an ice cold lysis buffer. Briefly, 150ul of RIPA buffer was added per mg muscle tissue dry weight, with 1ul protease/phosphatase inhibitors per 100ul lysis buffer. The samples were kept on ice at all times except for when manually disrupting to homogenise the tissue, and then incubated for 30 min on a rotator while on ice. Following the incubation, the samples were spun in the centrifuge for 40 min, at 15 000g at 4°C. Total protein was then diluted to 2ug/ul stock and aliquoted to run the samples in triplicates for protein concentration determination and for western blotting. Protein concentration was determined in a 1:10 dilution (Pierce Detergent Compatible Bradford Assay Reagent, Thermo Fisher Scientific, Oslo, Norway).

Western blots were done using the Jess Automated Western Blot System (Bio-Techne, Minneapolis, USA), according to the manufacturer's protocol, optimized for our lab. The stock solution (2ug/ul) was prepared with fluorescent master mix from the EZ standard pack (Bio-Techne) and 0.1X sample buffer (Bio-Techne) and thus diluted to 0.75 ug/ul for the assay. The assay samples were then briefly vortexed, denatured at 95°C, vortexed and spun down prior to loading the assay plate. To control for within-plate variation, we used a total protein normalization reagent (Bio-Techne) for all assays. A specific between-plate normalization was not used, however, all samples from one participant were analyzed on the same plate. Primary antibodies were purchased from Santa Cruz Biotechnology (Texas, USA): UBF 1:50, UBF F-9, sc-13125; and Cell Signaling Technology (Massachusetts, USA): c-Myc 1:15, c-Myc D84C12. Ready-to-use secondary antibodies (anti-mouse and anti-rabbit) were purchased from Bio-Techne. After loading, the assay plate was centrifuged at 2500 rpm for 5 min at room temperature. The assay was then run on the Jess system, with the chemiluminescence detection module.


### Oral Glucose Tolerance Test

The oral glucose tolerance test was completed in an overnight-fasted state, where participants ingested a bolus of 75g of glucose (Glucosum monohydricum, Merck KGaA, Darmstadt, Germany) mixed with 250ml water. Capillary blood was gathered from the participants' fingers and analyzed with in-house equipment (BIOSEN, C-Line, EKF diagnostic GmbH, Barleben). The glucose bolus was ingested after the baseline blood draw, and subsequent blood draws were taken at 15, 30, 45, 60, 90, and 120 minutes after glucose ingestion.


### Statistical analyses

All statistical analyses were performed in R (version 4.5.1; R Core Team, 2025) using the brms package (version 2.23.0; Bürkner, 2017) for Bayesian multilevel model fitting, with Stan (via rstan, version 2.32.7, and StanHeaders, version 2.32.10) as the sampling backend, and the marginaleffects package (version 0.32.0; Arel-Bundock, 2023) for posterior predictions and hypothesis testing. Primary models were fit with a fixed-effects structure of tx_sex (a combined age group × condition × sex factor) interacted with time, and a random-effects structure of (1 + time | participant) together with a limb-level random intercept (participant:leg or participant:arm, as appropriate) to account for the bilateral, within-participant sampling design. All continuous outcomes were log-transformed and standardized (mean-centered, scaled by their own standard deviation) before modelling, with weakly informative priors specified on the standardized scale. Model convergence was assessed using the Gelman-Rubin statistic (Rhat ≤ 1.01), effective sample size (bulk and tail ESS ≥ 400), and the number of divergent transitions after warm-up; models showing elevated divergences were re-fit with an increased target acceptance rate (adapt_delta up to 0.999) and additional sampling iterations, with final diagnostics reported alongside each model. Posterior contrasts were computed on the exponentiated (back-transformed) scale and are reported as ratios, interpretable as fold-changes, with a null value of 1. For each contrast we report the posterior mean, the 95% equal-tailed credible interval (ETI), and the posterior probability of direction (the proportion of posterior draws favouring an increase, denoted pos). A contrast was considered credible where the 95% ETI excluded the null value. Where models included both male and female participants, group-level estimates were computed as an equal-weighted average across sexes (i.e., not weighted by the observed male/female sample size within each group).

1RM leg press and bicep curl maximum voluntary contraction were modelled using the same Bayesian multilevel framework as the body composition outcomes described above, at pre-, mid-, and post-intervention timepoints. Isokinetic and isometric peak torque was modelled using a spline-based fixed-effects structure across the continuous angular velocity range (0–240°/s, normalized to 0–1), with condition-by-angular velocity and time-by-angular velocity interactions specified as population-level effects for allocation (intervention vs. control), and the full age × volume-condition group structure (tx_sex) included as a varying (partially pooled) effect together with its own angular velocity and time interactions. Random effects included participant-level intercepts and angular velocity slopes by time point, together with a participant:time:timerep random intercept to account for repeated within-session trials. Unlike the outcomes above, peak torque was not log-transformed; group-level contrasts are reported on the raw (additive) scale in Nm, with a null value of zero rather than one. The oral glucose tolerance test was modelled using a spline-based fixed-effects structure across sample time within the same Bayesian multilevel framework, with participant-level random splines. Total accumulated training volume load (kg × repetitions, summed across sets and sessions) was calculated descriptively by age group and condition and is reported as a manipulation check, without formal statistical modelling. Muscle volume was estimated from magnetic resonance imaging (MRI) at two time points (pre- and post-intervention). Muscle thickness was estimated from ultrasound imaging of m. vastus lateralis, measured at three time points (pre-, mid-, and post-intervention). Upper-body lean mass was estimated from dual-energy X-ray absorptiometry (DXA), also at three timepoints. All three body composition outcomes were modelled using the same Bayesian multilevel framework described above.

Western blot protein targets (UBF, c-Myc) were modelled using the same Bayesian multilevel approach described above, at the baseline and week 3 time points. Total RNA yield was modelled as a function of tissue mass rather than standardized to a fixed RNA-per-milligram ratio. The fixed-effects structure included the age group × allocation × training volume factor (tx) interacted with time (pre, week 3, post), together with standardized log-transformed tissue mass (mgstd) as an additive covariate, using an informed prior (mgstd ~ Normal(1, 1)) reflecting the consistently positive, sub-unity slope estimated across preliminary models. Unlike the other outcomes, RNA was modelled using tx rather than tx_sex, as sex was not modelled as a separate factor for this outcome. The random-effects structure comprised a participant-level random intercept and time slope ((1 + time | participant)), a leg-level random intercept (participant:leg) to account for the bilateral, within-participant sampling design, and a replicate-level random intercept (participant:sample) nested within participant, since RNA concentration was measured across multiple technical replicates per tissue sample. 

Gene-level RNA-sequencing counts (filtered for low-abundance targets using edgeR::filterByExpr [@edger], and normalized using the trimmed mean of M-values method [@robinson2010]) were modelled per-gene using a regularized negative binomial generalized linear mixed model (glmmTMB [@glmmmtb], nbinom2 family), with fixed effects for sex, age group, time (baseline and post-intervention), training volume, and their interactions, a participant-level random intercept, and the TMM-normalized effective library size included as an offset. Regularization was achieved through an empirical Bayes shrinkage procedure [@efron2010]: naive per-gene models were fit first using an in-house pipeline (seqwrap [@seqwrap]), and their parameter estimates used to construct informative priors — normally distributed for fixed effects, gamma-distributed for the participant-level variance, and derived from a local polynomial trend for the dispersion parameter — for a second, regularized model pass. Group-level contrasts were computed on the linear predictor scale using the marginaleffects package [@marginaleffects], with sex aggregated as an equal-weighted average across both levels, consistent with the approach used for the other outcomes described above. Gene-set enrichment analysis (Gene Ontology, Biological Process) was performed on the resulting per-gene contrast statistics using a rank-based method (clusterProfiler [@xu2024a]), with genes ranked by their contrast test statistic.