---
title: ""
format: html
bibliography: Relief-bibliography.bib
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

```
 


```{r}
library(here)
library(tidyverse)   
library(ggdist)      
library(patchwork)   
library(cowplot)     


```



## Results
<!--
Missing: 
- RNA seq analysis is under revision, results text/fig may need updating after. 
- Should we extend the volume load analysis? This is currently only a summary of the total load, and how much more mod lifted than low.
- Consider being way more strict with the usage of "increase" etc. Is it an increase/decrease/difference if its not credible?
-->
### Exercise Volume Load

Moderate volume exercise was estimated to ~2.8× higher than low in both age groups (young mod: 137438.7 ± 54491.7 kg, young low: 50693.5 ± 25707.7 kg, old mod: 118124.8 ± 38224.6 kg, old low: 42134.1 ± 14352.5 kg).

### Skeletal Muscle Hypertrophy

```{r}
#| label: fig-muscle-volume
#| fig-cap: "Posterior distribution of five between-group contrasts for the ratio (post/pre) of thigh muscle volume. Each row shows the full posterior density, with the point marking the mean ratio and the black bars the 50% and 95% CI. The dashed line marks no change (ratio = 1); the density is shaded blue where the ratio exceeds 1. The percentage label gives the posterior probability that the ratio exceeds 1 (pos), as a measure of directional confidence."
#| fig-width: 10
#| fig-height: 6

readRDS(here("figures", "fig-muscle-volume-combined.RDS"))

#use fig-muscle-volume.RDS for the manuscript version

```

_Thigh Muscle Volume_
Older adults in the intervention group (averaged across volume conditions) increased muscle volume by 3.7% (95% CI [1.8%, 5.8%], P(>0) > 99.9%), while there was no evidence of a change in the control group (0.0%, 95% CI [−3.0%, 3.1%], P(<0) = 50.0%). The difference between them was 3.8% (95% CI [0.0%, 7.6%], P(>0) = 97.6%). Young adults (averaged across volume conditions) increased by 6.0% (95% CI [3.9%, 8.1%], P(>0) > 99.9%), likely more than older adults, although the 95% CI included zero (2.2%, 95% CI [−0.5%, 5.0%], P(>0) = 94.0%). Within young adults, low-volume training increased muscle volume by 5.0% (95% CI [2.9%, 7.2%], P(>0) > 99.9%) and moderate-volume training by 7.0% (95% CI [4.9%, 9.2%], P(>0) > 99.9%), with a larger increase under moderate volume (1.9%, 95% CI [0.9%, 2.8%], P(>0) > 99.9%). In older adults, low-volume training increased muscle volume by 2.9% (95% CI [1.0%, 5.0%], P(>0) = 99.9%) and moderate-volume training by 4.6% (95% CI [2.6%, 6.6%], P(>0) > 99.9%), again with a larger increase under moderate volume (1.6%, 95% CI [0.6%, 2.5%], P(>0) = 99.9%). There was no evidence that the volume effect differed between age groups (0.3%, 95% CI [−1.0%, 1.6%], P(>0) = 68.5%).

```{r}
#| label: fig-muscle-thickness
#| fig-cap: ""
#| fig-width: 10
#| fig-height: 6

readRDS(here("figures", "fig-muscle-thickness.RDS"))

#this is only for the supplementary version

```

_M. Vastus Lateralis Thickness_
Older adults in the intervention group (averaged across volume conditions) increased muscle thickness by 7.6% from baseline to post (95% CI [3.5%, 11.9%], P(>0) > 99.9%), while there was only weak evidence of an increase in the control group (3.4%, 95% CI [−3.3%, 10.5%], P(>0) = 83.4%). There was also only weak evidence of a difference between them (4.2%, 95% CI [−3.7%, 12.4%], P(>0) = 84.8%). By the half-way point, the intervention group had increased by 5.3% (95% CI [1.2%, 9.4%], P(>0) = 99.7%), with weak evidence of an increase in the control group (3.6%, 95% CI [−3.7%, 11.0%], P(>0) = 82.7%) and no evidence of a difference between them (1.8%, 95% CI [−5.9%, 10.4%], P(>0) = 66.0%). Young adults (averaged across volume conditions) increased by 10.5% (95% CI [6.5%, 14.9%], P(>0) > 99.9%), with weak evidence of a larger increase than in older adults (2.8%, 95% CI [−2.7%, 8.6%], P(>0) = 83.8%). The same was true at the half-way point (young: 8.9%, 95% CI [4.7%, 13.2%], P(>0) > 99.9%; difference: 3.5%, 95% CI [−2.1%, 9.4%], P(>0) = 88.7%). Within young adults, low-volume training increased muscle thickness by 9.0% (95% CI [3.8%, 14.2%], P(>0) > 99.9%) and moderate-volume training by 12.1% (95% CI [7.0%, 17.5%], P(>0) > 99.9%), with weak evidence of a larger increase under moderate volume (2.9%, 95% CI [−2.6%, 8.7%], P(>0) = 83.9%). In older adults, moderate-volume training increased muscle thickness by 12.2% (95% CI [6.7%, 17.7%], P(>0) > 99.9%), whereas there was only weak evidence of an increase with low-volume training (3.2%, 95% CI [−1.8%, 8.2%], P(>0) = 89.8%). The increase was larger under moderate volume (8.8%, 95% CI [2.9%, 14.9%], P(>0) = 99.8%). The same pattern was present at the half-way point, with weak evidence of a larger increase under moderate volume in young adults (3.3%, 95% CI [−2.1%, 8.9%], P(>0) = 88.4%) and a larger increase under moderate volume in older adults (7.3%, 95% CI [1.6%, 13.0%], P(>0) = 99.7%). The volume effect was likely larger in older than in young adults at post, although the 95% CI included zero (−5.4%, 95% CI [−12.7%, 2.4%], P(<0) = 91.9%), with weak evidence at the half-way point (−3.6%, 95% CI [−10.7%, 3.8%], P(<0) = 84.0%).


```{r}
#| label: fig-ub-lean-mass
#| fig-cap: ""
#| fig-width: 10
#| fig-height: 6

readRDS(here("figures", "fig-ub-lean-mass.RDS"))

#this is only for the supplementary version

```

_Upper body lean mass_
Older adults in the intervention group (averaged across volume conditions) tended to increase upper-body lean mass by 1.2% from baseline to post (95% CI [−0.3%, 2.7%], P(>0) = 93.4%). There was no evidence of a change in the control group (0.8%, 95% CI [−1.9%, 3.6%], P(>0) = 72.8%) and no evidence of a difference between the groups (0.3%, 95% CI [−2.8%, 3.5%], P(>0) = 57.8%). At the half-way point, there was weak evidence of an increase in the intervention group (0.7%, 95% CI [−0.5%, 2.0%], P(>0) = 86.8%) and weak evidence of a decrease in the control group (−0.8%, 95% CI [−3.1%, 1.6%], P(<0) = 75.6%), with weak evidence of a difference between them (1.6%, 95% CI [−1.1%, 4.3%], P(>0) = 87.8%). Young adults (averaged across volume conditions) increased upper-body lean mass by 3.1% (95% CI [1.5%, 4.8%], P(>0) > 99.9%), likely more than older adults, although the 95% CI included zero (1.9%, 95% CI [−0.2%, 4.2%], P(>0) = 95.5%). At the half-way point, young adults tended to have increased (1.3%, 95% CI [−0.1%, 2.7%], P(>0) = 96.6%), with no evidence of a difference from older adults (0.6%, 95% CI [−1.3%, 2.5%], P(>0) = 74.1%). Within young adults, low-volume training increased upper-body lean mass by 3.3% (95% CI [1.3%, 5.3%], P(>0) > 99.9%) and moderate-volume training by 2.9% (95% CI [0.9%, 4.9%], P(>0) = 99.8%), with no evidence of a difference between them (−0.4%, 95% CI [−2.7%, 1.9%], P(<0) = 63.1%). In older adults, there was weak evidence of an increase with low-volume training (1.1%, 95% CI [−0.7%, 3.0%], P(>0) = 89.4%), and moderate-volume training likely increased it (1.2%, 95% CI [−0.6%, 3.1%], P(>0) = 90.0%), with no evidence of a difference between them (0.1%, 95% CI [−1.9%, 2.2%], P(>0) = 52.0%). There was no evidence of a volume difference at the half-way point in either young (−0.2%, 95% CI [−2.5%, 2.0%], P(<0) = 57.5%) or older adults (−0.6%, 95% CI [−2.7%, 1.4%], P(<0) = 71.8%). There was also no evidence that the volume effect differed between age groups (half-way point: 0.4%, 95% CI [−2.6%, 3.6%], P(>0) = 59.5%; post: −0.4%, 95% CI [−3.4%, 2.6%], P(<0) = 61.6%).


### Skeletal Muscle Strength

```{r}
#| label: fig-force-velocity
#| fig-cap: "Model-predicted peak isokinetic and isometric torque as a function of angular velocity (force–velocity relationship) in young and older adults across the resistance-training intervention. Lines show posterior median predictions from the Bayesian multilevel model, colored by training volume (low/moderate) and styled by timepoint (pre/mid/post). Top left: young intervention participants. Top right: older intervention participants. Bottom left: older passive control. Bottom right: pooled contrasts (intervention vs. control, and young vs. old, both pooled across volume condition) at the four tested angular velocities, shown as posterior mean change from pre-intervention (Δ, Nm) with 95% credible intervals, at the mid- and post-intervention timepoints."
#| fig-width: 11
#| fig-height: 12

readRDS(here("figures/fig-force-velocity.rds"))

```

_Isometric Force and Isometrick Peak Torque_
In older adults, peak torque in the intervention group (averaged across volume conditions) increased by 12.4%–19.0% from baseline to post, with larger increases at faster angular velocities (isometric: 12.4%, 95% CI [9.1%, 16.1%], P(>0) > 99.9%; 60°/s: 13.2%, 95% CI [9.3%, 17.5%], P(>0) > 99.9%; 120°/s: 14.6%, 95% CI [9.4%, 20.3%], P(>0) > 99.9%; 240°/s: 19.0%, 95% CI [11.0%, 27.6%], P(>0) > 99.9%). The control group also tended to increase (6.8%–9.2%), although the 95% CI excluded zero only for isometric torque (isometric: 8.6%, 95% CI [0.3%, 17.7%], P(>0) = 97.8%; 60°/s: 7.5%, 95% CI [−1.8%, 17.8%], P(>0) = 94.1%; 120°/s: 6.8%, 95% CI [−5.3%, 20.7%], P(>0) = 85.4%; 240°/s: 9.2%, 95% CI [−10.1%, 32.4%], P(>0) = 80.9%). There was only weak evidence that the intervention group increased more than the control group (3.7%–10.1%; isometric: 3.7%, 95% CI [−5.1%, 12.7%], P(>0) = 79.2%; 60°/s: 5.5%, 95% CI [−4.6%, 16.1%], P(>0) = 85.3%; 120°/s: 7.7%, 95% CI [−5.7%, 21.9%], P(>0) = 86.3%; 240°/s: 10.1%, 95% CI [−11.2%, 33.9%], P(>0) = 81.0%). By the half-way point, the intervention group had already increased by 9.9%–14.4% (isometric: 9.9%, 95% CI [6.6%, 13.4%], P(>0) > 99.9%; 60°/s: 11.0%, 95% CI [7.1%, 15.2%], P(>0) > 99.9%; 120°/s: 12.4%, 95% CI [7.2%, 17.9%], P(>0) > 99.9%; 240°/s: 14.4%, 95% CI [6.7%, 22.4%], P(>0) > 99.9%), whereas there was only weak evidence of a change in the control group (3.9%–7.6%; isometric: 3.9%, 95% CI [−4.1%, 12.4%], P(>0) = 82.4%; 60°/s: 4.9%, 95% CI [−4.4%, 14.9%], P(>0) = 85.1%; 120°/s: 6.1%, 95% CI [−5.9%, 19.9%], P(>0) = 83.4%; 240°/s: 7.6%, 95% CI [−11.2%, 30.3%], P(>0) = 76.5%). The intervention–control difference at this point (6.0%–7.3%) tended to favour the intervention for isometric torque, with weak or no evidence at faster velocities (isometric: 6.0%, 95% CI [−2.7%, 15.4%], P(>0) = 90.8%; 60°/s: 6.1%, 95% CI [−4.1%, 17.0%], P(>0) = 88.2%; 120°/s: 6.3%, 95% CI [−7.0%, 20.4%], P(>0) = 82.8%; 240°/s: 7.3%, 95% CI [−13.2%, 30.5%], P(>0) = 74.8%).

Young adults (averaged across volume conditions) increased peak torque by 14.2%–21.8% (isometric: 14.2%, 95% CI [10.7%, 18.0%], P(>0) > 99.9%; 60°/s: 15.4%, 95% CI [11.1%, 19.9%], P(>0) > 99.9%; 120°/s: 17.2%, 95% CI [11.6%, 23.2%], P(>0) > 99.9%; 240°/s: 21.8%, 95% CI [13.2%, 30.9%], P(>0) > 99.9%). There was only weak evidence that this increase was larger than in older adults (1.6%–2.5%; isometric: 1.6%, 95% CI [−1.2%, 5.0%], P(>0) = 84.7%; 60°/s: 1.9%, 95% CI [−1.7%, 6.0%], P(>0) = 83.9%; 120°/s: 2.3%, 95% CI [−2.4%, 7.5%], P(>0) = 82.4%; 240°/s: 2.5%, 95% CI [−4.4%, 10.1%], P(>0) = 75.2%). The pattern was similar at the half-way point, where young adults had increased by 11.6%–17.2% (isometric: 11.6%, 95% CI [8.2%, 15.2%], P(>0) > 99.9%; 60°/s: 13.1%, 95% CI [9.0%, 17.5%], P(>0) > 99.9%; 120°/s: 14.9%, 95% CI [9.5%, 20.6%], P(>0) > 99.9%; 240°/s: 17.2%, 95% CI [9.1%, 25.9%], P(>0) > 99.9%), with weak evidence of a larger increase than in older adults (1.5%–2.5%; isometric: 1.5%, 95% CI [−1.3%, 4.9%], P(>0) = 83.9%; 60°/s: 1.8%, 95% CI [−1.7%, 6.1%], P(>0) = 83.0%; 120°/s: 2.2%, 95% CI [−2.4%, 7.5%], P(>0) = 81.9%; 240°/s: 2.5%, 95% CI [−4.3%, 10.3%], P(>0) = 76.1%). The age difference was confined to the low-volume condition. At the half-way point, young adults' low-volume increase was larger than older adults' at isometric, 60°/s and 120°/s (3.6%–5.6%; isometric: 3.6%, 95% CI [0.3%, 7.9%], P(>0) = 98.3%; 60°/s: 4.6%, 95% CI [0.2%, 9.7%], P(>0) = 98.0%; 120°/s: 5.6%, 95% CI [0.1%, 12.1%], P(>0) = 97.7%), and likely larger at 240°/s (6.4%, 95% CI [−2.5%, 16.5%], P(>0) = 92.0%). By post, these differences remained likely, although all 95% CIs included zero (3.3%–6.1%; isometric: 3.3%, 95% CI [−0.1%, 7.4%], P(>0) = 97.1%; 60°/s: 4.2%, 95% CI [−0.2%, 9.1%], P(>0) = 97.0%; 120°/s: 5.2%, 95% CI [−0.4%, 11.4%], P(>0) = 96.6%; 240°/s: 6.1%, 95% CI [−2.5%, 15.8%], P(>0) = 91.5%). Under moderate volume, there was no evidence of an age difference at either time point (−1.2% to −0.1%, P(<0) ≤ 66.8%).

Within young adults, peak torque increased under both low volume (16.2%–26.7%; isometric: 16.2%, 95% CI [12.2%, 20.5%], P(>0) > 99.9%; 60°/s: 18.1%, 95% CI [13.3%, 23.2%], P(>0) > 99.9%; 120°/s: 20.8%, 95% CI [14.4%, 27.6%], P(>0) > 99.9%; 240°/s: 26.7%, 95% CI [16.7%, 37.7%], P(>0) > 99.9%) and moderate volume (12.2%–17.2%; isometric: 12.2%, 95% CI [8.6%, 16.0%], P(>0) > 99.9%; 60°/s: 12.8%, 95% CI [8.4%, 17.3%], P(>0) > 99.9%; 120°/s: 13.8%, 95% CI [8.1%, 19.8%], P(>0) > 99.9%; 240°/s: 17.2%, 95% CI [8.3%, 26.4%], P(>0) > 99.9%). The increase was smaller under moderate volume at every angular velocity (moderate vs. low: isometric: −3.4%, 95% CI [−6.0%, −1.1%], P(<0) = 99.8%; 60°/s: −4.5%, 95% CI [−7.3%, −1.6%], P(<0) = 99.9%; 120°/s: −5.7%, 95% CI [−9.4%, −2.2%], P(<0) > 99.9%; 240°/s: −7.4%, 95% CI [−13.7%, −1.2%], P(<0) = 98.8%), and this difference was already present at the half-way point (isometric: −4.0%, 95% CI [−6.6%, −1.6%], P(<0) > 99.9%; 60°/s: −5.1%, 95% CI [−8.1%, −2.2%], P(<0) > 99.9%; 120°/s: −6.3%, 95% CI [−10.0%, −2.8%], P(<0) > 99.9%; 240°/s: −7.5%, 95% CI [−13.9%, −0.8%], P(<0) = 98.5%). In older adults, low- and moderate-volume training produced similar increases (low: 12.5%–19.5%; isometric: 12.5%, 95% CI [9.0%, 16.3%], P(>0) > 99.9%; 60°/s: 13.4%, 95% CI [9.1%, 17.9%], P(>0) > 99.9%; 120°/s: 14.9%, 95% CI [9.3%, 20.8%], P(>0) > 99.9%; 240°/s: 19.5%, 95% CI [10.8%, 28.9%], P(>0) > 99.9%; moderate: 12.3%–18.5%; isometric: 12.3%, 95% CI [8.8%, 16.1%], P(>0) > 99.9%; 60°/s: 13.1%, 95% CI [8.9%, 17.5%], P(>0) > 99.9%; 120°/s: 14.4%, 95% CI [8.9%, 20.3%], P(>0) > 99.9%; 240°/s: 18.5%, 95% CI [10.1%, 27.9%], P(>0) > 99.9%). There was no evidence of a difference between them at any angular velocity, either at post (isometric: −0.2%, 95% CI [−2.3%, 2.0%], P(<0) = 56.7%; 60°/s: −0.3%, 95% CI [−2.9%, 2.3%], P(<0) = 58.6%; 120°/s: −0.4%, 95% CI [−3.7%, 2.9%], P(<0) = 60.3%; 240°/s: −0.7%, 95% CI [−6.4%, 5.1%], P(<0) = 60.3%) or at the half-way point (isometric: 0.1%, 95% CI [−1.9%, 2.3%], P(>0) = 54.5%; 60°/s: 0.1%, 95% CI [−2.4%, 2.7%], P(>0) = 52.3%; 120°/s: 0.0%, 95% CI [−3.2%, 3.3%], P(<0) = 50.8%; 240°/s: −0.5%, 95% CI [−6.4%, 5.5%], P(<0) = 56.4%).

The volume effect (moderate vs. low) differed between age groups at isometric, 60°/s and 120°/s. Relative to low volume, moderate volume was less favourable in young than in older adults, both at post (isometric: −3.3%, 95% CI [−6.5%, −0.2%], P(<0) = 98.2%; 60°/s: −4.2%, 95% CI [−8.1%, −0.4%], P(<0) = 98.5%; 120°/s: −5.3%, 95% CI [−10.1%, −0.5%], P(<0) = 98.4%) and at the half-way point (isometric: −4.1%, 95% CI [−7.5%, −1.1%], P(<0) = 99.6%; 60°/s: −5.1%, 95% CI [−9.0%, −1.4%], P(<0) = 99.6%; 120°/s: −6.3%, 95% CI [−11.0%, −1.5%], P(<0) = 99.6%). At 240°/s the difference was likely, although the 95% CI included zero (post: −6.7%, 95% CI [−14.8%, 2.0%], P(<0) = 94.1%; half-way: −7.0%, 95% CI [−15.2%, 2.1%], P(<0) = 94.3%).


```{r}
#| label: fig-leg-press
#| fig-cap: ""
#| fig-width: 10
#| fig-height: 6

readRDS(here("figures", "fig-leg-press.RDS"))

#this is only for the supplementary version

```

_1-Repetition Maximum Leg Press_
Older adults in the intervention group (averaged across volume conditions) increased 1RM leg press by 13.1% from baseline to post (95% CI [8.6%, 17.5%], P(>0) > 99.9%). The control group increased by a similar amount (11.9%, 95% CI [4.1%, 20.2%], P(>0) = 99.8%), and there was no evidence of a difference between them (1.2%, 95% CI [−6.8%, 9.7%], P(>0) = 60.6%). Both groups had already increased by the half-way point (intervention: 9.7%, 95% CI [6.1%, 13.4%], P(>0) > 99.9%; control: 7.2%, 95% CI [0.7%, 14.2%], P(>0) = 98.3%), again with no evidence of a difference (2.4%, 95% CI [−5.0%, 10.1%], P(>0) = 73.7%). Young adults (averaged across volume conditions) increased by 12.9% (95% CI [8.3%, 17.5%], P(>0) > 99.9%), with no evidence of a difference from older adults (−0.1%, 95% CI [−5.7%, 5.8%], P(<0) = 52.6%). The same held at the half-way point (young: 9.6%, 95% CI [5.7%, 13.6%], P(>0) > 99.9%; difference: −0.1%, 95% CI [−4.9%, 5.0%], P(<0) = 52.4%). Within young adults, low- and moderate-volume training increased 1RM by 12.2% (95% CI [7.1%, 17.5%], P(>0) > 99.9%) and 13.7% (95% CI [8.4%, 19.0%], P(>0) > 99.9%), respectively, with no evidence of a difference between them (1.3%, 95% CI [−3.2%, 6.0%], P(>0) = 70.4%). In older adults, low- and moderate-volume training increased 1RM by 13.0% (95% CI [8.0%, 18.1%], P(>0) > 99.9%) and 13.2% (95% CI [8.3%, 18.3%], P(>0) > 99.9%), respectively, again with no evidence of a difference (0.2%, 95% CI [−4.1%, 4.5%], P(>0) = 53.0%). At the half-way point, there was no evidence of a volume difference in young adults (0.4%, 95% CI [−4.1%, 4.9%], P(>0) = 56.5%) and only weak evidence in older adults (1.6%, 95% CI [−2.6%, 6.1%], P(>0) = 76.3%). There was no evidence that the volume effect differed between age groups, either at the half-way point (−1.1%, 95% CI [−7.1%, 4.9%], P(<0) = 65.4%) or at post (1.2%, 95% CI [−4.9%, 7.5%], P(>0) = 64.3%).


```{r}
#| label: fig-bicep-curl
#| fig-cap: ""
#| fig-width: 10
#| fig-height: 6

readRDS(here("figures", "fig-bicep-curl.RDS"))

#this is only for the supplementary version

```

_Bicep Curl Force Developement_
Older adults in the intervention group (averaged across volume conditions) increased bicep curl MVC by 5.3% from baseline to post (95% CI [1.6%, 9.1%], P(>0) = 99.8%), while the control group decreased (−8.0%, 95% CI [−14.2%, −1.9%], P(<0) = 99.3%). The resulting difference between groups (14.6%, 95% CI [6.2%, 23.6%], P(>0) > 99.9%) was already present at the half-way point (14.1%, 95% CI [6.0%, 22.9%], P(>0) = 99.9%), likely driven mainly by the control group's decline (−11.0%, 95% CI [−16.7%, −5.1%], P(<0) > 99.9%) rather than by a change in the intervention group, for which there was only weak evidence of an increase at that point (1.5%, 95% CI [−2.1%, 5.1%], P(>0) = 78.6%). Young adults (averaged across volume conditions) tended to increase by 3.8% (95% CI [−0.2%, 7.8%], P(>0) = 96.9%), with no evidence of a difference from older adults (−1.4%, 95% CI [−6.5%, 3.9%], P(<0) = 71.1%). At the half-way point, young adults tended to have increased by a similar amount (3.8%, 95% CI [−0.1%, 7.7%], P(>0) = 97.2%), with weak evidence of a larger increase than in older adults (2.3%, 95% CI [−2.9%, 7.8%], P(>0) = 80.3%). Within young adults, low-volume training increased MVC by 6.4% (95% CI [1.2%, 11.8%], P(>0) = 99.0%), whereas there was no evidence of an increase with moderate-volume training (1.3%, 95% CI [−4.0%, 6.5%], P(>0) = 68.5%); the increase was likely smaller under moderate volume, although the 95% CI included zero (moderate vs. low: −4.8%, 95% CI [−11.2%, 1.7%], P(<0) = 92.4%). In older adults, moderate-volume training increased MVC by 6.5% (95% CI [1.3%, 11.7%], P(>0) = 99.3%), and low-volume training tended to increase it by 4.2% (95% CI [−0.7%, 9.3%], P(>0) = 94.8%), with no evidence of a difference between them (2.3%, 95% CI [−4.2%, 9.1%], P(>0) = 74.7%). At the half-way point, there was weak evidence of a smaller increase under moderate volume in young adults (−3.3%, 95% CI [−9.6%, 3.4%], P(<0) = 84.2%) and no evidence of a volume difference in older adults (1.5%, 95% CI [−4.8%, 8.4%], P(>0) = 67.2%). The volume effect was likely less favourable to moderate volume in young than in older adults at post, although the 95% CI included zero (−6.8%, 95% CI [−15.1%, 2.2%], P(<0) = 93.2%), with weak evidence at the half-way point (−4.7%, 95% CI [−13.2%, 4.6%], P(<0) = 85.0%).

### Oral Glucose Tolerance Test

```{r}
#| label: fig-ogtt
#| fig-cap: "Oral glucose tolerance test (OGTT) responses before and after the training intervention, by age group. Top panels: blood glucose concentrations (mmol L⁻¹) across the 2-hour sampling period, shown for pre-intervention (left) and post-intervention (right) timepoints, in young (20–30 years) and old (70+ years) participants. Shaded ribbons represent the 10th, 25th, 75th, and 90th percentiles of the posterior distribution around the mean (solid line). Bottom panel: the estimated difference in training-induced change in blood glucose (post minus pre) between young and old participants, across the sampling period. Shaded ribbons represent the 2.5th, 10th, 25th, 75th, 90th, and 97.5th percentiles of the posterior distribution around the mean difference (solid line); the dashed horizontal line indicates no difference between age groups."
#| fig-width: 10
#| fig-height: 8

readRDS(here("figures/fig-ogtt.RDS"))

```

In older adults, the intervention group reduced its glucose response from 76 min after ingestion until the end of the sampling window, with the largest reduction at 120 min (−0.86 mmol/L, 95% CI [−1.28, −0.46], P(<0) > 99.9%). In the control group, the glucose response likely increased, although the 95% CI included zero (peak at 52 min: 0.46 mmol/L, 95% CI [−0.01, 0.93], P(>0) = 97.1%). The glucose response was lower in the intervention group than in the control group from 38 min through 120 min (at 120 min: −0.73 mmol/L, 95% CI [−1.48, −0.02], P(<0) = 97.7%; at 38 min: P(<0) = 97.5%). Young adults reduced their glucose response from 20 to 120 min after ingestion, with the largest reduction at 58 min (−1.05 mmol/L, 95% CI [−1.44, −0.67], P(<0) > 99.9%). This reduction was larger than in older adults in the intervention group between 30 and 78 min (P(<0) = 98.6% at 30 min and 98.2% at 78 min), with the largest difference at 54 min (−0.85 mmol/L, 95% CI [−1.34, −0.36], P(<0) > 99.9%). Before 30 min, the difference pointed in the same direction but its 95% CI included zero. After 78 min, the point estimate briefly changed sign, and there was no clear age difference by the end of the 2-h window.


### Molecular

```{r}
#| label: fig-protein-rna
#| fig-cap: "UBF and c-Myc (high peak, ~60 kDa) protein content, and total RNA yield, before and 3 weeks into (and after, where applicable) the training intervention. Panels A, C, and E: predicted values at each measured timepoint, corrected area (arbitrary units) for UBF and c-Myc, and total RNA (ng, adjusted for tissue weight). Error bars represent the 95% credible interval around the posterior mean (point and line). Panels B, D, and F: posterior distributions for five key between-group and between-condition contrasts — Old: Intervention vs. Control, Young vs. Old (intervention), Young: Mod vs. Low, Old: Mod vs. Low, and Volume effect: Young vs. Old, expressed as the ratio of change between the two groups or conditions being compared. For UBF and c-Myc, only the pre-to-week 3 change is shown, reflecting the timepoints measured. For total RNA, both the pre-to-week 3 change (grey) and the pre-to-post change (blue) are shown, per the legend. Points show the posterior mean ratio with 95% credible intervals; the dashed vertical line indicates no difference between groups (ratio = 1). Percentages indicate the posterior probability that the ratio exceeds 1 — i.e., the probability that the first-named group or condition in each comparison showed the larger increase."
#| fig-width: 16
#| fig-height: 10

readRDS(here("figures/fig-protein-rna.rds"))

```

__UBF Protein__
Older adults in the intervention group (averaged across volume conditions) tended to increase UBF protein content by week 3 (19.1%, 95% CI [−7.0%, 49.0%], P(>0) = 90.1%), as did the control group (24.9%, 95% CI [−15.7%, 64.3%], P(>0) = 90.8%), with no evidence of a difference between them (−2.6%, 95% CI [−33.7%, 38.1%], P(<0) = 58.6%). There was weak evidence of an increase in young adults (averaged across volume conditions; 15.7%, 95% CI [−8.0%, 44.2%], P(>0) = 88.5%), with no evidence of a difference from older adults (−1.5%, 95% CI [−29.0%, 33.8%], P(<0) = 56.8%). Within young adults, there was weak evidence of an increase with both low-volume (15.2%, 95% CI [−18.7%, 52.5%], P(>0) = 80.8%) and moderate-volume training (17.6%, 95% CI [−15.2%, 60.7%], P(>0) = 83.2%), with no evidence of a difference between them (4.4%, 95% CI [−34.0%, 54.8%], P(>0) = 53.9%). In older adults, there was weak evidence of an increase with low-volume training (27.1%, 95% CI [−13.3%, 74.7%], P(>0) = 88.9%) and no evidence with moderate-volume training (13.1%, 95% CI [−18.8%, 55.0%], P(>0) = 74.9%), again with no evidence of a difference between them (−8.2%, 95% CI [−44.1%, 41.9%], P(<0) = 67.8%). There was no evidence that the volume effect differed between age groups (20.1%, 95% CI [−39.8%, 112.3%], P(>0) = 67.2%).

__c-Myc Protein__
The difference in c-Myc (high peak) protein content between the old intervention group (averaged across volume conditions) and the control group was large by week 3 (99.8%, 95% CI [17.5%, 216.3%], P(>0) = 99.3%). This difference was driven mainly by a decrease in the control group (−34.5%, 95% CI [−56.0%, −5.5%], P(<0) = 98.6%), while the intervention group tended to increase (25.9%, 95% CI [−8.9%, 68.1%], P(>0) = 92.0%). There was weak evidence of an increase in young adults (averaged across volume conditions; 18.6%, 95% CI [−12.5%, 57.5%], P(>0) = 85.9%), with no evidence of a difference from older adults (−3.5%, 95% CI [−37.8%, 45.0%], P(<0) = 61.5%). Within young adults, there was no evidence of a change with low-volume training (4.4%, 95% CI [−30.1%, 49.1%], P(>0) = 55.2%), while moderate-volume training tended to increase c-Myc (37.3%, 95% CI [−12.2%, 106.2%], P(>0) = 91.5%). There was weak evidence of a larger increase under moderate volume (36.2%, 95% CI [−23.8%, 127.4%], P(>0) = 83.5%). The same pattern held in older adults: there was no evidence of a change with low-volume training (9.6%, 95% CI [−29.9%, 64.4%], P(>0) = 62.0%), moderate-volume training tended to increase c-Myc (47.9%, 95% CI [−4.7%, 118.4%], P(>0) = 95.5%), and there was weak evidence of a larger increase under moderate volume (41.0%, 95% CI [−24.3%, 138.0%], P(>0) = 84.5%). There was no evidence that the volume effect differed between age groups (5.2%, 95% CI [−55.8%, 114.3%], P(>0) = 46.7%).

_Total RNA_ <!-- total amount of RNA, accounting for tissue mass, but not normalizing to RNA/mg -->
Older adults in the intervention group (averaged across volume conditions) increased RNA yield by 14.1% from baseline to post (95% CI [5.3%, 23.5%], P(>0) = 99.9%). The control group tended to increase by a similar amount (13.5%, 95% CI [−4.1%, 34.2%], P(>0) = 93.2%), and there was no evidence of a difference between them (0.5%, 95% CI [−16.7%, 20.3%], P(>0) = 52.4%). The two groups differed in timing: the intervention group had already increased by week 3 (9.2%, 95% CI [1.2%, 18.2%], P(>0) = 98.6%), whereas the control group had not changed at that point (−0.9%, 95% CI [−15.1%, 15.2%], P(<0) = 55.0%). The difference between them at week 3 was still uncertain (10.6%, 95% CI [−7.0%, 30.8%], P(>0) = 87.6%). Young adults (averaged across volume conditions) increased by 11.6% by post (95% CI [2.6%, 21.4%], P(>0) = 99.6%), with no evidence of a difference from older adults (−2.4%, 95% CI [−12.6%, 9.5%], P(<0) = 65.3%). At week 3, however, young adults had increased by 17.4% (95% CI [7.8%, 28.0%], P(>0) > 99.9%), with weak evidence of a larger increase than older adults (7.4%, 95% CI [−4.5%, 20.8%], P(>0) = 88.9%). Between week 3 and post, young adults' RNA yield likely fell back relative to older adults' (−9.0%, 95% CI [−20.4%, 4.5%], P(<0) = 91.9%). Within young adults, both low-volume (15.7%, 95% CI [3.5%, 29.5%], P(>0) = 99.5%) and moderate-volume training (19.2%, 95% CI [6.4%, 33.5%], P(>0) = 99.9%) increased RNA yield by week 3. At post, the increase was 12.2% under moderate volume (95% CI [0.4%, 25.9%], P(>0) = 97.8%), while low volume tended to increase it by 10.8% (95% CI [−0.7%, 24.1%], P(>0) = 96.6%). There was no evidence of a difference between volume conditions at either time point (week 3: 2.9%, 95% CI [−10.8%, 19.9%], P(>0) = 65.7%; post: 1.1%, 95% CI [−12.6%, 17.1%], P(>0) = 56.0%). In older adults, low-volume (12.3%, 95% CI [1.0%, 24.6%], P(>0) = 98.3%) and moderate-volume training (15.8%, 95% CI [4.4%, 28.3%], P(>0) = 99.8%) both increased RNA yield by post, having tended to increase already by week 3 (low: 10.3%, 95% CI [−0.3%, 22.0%], P(>0) = 97.1%; moderate: 8.3%, 95% CI [−2.3%, 19.9%], P(>0) = 93.3%). There was no evidence of a difference between volume conditions at either time point (week 3: −1.9%, 95% CI [−14.0%, 12.3%], P(<0) = 60.4%; post: 3.1%, 95% CI [−9.9%, 17.8%], P(>0) = 67.2%). There was no evidence that the volume effect differed between age groups, either at week 3 (5.0%, 95% CI [−14.6%, 27.5%], P(>0) = 68.4%) or at post (−1.7%, 95% CI [−20.0%, 20.0%], P(<0) = 57.2%).

_RNA-sequencing_
```{r}
#| label: fig-rnaseq-combined
#| fig-cap: "Baseline and training-related transcriptional differences between age groups. (A) Baseline age differences (Young − Old): volcano plot of gene-level estimates (log₂ fold-change) against FDR (log₂ scale), with the dashed line marking FDR = 5%. Genes with the strongest evidence of differential expression are labeled. (B) Rank-based gene-set enrichment analysis (GSEA) of the baseline age contrast, showing the running enrichment score (top) and per-gene contrast effect size (z-score, bottom) across gene rank position, for three representative gene sets (oxidative phosphorylation, heterochromatin formation, complement activation, classical pathway). (C) Age × training-response difference (Young − Old): volcano plot of gene-level estimates for the interaction contrast, in the same format as (A). (D) GSEA of the age × training-response interaction contrast, in the same format as (B), for three representative gene sets (RNA splicing, angiogenesis, aerobic respiration)."
#| fig-width: 10
#| fig-height: 10


library(seqwrap)
m2r <- readRDS(here("data-out/m2-rnaseq.RDS"))
m2s <- seqwrap_summarise(m2r, verbose = FALSE)

n_diff_expressed <- m2s$summaries |> 
  dplyr::filter(term == "ageyoung") |> 
  dplyr::mutate(p.adj = p.adjust(p.value, method = "BH")) |> 
  dplyr::filter(p.adj < 0.05) |> 
  nrow()



n_sig_young_t3 <- m2s$summaries |> 
  dplyr::filter(term == "young_t3") |> 
  dplyr::mutate(p.adj = p.adjust(p.value, method = "BH")) |> 
  dplyr::filter(p.adj < 0.05) |> 
  nrow()

n_sig_old_t3 <- m2s$summaries |> 
  dplyr::filter(term == "old_t3") |> 
  dplyr::mutate(p.adj = p.adjust(p.value, method = "BH")) |> 
  dplyr::filter(p.adj < 0.05) |> 
  nrow()

n_sig_old_volume_t3 <- m2s$summaries |> 
  dplyr::filter(term == "old_volume_t3") |> 
  dplyr::mutate(p.adj = p.adjust(p.value, method = "BH")) |> 
  dplyr::filter(p.adj < 0.05) |> 
  nrow()



readRDS(here("figures/fig-rnaseq.RDS"))


```

Comparing gene-expression profiles between age groups at baseline (Young − Old) revealed marked differences, with `r n_diff_expressed` genes showing robust differences between the groups (FDR < 5%, Fig 6A/B). Rank-based gene-set enrichment analysis [@xu2024a] indicated that young participants had relatively higher expression of genes related to aerobic energy metabolism (oxidative phosphorylation) and relatively lower expression of genes related to immune system activation (complement activation, classical pathway) and epigenetic regulation of gene expression (heterochromatin formation). Training induced a large, broad transcriptional response in both age groups: `r n_sig_young_t3` genes changed from baseline to post in young participants, and `r n_sig_old_t3` in old participants (FDR < 5%, Fig 6C/D). Individual genes with the strongest training-induced changes were predominantly collagen and basement-membrane genes in both groups (e.g., COL4A1, COL4A2, LAMB1, PXDN in young; COL4A1, LAMA4, COL3A1, COL1A2, SPARC in old).

Rank-based gene-set enrichment analysis [@xu2024a] showed that the dominant biology of the training response was similar between age groups: 6 of the top 10 enriched categories were shared, with matching direction and comparable magnitude in both young and old (RNA splicing, ribonucleoprotein complex biogenesis, integrin-mediated signaling pathway, cell adhesion mediated by integrin, collagen metabolic process, and mRNA transport). Training up-regulated genes related to ECM/connective tissue remodeling (e.g., integrin-mediated signaling and adhesion, collagen metabolic process) and down-regulated genes related to RNA-processing and ribosome biogenesis (e.g., RNA splicing, ribonucleoprotein complex biogenesis, mRNA transport) in both groups. While essentially no genes (1 gene at 5% and 0 at 1%) showed a statistically significant age-dependent training response after FDR correction, gene-set enrichment analysis of the age × time interaction contrast revealed that the magnitude of the shared training response differed somewhat between age groups in specific processes: the down-regulation of RNA splicing and related mRNA-processing categories was more pronounced in young than old, while the up-regulation of angiogenesis and aerobic respiration/cellular respiration categories was more pronounced in young than old.

A smaller set of categories differed by age: young participants showed additional enrichment for cytoplasmic translation, mesenchymal cell migration, and angiogenesis, while old participants showed additional enrichment for coagulation- and vascular-related categories. <!--(regulation of coagulation, cellular extravasation, regulation of blood coagulation) — reflected in old participants' volcano plot by genes such as PECAM1, a canonical vascular/platelet-endothelial marker (@fig-training-volcano).-->In contrast to the broadly age-independent training effect, the effect of training volume (moderate vs. low) was age-dependent: a substantial number of genes differed by volume in old participants (`r n_sig_old_volume_t3` genes, FDR < 5%), while essentially no genes showed a volume effect in young participants. Among old participants, higher training volume was associated with relatively higher expression of genes related to mitochondrial biogenesis and oxidative metabolism (e.g., inner mitochondrial membrane organization, respiratory electron transport chain, purine-containing compound biosynthesis) and relatively lower expression of genes related to muscle contractile function and translation (e.g., skeletal muscle contraction, cytoplasmic translation, muscle organ development, striated muscle cell differentiation, ribosomal small subunit biogenesis). This volume effect was considerably more modest than the main training effect, both in the number of genes involved and in statistical strength. <!--(FDR values several orders of magnitude weaker than the training-effect contrasts).-->


