---
title: "Peak torque: force-velocity presentation options"
subtitle: "Responding to supervisor feedback — avoiding 4 comparisons per contrast"
format: html
editor_options: 
  chunk_output_type: console
---


```{r}
#| label: setup
#| message: false

library(brms)
library(tidyverse)
library(reliefdata)
library(marginaleffects)
library(collapse)
library(cowplot)

humac.dat <- reliefdata::relief_humac %>%
  mutate(participant = as.character(participant))

condition <- reliefdata::relief_volume %>%
  mutate(participant = as.character(participant))

participants <- reliefdata::relief_participants %>%
  mutate(age_group = if_else(age < 40, "yng", "old"))

humac_dat <- humac.dat %>%
  full_join(condition) %>%
  select(-arm) %>%
  full_join(participants) %>%
  filter(!participant %in% c("3004", "3005", "3010", "3033",
                              "3034", "3046", "3053", "3069",
                              "3092", "3010", "3019", "3026",
                              "3017", "3012")) %>%
  group_by(participant, time, timerep, leg, allocation, speed, condition, age_group, sex, weight) %>%
  summarise(pt = mean(pt, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    ystd = pt / max(pt, na.rm = TRUE),
    condition = if_else(allocation == "con", "con", condition),
    tx = paste0(allocation, "_", age_group, "_", condition),
    tx = factor(tx, levels = c("int_old_low", "int_old_mod", "int_yng_low", "int_yng_mod", "con_old_con")),
    tx_sex = paste0(tx, "_", sex),
    tx_sex = factor(tx_sex, levels = c(
      "int_old_low_female", "int_old_low_male",
      "int_old_mod_female", "int_old_mod_male",
      "int_yng_low_female", "int_yng_low_male",
      "int_yng_mod_female", "int_yng_mod_male",
      "con_old_con_female", "con_old_con_male"
    )),
    time = factor(time, levels = c("pre", "mid", "post")),
    sex_centered = if_else(sex == "female", -0.5, 0.5),
    speed = factor(speed, levels = c("0", "60", "120", "240")),
    sp = as.numeric(as.character(speed)),
    sp = sp / 240,
    sp2 = sp^2
  ) |>
  filter(!is.na(ystd))

max_pt <- max(humac_dat$pt, na.rm = TRUE)

# Reload the cached model 2 fit (tx_sex varying effects). file_refit = "on_change"
# means this will NOT recompute unless the data/formula/priors changed.
  form2 <- bf(
    ystd ~ 0 +
      # fixed: allocation-level curve at each time
      time:(allocation + allocation:sp + allocation:sp2) +
      # fixed: sex adjustment to the curve at each time (centered, so allocation curves stay sex-averaged)
      time:(sex_centered + sex_centered:sp + sex_centered:sp2) +
      # group level: persistent deviation + group-by-time deviation
      (1 + sp + sp2 | tx_sex:time) +
      # participant level: persistent deviation + participant-by-time deviation
      (1 + sp + sp2 | participant) +
      (1 + sp + sp2 | participant:time) +
      # session intercept: two sessions per participant per time, 8 rows each
      (1 | participant:time:timerep)
  )

priors2 <- get_prior(form2, data = humac_dat) |>
  filter(!(class %in% c("sd", "cor") & (group != "" | coef != ""))) |>
  mutate(prior = dplyr::case_when(
    class == "b" & stringr::str_detect(coef, "sex_centered")  ~ "normal(0, 0.25)",
    class == "b" & stringr::str_detect(coef, ":sp2$") ~ "normal(0.25, 0.2)",
    class == "b" & stringr::str_detect(coef, ":sp$")  ~ "normal(-0.5, 0.1)",
    class == "b" & coef != ""                          ~ "normal(0.5, 0.25)",
    class == "sd"                                      ~ "exponential(10)",
    class == "cor"                                     ~ "lkj(2)",
    class == "sigma"                                   ~ "exponential(1)",
    TRUE ~ prior))

# Check prior
# validate_prior(priors2, form2, data = humac_dat)


# A prior predictive for the mean response, how do these priors look?

data.frame(
  alpha = rnorm(100, 0.5, 0.25), 
  beta_sp = rnorm(100, -0.5, 0.1), 
  beta_sp2 = rnorm(100, 0.25, 0.2 ), 
  iter = 1:100) |> 
  expand_grid(sp = seq(from = 0, to = 1, by =.1)) |> 
  mutate(sp2 = sp^2, 
         y = alpha + beta_sp * sp + beta_sp2 * sp2) |>
  ggplot(aes(sp, y, group = iter)) + geom_line()
  




m2.post <- brm(form2,
                prior = priors2,
                data = humac_dat,
                cores = 4,
               chains = 4,
                iter = 3000,
                warmup = 1000,
                control = list(adapt_delta = 0.95),
                file = "fits/04-strength/m2-full",
                file_refit = "on_change")

m2.post



grid2 <- humac_dat |>
  distinct(tx_sex, time, allocation, sex_centered) |>
  expand_grid(sp = seq(min(humac_dat$sp), max(humac_dat$sp), length.out = 50)) |>
  mutate(sp2 = sp^2)

sex_weights <- humac_dat %>%
  distinct(participant, tx_sex) %>%
  distinct(tx_sex) %>%
  separate(tx_sex, into = c("allocation", "age", "condition", "sex"), remove = FALSE) %>%
  mutate(tx = paste(allocation, age, condition, sep = "_")) %>%
  select(tx_sex, tx)

contrast_defs <- list(
  "Volume (older)"                                        = c("int_old_mod", "int_old_low"),
  "Volume (younger)"                                      = c("int_yng_mod", "int_yng_low"),
  "Intervention vs control (low, older)"                  = c("int_old_low", "con_old_con"),
  "Intervention vs control (mod, older)"                  = c("int_old_mod", "con_old_con"),
  "Age (low volume)"                                      = c("int_old_low", "int_yng_low"),
  "Age (mod volume)"                                      = c("int_old_mod", "int_yng_mod"),
  "Age (pooled across volume)"                            = function(d) ((d$int_yng_mod + d$int_yng_low) / 2) - ((d$int_old_mod + d$int_old_low) / 2),
  "Intervention vs control (pooled across volume, older)" = function(d) ((d$int_old_mod + d$int_old_low) / 2) - d$con_old_con,
  "Volume \u00d7 age interaction"                         = function(d) (d$int_yng_mod - d$int_yng_low) - (d$int_old_mod - d$int_old_low)
)
```

Two small helpers used throughout, so the equal-weighted sex collapse and the six-contrast construction are only written once (previously duplicated per speed/quantity):

```{r}
#| label: helpers

weight_collapse <- function(draws_df, extra_group = character(0)) {
  group_cols <- c("drawid", "tx", "time", extra_group)
  draws_df |>
    left_join(sex_weights |> select(tx_sex, tx), by = "tx_sex") |>
    dplyr::group_by(dplyr::across(dplyr::all_of(group_cols))) |>
    fsummarise(draw = fmean(draw)) |>
    dplyr::ungroup()
}

pre_post_delta <- function(df, extra_group = character(0)) {
  df |>
    select(drawid, draw, tx, time, all_of(extra_group)) |>
    pivot_wider(names_from = time, values_from = draw) |>
    mutate(delta = post - pre) |>
    select(drawid, tx, all_of(extra_group), delta)
}


```

Instead of reading off the model at 4 snapped speeds, we keep every point on the continuous `sp` grid and let the ribbon carry the speed dependence. 

```{r}
#| label: section-a-curves
#| fig-height: 9

preds2 <- predictions(m2.post,
                       newdata = grid2,
                       re_formula = ~(1 + sp + sp2 | tx_sex:time),
                       by = c("tx_sex", "time", "sp"))

draws2 <- get_draws(preds2) |>
  weight_collapse(extra_group = "sp")

deltas2 <- draws2 |>
  pre_post_delta(extra_group = "sp")




# Handles both pairwise (character vector) and pooled/interaction (function)
# entries in contrast_defs — a plain pair-diff can't express e.g. "pooled
# across volume", which needs 3+ tx columns at once.
compute_contrast_curve <- function(delta_df, def, label) {
  wide <- delta_df |>
    pivot_wider(names_from = tx, values_from = delta)

  contrast_val <- if (is.function(def)) {
    def(wide)
  } else {
    wide[[def[1]]] - wide[[def[2]]]
  }

  wide |>
    mutate(contrast_val = contrast_val) |>
    summarise(.by = sp,
              m     = median(contrast_val),
              lwr90 = quantile(contrast_val, 0.05),  upr90 = quantile(contrast_val, 0.95),
              lwr75 = quantile(contrast_val, 0.125), upr75 = quantile(contrast_val, 0.875),
              lwr50 = quantile(contrast_val, 0.25),  upr50 = quantile(contrast_val, 0.75)) |>
    mutate(contrast = label, .before = sp)
}

contrast_curves <- purrr::imap_dfr(contrast_defs, function(def, label) {
  compute_contrast_curve(deltas2, def, label)
}) |>
  mutate(across(c(m, lwr90:upr50), ~ .x * max_pt))

# First, plot predictions
draws2 |>
  summarise(.by = c(sp, tx, time),
            m     = median(draw),
            lwr90 = quantile(draw, 0.05),  upr90 = quantile(draw, 0.95),
            lwr75 = quantile(draw, 0.125), upr75 = quantile(draw, 0.875),
            lwr50 = quantile(draw, 0.25),  upr50 = quantile(draw, 0.75)) |>
  ggplot(aes(sp, m, fill = tx, group = paste(tx, time))) +
  geom_ribbon(aes(ymin = lwr90, ymax = upr90), alpha = 0.15) +
  geom_ribbon(aes(ymin = lwr75, ymax = upr75), alpha = 0.25) +
  geom_ribbon(aes(ymin = lwr50, ymax = upr50), alpha = 0.35) +
  geom_line(aes(lty = time)) + facet_wrap(~ tx, ncol = 3)

ggplot(contrast_curves, aes(sp, m)) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey40") +
  geom_ribbon(aes(ymin = lwr90, ymax = upr90), alpha = 0.15, fill = "steelblue") +
  geom_ribbon(aes(ymin = lwr75, ymax = upr75), alpha = 0.25, fill = "steelblue") +
  geom_ribbon(aes(ymin = lwr50, ymax = upr50), alpha = 0.35, fill = "steelblue") +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ contrast, ncol = 2) +
  labs(x = "Speed (normalized, sp)", y = "\u0394 Peak torque (Nm), post-pre",
       title = "Contrast curves across the full speed range")


```



## Predictions at the four tested angular velocities

```{r}


## Same idea as grid2, but snapped to the four speeds actually tested
# (rather than a fine grid for plotting), carrying a speed_label for
# grouping/reporting.
target_speeds_df <- tibble::tibble(
  speed_label = factor(c("Isometric", "60\u00b0/s", "120\u00b0/s", "240\u00b0/s"),
                        levels = c("Isometric", "60\u00b0/s", "120\u00b0/s", "240\u00b0/s")),
  sp = c(0, 60 / 240, 120 / 240, 240 / 240)
)

grid_speeds <- humac_dat |>
  distinct(tx_sex, time, allocation, sex_centered) |>
  expand_grid(target_speeds_df) |>
  mutate(sp2 = sp^2)

preds_speeds <- predictions(m2.post,
                             newdata = grid_speeds,
                             re_formula = ~ (1 + sp + sp2 | tx_sex:time),
                             by = c("tx_sex", "time", "speed_label"))

draws_speeds <- get_draws(preds_speeds) |>
  weight_collapse(extra_group = "speed_label")

# Pre-mid-post: keep all three transitions (mid-pre, post-mid, post-pre),
# not just pre-post, so half-way-point results can be reported too.
build_speed_deltas <- function(draws_df) {
  draws_df |>
    select(drawid, draw, tx, time, speed_label) |>
    pivot_wider(names_from = time, values_from = draw) |>
    mutate(
      delta_mid     = mid  - pre,
      delta_midpost = post - mid,
      delta_post    = post - pre
    ) |>
    pivot_longer(starts_with("delta_"), names_to = "timepoint", values_to = "delta") |>
    mutate(timepoint = sub("delta_", "", timepoint))
}

build_speed_pct <- function(draws_df) {
  draws_df |>
    select(drawid, draw, tx, time, speed_label) |>
    pivot_wider(names_from = time, values_from = draw) |>
    mutate(
      pct_mid     = (mid  / pre - 1) * 100,
      pct_midpost = (post / mid  - 1) * 100,
      pct_post    = (post / pre - 1) * 100
    ) |>
    pivot_longer(starts_with("pct_"), names_to = "timepoint", values_to = "pct") |>
    mutate(timepoint = sub("pct_", "", timepoint))
}

deltas_speeds_full <- build_speed_deltas(draws_speeds)
pct_speeds_full    <- build_speed_pct(draws_speeds)

# Between-group: handles both pairwise (character vector) and pooled/interaction
# (function) entries in contrast_defs, same fix as compute_contrast_curve above.
summarise_contrasts_speed <- function(value_df, value_col, defs = contrast_defs, scale = 1) {
  wide <- value_df |>
    select(drawid, tx, speed_label, timepoint, value = all_of(value_col)) |>
    pivot_wider(names_from = tx, values_from = value)

  purrr::imap_dfr(defs, function(def, label) {
    contrast_val <- if (is.function(def)) {
      def(wide) * scale
    } else {
      (wide[[def[1]]] - wide[[def[2]]]) * scale
    }

    wide |>
      mutate(contrast_val = contrast_val) |>
      summarise(.by = c(speed_label, timepoint),
                mean    = mean(contrast_val),
                lower95 = quantile(contrast_val, 0.025),
                upper95 = quantile(contrast_val, 0.975),
                pos     = sum(contrast_val > 0) / n()) |>
      mutate(contrast = label, .before = mean)
  }) |>
    mutate(credible = ifelse(lower95 > 0 | upper95 < 0, "Yes", "No"))
}

# Within-group: does this group itself change, with no subtraction between
# groups involved
summarise_within_speed <- function(value_df, value_col, scale = 1) {
  value_df |>
    summarise(.by = c(tx, speed_label, timepoint),
              mean    = mean(.data[[value_col]]) * scale,
              lower95 = quantile(.data[[value_col]], 0.025) * scale,
              upper95 = quantile(.data[[value_col]], 0.975) * scale,
              pos     = sum(.data[[value_col]] > 0) / n()) |>
    mutate(credible = ifelse(lower95 > 0 | upper95 < 0, "Yes", "No"))
}

speed_table_nm   <- summarise_contrasts_speed(deltas_speeds_full, "delta", scale = max_pt) |>
  relocate(contrast, speed_label, timepoint, mean, lower95, upper95, pos, credible) |>
  arrange(contrast, speed_label, timepoint)

speed_table_pct  <- summarise_contrasts_speed(pct_speeds_full, "pct") |>
  relocate(contrast, speed_label, timepoint, mean, lower95, upper95, pos, credible) |>
  arrange(contrast, speed_label, timepoint)

within_table_nm  <- summarise_within_speed(deltas_speeds_full, "delta", scale = max_pt) |>
  relocate(tx, speed_label, timepoint, mean, lower95, upper95, pos, credible) |>
  arrange(tx, speed_label, timepoint)

within_table_pct <- summarise_within_speed(pct_speeds_full, "pct") |>
  relocate(tx, speed_label, timepoint, mean, lower95, upper95, pos, credible) |>
  arrange(tx, speed_label, timepoint)

speed_table_nm
speed_table_pct |>
  print(n = Inf)
within_table_nm
within_table_pct |>
  print(n = Inf)


```

