### Isometric and isokinetic peak torque - humac data

Discussion points: 1. m1/preds1 contrast block intent — Reference to m1/preds1 for the old_mod_vs_low-style contrast, but form1 (HUMAC) has no tx_sex term, so that object can't produce a group contrast. Is the contrast block meant to run against m2/preds2, and should re_formula be NA or tx_sex-inclusive.

2.  Sex-weighting for tx_sex → tx collapse — resolved for low/mod (inherently 50/50, bilateral design), but still open for the age/intervention/control contrasts more broadly: is sample-weighting (vs. equal-weighting) the right call here?

3.  Multiple comparisons framing — 6 contrasts reported from one model. Should they be treated as independent exploratory comparisons (no correction) or should a caveat/adjustment be applied given the growing count?

4.  Mid timepoint inclusion in the stacked hill-curve figures. Do we drop mid for more visual clarity?

Isometric MVC and isokinetic peak torque was measured by the humac dynamometer system. Four different angular velocities were measured, 0 (isometric), 60, 120 and 240°/s (isokinetic), before, half-way and after the intervention. 

With this data, we are interested in estimating:

1.  Within condition (low- and moderate-volume, and non-training control conditions) change from baseline.
2.  Differences between age-groups at baseline.
3.  Average between age-group differences in change from baseline (young vs. old in the intervention group)
4.  Average between control- vs. intervention-group changes from baseline.
5.  Average effect of condition in the intervention groups (low- vs. moderate volume)
6.  Differences between age-groups in effect of condition.

The first contrast will be built for descriptive purposes and serves as the building block for comparisons in contrast 3-4.

#### Load packages and data

```{r}

library(brms)
library(tidyverse)
library(reliefdata)
library(marginaleffects)

humac.dat <- reliefdata::relief_humac %>% 
  mutate(participant = as.character(participant))

condition <- reliefdata::relief_volume %>% 
  mutate(participant = as.character(participant))

participants <- reliefdata::relief_participants %>% 
  mutate(age_group = if_else(age < 40, "yng", "old"))


humac_dat <- humac.dat %>% 
  full_join(condition) %>% 
  select(-arm) %>% 
  full_join(participants) %>%  
  
  filter(!participant %in% c("3004", "3005", "3010", "3033",
                          "3034", "3046", "3053", "3069",
                          "3092", "3010", "3019", "3026",
                          "3017", "3012")) %>% # dropouts/missing all measurements
  #filter(!is.na(time), !is.na(speed)) %>%
  group_by(participant, time, timerep, leg, allocation, speed, condition, age_group, sex, weight) %>%
  summarise(pt = mean(pt, na.rm = TRUE), .groups = "drop") %>%
  mutate(
         ystd = pt / max(pt, na.rm = TRUE),  
         
         condition = if_else(allocation == "con", "con", condition),
         tx = paste0(allocation, "_", age_group, "_", condition),
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con")),
         tx_sex = paste0(tx, "_", sex), 
         tx_sex = factor(tx_sex, levels = c(
           "int_old_low_female", "int_old_low_male",
           "int_old_mod_female", "int_old_mod_male", 
           "int_yng_low_female", "int_yng_low_male",
           "int_yng_mod_female", "int_yng_mod_male", 
           "con_old_con_female", "con_old_con_male"
         )),
         time = factor(time, levels = c("pre", "mid", "post")), 
         sex_centered = if_else(sex == "female", -0.5, 0.5),
         speed = factor(speed, levels = c("0", "60", "120", "240")), 
         sp = as.numeric(as.character(speed)),
         sp = sp/240,
         sp2 = sp^2) |> 
  filter(!is.na(ystd)) |> 
  print()

# About using ystd = pt/max to rescale the data
# A global rescale by max still does something specific: it compresses everything into (0, ~1] range using one arbitrary anchor point (whichever observation happened to be the strongest). That's fine if the only goal is numerical convenience for priors/sampling (keeping ystd on a friendly 0–1 scale for prior specification, which is exactly what the prior predictive code suggests — priors like normal(0.5, 0.25) for the intercept assume a roughly 0–1 outcome). [Likely] If that's the actual purpose, it's a defensible, low-stakes choice and I'll drop it. But it does mean the model's fixed effects and group differences are all still in "fraction of the single strongest lift in the sample" units, not in any per-participant-relative sense — so when you write up results, you can't describe ystd as "percentage of individual capacity" or similar, only as a rescaled version of raw torque. Worth confirming that's how you and your supervisor plan to describe it, since that phrasing mistake is an easy one to make months from now when you're deep in the discussion section.




```

The priors can be tested for a population level trend easily like this:

```{r}

n <- 30 # 30 draws

data.frame(
  int = rnorm(n, 0.5, 0.25), 
  sp = rnorm(n, -0.5, 0.1), 
  sp2 = rnorm(n, 0.25, 0.2), 
  draw = 1:n
) |> 
  expand_grid(speed = seq(from = 0, to = 1, by = 0.1)) |> 
  mutate(speed2 = speed^2,
         # pt = peak torque scaled to 0-1 (xi / max(x))
         pt = int + sp * speed + sp2 * speed2) |> 
  ggplot(aes(speed, pt, group = draw)) + geom_line()



```

#### Models 

Next we set up priors for two different variations of the model. Model 1 is a fixed effects model for the groups with sex as a fixed offset over all time-points. Sex is scale -0.5 and 0.5 here so the fixed effects (predictions) will be the average between men and women.

Model 2 puts all condition groups in "varying effects". This means they will be regularized. Except for the allocation (control vs. intervention).

```{r}



form1 <- bf(ystd ~ 0 + time + time:sp + time:sp2 +
              sex_centered +
              (0 + time + time:sp + time:sp2 | participant) +
              (1 | participant:time:timerep))

form2 <- bf(ystd ~ 0 + time:(allocation + allocation:sp + allocation:sp2) +
              (0 + time + time:sp + time:sp2 | tx_sex) +
              (0 + time + time:sp + time:sp2 | participant) +
              (1 | participant:time:timerep))

# Adding timerep as a random intercept to fix pseudoreplication. It seems a better alternative to account for timerep here in the model, rather than averaging or taking the highest measurement. Not doing anything would mean risking posterior intervals looking tighter and more confident than they should be.


#form1 <- bf(ystd ~ 0 + time + time:sp + time:sp2 +
#              sex_centered +
#              (0 + time + time:sp + time:sp2 | participant))

#form2 <- bf(ystd ~ 0 + time:(allocation + allocation:sp + allocation:sp2) +
#              (0 + time + time:sp + time:sp2 | tx_sex) +
#              (0 + time + time:sp + time:sp2 | participant))


get_prior(form1, data = humac_dat)
get_prior(form2, data = humac_dat)


# Each tx group now has its own population-level intercept + speed curve
# at each timepoint, so there are many "b" coefficients. Rather than list each
# one, set the priors programmatically by matching the coefficient name:
#   * "sex_centered"  -> level-only sex effect, regularized toward 0 (no
#                        prior expectation of a male/female difference)
#   * ends in ":sp2"  -> quadratic speed term
#   * ends in ":sp"   -> linear speed term
#   * otherwise        -> intercept (level at sp = 0)
priors1 <- get_prior(form1, data = humac_dat) |>
  mutate(prior = dplyr::case_when(
    class == "b" & coef == "sex_centered"             ~ "normal(0, 0.1)",
    class == "b" & stringr::str_detect(coef, ":sp2$") ~ "normal(0.25, 0.2)",
    class == "b" & stringr::str_detect(coef, ":sp$")  ~ "normal(-0.5, 0.1)",
    class == "b" & coef != ""                          ~ "normal(0.5, 0.25)",
    class == "sd"                                      ~ "exponential(1)",
    class == "cor"                                     ~ "lkj(2)",
    class == "sigma"                                   ~ "exponential(1)",
    TRUE ~ prior)) 

priors2 <- get_prior(form2, data = humac_dat) |>
  mutate(prior = dplyr::case_when(
    class == "b" & stringr::str_detect(coef, ":sp2$") ~ "normal(0.25, 0.2)",
    class == "b" & stringr::str_detect(coef, ":sp$")  ~ "normal(-0.5, 0.1)",
    class == "b" & coef != ""                          ~ "normal(0.5, 0.25)",
    class == "sd"                                      ~ "exponential(1)",
    class == "cor"                                     ~ "lkj(2)",
    class == "sigma"                                   ~ "exponential(1)",
    TRUE ~ prior))  


m1.pp <- brm(form1,
             prior = priors1,
             data = humac_dat,
             file = "fits/04-strength/test-01",
             file_refit = "on_change",
             sample_prior = "only")



m2.pp <- brm(form2,
             prior = priors2,
             data = humac_dat,
             file = "fits/04-strength/test-02",
             file_refit = "on_change",
             sample_prior = "only")

# Is the prior for the newly added timerep doing as we intend? It should show up with "exponential(1)" in these checks
# prior_summary(m1.pp)
# prior_summary(m2.pp)
# Both checks out






```

```{r}


grid1 <- humac_dat |>
    distinct(tx, time) |>
    expand_grid(sp = seq(min(humac_dat$sp), max(humac_dat$sp), length.out = 50)) |>
    mutate(sp2 = sp^2,
           sex_centered = 0)   # predict the sex-averaged curve per tx

grid2 <- humac_dat |>
    distinct(tx_sex, time, allocation) |>
    expand_grid(sp = seq(min(humac_dat$sp), max(humac_dat$sp), length.out = 50)) |>
    mutate(sp2 = sp^2)   # predict the sex-averaged curve per tx




preds1 <- predictions(m1.pp,
            newdata = grid1,
            re_formula = NA,
            by = c("tx", "time", "sp"))

preds2 <- predictions(m2.pp,
            newdata = grid2,
            re_formula = NA,
            by = c("tx_sex", "time", "sp"))



# Display averages per group/condition

get_draws(preds1)  |>
  summarise(.by = c(tx, time, sp),
            m = mean(draw),
            lwr = quantile(draw, 0.05),
            upr = quantile(draw, 0.95)) |>
  ggplot(aes(sp, m, group = tx, color = tx)) + geom_line() +
  facet_wrap(~ time)
  



```

```{r}
### Get posterior distribution

m1.post <- update(m1.pp,
             sample_prior = "no",
             cores = 4,
             iter = 6000,
             warmup = 2000,
             control = list(adapt_delta = 0.95),
             file = "fits/04-strength/m1-full",
             file_refit = "on_change")

# Checking the model
# A model can run without errors, but that is not proof of convergence per se. rhat and neff_ratio are checks to make sure that the models actually converge and provide you with trustworthy answers. 

# rhat(m1.post)
# rhat essentially checks "did multiple independent attempts agree". A value of 1.0 means that the chains in the model are indistinguishable, they agree. Above 1.0 means the chains disagree, but the conventional cut-off is 1.01 - "close enough to indistinguishable". Checks out to all values being below 1.01 (even 1.005) - no red flags here.  

# neff_ratio(m1.post)
# neff_ratio essentially checks "how much independent information do you actually have". Neff = effective sample size, estimates how many independent draws would give you the same precision as your actual autocorrelated draws. Low ESS on a parameter can mean unreliable posterior summaries even with a clean looking rhat. Increasing the iterations from 4000 to 6000, and the warmup from 1500 to 2000 yielded a 5x higher effective sample size, a total of 1444 compared to 4000 iterations yielding 285, on the same parameter, comfortably clearing the conventional "at least 400" floor, with real headroom for stable 95% credible interval estimation.


m2.post <- update(m2.pp,
             sample_prior = "no",
             cores = 4,
             iter = 6000,
             warmup = 2000,
             control = list(adapt_delta = 0.95),
             file = "fits/04-strength/m2-full",
             file_refit = "on_change")

#rhat(m2.post)
# All at or below 1.003

#neff_ratio(m2.post)
# ESS: 1579


pp_check(m2.post, type = "dens_overlay_grouped",
         group = "tx_sex")

pp_check(m2.post)



```

```{r}



# Model 1

preds1 <- predictions(m1.post,
            newdata = grid1,
            re_formula = ~(0 + time + time:sp + time:sp2 | tx_sex),
            by = c("tx", "time", "sp"))


get_draws(preds1)  |>
  summarise(.by = c(tx, time, sp),
            m = mean(draw),
            lwr = quantile(draw, 0.05),
            upr = quantile(draw, 0.95)) |>

  mutate(tx2 = tx) |>
  separate(tx2, into = c("allocation", "age", "cond")) |>
  mutate(allocation_age = paste0(allocation, "_", age)) |>


  ggplot(aes(sp, m, group = tx, color = tx)) + geom_line() +
  facet_wrap(allocation_age ~ time)


# Calculate delta for each group

get_draws(preds1)  |>
  dplyr::select(drawid, draw, tx, time, sp) |>
  
  pivot_wider(names_from = time, values_from = draw) |>

  mutate(delta = post - pre) |>
 
  dplyr::select(drawid:sp, delta) |>
  pivot_wider(names_from = tx,
              values_from = delta) |>  
  
    # Calculate contrasts
  mutate(
    old_mod_vs_low = int_old_mod - int_old_low
    ) |>

  dplyr::select(drawid:sp, old_mod_vs_low) |>

  # Summarise the posterior of the contrast before plotting
  summarise(.by = sp,
            m   = median(old_mod_vs_low),
            lwr = quantile(old_mod_vs_low, 0.05),
            upr = quantile(old_mod_vs_low, 0.95)) |>

  ggplot(aes(sp, m)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2) +
  geom_line()
  

# Model 2

preds2 <- predictions(m2.post,
                      newdata = grid2,
                      re_formula = ~(0 + time + time:sp + time:sp2 | tx_sex),
                      by = c("tx_sex", "time", "sp"))

# Get actual sex count per tx_sex group to weight the collapse. This is done to use sample-weight, i.e., average using the actual observed male/female counts in each tx group - "what does the average participant in this group look like". The alternative would be to equal-weight the sexes, averaging the male and female posterior predictions 50/50, rather answering "what would the average of a hypothetical man and woman look like". Personally, Im not sure what is best, but my gut-instinct says actual sample-weight. 

sex_weights <- humac_dat %>% 
  distinct(participant, tx_sex) %>% 
  count(tx_sex, name = "n_participants") %>% 
  separate(tx_sex, into = c("allocation", "age", "condition", "sex"), remove = FALSE) %>% 
  mutate(tx = paste(allocation, age, condition, sep = "_")) %>% 
  select(tx_sex, tx, n_participants)


library(collapse)
draws2 <- get_draws(preds2) |>
  left_join(sex_weights, by = "tx_sex") |>
  fgroup_by(drawid, tx, time, sp) |>
  fsummarise(draw = fmean(draw, w = n_participants)) # also using fgroup and fsummarise from collapse as faster tools compared to dplyr

# Pre-post deltas
deltas2 <- draws2 |>
  select(drawid, draw, tx, time, sp) |>
  pivot_wider(names_from = time, values_from = draw) |>
  mutate(delta = post - pre) |>
  select(drawid, tx, sp, delta)

report_speeds <- c(0, 0.25, 0.5, 1)

# Pre-mid-post deltas
deltas2_full <- draws2 %>%
  select(drawid, draw, tx, time, sp) %>%
  pivot_wider(names_from = time, values_from = draw) %>%
  mutate(
    delta_mid     = mid  - pre,
    delta_midpost = post - mid,
    delta_post    = post - pre
  ) %>%
  select(drawid, tx, sp, delta_mid, delta_midpost, delta_post)



## Saving for fig plotting

# Snap to isometric (sp = 0) and 60°/s (sp = 60/240 = 0.25)
target_speeds <- c(isom = 0, sixty = 60/240)

snap_speed <- function(data, target) {
  actual_sp <- unique(data$sp)
  actual_sp[which.min(abs(actual_sp - target))]
}

# Predictions: tx-level, pre/post only, at each target speed 

max_pt <- max(humac_dat$pt, na.rm = TRUE)

make_pred_humac <- function(data, target_sp) {
  sp_use <- snap_speed(data, target_sp)
  data |>
    filter(sp == sp_use, time %in% c("pre", "post")) |>
    mutate(draw = draw * max_pt) |>
    summarise(.by = c(tx, time),
              m = mean(draw),
              lwr95 = quantile(draw, 0.025),
              upr95 = quantile(draw, 0.975))
}

pred_humac_isom <- make_pred_humac(draws2, target_speeds["isom"])
pred_humac_60   <- make_pred_humac(draws2, target_speeds["sixty"])

saveRDS(pred_humac_isom, "data/data-gen/pred_humac_isom.rds")
saveRDS(pred_humac_60,   "data/data-gen/pred_humac_60.rds")

# Contrasts: within-condition pre-post delta, at each target speed
make_contrast_humac <- function(data, target_sp) {
  sp_use <- snap_speed(data, target_sp)
  data |>
    filter(sp == sp_use) |>
    select(drawid, draw, tx, time) |>
    pivot_wider(names_from = time, values_from = draw) |>
    mutate(delta = (post - pre) * max_pt) |>
    summarise(.by = tx,
              mean = mean(delta),
              lower95 = quantile(delta, 0.025),
              upper95 = quantile(delta, 0.975),
              pos = sum(delta > 0) / n())
}


contrast_humac_isom <- make_contrast_humac(draws2, target_speeds["isom"])
contrast_humac_60   <- make_contrast_humac(draws2, target_speeds["sixty"])

saveRDS(contrast_humac_isom, "data/data-gen/contrast_humac_isom.rds")
saveRDS(contrast_humac_60,   "data/data-gen/contrast_humac_60.rds")


# Contrasts: deltas per timepoint and tx
summarise_delta <- function(data, delta_col, target_sp) {
  sp_use <- snap_speed(data, target_sp)
  data %>%
    filter(sp == sp_use) %>%
    summarise(.by = tx,
              mean = mean(.data[[delta_col]] * max_pt),
              lower95 = quantile(.data[[delta_col]] * max_pt, 0.025),
              upper95 = quantile(.data[[delta_col]] * max_pt, 0.975),
              pos = sum(.data[[delta_col]] > 0) / n()) %>%
    mutate(timepoint = sub("delta_", "", delta_col), .before = tx)
}

target_speeds <- c(isom = 0, sixty = 60/240, oneT = 120/240, twoforty = 240/240)

within_all <- purrr::map_dfr(target_speeds, function(ts) {
  bind_rows(
    summarise_delta(deltas2_full, "delta_mid",     ts),
    summarise_delta(deltas2_full, "delta_midpost", ts),
    summarise_delta(deltas2_full, "delta_post",    ts)
  )
}, .id = "speed_label")

print(within_all, n = Inf)

# Curve plot at tx level

draws2 %>% 
  summarise(.by = c(tx, time, sp),
            m = mean(draw),
            lwr = quantile(draw, 0.05),
            upr = quantile(draw, 0.95)) %>% 
  separate(tx, into = c("allocation", "age", "cond"), remove = FALSE) %>% 
  mutate(allocation_age = paste0(allocation, "_", age)) %>% 
  ggplot(aes(sp, m, group = tx, color = tx)) + geom_line() +
  facet_wrap(allocation_age ~ time)

## Delta contrast, same structure as m1 block
# Contrasts of interest:
# - int_old_mod - int_old_low (volume, older)
# - int_yng_mod - int_yng_low (volume, younger)
# - int_old_low - con_old_con (intervention vs control, low, older)
# - int_old_mod - con_old_con (intervention vs control, mod, older)
# - int_old_low - int_yng_low (age, low volume)
# - int_old_mod - int_yng_mod (age, mod volume)


draws2 |>
  select(drawid, draw, tx, time, sp) |>
  pivot_wider(names_from = time, values_from = draw) |>
  mutate(delta = post - pre) |>
  select(drawid:sp, delta) |>
  pivot_wider(names_from = tx, values_from = delta) |>
  mutate(old_mod_vs_low = int_old_mod - int_old_low) |>
  select(drawid:sp, old_mod_vs_low) |>
  summarise(.by = sp,
            m   = median(old_mod_vs_low),
            lwr = quantile(old_mod_vs_low, 0.05),
            upr = quantile(old_mod_vs_low, 0.95)) |>
 # mutate(across(c(m, lwr, upr), ~ .x * max_pt)) |>   # Rescaling ystd back to peak torque here
  ggplot(aes(sp, m)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2) +
  geom_line() +
  labs(y = "Δ ystd, post-pre, mod vs low (older adults)")
 # labs(y = "Δ peak torque (Nm), post-pre, mod vs low (older adults)")


## Posterior probability

# Pre-post only
compute_contrast <- function(data, group_a, group_b, label, targets = report_speeds) {
  actual_sp <- unique(data$sp)
  snapped <- sapply(targets, \(t) actual_sp[which.min(abs(actual_sp - t))])
  
  data |>
    filter(tx %in% c(group_a, group_b), sp %in% snapped) |>
    pivot_wider(names_from = tx, values_from = delta) |>
    mutate(contrast_val = .data[[group_a]] - .data[[group_b]]) |>
    summarise(.by = sp,
              mean    = mean(contrast_val),
              lower95 = quantile(contrast_val, 0.025),
              upper95 = quantile(contrast_val, 0.975),
              pos     = sum(contrast_val > 0) / n()) |>
    mutate(contrast = label,
           sp_requested = targets[match(sp, snapped)],
           .before = sp)
}


contrast_table <- bind_rows(
  compute_contrast(deltas2, "int_old_mod", "int_old_low", "Volume (older)"),
  compute_contrast(deltas2, "int_yng_mod", "int_yng_low", "Volume (younger)"),
  compute_contrast(deltas2, "int_old_low", "con_old_con", "Intervention vs control (low, older)"),
  compute_contrast(deltas2, "int_old_mod", "con_old_con", "Intervention vs control (mod, older)"),
  compute_contrast(deltas2, "int_old_low", "int_yng_low", "Age (low volume)"),
  compute_contrast(deltas2, "int_old_mod", "int_yng_mod", "Age (mod volume)")
) |>
  mutate(across(c(mean, lower95, upper95), ~ .x * max_pt)) |>
  mutate(credible = ifelse(lower95 > 0 | upper95 < 0, "Yes", "No")) |>
  relocate(contrast, sp_requested, sp, mean, lower95, upper95, pos, credible)

contrast_table

# Pre-mid-post

compute_contrast_full <- function(data, group_a, group_b, label, delta_col, targets = report_speeds) {
  actual_sp <- unique(data$sp)
  snapped <- sapply(targets, \(t) actual_sp[which.min(abs(actual_sp - t))])

  data %>%
    filter(tx %in% c(group_a, group_b), sp %in% snapped) %>%
    select(drawid, tx, sp, value = all_of(delta_col)) %>%
    pivot_wider(names_from = tx, values_from = value) %>%
    mutate(contrast_val = .data[[group_a]] - .data[[group_b]]) %>%
    summarise(.by = sp,
              mean = mean(contrast_val),
              lower95 = quantile(contrast_val, 0.025),
              upper95 = quantile(contrast_val, 0.975),
              pos = sum(contrast_val > 0) / n()) %>%
    mutate(contrast = label,
           timepoint = sub("delta_", "", delta_col),
           sp_requested = targets[match(sp, snapped)],
           .before = sp)
}

contrast_table_full <- bind_rows(
  lapply(c("delta_mid", "delta_midpost", "delta_post"), function(dc) {
    bind_rows(
      compute_contrast_full(deltas2_full, "int_old_mod", "int_old_low", "Volume (older)", dc),
      compute_contrast_full(deltas2_full, "int_yng_mod", "int_yng_low", "Volume (younger)", dc),
      compute_contrast_full(deltas2_full, "int_old_low", "con_old_con", "Intervention vs control (low, older)", dc),
      compute_contrast_full(deltas2_full, "int_old_mod", "con_old_con", "Intervention vs control (mod, older)", dc),
      compute_contrast_full(deltas2_full, "int_old_low", "int_yng_low", "Age (low volume)", dc),
      compute_contrast_full(deltas2_full, "int_old_mod", "int_yng_mod", "Age (mod volume)", dc)
    )
  })
) %>%
  mutate(across(c(mean, lower95, upper95), ~ .x * max_pt)) %>%
  mutate(credible = ifelse(lower95 > 0 | upper95 < 0, "Yes", "No")) %>%
  relocate(contrast, timepoint, sp_requested, sp, mean, lower95, upper95, pos, credible)

print(contrast_table_full, n = Inf)


# A tribble for overview

# ── Within-condition results (primary, per tx) ──────────────────────────────
results_humac_within <- within_all %>%
  mutate(
    credible = ifelse(lower95 > 0 | upper95 < 0, "Yes", "No"),
    comparison = "Within-condition"
  ) %>%
  rename(group = tx) %>%
  relocate(comparison, speed_label, timepoint, group, mean, lower95, upper95, pos, credible)

# ── Between-group results (secondary, named contrasts) ──────────────────────
results_humac_between <- contrast_table_full %>%
  mutate(speed_label = case_when(
    sp_requested == 0    ~ "isom",
    sp_requested == 0.25 ~ "sixty",
    sp_requested == 0.5  ~ "oneT",
    sp_requested == 1    ~ "twoforty"
  )) %>%
  rename(group = contrast) %>%
  select(-sp_requested, -sp) %>%
  relocate(comparison = group, speed_label, timepoint, mean, lower95, upper95, pos, credible) %>%
  mutate(comparison = "Between-group")  # move original label into `group` instead

# fix: keep the named contrast (e.g. "Volume (older)") as its own column,
# not overwritten by the generic "Between-group" label
results_humac_between <- contrast_table_full %>%
  mutate(speed_label = case_when(
    sp_requested == 0    ~ "isom",
    sp_requested == 0.25 ~ "sixty",
    sp_requested == 0.5  ~ "oneT",
    sp_requested == 1    ~ "twoforty"
  )) %>%
  rename(group = contrast) %>%
  mutate(comparison = "Between-group") %>%
  select(comparison, speed_label, timepoint, group, mean, lower95, upper95, pos, credible)

# ── Combined table, same shape for both parts ───────────────────────────────
results_humac <- bind_rows(results_humac_within, results_humac_between) %>%
  mutate(
    speed_label = factor(speed_label, levels = c("isom", "sixty", "oneT", "twoforty"),
                          labels = c("Isometric", "60°/s", "120°/s", "240°/s")),
    timepoint = factor(timepoint, levels = c("mid", "midpost", "post"),
                        labels = c("Pre-mid", "Mid-post", "Pre-post"))
  )

saveRDS(results_humac, ("./data/data-gen/results_humac.rds"))

## Saving 120 and 240 speeds

pred_humac_120  <- make_pred_humac(draws2, target_speeds["oneT"])
pred_humac_240  <- make_pred_humac(draws2, target_speeds["twoforty"])

saveRDS(pred_humac_120, ("./data/data-gen/pred_humac_120.rds"))
saveRDS(pred_humac_240, ("./data/data-gen/pred_humac_240.rds"))

contrast_humac_120 <- make_contrast_humac(draws2, target_speeds["oneT"])
contrast_humac_240 <- make_contrast_humac(draws2, target_speeds["twoforty"])

saveRDS(contrast_humac_120, ("./data/data-gen/contrast_humac_120.rds"))
saveRDS(contrast_humac_240, ("./data/data-gen/contrast_humac_240.rds"))


# For reporting:

results_humac %>% 
  filter(timepoint == "Pre-post") %>%   # or whichever timepoint(s) you're reporting
  select(comparison, speed_label, group, mean, lower95, upper95, pos) %>%
  print(n = Inf)

# Note: these data were never log-transformed, and is thus not a ratio, i.e., the null value is zero and not 1. 

```

```{r}

# Standard 5 contrasts — same construction as muscle volume/thickness/lean mass,
# built here as linear combinations of the within-tx pre-post (etc.) deltas.
# Dose-split contrasts above (contrast_table_full / results_humac_between)
# are kept as-is for the archive.

compute_standard_contrast <- function(data, delta_col, contrast_fn, label, targets = report_speeds) {
  actual_sp <- unique(data$sp)
  snapped <- sapply(targets, \(t) actual_sp[which.min(abs(actual_sp - t))])

  wide <- data |>
    filter(sp %in% snapped) |>
    select(drawid, tx, sp, value = all_of(delta_col)) |>
    pivot_wider(names_from = tx, values_from = value)

  wide |>
    mutate(contrast_val = contrast_fn(wide)) |>
    summarise(.by = sp,
              mean = mean(contrast_val),
              lower95 = quantile(contrast_val, 0.025),
              upper95 = quantile(contrast_val, 0.975),
              pos = sum(contrast_val > 0) / n()) |>
    mutate(contrast = label,
           timepoint = sub("delta_", "", delta_col),
           sp_requested = targets[match(sp, snapped)],
           .before = sp)
}

# Direction convention throughout: young - old, intervention - control,
# mod - low — matches the other outcomes exactly.
standard_contrast_defs <- list(
  yng_vs_old_delta = function(d) ((d$int_yng_mod + d$int_yng_low) / 2) - ((d$int_old_mod + d$int_old_low) / 2),
  old_vs_con_delta = function(d) ((d$int_old_mod + d$int_old_low) / 2) - d$con_old_con,
  yng_mod_vs_low   = function(d) d$int_yng_mod - d$int_yng_low,
  old_mod_vs_low   = function(d) d$int_old_mod - d$int_old_low,
  yng_vs_old_vol   = function(d) (d$int_yng_mod - d$int_yng_low) - (d$int_old_mod - d$int_old_low)
)

contrast_table_standard <- bind_rows(
  lapply(c("delta_mid", "delta_midpost", "delta_post"), function(dc) {
    bind_rows(lapply(names(standard_contrast_defs), function(nm) {
      compute_standard_contrast(deltas2_full, dc, standard_contrast_defs[[nm]], nm)
    }))
  })
) |>
  mutate(across(c(mean, lower95, upper95), ~ .x * max_pt)) |>
  mutate(credible = ifelse(lower95 > 0 | upper95 < 0, "Yes", "No")) |>
  relocate(contrast, timepoint, sp_requested, sp, mean, lower95, upper95, pos, credible)

results_humac_standard <- contrast_table_standard |>
  mutate(speed_label = case_when(
    sp_requested == 0    ~ "isom",
    sp_requested == 0.25 ~ "sixty",
    sp_requested == 0.5  ~ "oneT",
    sp_requested == 1    ~ "twoforty"
  )) |>
  rename(group = contrast) |>
  mutate(comparison = "Between-group (standard)") |>
  select(comparison, speed_label, timepoint, group, mean, lower95, upper95, pos, credible) |>
  mutate(
    speed_label = factor(speed_label, levels = c("isom", "sixty", "oneT", "twoforty"),
                          labels = c("Isometric", "60°/s", "120°/s", "240°/s")),
    timepoint = factor(timepoint, levels = c("mid", "midpost", "post"),
                        labels = c("Pre-mid", "Mid-post", "Pre-post"))
  )

saveRDS(results_humac_standard, "data/data-gen/results_humac_standard.rds")
# results_humac (dose-split, already saved above) stays untouched as the archive

```



```{r}

## Fig suggestion 1: Contrast curves across speed, by comparison
# This should be similar to the ogtt fig

max_pt <- max(humac_dat$pt, na.rm = TRUE)

# Requires re-computing the contrasts across the full sp grid (pre-snapping) with multiple quantile widths. This is to avoid producing a plot that looks like a single-CI ribbon plot repeated 6 times - we want a richer nested-uncertainty visualization.

compute_contrast_curve <- function(data, group_a, group_b, label) {
  data |>
    filter(tx %in% c(group_a, group_b)) |>
    pivot_wider(names_from = tx, values_from = delta) |>
    mutate(contrast_val = .data[[group_a]] - .data[[group_b]]) |>
    summarise(.by = sp,
              m     = median(contrast_val),
              lwr90 = quantile(contrast_val, 0.05),  upr90 = quantile(contrast_val, 0.95),
              lwr75 = quantile(contrast_val, 0.125),  upr75 = quantile(contrast_val, 0.875),
              lwr50 = quantile(contrast_val, 0.25),   upr50 = quantile(contrast_val, 0.75)) |>
    mutate(contrast = label, .before = sp)
}

contrast_curves <- bind_rows(
  compute_contrast_curve(deltas2, "int_old_mod", "int_old_low", "Volume (older)"),
  compute_contrast_curve(deltas2, "int_yng_mod", "int_yng_low", "Volume (younger)"),
  compute_contrast_curve(deltas2, "int_old_low", "con_old_con", "Intervention vs control (low, older)"),
  compute_contrast_curve(deltas2, "int_old_mod", "con_old_con", "Intervention vs control (mod, older)"),
  compute_contrast_curve(deltas2, "int_old_low", "int_yng_low", "Age (low volume)"),
  compute_contrast_curve(deltas2, "int_old_mod", "int_yng_mod", "Age (mod volume)")
) |>
  mutate(across(c(m, lwr90:upr50), ~ .x * max_pt))

ggplot(contrast_curves, aes(sp, m)) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey40") +
  geom_ribbon(aes(ymin = lwr90, ymax = upr90), alpha = 0.15, fill = "steelblue") +
  geom_ribbon(aes(ymin = lwr75, ymax = upr75), alpha = 0.25, fill = "steelblue") +
  geom_ribbon(aes(ymin = lwr50, ymax = upr50), alpha = 0.35, fill = "steelblue") +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ contrast, ncol = 2) +
  labs(x = "Speed (normalized, sp)", y = "Δ Peak torque (Nm), post-pre",
       title = "Contrast curves across speed, by comparison")

## Fig suggestion 2: Hill-curve

max_pt <- max(humac_dat$pt, na.rm = TRUE)

vol_colors <- c("Control" = "grey40", "Low volume" = "steelblue", "Moderate volume" = "darkorange")

all_data <- draws2 |>
  filter(tx %in% c("int_old_low", "int_old_mod", "int_yng_low", "int_yng_mod", "con_old_con")) |>
  summarise(.by = c(tx, time, sp), m = median(draw)) |>
  separate(tx, into = c("allocation", "age", "cond"), remove = FALSE) |>
  mutate(m = m * max_pt,
         cond_label = case_when(
           tx == "con_old_con" ~ "Control",
           cond == "low" ~ "Low volume",
           cond == "mod" ~ "Moderate volume"))

y_range <- range(all_data$m)

make_panel <- function(data, filter_expr, title) {
  data |>
    filter({{ filter_expr }}) |>
    ggplot(aes(sp, m, color = cond_label, linetype = time, group = interaction(tx, time))) +
    geom_line(linewidth = 0.8) +
    scale_color_manual(values = vol_colors, drop = TRUE) +
    scale_y_continuous(limits = y_range) +
    labs(x = "Speed (normalized, sp)", y = "Predicted peak torque (Nm)",
         title = title, color = "Group", linetype = "Timepoint")
}

p_young <- make_panel(all_data, age == "yng", "Young (intervention)")
p_old   <- make_panel(all_data, age == "old" & allocation == "int", "Old (intervention)")
p_oldvc <- make_panel(all_data, tx %in% c("int_old_low", "int_old_mod", "con_old_con"), "Old (intervention vs control)")

library(cowplot)
plot_grid(p_young, p_old, p_oldvc, ncol = 1, align = "v")




```

#### Results isometric/isokinetic peak torque

All intervention groups showed clear increases in peak torque from pre to post across every speed tested (pos > 99.7% throughout). In old low-volume, this ranged from 23.4 Nm at isometric (95% ETI [15.3, 31.6]) down to 14.7 Nm at 240°/s (95% ETI [7.5, 21.9]); a comparable pattern was observed in old moderate-volume (22.9 Nm isometric to 13.1 Nm at 240°/s), young low-volume (26.2 Nm to 16.0 Nm), and young moderate-volume (21.0 Nm to 11.1 Nm, pos = 99.7% at 240°/s — marginally lower than the other groups but still very strong evidence of an increase). Across all four groups, the magnitude of this pre-post increase attenuated progressively with speed, roughly halving between isometric and 240°/s.

The control group showed a smaller increase that did not reach the same level of confidence, though there was still moderate-to-strong evidence of some increase rather than no change: 12.3 Nm at isometric (95% ETI [−4.9, 29.7], pos = 91.8%), weakening somewhat at higher speeds (8.0 Nm at 60°/s, pos = 86.0%; 5.4 Nm at 120°/s, pos = 77.3%; 5.3 Nm at 240°/s, pos = 77.1%). <!-- This pattern — a real, if uncertain, increase in the control group alongside a strong, well-defined increase in the intervention groups — is worth keeping in mind when interpreting the intervention-vs-control contrasts below, since it suggests part of the raw intervention-group gains may not be training-specific.-->

Comparing volumes within the old individuals, there was little evidence of a difference between moderate- and low-volume training at any speed — if anything, the point estimates weakly favored low-volume (isometric: mean = −0.5 Nm, 95% ETI [−4.5, 3.5], pos = 39.6%; 60°/s: pos = 35.2%; 120°/s: pos = 30.4%; 240°/s: mean = −1.5 Nm, 95% ETI [−6.8, 3.6], pos = 28.0%).<!-- This is a notably different pattern from the volume-dose effect seen in muscle volume and thickness, where moderate-volume training showed a clear advantage over low-volume in older adults; here, for peak torque, the evidence leans the opposite direction, though weakly and without a confidently established difference.--> In young individuals, by contrast, low-volume training produced a consistently larger pre-post increase in peak torque than moderate-volume training across all four speeds — a substantially stronger and more directionally consistent effect than in the old group (isometric: mean = 5.2 Nm greater for low-volume, 95% ETI [0.5, 9.8], pos = 98.6% for low > mod; 60°/s: pos = 99.2%; 120°/s: pos = 99.4%; 240°/s: mean = 4.9 Nm greater, 95% ETI [−1.0, 10.6], pos = 95.1%). The credible interval only marginally included zero at the fastest speed, where the evidence was still strong (95.1%) despite the wider interval.

Comparing intervention to control, there was moderate-to-strong evidence that both low- and moderate-volume training produced larger pre-post increases than control in older adults, though the credible interval included zero at every speed. For low-volume vs. control: isometric mean = 11.1 Nm greater (95% ETI [−8.2, 30.0], pos = 87.7%), 60°/s pos = 91.8%, 120°/s pos = 92.0%, 240°/s mean = 9.4 Nm greater (95% ETI [−6.3, 24.9], pos = 88.6%). For moderate-volume vs. control: isometric mean = 10.6 Nm greater (95% ETI [−8.4, 29.4], pos = 86.5%), 60°/s pos = 90.4%, 120°/s pos = 90.2%, 240°/s mean = 7.8 Nm greater (95% ETI [−7.8, 23.1], pos = 84.3%).<!-- As noted above, this relatively muted evidence for an intervention-specific effect is consistent with the control group itself showing a real, if uncertain, increase over the same period — meaning part of the apparent gap between intervention and control reflects genuine training-independent change rather than an absence of a training effect.-->

Regarding age-related differences in training response, there was little evidence of a difference at low volume, with a weak lean toward older adults showing the larger increase (isometric: mean = −2.8 Nm, 95% ETI [−10.9, 4.9], pos = 76.4% for old > young; 60°/s: pos = 74.3%; 120°/s: pos = 71.7%; 240°/s: pos = 62.2%). At moderate volume, the evidence leaned weakly the other way, toward young adults showing the larger increase, and was consistently weaker than the low-volume age contrast (isometric: mean = 1.9 Nm, 95% ETI [−6.0, 9.9], pos = 69.0%; 60°/s: pos = 70.8%; 120°/s: pos = 71.6%; 240°/s: pos = 67.6%). Neither age contrast reached a level of evidence comparable to the volume-dose effects described above, and the direction of the age effect differed depending on training volume.
