---
title: "Strength analysis"
format: html
editor_options: 
  chunk_output_type: console
---



```{r, setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

```



## Relief exercise data analysis

This qmd is made with the purpose of analysing the exercise data, with the following question in mind:
1. Does young and old train at a similar training load?
2. Is there an actual difference between low and mod, as by design?

**Primary outcomes**
- Training volume per session? I think this will suffice for now. 


```{r, packages and data}

#--- Packages -------------------------------------------------------------------------------------

library(lme4); library(lmerTest); library(emmeans); library(tidyverse); library(ggrepel)


#--- Data -----------------------------------------------------------------------------------------

ex.dat <- reliefdata::relief_exercise

participants <- reliefdata::relief_participants %>%
  mutate(age_grp = if_else(age < 40, "yng", "old"))


#--- Data wrangling -------------------------------------------------------------------------------

# Checking what we're working with 

str(ex.dat) 

# Checking for duplicate labels

unique(ex.dat$exercise)
unique(ex.dat$condition)
unique(ex.dat$leg)


# Get age_grp lookup from bc_clean (or any cleaned outcome dataset)

age_lookup <- participants %>%
  distinct(participant, age_grp, sex) %>%
  mutate(participant = as.numeric(as.character(participant)))



#--- Exercise load computation --------------------------------------------------------------------

# Since these data are supposed to fit within a figure with at least muscle volume, -strength and the design, I will stick to computing the total volume per participant/condition at the moment.

vol_summary <- ex.dat %>%
  mutate(volume_load = kg * rep) %>%
  group_by(participant, condition) %>%
  summarise(total_vl = sum(volume_load, na.rm = TRUE), .groups = "drop") %>%
  left_join(age_lookup, by = "participant")

str(vol_summary)
vol_summary %>% count(age_grp, condition)




```


```{r, exercise model}

#--- Exercise model ------------------------------------------------------------------------------

# This runs a simple paired test within participants (since everyone has both conditions) nested by age group

# Together, this will give the age x condition interaction (i.e., was the low vs. mod difference bigger in one group than the other), and the pairwise contrasts will quantify the low vs. mod gap separately for young and old

mod_vl <- lmer(total_vl ~ age_grp * condition + sex +
                 (1 | participant),
               data = vol_summary)

anova(mod_vl)

emm_vl <- emmeans(mod_vl, ~ condition | age_grp)
pairs(emm_vl)

#--- The results ---------------------------------------------------------------------------------

# Condition — massive, unsurprising effect (p < .0001). Mod is roughly 76–87k kg more total volume load than low, which is exactly what you'd expect from a 3-set vs 1-set design across a full intervention.

# Age × condition (p = .13) — no significant interaction, meaning the relative difference between low and mod was comparable in both groups. Young adults had a slightly larger absolute gap (−86,745 vs −75,991 kg), but not significantly so.

# Age group (p = .009) — older adults accumulated less total volume load overall, which makes sense given they'd be lifting lower absolute loads.

# Sex (p = .001) — males accumulated more volume load, again expected.

# So the clean summary for figure 1 and the methods/results is: the low vs mod manipulation was successfully implemented in both age groups, with mod producing approximately 3× the volume load of low, and no significant age difference in the magnitude of that contrast.


#--- Relative load comparision young and old -----------------------------------------------------
# Since young typically lift more than old, it is also of interest to investigate if theres a difference in relative training volume/load between them. A common approach would be to express volume as a ratio or percentage, for instance on a log scale.

# On the log scale, low vs. mod contrasts becomes a log ratio, which you can exponentiate to get a fold-difference. If the contrasts are similar between young and old on the log scale, they achieved proportionally similar relative differences between conditions - arguably what matters most for dose-response studies. 

# Descriptive check before modelling 
# This gives the average mod/low fold-diff per group and tells you immediately whether the relative dose contrast was similar

vol_summary %>%
  group_by(participant, age_grp) %>%
  summarise(rel_ratio = total_vl[condition == "mod"] / total_vl[condition == "low"],
            .groups = "drop") %>%
  group_by(age_grp) %>%
  summarise(mean_ratio = mean(rel_ratio), sd_ratio = sd(rel_ratio))

# Here you can see a mean fold-diff of ca. 2.8x in both groups with nearly identical SDs, so the dose separation was essentially equivalent across age groups, where both exercised roughly at 3x the volume load in mod compared to low, with a similar variability. We can still run the log model as a formal test, though. 

# Model

vol_summary <- vol_summary %>%
  mutate(log_vl = log(total_vl))

saveRDS(vol_summary, "./data/data-gen/exercise_volume.rds")

mod_log <- lmer(log_vl ~ age_grp * condition + sex +
                  (1 | participant),
                data = vol_summary)

anova(mod_log)
emm_log <- emmeans(mod_log, ~ condition | age_grp)
pairs(emm_log)

#--- The results ---------------------------------------------------------------------------------

# The interaction is esentially zero, (F = 0.16, p = .69), and the log contrasts are nearly identical, -1.04 for old and -1.03 for young, which exponentiate to fold-diffs of e^1.04 ≈ 2.83 and e^1.03 ≈ 2.80, matching our descriptive ratios exactly.

# The age group main effect also disappears on the log scale (p = .32), which makes sense - young and old lifted different absolute loads, but proportionally their total volume was similar overall. 

# I.e., the conditions were dosed equivalently, relatively speaking, across age groups. 


## Extracting emmeans for fig

# Absolute emmeans

emm_vl_plot <- emmeans(mod_vl, ~ condition | age_grp)
emm_vl_df <- as.data.frame(emm_vl_plot) %>%
  mutate(age_grp = factor(age_grp, levels = c("yng", "old")))

str(emm_vl_df)

saveRDS(emm_vl_df, "./data/data-gen/emm_vl.rds")


# Relative emmeans

emm_log_df <- as.data.frame(emm_log) %>%
  mutate(across(c(emmean, lower.CL, upper.CL), exp))

saveRDS(emm_log_df, "./data/data-gen/emm_log_df.rds")

```


## Results structure

This section mostly serves as a manipulation check rather than a primary finding, so there's no need to overload it. The key things it establishes are: the doses were meaningfully different, the gap was consistent across age groups, and the baseline for interpreting everything else is sound.

You now have three results files covering all the primary and secondary outcomes. The logical next step would be to either merge them into a single results.qmd, or move on to the discussion where the real interpretive work happens — particularly around the speed-dependent attenuation pattern in isokinetics, the older adult catch-up in 1RM, and the contrast between leg and arm outcomes. That discussion section will need some literature to anchor it, so let me know how you want to approach it.

## Exercise Volume Load (absolute load)

Total volume load differed markedly between conditions (main effect of condition: F(1, 62) = 539.9, p \< .001), confirming that the volume manipulation was successfully implemented. The moderate volume condition produced approximately 76,000–87,000 kg greater total volume load than the low volume condition across the intervention, consistent with the intended three-set versus one-set design. Older adults accumulated less total volume load than young adults overall (main effect of age group: F(1, 61) = 7.27, p = .009), reflecting their lower absolute training loads, and males accumulated more than females (main effect of sex: F(1, 61) = 11.7, p = .001).

The Age × Condition interaction was non-significant (F(1, 62) = 2.36, p = .130), indicating that the relative difference between low and moderate volume was comparable across age groups. Estimated marginal means confirmed that both older adults (low: −75,991 kg vs. moderate, p \< .001) and young adults (low: −86,745 kg vs. moderate, p \< .001) showed large and significant condition differences of similar magnitude. These findings confirm that the contrast between training doses was implemented as intended in both age groups, providing a valid basis for interpreting the condition-related effects on hypertrophic and strength outcomes.

## Exercise Volume Load (relative load)

The moderate-volume condition produced approximately 2.8 times greater total volume load than the low-volume condition in both young (2.80 ± 0.28) and older adults (2.83 ± 0.29). The age group × condition interaction was non-significant on the log scale (F(1, 62) = 0.16, p = .69), confirming that the relative dose separation between conditions was equivalent across age groups.


