---
title: "Total Protein Analysis"
format: html
editor_options: 
  chunk_output_type: console
---


```{r, setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

```


## Relief Protein Concentration Analysis

This qmd is made with the purpose of analysing the total protein data, to investigate the change from baseline to 3w across age groups/conditions.

```{r, packages and data}

#--- Packages ------------------------------------------------------------------

library(lme4); library(lmerTest); library(emmeans); library(tidyverse)


#--- Data ----------------------------------------------------------------------

prot_dat <- reliefdata::relief_protconc

prot_qual <- reliefdata::relief_protqual

participants <- reliefdata::relief_participants %>%
  mutate(age_grp = if_else(age < 40, "yng", "old")) %>%
  select(-sex, -height, -weight)

condition <- reliefdata::relief_volume %>%
  mutate(participant = as.character(participant)) %>%
  select(-arm)


#--- Wrangling -----------------------------------------------------------------


# Running a quality check first. A cv% below 10 is usually considered clean, while anything above 15% is considered quite unreliable

# Flag high CV samples
prot_qc <- prot_qual %>%
  mutate(flag_cv = cv_p > 15)

# Check flagged
prot_qc %>% filter(flag_cv)

# Identify the 4 missing samples
all_samples <- prot_dat %>% distinct(sample)
missing_from_qc <- all_samples %>%
  anti_join(prot_qual, by = "sample")
missing_from_qc

# Inspecting flagged samples
prot_dat %>%
  filter(sample %in% c("4.1", "23.3")) %>%
  select(sample, participant, leg, time, dil_fac, well)
# The 23.3 sample from well D09 clearly deviates from the two others, in addition to having a CV% of 57.7, so something quite obviously went wrong here, and this one sample should be excluded to not greatly impact the avg of 23.3. Sample 4.1 is less obvious, since the CV% is borderline and the value itself does not deviate massively from the two others. Excluding the D09 sample from 23.3, but keeping all three from 4.1 for now. 


# Remove the single outlier well for 23.3
prot_dat_clean <- prot_dat %>%
  filter(!(sample == "23.3" & well == "D09"))

# Full cleanup
prot_final <- prot_dat %>%
  # Remove outlier well
  filter(!(sample == "23.3" & well == "D09")) %>%
  # Aggregate triplicates
  group_by(sample, participant, leg, time) %>%
  summarise(
    protein_conc = mean(dil_fac, na.rm = TRUE),
    n_reps = n(),
    cv_p = sd(dil_fac, na.rm = TRUE) / mean(dil_fac, na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  mutate(flag = case_when(
    cv_p > 15 ~ "high_cv",
    n_reps < 3 ~ "low_reps",
    TRUE ~ NA_character_
  )) %>%
  # Remove participants with insufficient data
  filter(!participant %in% c(3084, 3090)) %>%
  # Coerce participant to character for joining
  mutate(participant = as.character(participant))

# Build baseline covariate
pre_vals <- prot_final %>%
  filter(time == "pre") %>%
  select(participant, baseline_conc = protein_conc)

# Build model-ready df
prot_model <- prot_final %>%
  filter(time == "w3") %>%
  left_join(pre_vals, by = "participant") %>%
  left_join(participants %>% select(participant, age_grp, allocation), by = "participant") %>%
  left_join(condition, by = c("participant", "leg"))

# Verify
glimpse(prot_model)
prot_model %>% count(age_grp, condition)

prot_model %>%
  filter(age_grp == "old") %>%
  count(allocation)

# QC checks run during cleaning:
# prot_agg %>% filter(!is.na(flag))        # identified 4.1 (high_cv) and 23.3 (high_cv)
# prot_dat %>% filter(sample == "23.3")    # confirmed D09 as outlier well (7.78)
# prot_model %>% filter(is.na(baseline_conc)) # confirmed 3084 missing baseline

```


```{r, model}


#--- Set-up for modelling ------------------------------------------------------

# Since the pre value isnt leg-specific in a meaningful sense (assuming symmetry), this data will be treated as a participant-level baseline covariate, detached from the leg assignment and spread to both legs for that participant. Using an ANCOVA-style approach, assigning the single pre value to both legs of a participant and use it as a baseline covariate, then model w3 as the outcome.

# Model 1 — Intervention only: age_grp × condition
prot_int <- prot_model %>%
  filter(allocation == "int")

lmm_prot_int <- lmer(protein_conc ~ age_grp * condition + baseline_conc + (1|participant),
                     data = prot_int)
summary(lmm_prot_int)

# The results
# - No significant age × condition interaction (p = 0.132)
# - No main effect of condition (p = 0.209) — protein concentration did not differ between low and moderate volume at w3
# - No main effect of age group (p = 0.954)
# - Baseline concentration did not predict w3 values (p = 0.214)
# - Note: singular fit — participant random effect variance = 0; model behaved as ANCOVA

# Model 1 — intervention: condition × age_grp means
emm_prot_int <- emmeans(lmm_prot_int, ~ condition * age_grp)
emm_prot_int

saveRDS(emm_prot_int, ("./data/data-gen/emm_prot_int.rds"))

# Model 2 — Old intervention vs old control
prot_old <- prot_model %>%
  filter(age_grp == "old")

lmm_prot_old <- lmer(protein_conc ~ allocation + baseline_conc + (1|participant),
                     data = prot_old)
summary(lmm_prot_old)

# The results
# - No effect of allocation (p = 0.589) — old intervention participants did not differ from controls at w3
# - Baseline did not predict w3 (p = 0.647)

# Model 2 — old int vs con
emm_prot_old <- emmeans(lmm_prot_old, ~ allocation)
emm_prot_old

saveRDS(emm_prot_old, ("./data/data-gen/emm_prot_old.rds"))


```

Overall: No evidence of a training effect on skeletal muscle protein concentration at week 3, in either age group or relative to untrained controls. Estimates were tightly clustered (~3.7–4.1 across all groups) with overlapping confidence intervals throughout.
