---
title: "Strength analysis"
format: html
editor_options: 
  chunk_output_type: console
---



```{r, setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

```



## Relief strength analysis

This qmd is made with the purpose of analysing the muscle strength data, with the following question in mind:
1. Is there a difference between age and/or condition?
2. Does age influence the effect of the condition?

**Primary outcomes**
- Isometric MVC (peak torque/force)
- Isokinetic peak torque at 60, 120 and 240 d/s
- 1RM leg press
- Bicep Curl MVC

<!--
**Secondary outcomes**
- Time to peak torque (humac)
- Power (humac)
-->

**The flow of these analyses:**
- All packages and data are loaded in the "packages & data" chunk. All data is formatted for modelling there.
- Each outcome is treated separately (i.e. muscle volume, -thickness, upper body lean mass)
- For each outcome, you'll find the code chunk with the model, emmeans and contrasts (constrasts were ran only if necessary), then you'll find the results reported below the chunk.
- Comparisons between the young and old intervention group has a model per outcome, as do the comparison between old intervention and control.


### Data import and check 

The code chunk below imports all the needed data from the relief package. It also checks that the dataframes are sensible, i.e., contains the expected data, filters out NA values, and sets it up for the models.

```{r, packages and data}

#--- Packages ------------------------------------------------------------------------------------
library(lme4); library(lmerTest); library(emmeans); library(tidyverse); library(ggrepel)


#--- Data ----------------------------------------------------------------------------------------

iso.dat <- reliefdata::relief_humac %>%
  mutate(participant = as.character(participant))

keiser.dat <- reliefdata::relief_1rm 

bc.dat <- reliefdata::relief_isobc 

condition.l <- reliefdata::relief_volume %>%
  mutate(participant = as.character(participant)) %>%
  select(-arm)

condition.a <- reliefdata::relief_volume %>%
  mutate(participant = as.character(participant)) %>%
  select(-leg)

participants <- reliefdata::relief_participants %>%
  mutate(age_group = if_else(age < 40, "yng", "old"))

#--- Data wrangling-------------------------------------------------------------------------------


#--- Humac data ----------------------------------------------------------------------------------
iso.clean <- iso.dat %>%
  full_join(condition.l) %>%
  full_join(participants) %>%
  select(-age, -date, -height, age_grp = age_group)

# Having a look at the dataframe
glimpse(iso.clean)
names(iso.clean)

# Checking if the dataset is complete (does everyone have all the measurements etc)
iso.clean %>% 
  count(participant, time, leg, speed) %>%
  count(n, name = "n_combinations")

# A few missing 2 reps per timepoint, so looking further into it
iso.clean %>%
  count(participant, time, leg, speed) %>%
  filter(n == 1) %>%
  arrange(participant, speed, time)

# Inspection of strange looking rows
iso.clean %>% 
  filter(participant == "3012")

# Then aggregate to best rep per participant × leg × speed × time
# This could also be done by taking the mean at the respective timepoints, but it seems like taking the highest observed value is normal practice

iso.agg <- iso.clean %>%
  filter(!participant %in% c("3004", "3005", "3010", "3033",
                          "3034", "3046", "3053", "3069",
                          "3092", "3010", "3019", "3026",
                          "3017", "3012")) %>% # dropouts/missing all measurements
  filter(!is.na(time), !is.na(speed)) %>%
  group_by(participant, time, leg, allocation, speed, condition, age_grp, sex, weight) %>%
  summarise(pt = mean(pt, na.rm = TRUE), .groups = "drop")

# Isometric (speed == "0") - drop work/power, keep pt and pt_tt

iso_isom <- iso.agg %>%
  filter(speed == "0") 

saveRDS(iso_isom, "./data/data-gen/iso_isom.rds")

# Isokinetic (speed != "0") - keep everything

iso_isok <- iso.agg %>%
  filter(speed != "0") %>%
  mutate(speed = factor(speed, levels = c("60", "120", "240")))

saveRDS(iso_isok, "./data/data-gen/iso_isok.rds")

# Quick sanity check (checking n per speed, time, age_grp)

iso_isom %>%
  count(time, age_grp)

iso_isok %>%
  count(speed, time, age_grp)

# Before modelling, setting time and age_grp to factor instead of chr

# Factor and set reference levels
iso_isom_int <- iso_isom %>%
  mutate(
    time    = factor(time, levels = c("pre", "mid", "post")),
    age_grp = factor(age_grp, levels = c("yng", "old")),
    condition = factor(condition, levels = c("low", "mod"))
  ) %>%
  filter(allocation == "int")

iso_isom_con <- iso_isom %>%
  filter(age_grp == "old") %>%
  mutate(time = factor(time, levels = c("pre", "mid", "post")),
         trained = factor(allocation, levels = c("con", "int")))


iso_isok_int <- iso_isok %>%
  filter(allocation == "int") %>%
  mutate(
    time    = factor(time, levels = c("pre", "mid", "post")),
    age_grp = factor(age_grp, levels = c("yng", "old")),
    condition = factor(condition, levels = c("low", "mod"))
  )

iso_isok_con <- iso_isok %>%
  filter(age_grp == "old") %>%
  mutate(time = factor(time, levels = c("pre", "mid", "post")),
         trained = factor(allocation, levels = c("con", "int")))

# Splitting isokinetic data by speed

pt_60 <- iso_isok_int %>%
  filter(speed == 60) %>%
  mutate(participant = as.factor(participant))

pt_60_con <- iso_isok_con %>%
  filter(speed == 60) %>%
  mutate(participant = as.factor(participant))


pt_120 <- iso_isok_int %>%
  filter(speed == 120) %>%
  mutate(participant = as.factor(participant))

pt_120_con <- iso_isok_con %>%
  filter(speed == 120) %>%
  mutate(participant = as.factor(participant))


pt_240 <- iso_isok_int %>%
  filter(speed == 240) %>%
  mutate(participant = as.factor(participant))

pt_240_con <- iso_isok_con %>%
  filter(speed == 240) %>%
  mutate(participant = as.factor(participant))



#--- Keiser data ---------------------------------------------------------------------------------


# Adding training volume condition, fixing some variable names
keiser.dat2 <- keiser.dat %>%
  filter(timep != "famil") %>%
  full_join(condition.l) %>%
  select(-age, age_grp = age.grp, time = timep)
  
# Checking that dropouts are filtered

keiser.dat2 %>%
  group_by(participant, leg, time) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = time, values_from = n) %>%
  print(n = Inf)
  
  
keiser_clean <- keiser.dat2 %>%
  filter(!participant %in% c("3004", "3005", "3010", "3033",
                          "3034", "3046", "3053", "3069",
                          "3092", "3010", "3019", "3026",
                          "3017")) %>%
  mutate(
    leg = toupper(leg),
    time = case_when(
      time %in% c("mid1", "mid2") ~ "mid",
      time %in% c("post1", "post2") ~ "post",
      TRUE ~ time
    )
  ) %>%
  group_by(participant, leg, time, sex, allocation, age_grp, condition) %>%
  summarise(maxrm = mean(maxrm, na.rm = TRUE), .groups = "drop") %>%
  mutate(participant = as.factor(participant))


keiser_clean %>%
  mutate(leg = toupper(leg)) %>%
  group_by(participant, leg, time) %>%
  summarise(n = n(), .groups = "drop") %>%
  filter(n > 1)

# Then aggregate to best rep per participant × leg × time

keiser_clean <- keiser_clean %>%
  mutate(
    leg = toupper(leg),
    time = case_when(
      time %in% c("mid1", "mid2") ~ "mid",
      time %in% c("post1", "post2") ~ "post",
      TRUE ~ time
    )
  ) %>%
  group_by(participant, leg, time, sex, allocation, age_grp, condition) %>%
  summarise(maxrm = max(maxrm, na.rm = TRUE), .groups = "drop")

saveRDS(keiser_clean, "./data/data-gen/keiser_dat.rds")

# Final check

keiser_clean %>%
  group_by(participant, leg, time) %>%
  summarise(n = n(), .groups = "drop") %>%
  filter(n > 1) 


keiser_int <- keiser_clean %>%
  filter(allocation == "int") %>%
  mutate(time = factor(time, levels = c("pre", "mid", "post")),
         age_grp = factor(age_grp, levels = c("old", "yng")))
  

keiser_con <- keiser_clean %>%
  filter(age_grp == "old") %>%
  mutate(time = factor(time, levels = c("pre", "mid", "post")),
         trained = factor(allocation, levels = c("con", "int")))

#--- Bicep curl data -----------------------------------------------------------------------------
bc.raw <- bc.dat %>%
  full_join(condition.a) %>%
  select(-age, age_grp = age.grp, time = timep) %>%
  print()
  
  
# Checking the dataset

# Check for naming inconsistencies and structure
names(bc.raw)
str(bc.raw)

# Check unique time labels
unique(bc.raw$time)

# Check for lowercase leg entries
unique(bc.raw$arm)

# Check for duplicates
bc.raw %>%
  mutate(leg = toupper(arm)) %>%
  group_by(participant, arm, time) %>%
  summarise(n = n(), .groups = "drop") %>%
  filter(n > 1)

# Check missingness overview
bc.raw %>%
  mutate(leg = toupper(arm)) %>%
  group_by(participant, arm, time) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = time, values_from = n) %>%
  print(n = Inf)

# Cleanup check

bc_clean <- bc.dat %>%
  mutate(timep = case_when(
    timep == "famil" ~ "pre",
    TRUE ~ timep
  )) %>%
  mutate(mf3 = if_else(participant == "3021" & arm == "R" & timep == "pre", NA_real_, mf3)) %>%
  # excluding this single measurement since it was likely a typo (i.e., 100x greater than the other measures), and I dont know for sure what occured in that specific sampling.
  full_join(condition.a) %>%
  select(-age, age_grp = age.grp, time = timep) %>%
  mutate(
    arm = toupper(arm),
    time = case_when(
      time %in% c("mid1", "mid2") ~ "mid",
      time %in% c("post1", "post2", "post 1") ~ "post",
      TRUE ~ time
    )
  ) %>%
  filter(!is.na(time)) %>%
  pivot_longer(cols = c(mf1, mf2, mf3), names_to = "trial", values_to = "force") %>%
  group_by(participant, arm, time, sex, allocation, age_grp, condition) %>%
  summarise(mvc = mean(force, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    participant = as.factor(participant),
    time = factor(time, levels = c("pre", "mid", "post"))
  )

saveRDS(bc_clean, "./data/data-gen/bc_dat.rds")

# Sanity check (tells if the cell counts looks balanced across groupings)
str(bc_clean)
unique(bc_clean$time)
unique(bc_clean$arm)
bc_clean %>% count(age_grp, arm, time)


## Inspection due to massive residual variance and max scaled residual
#bc_clean %>%
#  arrange(desc(mvc)) %>%
#  head(20)

#bc_clean %>%
#  count(participant, arm, time) %>%
#  filter(n > 1)

# The above shows that participant 3021 has an mvc of 1284 N at pre, roughly 5x that of anyone else - thats a strong 93-year old!

# Further inspecting participant 3021
#bc.dat %>%
#  filter(participant == "3021") %>%
#  select(participant, arm, timep, mf1, mf2, mf3)

bc_int <- bc_clean %>%
  filter(allocation == "int") %>%
  mutate(time = factor(time, levels = c("pre", "mid", "post")),
         age_grp = factor(age_grp, levels = c("old", "yng")))

bc_con <- bc_clean %>%
  filter(age_grp == "old") %>%
  mutate(time = factor(time, levels = c("pre", "mid", "post")),
         trained = factor(allocation, levels = c("con", "int")))



```

### Modelling isometric MVC and isokinetic peak torque

The code chunks below contains the model for analysing isometric MVC and isokinetic peak torque at baseline, mid, and post. The data itself is described in the "reliefdata" package, but briefly, we are newton/newtonmeter values derived from the humac dynamometer software. A more thourough explanation of why mixed effects models were used can be found in the "bodycomp_analysis", under "Modelling Muscle Volume". In this section, the model tries to explain the mvc/peak torque observed by age_group, condition and time (fixed effects), with participant as a random effect. 
 

#### Model 1: Isometric MVC young/old intervention

This first model compares the young and old intervention groups.

```{r, model 1}

#--- Model ---------------------------------------------------------------------

mod_isom_pt_int <- lmer(pt ~ age_grp * condition * time + (1 | participant),
                    data = iso_isom_int)

summary(mod_isom_pt_int)
anova(mod_isom_pt_int)

saveRDS(iso_isom_int, "./data/data-gen/iso_isom_int.RDS")

#--- Emmeans -------------------------------------------------------------------

# Emmeans are extracted regardless for plotting, while pairwise contrasts are investigated when there's a significant interaction to decompose

# Nothing to decompose with contrasts here, so running emmeans for plotting

emm_isom_int <- emmeans(mod_isom_pt_int, ~ time * condition | age_grp)
saveRDS(emm_isom_int, "./data/data-gen/emm_isom.rds")

#--- Pre-post contrast for CI --------------------------------------------------

emmeans(mod_isom_pt_int, ~ age_grp * condition * time) %>%
  contrast("pairwise", by = c("age_grp", "condition")) %>%
  summary(infer = TRUE, adjust = "none")





```

**Result summary model 1: Isometric MVC young/old intervention**

Fixed effects (ANOVA):
- Age group:           F(1, 56.0)  = 13.56, p < .001
- Condition:           F(1, 276.0) = 0.35,  p = .552
- Time:                F(2, 276.1) = 54.80, p < .001
- Age × Condition:     F(1, 276.0) = 0.003, p = .959
- Age × Time:          F(2, 276.1) = 4.01,  p = .019
- Condition × Time:    F(2, 276.0) = 0.46,  p = .635
- Age × Cond × Time:   F(2, 276.0) = 0.57,  p = .564

Subgroup changes (emmeans, 95% CI):

Young, Low:
- Pre→Mid:  +31 Nm [21, 41],   p < .001
- Pre→Post: +34 Nm [24, 45],   p < .001
- Mid→Post: +4 Nm  [-7, 14],   p = .485
Young, Mod:
- Pre→Mid:  +22 Nm [12, 32],   p < .001
- Pre→Post: +27 Nm [17, 38],   p < .001
- Mid→Post: +5 Nm  [-5, 16],   p = .314
Old, Low:
- Pre→Mid:  +13 Nm [4, 23],    p = .006
- Pre→Post: +19 Nm [10, 29],   p < .001
- Mid→Post: +6 Nm  [-3, 15],   p = .210
Old, Mod:
- Pre→Mid:  +15 Nm [6, 25],    p = .002
- Pre→Post: +17 Nm [8, 27],    p < .001
- Mid→Post: +2 Nm  [-8, 11],   p = .699

Bottom line: Strong time effect across all groups. Age × Time significant (p = .019) — young gained more than old, particularly pre→mid. No dose-response. Gains plateau after mid in all groups (mid→post CIs all cross zero).

#### Model 2: Isometric MVC old intervention/control

This next model compares the old intervention and control groups.

```{r, model 2}

#--- Model ---------------------------------------------------------------------

mod_isom_pt_con <- lmer(pt ~ trained * time + (1 | participant), 
                        data = iso_isom_con)

summary(mod_isom_pt_con)
anova(mod_isom_pt_con)

saveRDS(iso_isom_con, "./data/data-gen/iso_isom_con.rds")

#--- Emmeans -------------------------------------------------------------------

# emmeans output
emm_isom_con <- emmeans(mod_isom_pt_con, ~ time | trained)
saveRDS(emm_isom_con, "./data/data-gen/emm_isom_con.rds")

# Contrast 1: Pre-timepoint contrast to check baseline imbalance
emm_isom_con_pre <- emmeans(mod_isom_pt_con, ~ trained | time)
contrast(emm_isom_con_pre, "pairwise", infer = TRUE)

#--- Pre-post contrast for CI --------------------------------------------------

emmeans(mod_isom_pt_con, ~ time | trained) %>%
  contrast("pairwise") %>%
  summary(infer = TRUE, adjust = "none")


```

**Result summary model 2: Isometric MVC old intervention/control**

Fixed effects (ANOVA):
- Trained:        F(1, 39.1) = 8.88,  p = .005
- Time:           F(2, 199.2) = 11.01, p < .001
- Trained × Time: F(2, 199.2) = 1.45,  p = .238

Pre→post change within group (emmeans, 95% CI):
- Control: Pre→Mid:  +3 Nm  [-8, 14],   p = .547
           Pre→Post: +13 Nm [1, 24],     p = .031
           Mid→Post: +9 Nm  [-2, 21],    p = .113
- Trained: Pre→Mid:  +14 Nm [8, 21],    p < .001
           Pre→Post: +18 Nm [12, 25],   p < .001
           Mid→Post: +4 Nm  [-2, 10],   p = .216

Trained vs. control at each timepoint:
- Pre:  Int > Con by 44 Nm [10, 78],    p = .013
- Mid:  Int > Con by 55 Nm [21, 89],    p = .002
- Post: Int > Con by 49 Nm [15, 84],    p = .006

Bottom line: Both groups gained strength over time, but trained improved more. Trained × Time non-significant (p = .238). Baseline imbalance present (trained already stronger at pre, p = .013) — complicates interpretation.


#### Model 3: Isokinetic peak torque at 60 d/s young/old intervention

This next model (3) compares the young and old intervention groups.

```{r, model 3}

#--- Model ---------------------------------------------------------------------

mod_isok_pt_60_int <- lmer(pt ~ age_grp * condition * time + (1 | participant/leg),
                         data = pt_60)

summary(mod_isok_pt_60_int)
anova(mod_isok_pt_60_int)

saveRDS(pt_60, "./data/data-gen/pt_60.rds")


#--- Emmeans -------------------------------------------------------------------

# Nothing to decompose with contrasts here, so running emmeans for plotting

emm_isok60_int <- emmeans(mod_isok_pt_60_int, ~ time * condition | age_grp)
saveRDS(emm_isok60_int, "./data/data-gen/emm_isok60.rds")


#--- Pre-post contrast for CI --------------------------------------------------

emmeans(mod_isok_pt_60_int, ~ age_grp * condition * time) %>%
  contrast("pairwise", by = c("age_grp", "condition")) %>%
  summary(infer = TRUE, adjust = "none")



```

**Result summary model 3: Isokinetic peak torque at 60 d/s young/old intervention**

Isokinetic Peak Torque 60°/s — Intervention model (mod_isok_pt_60_int)

Fixed effects (ANOVA):
- Age group:           F(1, 56.0)  = 12.22, p < .001
- Condition:           F(1, 56.6)  = 0.61,  p = .437
- Time:                F(2, 220.7) = 61.07, p < .001
- Age × Condition:     F(1, 56.6)  = 1.41,  p = .240
- Age × Time:          F(2, 220.7) = 0.76,  p = .469
- Condition × Time:    F(2, 222.0) = 1.09,  p = .339
- Age × Cond × Time:   F(2, 222.0) = 0.90,  p = .406

Subgroup changes (emmeans, 95% CI):
Young, Low:
- Pre→Mid:  +23 Nm [16, 30],   p < .001
- Pre→Post: +23 Nm [15, 30],   p < .001
- Mid→Post: +0 Nm  [-7, 8],    p = .930
Young, Mod:
- Pre→Mid:  +14 Nm [7, 21],    p < .001
- Pre→Post: +15 Nm [8, 23],    p < .001
- Mid→Post: +1 Nm  [-6, 9],    p = .696
Old, Low:
- Pre→Mid:  +14 Nm [8, 21],    p < .001
- Pre→Post: +17 Nm [10, 23],   p < .001
- Mid→Post: +2 Nm  [-5, 9],    p = .490
Old, Mod:
- Pre→Mid:  +14 Nm [7, 21],    p < .001
- Pre→Post: +17 Nm [10, 23],   p < .001
- Mid→Post: +3 Nm  [-4, 9],    p = .408

Bottom line: Strong time effect across all groups. Unlike isometric, Age × Time non-significant (p = .469) — young and old gained similarly at 60°/s. No dose-response. Gains largely occur pre→mid with minimal mid→post progression.


#### Model 4: Isokinetic peak torque at 60 d/s old intervention/control

This next model (4) compares the old intervention and control groups.

```{r, model 4}

#--- Model ---------------------------------------------------------------------

mod_isok_pt_60_con <- lmer(pt ~ trained * time + (1 | participant), data = pt_60_con)

summary(mod_isok_pt_60_con)
anova(mod_isok_pt_60_con)

saveRDS(pt_60_con, "./data/data-gen/pt_60_con.rds")


#--- Emmeans -------------------------------------------------------------------

# emmeans output
emm_isok60_con <- emmeans(mod_isok_pt_60_con, ~ time | trained)
saveRDS(emm_isok60_con, "./data/data-gen/emm_isok60_con.rds")

# The anova revealed different trajectories for int and con over time, worth checking with contrasts

emm_isok60_con_pre <- emmeans(mod_isok_pt_60_con, ~ trained | time)
contrast(emm_isok60_con_pre, "pairwise", infer = TRUE)

#--- Pre-post contrast for CI --------------------------------------------------

emmeans(mod_isok_pt_60_con, ~ time | trained) %>%
  contrast("pairwise") %>%
  summary(infer = TRUE, adjust = "none")


```

**Result summary model 4: Isokinetic peak torque at 60 d/s old intervention/control**

Isokinetic Peak Torque 60°/s — Control model (mod_isok_pt_60_con)

Fixed effects (ANOVA):
- Trained:        F(1, 39.1) = 6.70,  p = .013
- Time:           F(2, 199.3) = 8.88,  p < .001
- Trained × Time: F(2, 199.3) = 3.69,  p = .027

Pre→post change within group (emmeans, 95% CI):
- Control: Pre→Mid:  +2 Nm  [-11, 6],   p = .590
           Pre→Post: +4 Nm  [-13, 5],   p = .380
           Mid→Post: +2 Nm  [-11, 7],   p = .718
- Trained: Pre→Mid:  +14 Nm [9, 19],    p < .001
           Pre→Post: +17 Nm [12, 22],   p < .001
           Mid→Post: +3 Nm  [-2, 7],    p = .303

Trained vs. control at each timepoint:
- Pre:  Int > Con by 23 Nm [-2, 47],    p = .071
- Mid:  Int > Con by 34 Nm [10, 59],    p = .008
- Post: Int > Con by 35 Nm [11, 60],    p = .006

Bottom line: Trained older adults improved; controls did not. Trained × Time significant (p = .027). Borderline baseline difference (p = .071) — less severe than isometric but worth noting.


#### Model 5: Isokinetic peak torque at 120 d/s young/old intervention

This next model (5) compares the young and old intervention groups.

```{r, model 5}

#--- Model ---------------------------------------------------------------------

mod_isok_pt_120_int <- lmer(pt ~ age_grp * condition * time + (1 | participant/leg),
                         data = pt_120)

summary(mod_isok_pt_120_int)
anova(mod_isok_pt_120_int)

saveRDS(pt_120, "./data/data-gen/pt_120.rds")


#--- Emmeans -------------------------------------------------------------------

# Extracting emmeans for plotting
emm_isok120_int <- emmeans(mod_isok_pt_120_int, ~ time * condition | age_grp)
saveRDS(emm_isok120_int, "./data/data-gen/emm_isok120.rds")


#--- Pre-post contrasts for CI -------------------------------------------------

emmeans(mod_isok_pt_120_int, ~ age_grp * condition * time) %>%
  contrast("pairwise", by = c("age_grp", "condition")) %>%
  summary(infer = TRUE, adjust = "none")



```

**Result summary model 5: Isokinetic peak torque at 120 d/s young/old intervention**

Isokinetic Peak Torque 120°/s — Intervention model (mod_isok_pt_120_int)

Fixed effects (ANOVA):
- Age group:           F(1, 56.0)  = 11.01, p = .002
- Condition:           F(1, 56.3)  = 1.60,  p = .212
- Time:                F(2, 220.5) = 58.10, p < .001
- Age × Condition:     F(1, 56.3)  = 2.16,  p = .147
- Age × Time:          F(2, 220.5) = 0.009, p = .991
- Condition × Time:    F(2, 222.0) = 0.60,  p = .550
- Age × Cond × Time:   F(2, 222.0) = 0.04,  p = .965

Subgroup changes (emmeans, 95% CI):

Young, Low:
- Pre→Mid:  +16 Nm [9, 23],    p < .001
- Pre→Post: +19 Nm [12, 26],   p < .001
- Mid→Post: +3 Nm  [-4, 10],   p = .438
Young, Mod:
- Pre→Mid:  +13 Nm [6, 19],    p < .001
- Pre→Post: +14 Nm [8, 21],    p < .001
- Mid→Post: +2 Nm  [-5, 9],    p = .617
Old, Low:
- Pre→Mid:  +15 Nm [9, 21],    p < .001
- Pre→Post: +18 Nm [11, 24],   p < .001
- Mid→Post: +3 Nm  [-4, 9],    p = .412
Old, Mod:
- Pre→Mid:  +13 Nm [7, 19],    p < .001
- Pre→Post: +15 Nm [9, 21],    p < .001
- Mid→Post: +2 Nm  [-4, 8],    p = .520

Bottom line: Strong time effect, gains consistent across age groups — Age × Time flat (p = .991). This is the clearest velocity where young and old respond equivalently. No dose-response. All mid→post CIs cross zero.

#### Model 6: Isokinetic peak torque at 120 d/s old intervention/control

This next model (6) compares the old intervention and control groups.

```{r, model 6}

#--- Model ---------------------------------------------------------------------

mod_isok_pt_120_con <- lmer(pt ~ trained * time + (1 | participant), data = pt_120_con)

summary(mod_isok_pt_120_con)
anova(mod_isok_pt_120_con)

saveRDS(pt_120_con, "./data/data-gen/pt_120_con.rds")


#--- Emmeans -------------------------------------------------------------------

# Extracting emmeans for plotting
emm_isok120_con <- emmeans(mod_isok_pt_120_con, ~ time | trained)
saveRDS(emm_isok120_con, "./data/data-gen/emm_isok120_con.rds")


# Seeing the same pattern as for 60, but its worth confirming with contrasts here also
emm_isok120_con_pre <- emmeans(mod_isok_pt_120_con, ~ trained | time)
contrast(emm_isok120_con_pre, "pairwise", infer = TRUE)


#--- Pre-post contrasts for CI -------------------------------------------------

emmeans(mod_isok_pt_120_con, ~ time | trained) %>%
  contrast("pairwise") %>%
  summary(infer = TRUE, adjust = "none")


```

**Result summary model 6: Isokinetic peak torque at 120 d/s old intervention/control**

Isokinetic Peak Torque 120°/s — Control model (mod_isok_pt_120_con)

Fixed effects (ANOVA):
- Trained:        F(1, 39.1) = 7.04,  p = .011
- Time:           F(2, 199.2) = 14.03, p < .001
- Trained × Time: F(2, 199.2) = 4.71,  p = .010

Pre→post change within group (emmeans, 95% CI):
- Control: Pre→Mid:  +3 Nm  [-5, 10],   p = .461
           Pre→Post: +5 Nm  [-3, 13],   p = .173
           Mid→Post: +2 Nm  [-5, 10],   p = .512
- Trained: Pre→Mid:  +14 Nm [10, 18],   p < .001
           Pre→Post: +16 Nm [12, 20],   p < .001
           Mid→Post: +2 Nm  [-2, 6],    p = .262

Trained vs. control at each timepoint:
- Pre:  Int > Con by 20 Nm [-1, 41],    p = .067
- Mid:  Int > Con by 31 Nm [10, 52],    p = .005
- Post: Int > Con by 31 Nm [10, 52],    p = .006

Bottom line: Same pattern as 60°/s — trained improved, controls did not. Trained × Time significant (p = .010). Borderline baseline imbalance (p = .067).

#### Model 7: Isokinetic peak torque at 240 d/s young/old intervention

This next model (7) compares the young and old intervention groups.

```{r, model 7}

#--- Model ---------------------------------------------------------------------

mod_isok_pt_240_int <- lmer(pt ~ age_grp * condition * time + (1 | participant/leg),
                         data = pt_240)

summary(mod_isok_pt_240_int)
anova(mod_isok_pt_240_int)

saveRDS(pt_240, "./data/data-gen/pt_240.rds")


#--- Emmeans -------------------------------------------------------------------

# Extracting emmeans for plotting

emm_isok240_int <- emmeans(mod_isok_pt_240_int, ~ time * condition | age_grp)
saveRDS(emm_isok240_int, "./data/data-gen/emm_isok240.rds")


#--- Pre-post contrasts for CI -------------------------------------------------

emmeans(mod_isok_pt_240_int, ~ age_grp * condition * time) %>%
  contrast("pairwise", by = c("age_grp", "condition")) %>%
  summary(infer = TRUE, adjust = "none")



```

**Result summary model 7: Isokinetic peak torque at 240 d/s young/old intervention**

Isokinetic Peak Torque 240°/s — Intervention model (mod_isok_pt_240_int)

Fixed effects (ANOVA):
- Age group:           F(1, 56.0)  = 9.27,  p = .004
- Condition:           F(1, 56.5)  = 0.63,  p = .431
- Time:                F(2, 220.8) = 53.15, p < .001
- Age × Condition:     F(1, 56.5)  = 2.50,  p = .119
- Age × Time:          F(2, 220.8) = 0.82,  p = .443
- Condition × Time:    F(2, 222.6) = 0.72,  p = .489
- Age × Cond × Time:   F(2, 222.6) = 0.02,  p = .977

Subgroup changes (emmeans, 95% CI):

Young, Low:
- Pre→Mid:  +13 Nm [8, 19],    p < .001
- Pre→Post: +17 Nm [11, 22],   p < .001
- Mid→Post: +3 Nm  [-3, 9],    p = .271
Young, Mod:
- Pre→Mid:  +11 Nm [6, 17],    p < .001
- Pre→Post: +13 Nm [7, 19],    p < .001
- Mid→Post: +2 Nm  [-4, 7],    p = .594
Old, Low:
- Pre→Mid:  +10 Nm [5, 15],    p < .001
- Pre→Post: +14 Nm [8, 19],    p < .001
- Mid→Post: +4 Nm  [-2, 9],    p = .174
Old, Mod:
- Pre→Mid:  +8 Nm  [3, 13],    p = .003
- Pre→Post: +11 Nm [6, 16],    p < .001
- Mid→Post: +3 Nm  [-2, 8],    p = .272

Bottom line: Strong time effect, Age × Time non-significant (p = .443) — consistent with 120°/s. Young and old gained similarly at higher velocities. No dose-response. Confirms velocity-specific pattern: age differences in strength gains present at isometric but absent at 120 and 240°/s.

#### Model 8: Isokinetic peak torque at 240 d/s old intervention/control

This next model (8) compares the old intervention and control groups.

```{r, model 8}

#--- Model ---------------------------------------------------------------------

mod_isok_pt_240_con <- lmer(pt ~ trained * time + (1 | participant), data = pt_240_con)

summary(mod_isok_pt_240_con)
anova(mod_isok_pt_240_con)

saveRDS(pt_240_con, "./data/data-gen/pt_240_con.rds")


#--- Emmeans -------------------------------------------------------------------

# Extracting emmeans for plotting
emm_isok240_con <- emmeans(mod_isok_pt_240_con, ~ time | trained)
saveRDS(emm_isok240_con, "./data/data-gen/emm_isok240_con.rds")


# Contrasts (only a borderline diff, which diverges from 60 and 120)
emm_isok240_con_pre <- emmeans(mod_isok_pt_240_con, ~ trained | time)
contrast(emm_isok240_con_pre, "pairwise", infer = TRUE)


#--- Pre-post contrasts for CI -------------------------------------------------

emmeans(mod_isok_pt_240_con, ~ time | trained) %>%
  contrast("pairwise") %>%
  summary(infer = TRUE, adjust = "none")


```

**Result summary model 8: Isokinetic peak torque at 240 d/s old intervention/control**

Isokinetic Peak Torque 240°/s — Control model (mod_isok_pt_240_con)

Fixed effects (ANOVA):
- Trained:        F(1, 39.1) = 7.39,  p = .010
- Time:           F(2, 199.3) = 10.66, p < .001
- Trained × Time: F(2, 199.3) = 3.34,  p = .037

Pre→post change within group (emmeans, 95% CI):
- Control: Pre→Mid:  +2 Nm  [-4, 9],    p = .427
           Pre→Post: +4 Nm  [-3, 10],   p = .262
           Mid→Post: +1 Nm  [-5, 7],    p = .719
- Trained: Pre→Mid:  +9 Nm  [6, 12],    p < .001
           Pre→Post: +12 Nm [9, 16],    p < .001
           Mid→Post: +3 Nm  [0, 7],     p = .058

Trained vs. control at each timepoint:
- Pre:  Int > Con by 15 Nm [0, 31],     p = .056
- Mid:  Int > Con by 22 Nm [6, 37],     p = .007
- Post: Int > Con by 24 Nm [8, 40],     p = .004

Bottom line: Trained improved; controls did not. Trained × Time significant (p = .037). Pre difference borderline (p = .056) — consistent pattern with other isokinetic velocities.

Of note: The velocity-specific pattern holds cleanly — Age × Time significant at isometric (p = .019), absent at 120°/s (p = .991) and 240°/s (p = .443), and 60°/s sits in between (p = .469). That's a coherent narrative worth making explicit.

### Modelling keiser 1RM leg press

The code chunks below contains the model for analysing 1RM leg press at baseline, mid, and post. The data itself is described in the "reliefdata" package, but briefly, we are using kg values derived from the keiser software. The model uses the same logic as the ones for the isometric and isokinetic data. 

#### Model 9: 1RM leg press young/old intervention

This next model (9) compares the young and old intervention groups.

```{r, model 9}

#--- Model ---------------------------------------------------------------------

mod_keiser_int <- lmer(maxrm ~ age_grp * condition * time + (1 | participant),
                   data = keiser_int)

summary(mod_keiser_int)
anova(mod_keiser_int)

saveRDS(keiser_int, "./data/data-gen/keiser_int.rds")


#--- Emmeans -------------------------------------------------------------------

# Extracting emmeans for plotting
emm_keiser_int <- emmeans(mod_keiser_int, ~ time * condition | age_grp)
saveRDS(emm_keiser_int, "./data/data-gen/emm_keiser.rds")

#--- Pre-post contrasts for CI -------------------------------------------------

emmeans(mod_keiser_int, ~ age_grp * condition * time) %>%
  contrast("pairwise", by = c("age_grp", "condition")) %>%
  summary(infer = TRUE, adjust = "none")



```

**Result summary model 9: 1RM leg press young/old intervention**

Keiser Leg Press 1RM — Intervention model (mod_keiser_int)

Fixed effects (ANOVA):
- Age group:           F(1, 57.0)  = 7.65,  p = .008
- Condition:           F(1, 277.0) = 0.30,  p = .587
- Time:                F(2, 277.1) = 83.75, p < .001
- Age × Condition:     F(1, 277.0) = 2.58,  p = .109
- Age × Time:          F(2, 277.1) = 1.73,  p = .179
- Condition × Time:    F(2, 277.0) = 0.06,  p = .942
- Age × Cond × Time:   F(2, 277.0) = 0.16,  p = .854

Subgroup changes (emmeans, 95% CI):

Old, Low:
- Pre→Mid:  +17 Nm [8, 26],    p < .001
- Pre→Post: +28 Nm [19, 37],   p < .001
- Mid→Post: +11 Nm [2, 20],    p = .014
Old, Mod:
- Pre→Mid:  +17 Nm [8, 26],    p < .001
- Pre→Post: +27 Nm [18, 36],   p < .001
- Mid→Post: +10 Nm [1, 19],    p = .029
Young, Low:
- Pre→Mid:  +25 Nm [16, 35],   p < .001
- Pre→Post: +30 Nm [21, 40],   p < .001
- Mid→Post: +5 Nm  [-5, 14],   p = .335
Young, Mod:
- Pre→Mid:  +26 Nm [17, 36],   p < .001
- Pre→Post: +34 Nm [24, 44],   p < .001
- Mid→Post: +8 Nm  [-2, 17],   p = .108

Bottom line: Strong time effect in all groups. Age × Time non-significant (p = .179) — both age groups gained meaningfully. No dose-response. Unlike isokinetic outcomes, mid→post progression significant in old (both conditions), suggesting continued adaptation. Familiarisation artefact possible for controls — interpret control model with caution.

#### Model 10: 1RM leg press old intervention/control

This next model (10) compares the old intervention and control groups.

```{r, model 10}

#--- Model ---------------------------------------------------------------------

mod_keiser_con <- lmer(maxrm ~ trained * time + (1 | participant), data = keiser_con)

summary(mod_keiser_con)
anova(mod_keiser_con)

saveRDS(keiser_con, "./data/data-gen/keiser_con.rds")
### Results from the control anova

# Trained: F(1, 39.0) = 0.43, p = .518
# Time: F(2, 195.1) = 24.53, p < .001
# Trained × Time: F(2, 195.1) = 0.91, p = .405

### Keiser checkup
# How many participants per group?
#keiser_con %>% count(trained)

# Any extreme values?
#keiser_con %>% 
#  group_by(trained, time) %>% 
#  summarise(mean = mean(maxrm, na.rm = TRUE),
#            sd = sd(maxrm, na.rm = TRUE),
#            min = min(maxrm, na.rm = TRUE),
#            max = max(maxrm, na.rm = TRUE))

# Individual trajectories — do any controls improve dramatically?
#keiser_con %>%
#  filter(trained == "con") %>%
#  ggplot(aes(x = time, y = maxrm, group = participant)) +
#  geom_line(alpha = 0.5) +
#  geom_point()

# The plot and summary table tell a clear story — the control group is genuinely improving on the Keiser across all three timepoints (167 → 174 → 182 kg), and there are no obvious outliers or data entry errors. The trajectories look smooth and plausible, no wild jumps.

# So this is likely a real phenomenon rather than a data quality issue. A few possible explanations worth considering:

# Learning/familiarisation effect — the Keiser air resistance system has an unusual feel compared to free weights or standard machines. Controls coming in for repeated testing may simply be getting better at the task across three visits, even without training. This is a known issue with 1RM testing on novel equipment.

# The intervention group n is 184 vs. 56 for controls — that's worth double-checking. With your design you'd expect roughly 2:1 (intervention has two legs, control has one score per timepoint), but 184/56 = 3.3. Run a quick check:

#keiser_con %>% 
#  count(trained, participant) %>% 
#  count(trained, name = "n_participants")

# 10 control participants × 3 timepoints × ~2 legs would give ~60 rows, which matches the 56 you have (a few missing). And 31 intervention participants × 3 timepoints × 2 legs = 186, close to the 184. So the structure is correct.

# The improvement in controls is therefore genuine and the model is fine as-is. It most likely reflects the familiarisation/learning effect on the Keiser — 10 older adults being tested three times on an air resistance leg press will show some improvement just from task familiarity, and with only 10 controls the variance is large enough (SD ~60 kg) that the model can't distinguish their trajectory from the intervention group's.

# This is worth a sentence in the discussion — something along the lines of: the absence of a significant trained × time interaction for 1RM leg press may reflect a familiarisation effect with the Keiser air resistance system in the control group, which could obscure a genuine training effect. The isokinetic and isometric results, which showed consistent training effects, are less susceptible to this since dynamometry testing is more standardised across visits.


#--- Emmeans -------------------------------------------------------------------

# Extracting emmeans for plotting
emm_keiser_con <- emmeans(mod_keiser_con, ~ time | trained)
saveRDS(emm_keiser_con, "./data/data-gen/emm_keiser_con.rds")


#--- Pre-post contrasts for CI -------------------------------------------------

emmeans(mod_keiser_con, ~ time | trained) %>%
  contrast("pairwise") %>%
  summary(infer = TRUE, adjust = "none")


```

**Result summary model 10: 1RM leg press old intervention/control**

Keiser Leg Press 1RM — Control model (mod_keiser_con)

Fixed effects (ANOVA):
- Trained:        F(1, 39.0) = 0.43,  p = .518
- Time:           F(2, 195.1) = 24.53, p < .001
- Trained × Time: F(2, 195.1) = 0.91,  p = .405

Pre→post change within group (emmeans, 95% CI):
- Control: Pre→Mid:  +12 kg [0, 24],    p = .048
           Pre→Post: +19 kg [7, 30],    p = .002
           Mid→Post: +7 kg  [-5, 18],   p = .255
- Trained: Pre→Mid:  +17 kg [11, 23],   p < .001
           Pre→Post: +28 kg [21, 34],   p < .001
           Mid→Post: +11 kg [4, 17],    p = .001

Trained vs. control at each timepoint:
- Pre:  groups did not differ (p = .617, baseline balanced)

Bottom line: Both trained and control older adults increased leg press over time. Trained × Time non-significant (p = .405) — no differential training response. Controls improving substantially (pre→post +19 kg) is the familiarisation artefact flagged previously. No baseline imbalance (p = .617).


### Modelling isometric bicep curl

The code chunks below contains the model for analysing isometric bicep curl at baseline, mid, and post. The data itself is described in the "reliefdata" package, but briefly, we are using max force values derived from the MuscleLab software. The model uses the same logic as the ones for the isometric and isokinetic data.

#### Model 11: Isometric bicep curl young/old intervention

This next model (11) compares the young and old intervention groups.

```{r, model 11}

#--- Model ---------------------------------------------------------------------

mod_bc_int <- lmer(mvc ~ age_grp * condition * time + (1 | participant),
                   data = bc_int)

summary(mod_bc_int)
anova(mod_bc_int)

saveRDS(bc_int, "./data/data-gen/bc_int.rds")


#--- Emmeans -------------------------------------------------------------------

# Extracting emmeans for plotting
emm_bc_int <- emmeans(mod_bc_int, ~ time * condition | age_grp)
saveRDS(emm_bc_int, "./data/data-gen/emm_bc.rds")


# Contrasts (given the age x condition sig)
emmeans(mod_bc_int, ~ condition | age_grp) %>%
  contrast("pairwise", infer = TRUE)


#--- Pre-post contrasts for CI -------------------------------------------------

emmeans(mod_bc_int, ~ age_grp * condition * time) %>%
  contrast("pairwise", by = c("age_grp", "condition")) %>%
  summary(infer = TRUE, adjust = "none")



```

**Result summary model 11: isometric bicep curl young/old intervention**

Bicep Curl MVC — Intervention model (mod_bc_int)

Fixed effects (ANOVA):
- Age group:           F(1, 69.2)  = 0.63,  p = .430
- Condition:           F(1, 299.2) = 0.68,  p = .410
- Time:                F(2, 300.0) = 4.22,  p = .016
- Age × Condition:     F(1, 299.2) = 4.49,  p = .035
- Age × Time:          F(2, 300.0) = 1.54,  p = .215
- Condition × Time:    F(2, 299.2) = 0.48,  p = .616
- Age × Cond × Time:   F(2, 299.2) = 0.34,  p = .711

Subgroup changes (emmeans, 95% CI):

Old, Low:
- Pre→Mid:  +0 Nm  [-5, 5],    p = .906
- Pre→Post: +4 Nm  [-1, 10],   p = .093
- Mid→Post: +4 Nm  [-1, 9],    p = .124
Old, Mod:
- Pre→Mid:  +0 Nm  [-5, 5],    p = .895
- Pre→Post: +4 Nm  [-1, 10],   p = .099
- Mid→Post: +5 Nm  [0, 11],    p = .080
Young, Low:
- Pre→Mid:  +6 Nm  [0, 12],    p = .037
- Pre→Post: +6 Nm  [0, 12],    p = .044
- Mid→Post: +0 Nm  [-6, 6],    p = .966
Young, Mod:
- Pre→Mid:  +2 Nm  [-4, 8],    p = .511
- Pre→Post: +2 Nm  [-4, 8],    p = .549
- Mid→Post: +0 Nm  [-6, 6],    p = .961

Age × Condition contrast (averaged over time):
- Old:   Low vs. Mod: -1.5 Nm [-4.4, 1.5],  p = .334
- Young: Low vs. Mod: +3.3 Nm [0.02, 6.6],  p = .049

Bottom line: Significant Age × Condition interaction (p = .035). Young adults on low volume outperformed young adults on mod — an unexpected reversal. Old adults showed no condition difference. Time effect present (p = .016) but modest. Young Low the only group with significant pre→post gains; Young Mod and both Old groups CIs largely cross zero. This outcome warrants careful interpretation — possible regression to mean or randomisation imbalance in young mod.

#### Model 12: Isometric bicep curl old intervention/control

This next model (12) compares the old intervention and control groups.

```{r, model 12}

#--- Model ---------------------------------------------------------------------

mod_bc_con <- lmer(mvc ~ trained * time + (1 | participant), data = bc_con)

summary(mod_bc_con)
anova(mod_bc_con)

saveRDS(bc_con, "./data/data-gen/bc_con.rds")


#--- Emmeans -------------------------------------------------------------------

# Extracting emmeans for plotting
emm_bc_con <- emmeans(mod_bc_con, ~ time | trained)
saveRDS(emm_bc_con, "./data/data-gen/emm_bc_con.rds")


# Contrasts
emm_bc_con_pre <- emmeans(mod_bc_con, ~ trained | time)
contrast(emm_bc_con_pre, "pairwise", infer = TRUE)

#--- Pre-post contrasts for CI -------------------------------------------------

emmeans(mod_bc_con, ~ time | trained) %>%
  contrast("pairwise") %>%
  summary(infer = TRUE, adjust = "none")


```

**Result summary model 12: isometric bicep curl old intervention/control**

Bicep Curl MVC — Control model (mod_bc_con)

Fixed effects (ANOVA):
- Trained:        F(1, 46.9) = 5.00,  p = .030
- Time:           F(2, 217.3) = 2.38,  p = .095
- Trained × Time: F(2, 217.3) = 3.76,  p = .025

Pre→post change within group (emmeans, 95% CI):
- Control: Pre→Mid:  -9 Nm  [-16, -2],  p = .018
           Pre→Post: -7 Nm  [-14, 1],   p = .080
           Mid→Post: +2 Nm  [-5, 9],    p = .535
- Trained: Pre→Mid:  +0 Nm  [-4, 4],    p = .998
           Pre→Post: +4 Nm  [0, 9],     p = .033
           Mid→Post: +4 Nm  [0, 9],     p = .033

Trained vs. control at each timepoint:
- Pre:  Int > Con by 28 Nm [-4, 60],    p = .080
- Mid:  Int > Con by 37 Nm [5, 69],     p = .023
- Post: Int > Con by 39 Nm [8, 71],     p = .016

Bottom line: Trained × Time significant (p = .025). Trained older adults gained modestly pre→post; controls declined pre→mid before recovering. The control decline is notable — possible fatigue, learning, or measurement artefact. Borderline baseline imbalance (p = .080). Interpret with caution.

**Comments**

The baseline imbalance in control models is consistent across all isokinetic outcomes and isometric — trained older adults were already stronger at pre. The Keiser is the exception with no baseline difference, which is also why the familiarisation artefact interpretation is cleanest there.

The bicep curl control group declining pre→mid is unusual and worth flagging — it's not a pattern seen elsewhere and could affect how you interpret the trained × time interaction in that model.

## TODO

- Daniel, you might want to look over these.
- Use change scores instead? Thats what I did for the ribose paper.

