---
title: "Body composition"
format: html
editor: source
editor_options: 
  chunk_output_type: console
---


```{r}
#| label: packages-and-settings
#| eval: true
#| message: false
#| warning: false
#| echo: true

# Loading packages
library(brms)
library(marginaleffects)
library(tidyverse)
library(reliefdata)


# This setting disables a safety warning message for the marginaleffects package
# we should still confirm the order of effects for the contrasts
options(marginaleffects_safe = FALSE)


```

## Muscle volume
<!-- edit this text -->
Muscle volume is estimate from magnetic resonance imaging (MRI). In the data set we have estimates of muscle, intramuscular fat, subcutaneous fat, and bone volumes. MRI measurements were conducted before and after the intervention (two time-points).

We will start by modelling the muscle volumes and we are interested in estimating:

1.  Within condition (low- and moderate-volume, and non-training control conditions) change from baseline.
2.  Differences between age-groups at baseline.
3.  Average between age-group differences in change from baseline (young vs. old in the intervention group)
4.  Average between control- vs. intervention-group changes from baseline.
5.  Average effect of condition in the intervention groups (low- vs. moderate volume)
6.  Differences between age-groups in effect of condition.

The first contrast will be built for descriptive purposes and serves as the building block for comparisons in contrast 3-4.

### Data
<!-- edit this text -->
In the code chunk below we plot the data grouped by participant leg.


<!-- edit this text -->
The raw data indicate separation between sex and age in muscle volume across time-points. Large variation in muscle volumes between participants suggest that a relevant comparison is the relative change in muscle volume.

### Models 
<!-- edit this text -->
We will construct a set of models for the outcome variable muscle volume after transformation to the log scale and standardization. The models will assume a Normal likelihood $y \sim \operatorname{Normal}(\mu_i, \sigma)$.

In model 1

$$
\begin{align}
\mu_i &= 
\end{align}
$$

### Isometric force and peak torque

```{r}

# Data


humac.dat <- reliefdata::relief_humac %>% 
  mutate(participant = as.character(participant))

condition <- reliefdata::relief_volume %>% 
  mutate(participant = as.character(participant))

participants <- reliefdata::relief_participants %>% 
  mutate(age_group = if_else(age < 40, "yng", "old"))


humac_dat <- humac.dat %>% 
  full_join(condition) %>% 
  select(-arm) %>% 
  full_join(participants) %>%  
  
  filter(!participant %in% c("3004", "3005", "3010", "3033",
                          "3034", "3046", "3053", "3069",
                          "3092", "3010", "3019", "3026",
                          "3017", "3012")) %>% # dropouts/missing all measurements
  #filter(!is.na(time), !is.na(speed)) %>%
  group_by(participant, time, timerep, leg, allocation, speed, condition, age_group, sex, weight) %>%
  summarise(pt = mean(pt, na.rm = TRUE), .groups = "drop") %>%
  mutate(
         ystd = pt / max(pt, na.rm = TRUE),  
         
         condition = if_else(allocation == "con", "con", condition),
         tx = paste0(allocation, "_", age_group, "_", condition),
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con")),
         tx_sex = paste0(tx, "_", sex), 
         tx_sex = factor(tx_sex, levels = c(
           "int_old_low_female", "int_old_low_male",
           "int_old_mod_female", "int_old_mod_male", 
           "int_yng_low_female", "int_yng_low_male",
           "int_yng_mod_female", "int_yng_mod_male", 
           "con_old_con_female", "con_old_con_male"
         )),
         time = factor(time, levels = c("pre", "mid", "post")), 
         sex_centered = if_else(sex == "female", -0.5, 0.5),
         speed = factor(speed, levels = c("0", "60", "120", "240"))) 



# A preliminary figure of observed data

humac_dat %>% 
  # Change factor names for plotting
  mutate(time = factor(time, labels = c("Pre", "Mid", "Post")),
         age_group = factor(age_group, labels = c("Old", "Young")),
         allocation = factor(allocation, labels = c("Control", "Intervention")),
         sex = factor(sex, labels = c("Female", "Male")),
         speed = factor(speed, labels = c("0", "60", "120", "240"))) %>% 
  # Summarising the timereps for the visualisation only
  group_by(participant, time, leg, allocation, speed, condition, age_group, sex, tx, tx_sex) %>% 
  summarise(pt = mean(pt, na.rm = TRUE)) %>% 
  ggplot(aes(time, pt,
             color = sex,
             group = paste(participant, leg))) +
  geom_line() +
  facet_wrap(~ paste(age_group, allocation, speed)) +
  labs(x = "", y = "Force/Peak Torque (N/Nm)") +
  theme(axis.title.y = ggtext::element_markdown())
  


## Save sd and mean for back transform
mean_sd_pt <- humac_dat %>% 
  summarise(mean_pt = mean(pt, na.rm = TRUE),
         sd_pt = sd(pt, na.rm = TRUE), 
         max = max(pt, na.rm = TRUE))





```

### Baseline only

```{r}

# Starting with a baseline check
humac_dat2 <- humac_dat %>% 
    filter(time == "pre") %>% 
  mutate(sp = as.numeric(as.character(speed)),
         sp = sp / 240,
         sp2 = sp^2) %>% 
  filter(pt > 0) 



form0 <- bf(ystd ~ allocation + allocation:sp + allocation:sp2 + 
             (1 + sp + sp2|participant) +
             (1 + sp + sp2|tx_sex))


# Priors

get_prior(form0, data = humac_dat2)


pri0 <- c(
  
  # Allocation main effect (int vs con baseline difference; con is reference, absorbed into Intercept)
  prior(normal(0.6, 0.4), class = "b", coef = "allocationint"),
  
  # Allocation:sp interactions
  prior(normal(0.6, 0.4), class = "b", coef = "allocationcon:sp"),
  prior(normal(0.6, 0.4), class = "b", coef = "allocationint:sp"),
  
  # Allocation:sp2 interactions
  prior(normal(0.6, 0.4), class = "b", coef = "allocationcon:sp2"),
  prior(normal(0.6, 0.4), class = "b", coef = "allocationint:sp2"),
  
  # Random effects — participant, split by term
  prior(exponential(1), class = "sd", group = "participant", coef = "Intercept"),
  prior(exponential(1), class = "sd", group = "participant", coef = "sp"),
  prior(exponential(1), class = "sd", group = "participant", coef = "sp2"),
  
  # Random effects — tx_sex, split by term
  prior(exponential(1), class = "sd", group = "tx_sex", coef = "Intercept"),
  prior(exponential(1), class = "sd", group = "tx_sex", coef = "sp"),
  prior(exponential(1), class = "sd", group = "tx_sex", coef = "sp2"),
  
  # Residual SD
  prior(exponential(1), class = "sigma")
)

m0.pp <- brm(form0,
             prior = pri0,
             data = humac_dat2,
             file = "fits/04-strength/m0",
             sample_prior = "only")

# Set up predictions/contrasts for model 0


m0_pred_contr <- function(m, nd1, nd2) {
  grid <- humac_dat2 %>% 
    distinct(tx_sex, allocation) %>% 
    expand_grid(sp = seq(min(humac_dat2$sp), max(humac_dat2$sp), length.out = 50)) %>%
    mutate(sp2 = sp^2)
  
  contr <- predictions(
    m,
    newdata = grid,
    re_formula = ~ (sp + sp2 | tx_sex),
    ndraws = nd1,
    by = "tx_sex"
  )
  
  pred <- predictions(
    m,
    ndraws = nd2,
    newdata = grid,
    re_formula = ~ (sp + sp2 | tx_sex))
  
  return(list(contr = contr, pred = pred))
}


m0_preds <- m0_pred_contr(m0.pp, nd1 = 1000, nd2 = 50)

# Prior predictive simulations

posterior_draws(m0_preds$pred) %>% 
  data.frame() %>% 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"),
         tx = gsub("_female", "", tx_sex),
         tx = gsub("_male", "", tx),
         tx = factor(tx, levels = c("int_old_low", "int_old_mod", "int_yng_low",
                                    "int_yng_mod", "con_old_con"))) %>% 
  mutate(model = "m0") %>% 
  ggplot(aes(sp, draw, group = paste(drawid, tx, sex),
             color = sex)) +
  geom_line() +
  facet_grid(model ~ tx)


posterior_draws(m0_preds$contr) %>% 
  mutate(model = "m0") %>% 
  ggplot(aes(draw, color = model)) +
  geom_line(stat = "density") +
  #facet_wrap()
  geom_vline(xintercept = 0, lty = 2) +
  coord_cartesian(xlim = c(-5, 5))




```

```{r}

### Get the posterior distribution

m0.post <- update(m0.pp,
                  sample_prior = "no",
                  iter = 4000)

# this produces:
#Warning messages:
#1: There were 4 divergent transitions after warmup. See
#https://mc-stan.org/misc/warnings.html#divergent-transitions-after-warmup
#to find out why this is a problem and how to eliminate them. 
#2: There were 6 transitions after warmup that exceeded the maximum treedepth. Increase max_treedepth above #10. See
#https://mc-stan.org/misc/warnings.html#maximum-treedepth-exceeded 
#3: Examine the pairs() plot to diagnose sampling problems

pp_check(m0.post, type = "dens_overlay_grouped",
         group = "tx_sex")

pp_check(m0.post)


```

```{r}

# Group comparisons

m0_preds <- m0_pred_contr(m0.post, nd1 = 4000, nd2 = 50)

posterior_draws(m0_preds$pred) %>% 
  data.frame() %>% 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"),
         tx = gsub("_female", "", tx_sex),
         tx = gsub("_male", "", tx),
         tx = factor(tx, levels = c("int_old_low", "int_old_mod", "int_yng_low",
                                    "int_yng_mod", "con_old_con"))) %>% 
  mutate(model = "m0") %>% 
  ggplot(aes(sp, draw, group = paste(drawid, tx, sex),
             color = sex)) +
  geom_line() +
  facet_grid(model ~ tx)


posterior_draws(m0_preds$contr) %>% 
  mutate(model = "m0") %>% 
  ggplot(aes(draw, color = model)) +
  geom_line(stat = "density") +
  facet_wrap(~ tx_sex) +
  geom_vline(xintercept = 0, lty = 2) +
  coord_cartesian(xlim = c(-0.5, 0.5))


####################### Predictions for plotting ###############################


p1 <- posterior_draws(m0_preds$pred) %>% 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) %>% 
  mutate(draw = draw * mean_sd_pt$max,
         sp = sp * 240) %>% # 240 is the highest observed angular velocity 

 summarise(.by = c(tx, sex, sp), 
           m = mean(draw), 
           lwr50 = quantile(draw, 0.25), 
           upr50 = quantile(draw, 0.75),  
           lwr90 = quantile(draw, 0.05), 
           upr90 = quantile(draw, 0.95),
           lwr95 = quantile(draw, 0.025), 
           upr95 = quantile(draw, 0.975)) %>%
 
 ggplot(aes(sp, m, color = sex, group = paste(tx, sex))) +
  
  geom_ribbon(aes(ymin = lwr95, ymax = upr95, fill = sex, color = NULL), alpha = 0.15) +
  geom_ribbon(aes(ymin = lwr90, ymax = upr90, fill = sex, color = NULL), alpha = 0.2) +
  geom_ribbon(aes(ymin = lwr50, ymax = upr50, fill = sex, color = NULL), alpha = 0.3) +
  geom_line(linewidth = 1) +
  
  facet_wrap(~ tx)

# The plot above shows:
# Facets by age group and training condition
# Within each facet the predicted peak torque of male vs. female is shown across the full range of angular velocity(from sp, rescaled 0-240d/s)
# The bands (50/90/95%) are posterior uncertainty in the mean prediction at each speed - not raw individual variability. A narrower band thus means higher confidence on where that group average sits.


p2 <- posterior_draws(m0_preds$contr) %>%  
 
  
  mutate(draw = draw * mean_sd_pt$sd_pt) %>%  
  
  ggplot(aes(draw)) + geom_line(stat = "density") + 
  facet_wrap(~ tx_sex) +
  geom_vline(xintercept = 1, lty = 2)


```

### All time points


```{r}

humac_dat2 <- humac_dat %>% 
  mutate(sp = as.numeric(as.character(speed)),
         sp = sp / 240,
         sp2 = sp^2) %>% 
  filter(pt > 0) 




form1 <- bf(ystd ~ 0 + tx_sex + tx_sex:time + tx_sex:sp + tx_sex:sp2 +
                    (1 + time + sp + sp2 | participant) +
                    (1 | participant:leg) +
                    (1 | participant:time:timerep))



# Priors

get_prior(form1, data = humac_dat2)

# The below code was suggested by claude, but it contains nothing for the random effects as we've had in previous priors - start by checking this. Get back to the claude convo by accessing the "posterior draws visualization..." chat. 

pri1 <- c(
  # Intercept-equivalent: tx_sex main effects (this is the level at sp=0, time=pre)
  set_prior("normal(0.4, 0.3)", class = "b", coef = paste0("tx_sex", levels(humac_dat2$tx_sex))),
  
  # sp slope - expect negative (torque declines with speed), moderate magnitude
  set_prior("normal(-0.5, 0.5)", class = "b", 
             coef = paste0("tx_sex", levels(humac_dat2$tx_sex), ":sp")),
  
  # sp2 - smaller curvature term, less certain of sign in general but anchor near test
  set_prior("normal(0.2, 0.5)", class = "b", 
             coef = paste0("tx_sex", levels(humac_dat2$tx_sex), ":sp2")),
  
  # time effects - mid/post vs pre, expect small shifts on 0-1 scale, centered at 0
  set_prior("normal(0, 0.2)", class = "b", 
             coef = paste0("tx_sex", levels(humac_dat2$tx_sex), ":timemid")),
  set_prior("normal(0, 0.2)", class = "b", 
             coef = paste0("tx_sex", levels(humac_dat2$tx_sex), ":timepost"))
)


############### OLD ###########################

pri0 <- c(
  
  # Allocation main effect (int vs con baseline difference; con is reference, absorbed into Intercept)
  prior(normal(0.6, 0.4), class = "b", coef = "allocationint"),
  
  # Allocation:sp interactions
  prior(normal(0.6, 0.4), class = "b", coef = "allocationcon:sp"),
  prior(normal(0.6, 0.4), class = "b", coef = "allocationint:sp"),
  
  # Allocation:sp2 interactions
  prior(normal(0.6, 0.4), class = "b", coef = "allocationcon:sp2"),
  prior(normal(0.6, 0.4), class = "b", coef = "allocationint:sp2"),
  
  # Random effects — participant, split by term
  prior(exponential(1), class = "sd", group = "participant", coef = "Intercept"),
  prior(exponential(1), class = "sd", group = "participant", coef = "sp"),
  prior(exponential(1), class = "sd", group = "participant", coef = "sp2"),
  
  # Random effects — tx_sex, split by term
  prior(exponential(1), class = "sd", group = "tx_sex", coef = "Intercept"),
  prior(exponential(1), class = "sd", group = "tx_sex", coef = "sp"),
  prior(exponential(1), class = "sd", group = "tx_sex", coef = "sp2"),
  
  # Residual SD
  prior(exponential(1), class = "sigma")
)

m0.pp <- brm(form0,
             prior = pri0,
             data = humac_dat2,
             file = "fits/04-strength/m0",
             sample_prior = "only")

# Set up predictions/contrasts for model 0


m0_pred_contr <- function(m, nd1, nd2) {
  grid <- humac_dat2 %>% 
    distinct(tx_sex, allocation) %>% 
    expand_grid(sp = seq(min(humac_dat2$sp), max(humac_dat2$sp), length.out = 50)) %>%
    mutate(sp2 = sp^2)
  
  contr <- predictions(
    m,
    newdata = grid,
    re_formula = ~ (sp + sp2 | tx_sex),
    ndraws = nd1,
    by = "tx_sex"
  )
  
  pred <- predictions(
    m,
    ndraws = nd2,
    newdata = grid,
    re_formula = ~ (sp + sp2 | tx_sex))
  
  return(list(contr = contr, pred = pred))
}


m0_preds <- m0_pred_contr(m0.pp, nd1 = 1000, nd2 = 50)

# Prior predictive simulations

posterior_draws(m0_preds$pred) %>% 
  data.frame() %>% 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"),
         tx = gsub("_female", "", tx_sex),
         tx = gsub("_male", "", tx),
         tx = factor(tx, levels = c("int_old_low", "int_old_mod", "int_yng_low",
                                    "int_yng_mod", "con_old_con"))) %>% 
  mutate(model = "m0") %>% 
  ggplot(aes(sp, draw, group = paste(drawid, tx, sex),
             color = sex)) +
  geom_line() +
  facet_grid(model ~ tx)


posterior_draws(m0_preds$contr) %>% 
  mutate(model = "m0") %>% 
  ggplot(aes(draw, color = model)) +
  geom_line(stat = "density") +
  #facet_wrap()
  geom_vline(xintercept = 0, lty = 2) +
  coord_cartesian(xlim = c(-5, 5))




```

```{r}

### Get the posterior distribution

m0.post <- update(m0.pp,
                  sample_prior = "no",
                  iter = 4000)

# this produces:
#Warning messages:
#1: There were 4 divergent transitions after warmup. See
#https://mc-stan.org/misc/warnings.html#divergent-transitions-after-warmup
#to find out why this is a problem and how to eliminate them. 
#2: There were 6 transitions after warmup that exceeded the maximum treedepth. Increase max_treedepth above #10. See
#https://mc-stan.org/misc/warnings.html#maximum-treedepth-exceeded 
#3: Examine the pairs() plot to diagnose sampling problems

pp_check(m0.post, type = "dens_overlay_grouped",
         group = "tx_sex")

pp_check(m0.post)


```

```{r}

# Group comparisons

m0_preds <- m0_pred_contr(m0.post, nd1 = 4000, nd2 = 50)

posterior_draws(m0_preds$pred) %>% 
  data.frame() %>% 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"),
         tx = gsub("_female", "", tx_sex),
         tx = gsub("_male", "", tx),
         tx = factor(tx, levels = c("int_old_low", "int_old_mod", "int_yng_low",
                                    "int_yng_mod", "con_old_con"))) %>% 
  mutate(model = "m0") %>% 
  ggplot(aes(sp, draw, group = paste(drawid, tx, sex),
             color = sex)) +
  geom_line() +
  facet_grid(model ~ tx)


posterior_draws(m0_preds$contr) %>% 
  mutate(model = "m0") %>% 
  ggplot(aes(draw, color = model)) +
  geom_line(stat = "density") +
  facet_wrap(~ tx_sex) +
  geom_vline(xintercept = 0, lty = 2) +
  coord_cartesian(xlim = c(-0.5, 0.5))


####################### Predictions for plotting ###############################


p1 <- posterior_draws(m0_preds$pred) %>% 
  mutate(sex = if_else(grepl("_female", tx_sex), "female", "male"), 
         tx = gsub("_female", "", tx_sex), 
         tx = gsub("_male", "", tx), 
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con"))) %>% 
  mutate(draw = draw * mean_sd_pt$max,
         sp = sp * 240) %>% # 240 is the highest observed angular velocity 

 summarise(.by = c(tx, sex, sp), 
           m = mean(draw), 
           lwr50 = quantile(draw, 0.25), 
           upr50 = quantile(draw, 0.75),  
           lwr90 = quantile(draw, 0.05), 
           upr90 = quantile(draw, 0.95),
           lwr95 = quantile(draw, 0.025), 
           upr95 = quantile(draw, 0.975)) %>%
 
 ggplot(aes(sp, m, color = sex, group = paste(tx, sex))) +
  
  geom_ribbon(aes(ymin = lwr95, ymax = upr95, fill = sex, color = NULL), alpha = 0.15) +
  geom_ribbon(aes(ymin = lwr90, ymax = upr90, fill = sex, color = NULL), alpha = 0.2) +
  geom_ribbon(aes(ymin = lwr50, ymax = upr50, fill = sex, color = NULL), alpha = 0.3) +
  geom_line(linewidth = 1) +
  
  facet_wrap(~ tx)

# The plot above shows:
# Facets by age group and training condition
# Within each facet the predicted peak torque of male vs. female is shown across the full range of angular velocity(from sp, rescaled 0-240d/s)
# The bands (50/90/95%) are posterior uncertainty in the mean prediction at each speed - not raw individual variability. A narrower band thus means higher confidence on where that group average sits.


p2 <- posterior_draws(m0_preds$contr) %>%  
 
  
  mutate(draw = draw * mean_sd_pt$sd_pt) %>%  
  
  ggplot(aes(draw)) + geom_line(stat = "density") + 
  facet_wrap(~ tx_sex) +
  geom_vline(xintercept = 1, lty = 2)


```


































