---
title: "RNA x mg analysis"
format: html
editor: source
editor_options: 
  chunk_output_type: console
---


```{r}
#| label: packages-and-settings
#| eval: true
#| message: false
#| warning: false
#| echo: true

# Loading packages
library(brms)
library(marginaleffects)
library(tidyverse)
library(reliefdata)



# This setting disables a safety warning message for the marginaleffects package
# we should still confirm the order of effects for the contrasts
options(marginaleffects_safe = FALSE)


```


### TLDR

Built three models with RNA yield and mg log transformed, mean centered and standardize it by scaling via sd(rna). 

form0: ystd ~ mgstd
model 1 simply tests the baseline relationship by regressing standardized log-RNA on standardized log-mg alone, no grouping structure. This establishes B1 (the mg-RNA slope) in its rawest form, but treats all 388 observations as independent, ignoring that multiple observations come from the same participant. Results show that mgstd = 0.85, credibly below 1.

form1: ystd ~ mgstd + (1|participant)
Same fixed relationship as form0, but adds a random intercept for participant to account for repeated measures (multiple leg x time observations per person). Separates between-participant variance from residual variance rather than lumping both into one error term. Confirms the mg-RNA slope isn't an artifact of pooling across participants, B1 stays stable at 0.84.

form2: ystd ~ mgstd + tx + time + (1|participant)
Adds two fixed effects, both purely additive (no interactions): tx (the 5-level group factor age x allocation x volume combined) and time (pre/w3/post). Tests whether RNA level shifts by group (young_int, old_int, old_con, B2) and by time (B3), independent of tissue mass and independent of each other. Using tx as a stand-in for simply "age" here, to prevent collapsing old_int and old_con together as "old". 

form3: ystd ~ mgstd + tx:time + (1 + time|participant) + (1|participant:leg)
The full interaction model. tx:time expands to tx + time + tx:time, letting the shape of the time trajectory differ by group, not just the average level. This tests whether e.g., the control group's RNA rises differently over time than the intervention groups. Random-effects side also grows: (1 + time|participant) allows each participant's trajectory to vary around the population trend (not just their baseline level), and (1|participant:leg) accounts for leg-level variation nested within participant, relevant since volume condition is assigned at the leg level for intervention participants. The interaction terms were all null - no credible evidence that trajectories differ by group, on top of the m2 finding that controls sit at a persistently lower RNA level throughout. 

![Total RNA yield increases with tissue weight on a log-log scale, but with a slope below 1, indicating a sub-proportional relationship rather than the fixed ratio assumed by RNA/mg standardization. Lines are colored by timepoint (pre, week 3, post); the slight separation hints at a time-dependent shift in RNA yield at a given tissue mass, later confirmed formally in the model results.](figures/rna_mg_relationship.png)


__Questions:__

1. Was B2 × age intended as the full tx group structure (used here) or a simpler two-level age contrast (deferred to later)?

2. Given the null interactions, is the added complexity of tx:time worth retaining over the simpler additive m2, or is establishing that null itself the point?


### The relationship between total RNA and mg tissue

The current RELIEF RNA analysis pipeline standardizes RNA yield using a fixed
RNA/mg ratio, which implicitly assumes a strictly proportional relationship
(slope = 1, intercept = 0) between tissue mass and RNA yield, and treats
mg weight as measured without error. In practice, weighing small tissue
pieces carries real measurement error, which is amplified when dividing by
small mg values — potentially contributing to the sampling instability seen
in the UBF and c-Myc models.

This document explores an alternative: rather than pre-computing a fixed
ratio, we model the relationship between measured RNA and mg weight
directly (on the log-log scale, consistent with the project's existing
log + standardization convention). The aim is to let the mg-RNA
relationship be estimated from the data rather than assumed, which should:

- allow uncertainty to scale appropriately for extreme mg observations,
  rather than being artificially inflated or deflated by the ratio, and
- yield more precise estimates in the well-populated middle of the mg
  distribution, where the fixed-ratio approach was not actually using the
  available information efficiently.


#### Data

In the code chunk below we plot the data grouped by participant leg. Instead of making a ratio between total rna and mg here, we log-transform the data, then mean center it (rna-mean(rna)) and then standardising it by scaling (/sd(rna)). This will be applied to both the logged RNA yield and mg, which will be regressed like "ystd ~ mgstd" = standardised slope (B1)/"elasticity-lik" quantity saying: "for every 1 SD increase in log(tissue mass), how many SDs does log(RNA) shift". So, its no longer a raw biological unit like ng RNA per mg tissue, but it is directly comparable in scale to other standardised coefficients in our other brms models. And of course, also back-transformable via exp(draw x sd_rna) logic, which will be applied for reporting the results.  

```{r}

# Data

conc.dat <- reliefdata::relief_rnaconc |> 
  mutate(participant = as.character(participant)) |> 
  mutate(participant = ifelse(is.na(participant), "blank", participant),
         leg = ifelse(is.na(leg), "blank", leg),
         time = ifelse(is.na(time), "blank", time),
         weight = ifelse(is.na(weight), "0", weight)) |> 
  mutate(total.rna = conc * 4 * 25) |> 
  dplyr::select(participant, leg, time, weight, well, plate, sample, total.rna) |> 
  filter(participant != "blank") |> 
  mutate(weight = as.numeric(weight)) |> 
  print()

condition <- reliefdata::relief_volume |> 
  mutate(participant = as.character(participant))

participants <- reliefdata::relief_participants |> 
  mutate(age_group = if_else(age < 40, "yng", "old")) |> 
  dplyr::select(participant, age, sex, height, bodyweight = weight, allocation, age_group)


rna.dat <- conc.dat |> 
  full_join(condition) |> 
  select(-arm) |> 
  full_join(participants) |> 
  group_by(participant, leg, time, age_group, allocation, condition, sex) |> 
  summarise(
    mean_rna = mean(total.rna, na.rm = TRUE), # average technical replicates (conc side)
    weight = first(weight)                    # one weight measurement per sample
  ) |> 
  ungroup() |> 
  mutate(
    y     = log(mean_rna),
    ystd  = (y - mean(y, na.rm = TRUE)) / sd(y, na.rm = TRUE),
    mg    = log(weight),
    mgstd = (mg - mean(mg, na.rm = TRUE)) / sd(mg, na.rm = TRUE),
    condition = if_else(allocation == "con", "con", condition),
    tx = paste0(allocation, "_", age_group, "_", condition),
    tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con")),
    tx_sex = paste0(tx, "_", sex),
    tx_sex = factor(tx_sex, levels = c(
      "int_old_low_female", "int_old_low_male",
      "int_old_mod_female", "int_old_mod_male", 
      "int_yng_low_female", "int_yng_low_male",
      "int_yng_mod_female", "int_yng_mod_male", 
      "con_old_con_female", "con_old_con_male"
    )),
    time = factor(time, levels = c("pre", "w3", "post")),
    sex_centered = if_else(sex == "female", -0.5, 0.5)
  ) |> 
  filter(!is.na(y)) |> 
  filter(!is.na(mg)) |> 
  filter(!is.na(time)) |> 
  print()
    
# Checking for mg = 0, which would produce inf
#sum(is.infinite(rna.dat$mg))
# zero
  

## Save sd and mean for back transformation
mean_rna <- mean(rna.dat$y, na.rm = TRUE)
sd_rna <- sd(rna.dat$y, na.rm = TRUE)
mean_mg <- mean(rna.dat$mg, na.rm = TRUE)
sd_mg <- sd(rna.dat$mg, na.rm = TRUE)

# A preliminary figure of observed data
# RNA plotted against weight - the relationship to be estimated in the moodel

rna_mg_plot <- rna.dat %>% 
  ggplot(aes(weight, mean_rna, color = time)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Tissue weight (mg, log scale)", 
       y = "Total RNA (ng, log scale)")


rna.dat %>% 
  ggplot(aes(mgstd, ystd, color = time)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +

  labs(x = "Tissue weight (mg, log scale, standardized)", 
       y = "Total RNA (ng, log scale, standardized)")



#ggsave("figures/rna_mg_relationship.png", 
#       plot = rna_mg_plot,
#       width = 7, height = 5, dpi = 300)


## Data for model 4
# Model 4 will have a random intercept per biopsy replicate, so Im making a separate data frame here without summarising to mean_rna per biopsy.


rna_rep.dat <- conc.dat |> 
  full_join(condition) |> 
  select(-arm) |> 
  full_join(participants) |> 
  mutate(
    y     = log(total.rna),
    ystd  = (y - mean(y, na.rm = TRUE)) / sd(y, na.rm = TRUE),
    mg    = log(weight),
    mgstd = (mg - mean(mg, na.rm = TRUE)) / sd(mg, na.rm = TRUE),
    condition = if_else(allocation == "con", "con", condition),
    tx = paste0(allocation, "_", age_group, "_", condition),
    tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con")),
    tx_sex = paste0(tx, "_", sex),
    tx_sex = factor(tx_sex, levels = c(
      "int_old_low_female", "int_old_low_male",
      "int_old_mod_female", "int_old_mod_male", 
      "int_yng_low_female", "int_yng_low_male",
      "int_yng_mod_female", "int_yng_mod_male", 
      "con_old_con_female", "con_old_con_male"
    )),
    time = factor(time, levels = c("pre", "w3", "post")),
    sex_centered = if_else(sex == "female", -0.5, 0.5)
  ) |> 
  filter(!is.na(y)) |> 
  filter(!is.na(mg)) |> 
  filter(!is.na(time)) |> 
  print()

mean_rna_rep <- mean(rna_rep.dat$y, na.rm = TRUE)
sd_rna_rep   <- sd(rna_rep.dat$y, na.rm = TRUE)
mean_mg_rep  <- mean(rna_rep.dat$mg, na.rm = TRUE)
sd_mg_rep    <- sd(rna_rep.dat$mg, na.rm = TRUE)

```

#### Models

- Update priors to positive slope for mg tissue



```{r}
# Setting up formulas
form0 <- bf(ystd ~ mgstd) # No grouping, a clean look at the relationship

form1 <- bf(ystd ~ mgstd + (1|participant)) # Now adding a random intercept for participant, to account for repeated measures
# this also means that the model now separates between-participant variance (sd(Intercept)) from residual variance(sigma), i.e., sigma should drop compared to m0.

form2 <- bf(ystd ~ mgstd + tx + time + (1|participant))
# this model has tx and time added, to also explore age, allocation and time effects on the relationship between RNA yield and mg. The initial idea was to only check age and time, but using age_group here would collapse int_old and con_old, which should stay separate. "tx" fixes that, but also adds the allocation and condition. An alternative would be to create a new factor that includes only age group and allocation (int_young, int_old, con_old). Written as is, this model is additive, and tests "does group shift RNA level" and "does time shift RNA level".

# What model 2 cannot tell us, is whether the time trajectory itself differs by group - does the control group's RNA stay flat across time while interventions group rise? We can test this by adding tx:time. Probably also time to expand the random effects, with time and tx in the model. Adding time lets each participant's time trajectory vary, and one for leg lets the model capture if a person's low-volume leg runs differently from the moderate-volume leg.

form3 <- bf(ystd ~ mgstd + tx * time + (1 + time | participant) + (1 | participant:leg))

form4 <- bf(ystd ~ mgstd + tx * time + (1 + time | participant) + (1 | participant:leg) + 
  (1|participant:sample))

# This model expands on m3, adding (1|participant:sample). Using participant:sample here instead of "just" sample, for the nesting structure to be explicit and self-documented (i.e., "sample nested within participant").

# Priors
get_prior(form0, data = rna.dat)
get_prior(form1, data = rna.dat)
get_prior(form2, data = rna.dat)
get_prior(form3, data = rna.dat)
get_prior(form4, data = rna_rep.dat)


pri0 <- c(
  prior(normal(0, 1), class = "Intercept"),
  prior(normal(1, 1), class = "b", coef = "mgstd"),
  prior(exponential(1), class = "sigma")
) 
# since both ystd and mgstd are standardized (mean 0, SD 1), a normal(0, 1) prior on the slope is a reasonable weakly-informative choice - meaning "we don't strongly expect the standardized slope to exceed roughly ±2"

# Intercept normal(0, 1) - since ystd is standardized (mean 0), the intercept represents predicted ystd when mgstd = 0 (i.e., at average log-mg). We'd expect this to be close to 0 if mg fully explains the centering, so a weakly-informative prior centered at 0 is sensible, consistent with how we've set priors in previous brms

# b for mgstd, normal(0, 1) - same logic as before: standardized slope, so ±2 SD as a generous plausible range

# sigma, exponential(1) - matches our convention in prior brms work


pri1 <- c(
  prior(normal(0, 1), class = "Intercept"),
  prior(normal(0, 1), class = "b", coef = "mgstd"),
  prior(exponential(1), class = "sd", group = "participant"),
  prior(exponential(1), class = "sigma")
)

# sd gets added as a variance component for participant



pri2 <- c(
  prior(normal(0, 1), class = "Intercept"),
  prior(normal(0, 1), class = "b", coef = "mgstd"),
  prior(normal(0, 1), class = "b", coef = "timew3"),
  prior(normal(0, 1), class = "b", coef = "timepost"),
  prior(normal(0, 1), class = "b", coef = "txcon_old_con"),
  prior(normal(0, 1), class = "b", coef = "txint_old_mod"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_low"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod"),
  prior(exponential(1), class = "sd", group = "participant"),
  prior(exponential(1), class = "sigma")
)

# the reference level here is int_old_low (absorbed into the intercept), since its the first level in the tx factor order. So the intercept now represents predicted ystd for an int_old_low participant at time = pre (pre being the reference level of time) and average mg. Each tx coefficient is then a shift relative to int_old_low, and con_olc_con specifically gives the intervention-vs-control directly as a single coefficient.

# currently no pred/contr functions, since none of the models has any interactions, they're all single coefficients with e.g. int_old_low/pre (m2) as the reference. 



pri3 <- c(
  prior(normal(0, 1), class = "Intercept"),
  prior(normal(1, 1), class = "b", coef = "mgstd"),
  prior(normal(0, 1), class = "b", coef = "timew3"),
  prior(normal(0, 1), class = "b", coef = "timepost"),
  prior(normal(0, 1), class = "b", coef = "txcon_old_con"),
  prior(normal(0, 1), class = "b", coef = "txint_old_mod"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_low"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod"),
  prior(normal(0, 1), class = "b", coef = "txcon_old_con:timew3"),
  prior(normal(0, 1), class = "b", coef = "txcon_old_con:timepost"),
  prior(normal(0, 1), class = "b", coef = "txint_old_mod:timew3"),
  prior(normal(0, 1), class = "b", coef = "txint_old_mod:timepost"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_low:timew3"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_low:timepost"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod:timew3"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod:timepost"),
  prior(lkj(2), class = "cor", group = "participant"),
  prior(exponential(1), class = "sd", group = "participant"),
  prior(exponential(1), class = "sd", group = "participant:leg"),
  prior(exponential(1), class = "sigma")
)


pri4 <- c(
  prior(normal(0, 1), class = "Intercept"),
  prior(normal(1, 1), class = "b", coef = "mgstd"),  
  prior(normal(0, 1), class = "b", coef = "timew3"),
  prior(normal(0, 1), class = "b", coef = "timepost"),
  prior(normal(0, 1), class = "b", coef = "txcon_old_con"),
  prior(normal(0, 1), class = "b", coef = "txint_old_mod"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_low"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod"),
  prior(normal(0, 1), class = "b", coef = "txcon_old_con:timew3"),
  prior(normal(0, 1), class = "b", coef = "txcon_old_con:timepost"),
  prior(normal(0, 1), class = "b", coef = "txint_old_mod:timew3"),
  prior(normal(0, 1), class = "b", coef = "txint_old_mod:timepost"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_low:timew3"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_low:timepost"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod:timew3"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod:timepost"),
  prior(lkj(2), class = "cor", group = "participant"),
  prior(exponential(1), class = "sd", group = "participant"),
  prior(exponential(1), class = "sd", group = "participant:leg"),
  prior(exponential(1), class = "sd", group = "participant:sample"),
  prior(exponential(1), class = "sigma")
)



# adding lkj(2) for the correlation matrix between participant-level random intercept/slopes


m0.pp <- brm(form0,
             prior = pri0,
             data = rna.dat,
             file = "fits/XX/m0.pp",
             file_refit = "on_change",
             sample_prior = "only")
#pp_check(m0.pp, ndraws = 100)

# Prior predictive simulation 
as_draws_df(m0.pp) |> 
  filter(.draw %in% 1:100) |> 
  expand_grid(mgstd = seq(from = -3, to = 3, by = 0.2)) |> 
  # predict
  mutate(y = b_Intercept + b_mgstd * mgstd) |> 
  ggplot(aes(mgstd, y, group = .draw)) + 
  geom_line(alpha = 0.2) +
  
  geom_point(data = rna.dat, 
             aes(mgstd, ystd, group = NULL))
  
 
# Update model to the posterior

m0.post <- update(m0.pp,
                  sample_prior = "no",
                  iter = 4000,
                  file = "fits/XX/m0.post",
                  file_refit = "on_change")

# Posterior predictive simulation 
as_draws_df(m0.post) |> 
  filter(.draw %in% 1:100) |> 
  expand_grid(mgstd = seq(from = -4, to = 3, by = 0.5)) |> 
  # predict
  mutate(y = b_Intercept + b_mgstd * mgstd) |> 
  ggplot(aes(mgstd, y, group = .draw)) + 
  geom_line(alpha = 0.2) +
  
  geom_point(data = rna.dat, 
             aes(mgstd, ystd, group = NULL))



  



m1.pp <- brm(form1,
             prior = pri1,
             data = rna.dat,
             file = "fits/XX/m1.pp",
             file_refit = "on_change",
             sample_prior = "only")
#pp_check(m1.pp, ndraws = 100)

as_draws_df(m1.pp) |> 
  filter(.draw %in% 1:100) |> 
  expand_grid(mgstd = seq(from = -3, to = 3, by = 0.2)) |> 
  # predict
  mutate(y = b_Intercept + b_mgstd * mgstd) |> 
  ggplot(aes(mgstd, y, group = .draw)) + 
  geom_line(alpha = 0.2) +
  
  geom_point(data = rna.dat, 
             aes(mgstd, ystd, group = NULL))





m2.pp <- brm(form2,
             prior = pri2,
             data = rna.dat,
             file = "fits/XX/m2.pp",
             file_refit = "on_change",
             sample_prior = "only")
#pp_check(m2.pp, ndraws = 100)


m3.pp <- brm(form3,
             prior = pri3,
             data = rna.dat,
             file = "fits/XX/m3.pp",
             file_refit = "on_change",
             sample_prior = "only")
#pp_check(m3.pp, ndraws = 100)



m4.pp <- brm(form4,
             prior = pri4,
             data = rna_rep.dat,
             file = "fits/XX/m4.pp",
             file_refit = "on_change",
             sample_prior = "only")













```

```{r}
### Posteriors

## Model 3

pred <- predictions(
    m3.pp,
    type = "prediction",
    newdata = datagrid(tx = unique,
                        time = unique),
    ndraws = 1000) 



prior <- get_draws(pred) |> 
  dplyr::select(drawid, draw, participant, tx, time) |> 
  mutate(fit = "prior")




m3.post <- update(m3.pp,
                   sample_prior = "no",
                   cores = 4,
                   control = list(adapt_delta = 0.95,
                                  max_treedepth = 15),
                   iter = 4000,
                   file = "fits/XX/m3.post",
                   file_refit = "on_change")
pp_check(m3.post, ndraws = 100)
pred.post <- predictions(
    m3.post,
    type = "prediction",
    newdata = datagrid(tx = unique,
                        time = unique),
    ndraws = 4000) 


post <- get_draws(pred.post) |> 
  dplyr::select(drawid, draw, participant, tx, time) |> 
  mutate(fit = "posterior")

bind_rows(prior, 
          post) |> 
  ggplot(aes(draw, color = fit)) + 
  geom_line(stat = "density") +
  facet_wrap(tx ~ time)



# Generate predictions (swapping from type = "predictions" to "response")
pred.post <- predictions(
    m3.post,
    type = "response",   
    newdata = datagrid(tx = unique,
                        time = unique),
    ndraws = 4000)

# Setting up contrasts for plotting
all_contrasts <- get_draws(pred.post) |> 
  dplyr::select(drawid, draw, participant, tx, time) |> 
  mutate(grp = paste0(tx, time)) |> 
  dplyr::select(-tx, -time) |> 
  pivot_wider(names_from = grp, values_from = draw) |> 
  mutate(
    # Within-group time deltas
    int_old_low_w3      = int_old_loww3  - int_old_lowpre,
    int_old_low_post     = int_old_lowpost - int_old_lowpre,
    int_old_low_w3post   = int_old_lowpost - int_old_loww3,
    
    int_old_mod_w3       = int_old_modw3  - int_old_modpre,
    int_old_mod_post      = int_old_modpost - int_old_modpre,
    int_old_mod_w3post    = int_old_modpost - int_old_modw3,
    
    int_yng_low_w3        = int_yng_loww3  - int_yng_lowpre,
    int_yng_low_post       = int_yng_lowpost - int_yng_lowpre,
    int_yng_low_w3post     = int_yng_lowpost - int_yng_loww3,
    
    int_yng_mod_w3         = int_yng_modw3  - int_yng_modpre,
    int_yng_mod_post        = int_yng_modpost - int_yng_modpre,
    int_yng_mod_w3post      = int_yng_modpost - int_yng_modw3,
    
    con_old_con_w3           = con_old_conw3  - con_old_conpre,
    con_old_con_post          = con_old_conpost - con_old_conpre,
    con_old_con_w3post        = con_old_conpost - con_old_conw3,
    
    # Volume effect diffs (mod vs low), within age group
    old_mod_vs_low_w3      = int_old_mod_w3 - int_old_low_w3,
    old_mod_vs_low_post     = int_old_mod_post - int_old_low_post,
    old_mod_vs_low_w3post   = int_old_mod_w3post - int_old_low_w3post,
    
    yng_mod_vs_low_w3       = int_yng_mod_w3 - int_yng_low_w3,
    yng_mod_vs_low_post      = int_yng_mod_post - int_yng_low_post,
    yng_mod_vs_low_w3post    = int_yng_mod_w3post - int_yng_low_w3post,
    
    # Average group diffs: young vs old (averaging low/mod within age)
    yng_vs_old_w3         = ((int_yng_low_w3 + int_yng_mod_w3) / 2) - ((int_old_low_w3 + int_old_mod_w3) / 2),
    yng_vs_old_post         = ((int_yng_low_post + int_yng_mod_post) / 2) - ((int_old_low_post + int_old_mod_post) / 2),
    yng_vs_old_w3post        = ((int_yng_low_w3post + int_yng_mod_w3post) / 2) - ((int_old_low_w3post + int_old_mod_w3post) / 2),
    
    # Old intervention vs control
    old_vs_con_w3          = ((int_old_low_w3 + int_old_mod_w3) / 2) - con_old_con_w3,
    old_vs_con_post          = ((int_old_low_post + int_old_mod_post) / 2) - con_old_con_post,
    old_vs_con_w3post         = ((int_old_low_w3post + int_old_mod_w3post) / 2) - con_old_con_w3post,
    
    # Volume effect diffs between age groups
    yng_vs_old_vol_w3      = yng_mod_vs_low_w3 - old_mod_vs_low_w3,
    yng_vs_old_vol_post      = yng_mod_vs_low_post - old_mod_vs_low_post,
    yng_vs_old_vol_w3post     = yng_mod_vs_low_w3post - old_mod_vs_low_w3post
  ) |> 
  dplyr::select(drawid, 
                int_old_low_w3:int_old_low_w3post, int_old_mod_w3:int_old_mod_w3post,
                int_yng_low_w3:int_yng_low_w3post, int_yng_mod_w3:int_yng_mod_w3post,
                con_old_con_w3:con_old_con_w3post,
                old_mod_vs_low_w3:old_mod_vs_low_w3post, yng_mod_vs_low_w3:yng_mod_vs_low_w3post,
                yng_vs_old_w3:yng_vs_old_w3post, old_vs_con_w3:old_vs_con_w3post,
                yng_vs_old_vol_w3:yng_vs_old_vol_w3post) |> 
  pivot_longer(names_to = "contrast", values_to = "value", cols = -drawid) |> 
  mutate(
    delta = case_when(
      grepl("w3post$", contrast) ~ "w3 to post",
      grepl("w3$", contrast) ~ "pre to w3",
      grepl("post$", contrast) ~ "pre to post"
    ),
    contrast_group = case_when(
      grepl("^int_|^con_old_con", contrast) ~ "within-group",   # fixed: con_old_con now included
      grepl("_mod_vs_low", contrast) ~ "volume effect",
      grepl("^yng_vs_old_vol", contrast) ~ "volume x age",
      grepl("^yng_vs_old_w3|^yng_vs_old_post", contrast) ~ "age effect",
      grepl("^old_vs_con", contrast) ~ "intervention vs control"
    )
  )

## Back-transformation for reporting
# Prior to this back-transformation, the values of ystd and mgstd are log-transformed, mean-centered and sd-standardized - this could indeed be reported as a diff in SD's, but its not very intuitive. Therefore, the below code back-transforms ystd to fold-change. 
# For within-group deltas (e.g. int_old_low_w3) this is a straightforward "RNA at w3 is fc_est x RNA at pre", e.g. 1.15 = 15% higher
# For between group/comparison contrasts (e.g., old_vs_con_w3) the value being exponentiated is already a difference of differences (a diff in log-fold-changes), so the resulting fold_change means "the intervention group's fold-change in is fc_est x the control group's fold-change" - a ratio of ratios

all_contrasts_fc <- all_contrasts |> 
  mutate(fold_change = exp(value * sd_rna))

saveRDS(all_contrasts_fc, "./data/data-gen/rna_mg_contrasts_fc.rds")

contrast_summary <- all_contrasts_fc |> 
  group_by(contrast, contrast_group, delta) |> 
  summarise(
    ystd_est   = median(value),
    ystd_lower = quantile(value, 0.025),
    ystd_upper = quantile(value, 0.975),
    fc_est     = median(fold_change),
    fc_lower   = quantile(fold_change, 0.025),
    fc_upper   = quantile(fold_change, 0.975),
    pos        = sum(fold_change > 1) / n(),   
    .groups = "drop"
  ) |> 
  arrange(contrast_group, contrast, delta)

saveRDS(contrast_summary, "data/data-gen/rna_mg_contrast_summary.rds")

print(contrast_summary, n = 30)




## Model 4

pred_rep <- predictions(
  m4.pp,
  type = "prediction",
  newdata = datagrid(tx = unique,
                     time = unique),
  ndraws = 1000)

prior_rep <- get_draws(pred4) |> 
  select(drawid, draw, participant, tx, time) |> 
  mutate(fit = "prior")

# Fit the posterior

m4.post <- update(m4.pp,
                  sample_prior = "no",
                  cores = 4,
                  control = list(adapt_delta = 0.95,
                                 max_treedepth = 15),
                  iter = 4000,
                  file = "fits/XX/m4.post",
                  file_refit = "on_change")
pp_check(m4.post, ndraws = 100)


pred.post_rep <- predictions(
    m4.post,
    type = "response",   
    newdata = datagrid(tx = unique,
                        time = unique),
    re_formula = NA,
    ndraws = 4000)


post_rep <- get_draws(pred.post_rep) |> 
  select(drawid, draw, participant, tx, time) |> 
  mutate(fit = "posterior")


bind_rows(prior_rep, post_rep) |> 
  ggplot(aes(draw, color = fit)) +
  geom_line(stat = "density") +
  facet_wrap(tx ~ time)


# Setting up contrasts for plotting

all_contrasts_rep <- get_draws(pred.post_rep) |> 
  dplyr::select(drawid, draw, participant, tx, time) |> 
  mutate(grp = paste0(tx, time)) |> 
  dplyr::select(-tx, -time) |> 
  pivot_wider(names_from = grp, values_from = draw) |> 
  mutate(
    # Within-group time deltas
    int_old_low_w3      = int_old_loww3  - int_old_lowpre,
    int_old_low_post     = int_old_lowpost - int_old_lowpre,
    int_old_low_w3post   = int_old_lowpost - int_old_loww3,
    
    int_old_mod_w3       = int_old_modw3  - int_old_modpre,
    int_old_mod_post      = int_old_modpost - int_old_modpre,
    int_old_mod_w3post    = int_old_modpost - int_old_modw3,
    
    int_yng_low_w3        = int_yng_loww3  - int_yng_lowpre,
    int_yng_low_post       = int_yng_lowpost - int_yng_lowpre,
    int_yng_low_w3post     = int_yng_lowpost - int_yng_loww3,
    
    int_yng_mod_w3         = int_yng_modw3  - int_yng_modpre,
    int_yng_mod_post        = int_yng_modpost - int_yng_modpre,
    int_yng_mod_w3post      = int_yng_modpost - int_yng_modw3,
    
    con_old_con_w3           = con_old_conw3  - con_old_conpre,
    con_old_con_post          = con_old_conpost - con_old_conpre,
    con_old_con_w3post        = con_old_conpost - con_old_conw3,
    
    # Volume effect diffs (mod vs low), within age group
    old_mod_vs_low_w3      = int_old_mod_w3 - int_old_low_w3,
    old_mod_vs_low_post     = int_old_mod_post - int_old_low_post,
    old_mod_vs_low_w3post   = int_old_mod_w3post - int_old_low_w3post,
    
    yng_mod_vs_low_w3       = int_yng_mod_w3 - int_yng_low_w3,
    yng_mod_vs_low_post      = int_yng_mod_post - int_yng_low_post,
    yng_mod_vs_low_w3post    = int_yng_mod_w3post - int_yng_low_w3post,
    
    # Average group diffs: young vs old (averaging low/mod within age)
    yng_vs_old_w3         = ((int_yng_low_w3 + int_yng_mod_w3) / 2) - ((int_old_low_w3 + int_old_mod_w3) / 2),
    yng_vs_old_post         = ((int_yng_low_post + int_yng_mod_post) / 2) - ((int_old_low_post + int_old_mod_post) / 2),
    yng_vs_old_w3post        = ((int_yng_low_w3post + int_yng_mod_w3post) / 2) - ((int_old_low_w3post + int_old_mod_w3post) / 2),
    
    # Old intervention vs control
    old_vs_con_w3          = ((int_old_low_w3 + int_old_mod_w3) / 2) - con_old_con_w3,
    old_vs_con_post          = ((int_old_low_post + int_old_mod_post) / 2) - con_old_con_post,
    old_vs_con_w3post         = ((int_old_low_w3post + int_old_mod_w3post) / 2) - con_old_con_w3post,
    
    # Volume effect diffs between age groups
    yng_vs_old_vol_w3      = yng_mod_vs_low_w3 - old_mod_vs_low_w3,
    yng_vs_old_vol_post      = yng_mod_vs_low_post - old_mod_vs_low_post,
    yng_vs_old_vol_w3post     = yng_mod_vs_low_w3post - old_mod_vs_low_w3post
  ) |> 
  dplyr::select(drawid, 
                int_old_low_w3:int_old_low_w3post, int_old_mod_w3:int_old_mod_w3post,
                int_yng_low_w3:int_yng_low_w3post, int_yng_mod_w3:int_yng_mod_w3post,
                con_old_con_w3:con_old_con_w3post,
                old_mod_vs_low_w3:old_mod_vs_low_w3post, yng_mod_vs_low_w3:yng_mod_vs_low_w3post,
                yng_vs_old_w3:yng_vs_old_w3post, old_vs_con_w3:old_vs_con_w3post,
                yng_vs_old_vol_w3:yng_vs_old_vol_w3post) |> 
  pivot_longer(names_to = "contrast", values_to = "value", cols = -drawid) |> 
  mutate(
    delta = case_when(
      grepl("w3post$", contrast) ~ "w3 to post",
      grepl("w3$", contrast) ~ "pre to w3",
      grepl("post$", contrast) ~ "pre to post"
    ),
    contrast_group = case_when(
      grepl("^int_|^con_old_con", contrast) ~ "within-group",   # fixed: con_old_con now included
      grepl("_mod_vs_low", contrast) ~ "volume effect",
      grepl("^yng_vs_old_vol", contrast) ~ "volume x age",
      grepl("^yng_vs_old_w3|^yng_vs_old_post", contrast) ~ "age effect",
      grepl("^old_vs_con", contrast) ~ "intervention vs control"
    )
  )


# Back-transformation

all_contrasts_fc_rep <- all_contrasts_rep |> 
  mutate(fold_change = exp(value * sd_rna_rep))


saveRDS(all_contrasts_fc_rep, "./data/data-gen/rna_mg_contrasts_fc_m4.rds")


contrast_summary_rep <- all_contrasts_fc_rep |> 
  group_by(contrast, contrast_group, delta) |> 
  summarise(
    ystd_est   = median(value),
    ystd_lower = quantile(value, 0.025),
    ystd_upper = quantile(value, 0.975),
    fc_est     = median(fold_change),
    fc_lower   = quantile(fold_change, 0.025),
    fc_upper   = quantile(fold_change, 0.975),
    pos        = sum(fold_change > 1) / n(),   
    .groups = "drop"
  ) |>
  arrange(contrast_group, contrast, delta)


saveRDS(contrast_summary_rep, "./data/data-gen/rna_mg_contrast_summary_m4.rds")

print(contrast_summary_rep, n = 30)


# The replicate-level random effect tightened the estimates without changing the substantive conclusions, which is what it was supposed to do. 

summary(m4.post)

```




## Results

Total RNA yield increased sub-proportionally with tissue mass on the log-log scale (β = 0.83–0.86 across all model iterations, 95% CrI excluding 1 throughout), rather than the strictly proportional relationship assumed by a fixed RNA/mg ratio. This slope was stable regardless of model complexity or whether technical replicate structure was accounted for, indicating it reflects a genuine property of the tissue rather than an artifact of model specification.

<!--
Consider for discussion/methods: Accounting for technical replicates. To separate genuine biological/sample-level variability from technical noise arising from repeated concentration readings of the same tissue extract, a random effect for replicate (nested within participant) was added to the model, fit on the full replicate-level dataset (1552 observations) rather than pre-averaged values. This partitioning showed that approximately 78% of the variance previously absorbed into the residual term reflected genuine sample-to-sample biological variability, with the remaining ~22% attributable to technical replicate noise. Accounting for this structure sharpened estimates throughout without altering the underlying pattern of results — several contrasts that were moderate-to-strong in the summarised-replicate model became strong once technical noise was properly partitioned out.
-->

Within-group changes over time. After accounting for tissue mass and technical replicate variance, RNA yield increased from pre to week 3 and from pre to post in nearly all groups, with strong evidence (pos = 0.86–1.00) across every intervention arm. The control group showed a strong increase from pre to post (fold-change = 1.13, pos = 0.936) and from w3 to post (fold-change = 1.15, pos = 0.924), but no change from pre to w3 (fold-change = 0.99, pos = 0.450) — a genuine null. Overall, the time-dependent RNA increase was broadly consistent across groups, in line with previously reported RNA kinetics in this cohort.

Age effect. The direction of the age-related difference in RNA trajectory reversed across the study period. From pre to week 3, younger participants showed a strong lean toward a larger increase than older participants (fold-change = 1.07, pos = 0.874), but from week 3 to post this pattern reversed, with older participants showing a weak-to-moderate lean toward a larger late-phase increase (fold-change = 0.91 for young vs. old, pos = 0.089, i.e. pos = 0.911 in the old-favoring direction). The net pre-to-post age difference was a genuine null (fold-change = 0.98, pos = 0.334). This suggests a difference in the timing of the RNA response between age groups rather than a difference in its total magnitude.

Intervention vs. control. A similar timing-dependent pattern emerged when comparing the old intervention groups against the old passive control. From pre to week 3, the intervention groups showed a strong lean toward a larger increase than control (fold-change = 1.10, pos = 0.874), while from week 3 to post this reversed, with control showing a moderate-to-strong lean toward catching up (fold-change = 0.91 for old vs. control, pos = 0.196, i.e. pos = 0.804 in the control-favoring direction). The net pre-to-post comparison was a genuine null (fold-change = 1.01, pos = 0.532), indicating that intervention and control groups reached a similar overall RNA increase by the end of the study, achieved via different time-courses.

Volume and volume × age. No credible evidence was found for an effect of training volume (moderate vs. low) on RNA trajectory in either age group, at any time window (pos = 0.40–0.73), nor for the volume effect differing by age (volume × age, pos = 0.28–0.68).

<!--
Summary. Once tissue mass was properly accounted for via direct modeling, and technical replicate variance was separated from genuine sample-level biological variability, RNA yield showed a robust, near-universal increase across the training period in essentially all groups. Age and intervention status did not alter the total magnitude of this increase, but did modulate its timing — younger and intervention participants tended to rise earlier, with older and control participants showing comparatively more of their increase later in the protocol. Training volume showed no credible influence at any stage.
-->





















----------------------------- OLD OUTDATED CODE --------------------------------

```{r}

# Original contrast plot
#get_draws(pred.post) |> 
#  dplyr::select(drawid, draw, participant, tx, time) |> 
#  mutate(grp = paste0(tx, time)) |> 
#dplyr::select(-tx, -time) |> 
#  pivot_wider(names_from = grp, values_from = draw) |> 


#  mutate(
#    low = int_yng_loww3  - int_yng_lowpre ,
#    mod = int_yng_modw3 - int_yng_modpre, 
#    low_vs_mod = mod - low
#  ) |> 
#  dplyr::select(low, mod,low_vs_mod, drawid) |>
#  pivot_longer(names_to = "contrast", 
#               values_to = "value", 
#               cols = low:low_vs_mod) |> 
 
#  ggplot(aes(value, group = contrast, color = contrast)) + 
#  geom_line(stat = "density") + 
#  scale_x_continuous(breaks = c(0, 1, 2, 3, 4))




m3_pred_contr <- function(m, nd1, nd2) {
  
  contr <- predictions(
    m,
    newdata = datagrid(tx = unique,
                        time = unique),
    ndraws = nd1,
    hypothesis = c(
      "int_old_low_delta_w3"     = "b2 - b1 = 0",
      "int_old_low_delta_post"   = "b3 - b1 = 0",
      "int_old_low_delta_w3post" = "b3 - b2 = 0",
      "int_old_mod_delta_w3"     = "b5 - b4 = 0",
      "int_old_mod_delta_post"   = "b6 - b4 = 0",
      "int_old_mod_delta_w3post" = "b6 - b5 = 0",
      "int_yng_low_delta_w3"     = "b8 - b7 = 0",
      "int_yng_low_delta_post"   = "b9 - b7 = 0",
      "int_yng_low_delta_w3post" = "b9 - b8 = 0",
      "int_yng_mod_delta_w3"     = "b11 - b10 = 0",
      "int_yng_mod_delta_post"   = "b12 - b10 = 0",
      "int_yng_mod_delta_w3post" = "b12 - b11 = 0",
      "con_old_con_delta_w3"     = "b14 - b13 = 0",
      "con_old_con_delta_post"   = "b15 - b13 = 0",
      "con_old_con_delta_w3post" = "b15 - b14 = 0",

      # Average group diffs pre-mid
      "yng_vs_old_delta_w3" = "(((b11 - b10) + (b8 - b7)) / 2) - (((b2 - b1) + (b5 - b4)) / 2) = 0",
      "old_vs_con_delta_w3" = "((b2 - b1) + (b5 - b4)) / 2 - (b14 - b13) = 0",

      # Average group diffs mid-post
      "yng_vs_old_delta_w3post" = "(((b12 - b11) + (b9 - b8)) / 2) - (((b3 - b2) + (b6 - b5)) / 2) = 0",
      "old_vs_con_delta_w3post" = "((b3 - b2) + (b6 - b5)) / 2 - (b15 - b14) = 0",

      # Average group diffs pre-post
      "yng_vs_old_delta_post" = "(((b12 - b10) + (b9 - b7)) / 2) - (((b3 - b1) + (b6 - b4)) / 2) = 0",
      "old_vs_con_delta_post" = "((b3 - b1) + (b6 - b4)) / 2 - (b15 - b13) = 0",

      # Volume effect diffs pre-mid
      "yng_mod_vs_low_w3" = "(b11 - b10) - (b8 - b7) = 0",
      "old_mod_vs_low_w3" = "(b5 - b4) - (b2 - b1) = 0",

      # Volume effect diffs mid-post
      "yng_mod_vs_low_w3post" = "(b12 - b11) - (b9 - b8) = 0",
      "old_mod_vs_low_w3post" = "(b6 - b5) - (b3 - b2) = 0",

      # Volume effect diffs pre-post
      "yng_mod_vs_low_post" = "(b12 - b10) - (b9 - b7) = 0",
      "old_mod_vs_low_post" = "(b6 - b4) - (b3 - b1) = 0",

      # Volume effect diffs between age groups
      "yng_vs_old_vol_w3"     = "((b11 - b10) - (b8 - b7)) - ((b5 - b4) - (b2 - b1)) = 0",
      "yng_vs_old_vol_w3post" = "((b12 - b11) - (b9 - b8)) - ((b6 - b5) - (b3 - b2)) = 0",
      "yng_vs_old_vol_post"   = "((b12 - b10) - (b9 - b7)) - ((b6 - b4) - (b3 - b1)) = 0"
    ))

  pred <- predictions(
    m,
    ndraws = nd2,
    newdata = datagrid(tx = unique,
                        time = unique),
    re_formula = NA)

  return(list(contr = contr, pred = pred))
}


m3_preds_pp  <- m3_pred_contr(m3.pp,   nd1 = 1000, nd2 = 50)













### Get the posterior distribution

m0.post <- update(m0.pp,
                  sample_prior = "no",
                  iter = 4000,
                  file = "fits/XX/m0.post",
                  file_refit = "on_change")

#summary(m0.post)
#pp_check(m0.post)

# mgstd slope = 0.95, 95% CI [0.79, 0.90] - strongly positive (more tissue, more RNA), but the entire credible interval sits below 1 - i.e., the RNA-mg relationship is sub-proportional, a 1 SD increase in log(mg) predicts less than 1 SD increase in log(RNA), not a 1:1 match. This falls in line with the flatter line observed in the preliminary plot.

# intercept = 0 - as expected since both variables are centered

# sigma = 0.53 - the risidual SD on the standardized scale. Since ystd has SD = 1 by construction, sigma = 0.53 means mg alone leaves a fair amount of variance unexplained, where we will try to add explanatory power using time, age_group, allocation etc

m1.post <- update(m1.pp,
                  sample_prior = "no",
                  iter = 4000,
                  file = "fits/XX/m1.post",
                  file_refit = "on_change")

#summary(m1.post)
#pp_check(m1.post)

# mgstd slope essentially identical to m0 (0.84 [0.79, 0.90]), so mg-RNA relationship is not an artifact of pooling across participants who differ systematically in both variables.

# sd(Intercept) for participant (0.26, [0.19, 0.34]), quantifies genuine between-participant variability in baseline (mg-adjusted) RNA expression; some people run consistently higher or lower RNA than others, independent of tissue mass (variance previously folded into sigma in m0). 

# sigma dropped as suspected (0.53 -> 0.47) - variance picked up by the new participant-level term


m2.post <- update(m2.pp,
                  sample_prior = "no",
                  iter = 4000,
                  file = "fits/XX/m2.post",
                  file_refit = "on_change")

#summary(m2.post)
#pp_check(m2.post)
pp_check(m2.post, type = "dens_overlay_grouped", 
         group = "tx")


# mgstd slop holds well, essentially unchanged again (0.86 [0.81, 0.91]) - a sign that no matter what else gets added to the model, the relationship keeps holding steady

# txcon_old_con = -0.48 [-0.70, -0.27] - entire CI is comfortably below 0, meaning the old control group has credibly lower RNA yield (at a given mg and timepoint) than int_old_low. This is quite different to the first total RNA analysis, where controls showed a near-identical increase to the intervention - so, here its showing up as a level difference rather than a time-trend difference

# txint_old_mod, txint_yng_low, txint_yng_mod - all have CIs comfortably straddling 0, i.e. no credible baseline diff from int_old_low among the intervention arms themselves. Age and volume dont seem to be driving big level shifts; its specifically the control group that stands out

# timew3 = 0.24 [0.13, 0.34]and timepost = 0.27 [0.15, 0.38] - both credibly positive, so pre -> w3/pre -> post shift seen in the preliminary plot. They are almost equal in magnitude, suggesting most of the increase happens early - no surprise there. 

m3.post <- update(m3.pp,
                   sample_prior = "no",
                   cores = 4,
                   control = list(adapt_delta = 0.95,
                                  max_treedepth = 15),
                   iter = 4000,
                   file = "fits/XX/m3.post",
                   file_refit = "on_change")

summary(m3.post)
pp_check(m3.post)

# mgstd holds up (0.85 [0.80, 0.90]) across all models

# 

```

```{r}

# Group comparisons

m3_preds_post <- m3_pred_contr(m3.post, nd1 = 1000, nd2 = 50)



```

#### Results

Model 3 summary: ystd ~ mgstd + tx * time + (1 + time | participant) + (1 | participant:leg)

Model diagnostics: Clean sampling (Rhat ≤ 1.01, all ESS > 1000, adapt_delta = 0.95, no divergences); posterior predictive checks track observed distribution well.

Core mg-RNA relationship (B1):

mgstd = 0.85 [0.80, 0.90] — stable across all four models tested (m0–m3: 0.85, 0.84, 0.86, 0.85).
Slope is credibly below 1, indicating RNA yield increases sub-proportionally with tissue mass — direct evidence against the fixed-ratio standardization's implicit assumption of a 1:1 relationship.

Group-level effects (B2, tx):

txcon_old_con = -0.40 [-0.70, -0.10] — old control group sits at a credibly lower RNA level than int_old_low (reference), holding mg and time constant.
No credible baseline differences among intervention arms (int_old_mod, int_yng_low, int_yng_mod vs. reference) — age and volume don't show baseline-level effects on their own.
In absolute terms (pred): con_old_con sits credibly below zero at pre and w3 (CIs excluding 0), while all intervention groups straddle zero at every timepoint.

Time effects (B3):

timew3 = 0.22 [0.00, 0.43], timepost = 0.26 [0.04, 0.49] — both credibly positive, similar magnitude to each other, suggesting most of the increase happens by w3 and plateaus into post. Consistent with existing literature.

Interactions (tx:time) — the key new finding:

Every tx:time interaction term's CI comfortably includes 0 — no credible evidence that the control group's (or any group's) time-trajectory differs from the reference (int_old_low).
Derived contrasts confirm this: old_vs_con_delta_w3 = 0.19 [-0.18, 0.55], old_vs_con_delta_post = 0.02 [-0.38, 0.44] — both null.
Volume contrasts (mod vs. low, within and across age groups) are also all null across every time window.
One borderline exception: yng_vs_old_delta_w3post = -0.21 [-0.50, 0.08] — leans negative, CI just touches 0 (weak lean by your three-tier convention, not a stronger signal).

Bottom line: Controls show a lower overall RNA level than interventions, but their RNA rises over time in step with intervention groups — the increase looks more like a general procedural/biopsy response than a training-specific one. No credible volume (low vs. moderate) or age-dependent differences in either RNA level or trajectory shape.

