---
title: "Notes on brms-version of full isometric-isokinteic model"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

Load packages and data

Discussion points:
1. m1/preds1 contrast block intent — Reference to m1/preds1 for the old_mod_vs_low-style contrast, but form1 (HUMAC) has no tx_sex term, so that object can't produce a group contrast. Is the contrast block meant to run against m2/preds2, and should re_formula be NA or tx_sex-inclusive.

2. Sex-weighting for tx_sex → tx collapse — resolved for low/mod (inherently 50/50, bilateral design), but still open for the age/intervention/control contrasts more broadly: is sample-weighting (vs. equal-weighting) the right call here?

3. Multiple comparisons framing — 6 contrasts reported from one model. Should they be treated as independent exploratory comparisons (no correction) or should a caveat/adjustment be applied given the growing count?

4. Mid timepoint inclusion in the stacked hill-curve figures. Do we drop mid for more visual clarity?


```{r}

library(brms)
library(tidyverse)
library(reliefdata)
library(marginaleffects)

humac.dat <- reliefdata::relief_humac %>% 
  mutate(participant = as.character(participant))

condition <- reliefdata::relief_volume %>% 
  mutate(participant = as.character(participant))

participants <- reliefdata::relief_participants %>% 
  mutate(age_group = if_else(age < 40, "yng", "old"))


humac_dat <- humac.dat %>% 
  full_join(condition) %>% 
  select(-arm) %>% 
  full_join(participants) %>%  
  
  filter(!participant %in% c("3004", "3005", "3010", "3033",
                          "3034", "3046", "3053", "3069",
                          "3092", "3010", "3019", "3026",
                          "3017", "3012")) %>% # dropouts/missing all measurements
  #filter(!is.na(time), !is.na(speed)) %>%
  group_by(participant, time, timerep, leg, allocation, speed, condition, age_group, sex, weight) %>%
  summarise(pt = mean(pt, na.rm = TRUE), .groups = "drop") %>%
  mutate(
         ystd = pt / max(pt, na.rm = TRUE),  
         
         condition = if_else(allocation == "con", "con", condition),
         tx = paste0(allocation, "_", age_group, "_", condition),
         tx = factor(tx, levels = c("int_old_low","int_old_mod", "int_yng_low","int_yng_mod", "con_old_con")),
         tx_sex = paste0(tx, "_", sex), 
         tx_sex = factor(tx_sex, levels = c(
           "int_old_low_female", "int_old_low_male",
           "int_old_mod_female", "int_old_mod_male", 
           "int_yng_low_female", "int_yng_low_male",
           "int_yng_mod_female", "int_yng_mod_male", 
           "con_old_con_female", "con_old_con_male"
         )),
         time = factor(time, levels = c("pre", "mid", "post")), 
         sex_centered = if_else(sex == "female", -0.5, 0.5),
         speed = factor(speed, levels = c("0", "60", "120", "240")), 
         sp = as.numeric(as.character(speed)),
         sp = sp/240,
         sp2 = sp^2) |> 
  filter(!is.na(ystd)) |> 
  print()

# About using ystd = pt/max to rescale the data
# A global rescale by max still does something specific: it compresses everything into (0, ~1] range using one arbitrary anchor point (whichever observation happened to be the strongest). That's fine if the only goal is numerical convenience for priors/sampling (keeping ystd on a friendly 0–1 scale for prior specification, which is exactly what the prior predictive code suggests — priors like normal(0.5, 0.25) for the intercept assume a roughly 0–1 outcome). [Likely] If that's the actual purpose, it's a defensible, low-stakes choice and I'll drop it. But it does mean the model's fixed effects and group differences are all still in "fraction of the single strongest lift in the sample" units, not in any per-participant-relative sense — so when you write up results, you can't describe ystd as "percentage of individual capacity" or similar, only as a rescaled version of raw torque. Worth confirming that's how you and your supervisor plan to describe it, since that phrasing mistake is an easy one to make months from now when you're deep in the discussion section.




```

The priors can be tested for a population level trend easily like this:

```{r}

n <- 30 # 30 draws

data.frame(
  int = rnorm(n, 0.5, 0.25), 
  sp = rnorm(n, -0.5, 0.1), 
  sp2 = rnorm(n, 0.25, 0.2), 
  draw = 1:n
) |> 
  expand_grid(speed = seq(from = 0, to = 1, by = 0.1)) |> 
  mutate(speed2 = speed^2,
         # pt = peak torque scaled to 0-1 (xi / max(x))
         pt = int + sp * speed + sp2 * speed2) |> 
  ggplot(aes(speed, pt, group = draw)) + geom_line()



```

Next we set up priors for two different variations of the model. Model 1 is a fixed effects model for the groups with sex as a fixed offset over all time-points. Sex is scale -0.5 and 0.5 here so the fixed effects (predictions) will be the average between men and women.

Model 2 puts all condition groups in "varying effects". This means they will be regularized. Except for the allocation (control vs. intervention).

```{r}



form1 <- bf(ystd ~ 0 + time + time:sp + time:sp2 +
              sex_centered +
              (0 + time + time:sp + time:sp2 | participant) +
              (1 | participant:time:timerep))

form2 <- bf(ystd ~ 0 + time:(allocation + allocation:sp + allocation:sp2) +
              (0 + time + time:sp + time:sp2 | tx_sex) +
              (0 + time + time:sp + time:sp2 | participant) +
              (1 | participant:time:timerep))

# Adding timerep as a random intercept to fix pseudoreplication. It seems a better alternative to account for timerep here in the model, rather than averaging or taking the highest measurement. Not doing anything would mean risking posterior intervals looking tighter and more confident than they should be.


#form1 <- bf(ystd ~ 0 + time + time:sp + time:sp2 +
#              sex_centered +
#              (0 + time + time:sp + time:sp2 | participant))

#form2 <- bf(ystd ~ 0 + time:(allocation + allocation:sp + allocation:sp2) +
#              (0 + time + time:sp + time:sp2 | tx_sex) +
#              (0 + time + time:sp + time:sp2 | participant))


get_prior(form1, data = humac_dat)
get_prior(form2, data = humac_dat)


# Each tx group now has its own population-level intercept + speed curve
# at each timepoint, so there are many "b" coefficients. Rather than list each
# one, set the priors programmatically by matching the coefficient name:
#   * "sex_centered"  -> level-only sex effect, regularized toward 0 (no
#                        prior expectation of a male/female difference)
#   * ends in ":sp2"  -> quadratic speed term
#   * ends in ":sp"   -> linear speed term
#   * otherwise        -> intercept (level at sp = 0)
priors1 <- get_prior(form1, data = humac_dat) |>
  mutate(prior = dplyr::case_when(
    class == "b" & coef == "sex_centered"             ~ "normal(0, 0.1)",
    class == "b" & stringr::str_detect(coef, ":sp2$") ~ "normal(0.25, 0.2)",
    class == "b" & stringr::str_detect(coef, ":sp$")  ~ "normal(-0.5, 0.1)",
    class == "b" & coef != ""                          ~ "normal(0.5, 0.25)",
    class == "sd"                                      ~ "exponential(1)",
    class == "cor"                                     ~ "lkj(2)",
    class == "sigma"                                   ~ "exponential(1)",
    TRUE ~ prior)) 

priors2 <- get_prior(form2, data = humac_dat) |>
  mutate(prior = dplyr::case_when(
    class == "b" & stringr::str_detect(coef, ":sp2$") ~ "normal(0.25, 0.2)",
    class == "b" & stringr::str_detect(coef, ":sp$")  ~ "normal(-0.5, 0.1)",
    class == "b" & coef != ""                          ~ "normal(0.5, 0.25)",
    class == "sd"                                      ~ "exponential(1)",
    class == "cor"                                     ~ "lkj(2)",
    class == "sigma"                                   ~ "exponential(1)",
    TRUE ~ prior))  


m1.pp <- brm(form1,
             prior = priors1,
             data = humac_dat,
             file = "fits/04-strength/test-01",
             sample_prior = "only")

m2.pp <- brm(form2,
             prior = priors2,
             data = humac_dat,
             file = "fits/04-strength/test-02",
             sample_prior = "only")

# Is the prior for the newly added timerep doing as we intend? It should show up with "exponential(1)" in these checks
# prior_summary(m1.pp)
# prior_summary(m2.pp)
# Both checks out






```

```{r}


grid1 <- humac_dat |>
    distinct(tx, time) |>
    expand_grid(sp = seq(min(humac_dat$sp), max(humac_dat$sp), length.out = 50)) |>
    mutate(sp2 = sp^2,
           sex_centered = 0)   # predict the sex-averaged curve per tx

grid2 <- humac_dat |>
    distinct(tx_sex, time, allocation) |>
    expand_grid(sp = seq(min(humac_dat$sp), max(humac_dat$sp), length.out = 50)) |>
    mutate(sp2 = sp^2)   # predict the sex-averaged curve per tx




preds1 <- predictions(m1.pp,
            newdata = grid1,
            re_formula = NA,
            by = c("tx", "time", "sp"))

preds2 <- predictions(m2.pp,
            newdata = grid2,
            re_formula = NA,
            by = c("tx_sex", "time", "sp"))



# Display averages per group/condition

get_draws(preds1)  |>
  summarise(.by = c(tx, time, sp),
            m = mean(draw),
            lwr = quantile(draw, 0.05),
            upr = quantile(draw, 0.95)) |>
  ggplot(aes(sp, m, group = tx, color = tx)) + geom_line() +
  facet_wrap(~ time)
  



```

```{r}
### Get posterior distribution

m1.post <- update(m1.pp,
             sample_prior = "no",
             cores = 4,
             iter = 6000,
             warmup = 2000,
             control = list(adapt_delta = 0.95),
             file = "fits/04-strength/m1-full",
             file_refit = "on_change")

# Checking the model
# A model can run without errors, but that is not proof of convergence per se. rhat and neff_ratio are checks to make sure that the models actually converge and provide you with trustworthy answers. 

# rhat(m1.post)
# rhat essentially checks "did multiple independent attempts agree". A value of 1.0 means that the chains in the model are indistinguishable, they agree. Above 1.0 means the chains disagree, but the conventional gutoff is 1.01 - "close enough to indistinguishable". Checks out to all values being below 1.01 (even 1.005) - no red flags here.  

# neff_ratio(m1.post)
# neff_ratio essentially checks "how much independent information do you actually have". Neff = effective sample size, estimates how many independent draws would give you the same precision as your actual autocorrelated draws. Low ESS on a parameter can mean unreliable posterior summaries even with a clean looking rhat. Increasing the iterations from 4000 to 6000, and the warmup from 1500 to 2000 yielded a 5x higher effective sample size, a total of 1444 compared to 4000 iterations yielding 285, on the same parameter, comfortably clearing the conventional "at least 400" floor, with real headroom for stable 95% credible interval estimation.


m2.post <- update(m2.pp,
             sample_prior = "no",
             cores = 4,
             iter = 6000,
             warmup = 2000,
             control = list(adapt_delta = 0.95),
             file = "fits/04-strength/m2-full",
             file_refit = "on_change")

#rhat(m2.post)
# All at or below 1.003

#neff_ratio(m2.post)
# ESS: 1579


pp_check(m2.post, type = "dens_overlay_grouped",
         group = "tx_sex")

pp_check(m2.post)



```

```{r}



# Model 1

preds1 <- predictions(m1.post,
            newdata = grid1,
            re_formula = ~(0 + time + time:sp + time:sp2 | tx_sex),
            by = c("tx", "time", "sp"))


get_draws(preds1)  |>
  summarise(.by = c(tx, time, sp),
            m = mean(draw),
            lwr = quantile(draw, 0.05),
            upr = quantile(draw, 0.95)) |>

  mutate(tx2 = tx) |>
  separate(tx2, into = c("allocation", "age", "cond")) |>
  mutate(allocation_age = paste0(allocation, "_", age)) |>


  ggplot(aes(sp, m, group = tx, color = tx)) + geom_line() +
  facet_wrap(allocation_age ~ time)


# Calculate delta for each group

get_draws(preds1)  |>
  dplyr::select(drawid, draw, tx, time, sp) |>
  
  pivot_wider(names_from = time, values_from = draw) |>

  mutate(delta = post - pre) |>
 
  dplyr::select(drawid:sp, delta) |>
  pivot_wider(names_from = tx,
              values_from = delta) |>  
  
    # Calculate contrasts
  mutate(
    old_mod_vs_low = int_old_mod - int_old_low
    ) |>

  dplyr::select(drawid:sp, old_mod_vs_low) |>

  # Summarise the posterior of the contrast before plotting
  summarise(.by = sp,
            m   = median(old_mod_vs_low),
            lwr = quantile(old_mod_vs_low, 0.05),
            upr = quantile(old_mod_vs_low, 0.95)) |>

  ggplot(aes(sp, m)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2) +
  geom_line()
  

# Model 2

preds2 <- predictions(m2.post,
                      newdata = grid2,
                      re_formula = ~(0 + time + time:sp + time:sp2 | tx_sex),
                      by = c("tx_sex", "time", "sp"))

# Get actual sex count per tx_sex group to weight the collapse. This is done to use sample-weight, i.e., average using the actual observed male/female counts in each tx group - "what does the average participant in this group look like". The alternative would be to equal-weight the sexes, averaging the male and female posterior predictions 50/50, rather answering "what would the average of a hypothetical man and woman look like". Personally, Im not sure what is best, but my gut-instinct says actual sample-weight. 

sex_weights <- humac_dat %>% 
  distinct(participant, tx_sex) %>% 
  count(tx_sex, name = "n_participants") %>% 
  separate(tx_sex, into = c("allocation", "age", "condition", "sex"), remove = FALSE) %>% 
  mutate(tx = paste(allocation, age, condition, sep = "_")) %>% 
  select(tx_sex, tx, n_participants)


library(collapse)
draws2 <- get_draws(preds2) |>
  left_join(sex_weights, by = "tx_sex") |>
  fgroup_by(drawid, tx, time, sp) |>
  fsummarise(draw = fmean(draw, w = n_participants)) # also using fgroup and fsummarise from collapse as faster tools compared to dplyr

# Curve plot at tx level

draws2 %>% 
  summarise(.by = c(tx, time, sp),
            m = mean(draw),
            lwr = quantile(draw, 0.05),
            upr = quantile(draw, 0.95)) %>% 
  separate(tx, into = c("allocation", "age", "cond"), remove = FALSE) %>% 
  mutate(allocation_age = paste0(allocation, "_", age)) %>% 
  ggplot(aes(sp, m, group = tx, color = tx)) + geom_line() +
  facet_wrap(allocation_age ~ time)

## Delta contrast, same structure as m1 block
# Contrasts of interest:
# - int_old_mod - int_old_low (volume, older)
# - int_yng_mod - int_yng_low (volume, younger)
# - int_old_low - con_old_con (intervention vs control, low, older)
# - int_old_mod - con_old_con (intervention vs control, mod, older)
# - int_old_low - int_yng_low (age, low volume)
# - int_old_mod - int_yng_mod (age, mod volume)

max_pt <- max(humac_dat$pt, na.rm = TRUE)

draws2 |>
  select(drawid, draw, tx, time, sp) |>
  pivot_wider(names_from = time, values_from = draw) |>
  mutate(delta = post - pre) |>
  select(drawid:sp, delta) |>
  pivot_wider(names_from = tx, values_from = delta) |>
  mutate(old_mod_vs_low = int_old_mod - int_old_low) |>
  select(drawid:sp, old_mod_vs_low) |>
  summarise(.by = sp,
            m   = median(old_mod_vs_low),
            lwr = quantile(old_mod_vs_low, 0.05),
            upr = quantile(old_mod_vs_low, 0.95)) |>
 # mutate(across(c(m, lwr, upr), ~ .x * max_pt)) |>   # Rescaling ystd back to peak torque here
  ggplot(aes(sp, m)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2) +
  geom_line() +
  labs(y = "Δ ystd, post-pre, mod vs low (older adults)")
 # labs(y = "Δ peak torque (Nm), post-pre, mod vs low (older adults)")

# Computing all contrasts and making a table

# Compute post-pre delta once, at tx level, for all groups
deltas2 <- draws2 |>
  select(drawid, draw, tx, time, sp) |>
  pivot_wider(names_from = time, values_from = draw) |>
  mutate(delta = post - pre) |>
  select(drawid, tx, sp, delta)

# Define the speeds you want reported
report_speeds <- c(0, 0.25, 0.5, 1)

# Contrast function, operating on deltas2
report_speeds <- c(0, 0.25, 0.5, 1)

compute_contrast <- function(data, group_a, group_b, label, targets = report_speeds) {
  # snap each requested speed to the nearest actual grid value
  actual_sp <- unique(data$sp)
  snapped <- sapply(targets, \(t) actual_sp[which.min(abs(actual_sp - t))])
  
  data |>
    filter(tx %in% c(group_a, group_b), sp %in% snapped) |>
    pivot_wider(names_from = tx, values_from = delta) |>
    mutate(contrast_val = .data[[group_a]] - .data[[group_b]]) |>
    summarise(.by = sp,
              m   = median(contrast_val),
              lwr = quantile(contrast_val, 0.05),
              upr = quantile(contrast_val, 0.95)) |>
    mutate(contrast = label,
           sp_requested = targets[match(sp, snapped)],  # keep a record of intended vs actual
           .before = sp)
}

# Run all 6 and bind
contrast_table <- bind_rows(
  compute_contrast(deltas2, "int_old_mod", "int_old_low", "Volume (older)"),
  compute_contrast(deltas2, "int_yng_mod", "int_yng_low", "Volume (younger)"),
  compute_contrast(deltas2, "int_old_low", "con_old_con", "Intervention vs control (low, older)"),
  compute_contrast(deltas2, "int_old_mod", "con_old_con", "Intervention vs control (mod, older)"),
  compute_contrast(deltas2, "int_old_low", "int_yng_low", "Age (low volume)"),
  compute_contrast(deltas2, "int_old_mod", "int_yng_mod", "Age (mod volume)")
) |>
  mutate(across(c(m, lwr, upr), ~ .x * max_pt)) |>   # remove/comment if staying on ystd scale, per your earlier decision
  relocate(contrast, sp, m, lwr, upr)

contrast_table

# "Volume (younger)" is the only contrast whose CI excludes zero at every single reported speed — [-9.14, -1.28] at sp=0, holding a negative-only interval all the way to [-9.68, -0.04] at sp=1. Every other contrast's interval straddles zero somewhere here. I.e., younger adults show a credible mod > low volume effect on post-pre torque change, while older adults don't (row 1-4, CI spans zero throughout). That's an age-dependent dose-response result, quite interesting. And mod actually produce less pre-post change than low. 



```

```{r}

## Fig suggestion 1: Contrast curves across speed, by comparison
# This should be similar to the ogtt fig

max_pt <- max(humac_dat$pt, na.rm = TRUE)

# Requires re-computing the contrasts across the full sp grid (pre-snapping) with multiple quantile widths. This is to avoid producing a plot that looks like a single-CI ribbon plot repeated 6 times - we want a richer nested-uncertainty visualization.

compute_contrast_curve <- function(data, group_a, group_b, label) {
  data |>
    filter(tx %in% c(group_a, group_b)) |>
    pivot_wider(names_from = tx, values_from = delta) |>
    mutate(contrast_val = .data[[group_a]] - .data[[group_b]]) |>
    summarise(.by = sp,
              m     = median(contrast_val),
              lwr90 = quantile(contrast_val, 0.05),  upr90 = quantile(contrast_val, 0.95),
              lwr75 = quantile(contrast_val, 0.125),  upr75 = quantile(contrast_val, 0.875),
              lwr50 = quantile(contrast_val, 0.25),   upr50 = quantile(contrast_val, 0.75)) |>
    mutate(contrast = label, .before = sp)
}

contrast_curves <- bind_rows(
  compute_contrast_curve(deltas2, "int_old_mod", "int_old_low", "Volume (older)"),
  compute_contrast_curve(deltas2, "int_yng_mod", "int_yng_low", "Volume (younger)"),
  compute_contrast_curve(deltas2, "int_old_low", "con_old_con", "Intervention vs control (low, older)"),
  compute_contrast_curve(deltas2, "int_old_mod", "con_old_con", "Intervention vs control (mod, older)"),
  compute_contrast_curve(deltas2, "int_old_low", "int_yng_low", "Age (low volume)"),
  compute_contrast_curve(deltas2, "int_old_mod", "int_yng_mod", "Age (mod volume)")
) |>
  mutate(across(c(m, lwr90:upr50), ~ .x * max_pt))

ggplot(contrast_curves, aes(sp, m)) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey40") +
  geom_ribbon(aes(ymin = lwr90, ymax = upr90), alpha = 0.15, fill = "steelblue") +
  geom_ribbon(aes(ymin = lwr75, ymax = upr75), alpha = 0.25, fill = "steelblue") +
  geom_ribbon(aes(ymin = lwr50, ymax = upr50), alpha = 0.35, fill = "steelblue") +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ contrast, ncol = 2) +
  labs(x = "Speed (normalized, sp)", y = "Δ Peak torque (Nm), post-pre",
       title = "Contrast curves across speed, by comparison")

## Fig suggestion 2: Hill-curve

max_pt <- max(humac_dat$pt, na.rm = TRUE)

vol_colors <- c("Control" = "grey40", "Low volume" = "steelblue", "Moderate volume" = "darkorange")

all_data <- draws2 |>
  filter(tx %in% c("int_old_low", "int_old_mod", "int_yng_low", "int_yng_mod", "con_old_con")) |>
  summarise(.by = c(tx, time, sp), m = median(draw)) |>
  separate(tx, into = c("allocation", "age", "cond"), remove = FALSE) |>
  mutate(m = m * max_pt,
         cond_label = case_when(
           tx == "con_old_con" ~ "Control",
           cond == "low" ~ "Low volume",
           cond == "mod" ~ "Moderate volume"))

y_range <- range(all_data$m)

make_panel <- function(data, filter_expr, title) {
  data |>
    filter({{ filter_expr }}) |>
    ggplot(aes(sp, m, color = cond_label, linetype = time, group = interaction(tx, time))) +
    geom_line(linewidth = 0.8) +
    scale_color_manual(values = vol_colors, drop = TRUE) +
    scale_y_continuous(limits = y_range) +
    labs(x = "Speed (normalized, sp)", y = "Predicted peak torque (Nm)",
         title = title, color = "Group", linetype = "Timepoint")
}

p_young <- make_panel(all_data, age == "yng", "Young (intervention)")
p_old   <- make_panel(all_data, age == "old" & allocation == "int", "Old (intervention)")
p_oldvc <- make_panel(all_data, tx %in% c("int_old_low", "int_old_mod", "con_old_con"), "Old: intervention vs control")

library(cowplot)
plot_grid(p_young, p_old, p_oldvc, ncol = 1, align = "v")



```

