---
title: "Muscle thickness — image-level modelling (VL only)"
format: html
editor: source
editor_options: 
  chunk_output_type: console
---

Kept separate from `03-muscle.qmd` so the working pipeline there is untouched.
Two changes from the working version:

1.  Vastus lateralis (VL) only.
2.  No averaging of the 3 images per leg/time-point before modelling — all
    image-level rows are kept, and an occasion-level random effect
    (`participant:leg:time`) is added to every candidate model so the 3
    images per occasion are treated as repeated measures of one true value,
    not as independent observations.

Otherwise this mirrors the working thickness section exactly: same form0-3
candidate structure, same contrast functions (unchanged — the coefficient
structure of the fixed effects doesn't change, only the data and the added
random effect do), same model-selection logic of settling on model 2 for
results.

```{r}
#| label: packages-and-settings
#| eval: true
#| message: false
#| warning: false
#| echo: true

library(brms)
library(marginaleffects)
library(tidyverse)
library(reliefdata)

options(marginaleffects_safe = FALSE)
```

### Data — image level, VL only

```{r}
#| label: data-image-level

us.dat <- reliefdata::relief_thickness |> 
  mutate(participant = as.character(participant))

condition <- reliefdata::relief_volume |> 
  mutate(participant = as.character(participant))

participants <- reliefdata::relief_participants |> 
  mutate(age_group = if_else(age < 40, "yng", "old"))

# Sanity check: how many images per participant/leg/time?
us.dat |> 
  count(participant, leg, time, name = "n_images") |> 
  count(n_images)

# Completeness check: any participant/leg missing a whole occasion
# (as opposed to just missing one of the 3 images within an occasion)?
us.dat |> 
  distinct(participant, leg, time) |> 
  complete(nesting(participant, leg), time) |> 
  mutate(present = !is.na(time) & paste(participant, leg, time) %in%
           paste(us.dat$participant, us.dat$leg, us.dat$time)) |> 
  summarise(.by = c(participant, leg),
            n_occasions = sum(present),
            missing = paste(time[!present], collapse = ", ")) |> 
  filter(n_occasions < 3) |> 
  arrange(participant)

thickness_img <- us.dat |>
  full_join(condition) |>
  select(-arm) |> 
  full_join(participants) |>
  select(-VIthick) |>   # VI dropped per decision above
  filter(!is.na(VLthick)) |> 
  mutate(y = log(VLthick),
         ystd = (y - mean(y, na.rm = TRUE)) / sd(y, na.rm = TRUE),
         condition = if_else(allocation == "con", "con", condition),
         tx = paste0(allocation, "_", age.grp, "_", condition),
         tx = factor(tx, levels = c("int_old_low", "int_old_mod",
                                     "int_yng_low", "int_yng_mod",
                                     "con_old_con")),
         tx_sex = paste0(tx, "_", sex),
         tx_sex = factor(tx_sex, levels = c(
           "int_old_low_female", "int_old_low_male",
           "int_old_mod_female", "int_old_mod_male",
           "int_yng_low_female", "int_yng_low_male",
           "int_yng_mod_female", "int_yng_mod_male",
           "con_old_con_female", "con_old_con_male"
         )),
         time = factor(time, levels = c("pre", "mid", "post")),
         sex_centered = if_else(sex == "female", -0.5, 0.5)) |> 
  filter(!is.na(y)) |> 
  filter(!is.na(time))

mean_us_img <- mean(thickness_img$y, na.rm = TRUE)
sd_us_img   <- sd(thickness_img$y, na.rm = TRUE)
```

#### Models

Same four candidates as the working pipeline, each with `(1|participant:leg:time)`
added for the image-level occasion structure. form0 didn't have a
`participant:leg` term in the working version either, but it needs the
occasion-level term now regardless — without it, the 3 images per occasion
would be treated as independent, artificially inflating N and understating
uncertainty.

```{r}
#| label: model-formulas

form0 <- bf(ystd ~ 0 + tx + tx:time + sex +
              (1 | participant) +
              (1 | participant:leg:time))

form1 <- bf(ystd ~ time + allocation + allocation:time +
              (1 | participant) +
              (1 | participant:leg) +
              (1 | participant:leg:time) +
              (time | tx_sex))

form2 <- bf(ystd ~ 0 + tx_sex + tx_sex:time +
              (1 + time | participant) +
              (1 | participant:leg) +
              (1 | participant:leg:time))

form3 <- bf(ystd ~ time + allocation + allocation:time +
              (1 + time | participant) +
              (1 | participant:leg) +
              (1 | participant:leg:time) +
              (time | tx_sex))

get_prior(form0, data = thickness_img)
get_prior(form1, data = thickness_img)
get_prior(form2, data = thickness_img)
get_prior(form3, data = thickness_img)

pri0 <- c(
  prior(normal(0, 1), class = "b", coef = "sexmale"),
  prior(normal(0, 1), class = "b", coef = "txcon_old_con"),
  prior(normal(0, 1), class = "b", coef = "txint_old_low"),
  prior(normal(0, 1), class = "b", coef = "txint_old_mod"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_low"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod"),
  prior(normal(0, 1), class = "b", coef = "txcon_old_con:timemid"),
  prior(normal(0, 1), class = "b", coef = "txint_old_low:timemid"),
  prior(normal(0, 1), class = "b", coef = "txint_old_mod:timemid"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_low:timemid"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod:timemid"),
  prior(normal(0, 1), class = "b", coef = "txcon_old_con:timepost"),
  prior(normal(0, 1), class = "b", coef = "txint_old_low:timepost"),
  prior(normal(0, 1), class = "b", coef = "txint_old_mod:timepost"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_low:timepost"),
  prior(normal(0, 1), class = "b", coef = "txint_yng_mod:timepost"),
  prior(exponential(1), class = sd),
  prior(exponential(1), class = sigma)
)

pri1 <- c(
  prior(normal(0, 1), class = "Intercept"),
  prior(normal(0, 1), class = "b"),
  prior(exponential(1), class = sd),
  prior(exponential(1), class = sigma)
)

pri2 <- c(
  prior(normal(0, 1), class = "b"),
  prior(lkj(2), class = "cor", group = "participant"),
  prior(exponential(1), class = sd, group = "participant"),
  prior(exponential(1), class = sd, group = "participant:leg"),
  prior(exponential(1), class = sd, group = "participant:leg:time"),
  prior(exponential(1), class = sigma)
)

pri3 <- c(
  prior(normal(0, 1), class = "Intercept"),
  prior(normal(0, 1), class = "b"),
  prior(lkj(2), class = "cor", group = "participant"),
  prior(exponential(1), class = sd),
  prior(exponential(1), class = sigma)
)

mt0.pp <- brm(form0, 
              prior = pri0, 
              data = thickness_img,
              file = "fits/XX/mt0",
              file_refit = "on_change", 
              sample_prior = "only")

mt1.pp <- brm(form1, 
              prior = pri1, 
              data = thickness_img,
              file = "fits/XX/mt1",
              file_refit = "on_change", 
              sample_prior = "only")

mt2.pp <- brm(form2, 
              prior = pri2, 
              data = thickness_img,
              file = "fits/XX/mt2",
              file_refit = "on_change", 
              sample_prior = "only")

mt3.pp <- brm(form3, 
              prior = pri3, 
              data = thickness_img,
              file = "fits/XX/mt3",
              file_refit = "on_change", 
              sample_prior = "only")




```

#### Contrast functions — unchanged from the working pipeline

Fixed-effect coefficient order is identical to the averaged-data models
(the added term is random-effects only), so `mt0_pred_contr()` through
`mt3_pred_contr()` are copied verbatim from `03-muscle.qmd`.

```{r}
#| label: contrast-functions

mt0_pred_contr <- function(m, nd1, nd2) {
  contr <- predictions(
    m,
    newdata = datagrid(tx = unique, time = unique),
    ndraws = nd1,
    hypothesis = c(
      "int_old_low_delta_mid" = "b2 - b1 = 0",
      "int_old_low_delta_post" = "b3 - b1 = 0",
      "int_old_low_delta_midpost" = "b3 - b2 = 0",
      "int_old_mod_delta_mid" = "b5 - b4 = 0",
      "int_old_mod_delta_post" = "b6 - b4 = 0",
      "int_old_mod_delta_midpost" = "b6 - b5 = 0",
      "int_yng_low_delta_mid" = "b8 - b7 = 0",
      "int_yng_low_delta_post" = "b9 - b7 = 0",
      "int_yng_low_delta_midpost" = "b9 - b8 = 0",
      "int_yng_mod_delta_mid" = "b11 - b10 = 0",
      "int_yng_mod_delta_post" = "b12 - b10 = 0",
      "int_yng_mod_delta_midpost" = "b12 - b11 = 0",
      "con_old_con_delta_mid" = "b14 - b13 = 0",
      "con_old_con_delta_post" = "b15 - b13 = 0",
      "con_old_con_delta_midpost" = "b15 - b14 = 0",
      "yng_vs_old_delta_mid" = "(((b11 - b10) + (b8 - b7)) / 2)  - (((b2 - b1) + (b5 - b4))/2) = 0",
      "old_vs_con_delta_mid" = "((b2 - b1) + (b5 - b4))/2  - (b14 - b13) = 0",
      "yng_vs_old_delta_midpost" = "(((b12 - b11) + (b9 - b8)) / 2)  - (((b3 - b2) + (b6 - b5))/2) = 0",
      "old_vs_con_delta_midpost" = "((b3 - b2) + (b6 - b5))/2  - (b15 - b14) = 0",
      "yng_vs_old_delta_post" = "(((b12 - b10) + (b9 - b7)) / 2)  - (((b3 - b1) + (b6 - b4))/2) = 0",
      "old_vs_con_delta_post" = "((b3 - b1) + (b6 - b4))/2  - (b15 - b13) = 0",
      "yng_mod_vs_low_mid" = "(b11 - b10) - (b8 - b7) = 0",
      "old_mod_vs_low_mid" = "(b5 - b4) - (b2 - b1) = 0",
      "yng_mod_vs_low_midpost" = "(b12 - b11) - (b9 - b8) = 0",
      "old_mod_vs_low_midpost" = "(b6 - b5) - (b3 - b2) = 0",
      "yng_mod_vs_low_post" = "(b12 - b10) - (b9 - b7) = 0",
      "old_mod_vs_low_post" = "(b6 - b4) - (b3 - b1) = 0",
      "yng_vs_old_vol_mid" = "((b11 - b10) - (b8 - b7)) - ((b5 - b4) - (b2 - b1)) = 0",
      "yng_vs_old_vol_midpost" = "((b12 - b11) - (b9 - b8)) - ((b6 - b5) - (b3 - b2)) = 0",
      "yng_vs_old_vol_post" = "((b12 - b10) - (b9 - b7)) - ((b6 - b4) - (b3 - b1)) = 0"
    ))
  pred <- predictions(
    m, ndraws = nd2,
    newdata = datagrid(tx = unique, time = unique, sex = unique),
    re_formula = NA)
  list(contr = contr, pred = pred)
}

mt1_pred_contr <- function(m, nd1, nd2) {
  grid <- thickness_img %>%
    distinct(tx_sex, allocation) %>%
    expand_grid(time = unique(thickness_img$time))
  contr <- predictions(
    m, newdata = grid, re_formula = ~ (time | tx_sex),
    ndraws = nd1, by = c("tx_sex", "time"),
    hypothesis = c(
      "int_old_low_delta_mid" = "b2 - b1 = 0",
      "int_old_low_delta_post" = "b3 - b1 = 0",
      "int_old_low_delta_midpost" = "b3 - b2 = 0",
      "int_old_mod_delta_mid" = "b5 - b4 = 0",
      "int_old_mod_delta_post" = "b6 - b4 = 0",
      "int_old_mod_delta_midpost" = "b6 - b5 = 0",
      "int_yng_low_delta_mid" = "b8 - b7 = 0",
      "int_yng_low_delta_post" = "b9 - b7 = 0",
      "int_yng_low_delta_midpost" = "b9 - b8 = 0",
      "int_yng_mod_delta_mid" = "b11 - b10 = 0",
      "int_yng_mod_delta_post" = "b12 - b10 = 0",
      "int_yng_mod_delta_midpost" = "b12 - b11 = 0",
      "con_old_con_delta_mid" = "b14 - b13 = 0",
      "con_old_con_delta_post" = "b15 - b13 = 0",
      "con_old_con_delta_midpost" = "b15 - b14 = 0",
      "yng_vs_old_delta_mid" = "(((b11 - b10) + (b8 - b7)) / 2)  - (((b2 - b1) + (b5 - b4))/2) = 0",
      "old_vs_con_delta_mid" = "((b2 - b1) + (b5 - b4))/2  - (b14 - b13) = 0",
      "yng_vs_old_delta_midpost" = "(((b12 - b11) + (b9 - b8)) / 2)  - (((b3 - b2) + (b6 - b5))/2) = 0",
      "old_vs_con_delta_midpost" = "((b3 - b2) + (b6 - b5))/2  - (b15 - b14) = 0",
      "yng_vs_old_delta_post" = "(((b12 - b10) + (b9 - b7)) / 2)  - (((b3 - b1) + (b6 - b4))/2) = 0",
      "old_vs_con_delta_post" = "((b3 - b1) + (b6 - b4))/2  - (b15 - b13) = 0",
      "yng_mod_vs_low_mid" = "(b11 - b10) - (b8 - b7) = 0",
      "old_mod_vs_low_mid" = "(b5 - b4) - (b2 - b1) = 0",
      "yng_mod_vs_low_midpost" = "(b12 - b11) - (b9 - b8) = 0",
      "old_mod_vs_low_midpost" = "(b6 - b5) - (b3 - b2) = 0",
      "yng_mod_vs_low_post" = "(b12 - b10) - (b9 - b7) = 0",
      "old_mod_vs_low_post" = "(b6 - b4) - (b3 - b1) = 0",
      "yng_vs_old_vol_mid" = "((b11 - b10) - (b8 - b7)) - ((b5 - b4) - (b2 - b1)) = 0",
      "yng_vs_old_vol_midpost" = "((b12 - b11) - (b9 - b8)) - ((b6 - b5) - (b3 - b2)) = 0",
      "yng_vs_old_vol_post" = "((b12 - b10) - (b9 - b7)) - ((b6 - b4) - (b3 - b1)) = 0"
    ))
  pred <- predictions(m, ndraws = nd2, newdata = grid, re_formula = ~ (time | tx_sex))
  list(contr = contr, pred = pred)
}

mt2_pred_contr <- function(m, nd1, nd2) {
  contr <- predictions(
    m, newdata = datagrid(tx_sex = unique, time = unique), re_formula = NA,
    ndraws = nd1,
    hypothesis = c(
      "int_old_low_delta_mid" = "((b2-b1)+(b5-b4))/2 = 0",
      "int_old_mod_delta_mid" = "((b8-b7)+(b11-b10))/2 = 0",
      "int_yng_low_delta_mid" = "((b14-b13)+(b17-b16))/2 = 0",
      "int_yng_mod_delta_mid" = "((b20-b19)+(b23-b22))/2  = 0",
      "con_old_con_delta_mid" = "((b26-b25)+(b29-b28))/2 = 0",
      "int_old_low_delta_midpost" = "((b3-b2)+(b6-b5))/2 = 0",
      "int_old_mod_delta_midpost" = "((b9-b8)+(b12-b11))/2 = 0",
      "int_yng_low_delta_midpost" = "((b15-b14)+(b18-b17))/2 = 0",
      "int_yng_mod_delta_midpost" = "((b21-b20)+(b24-b23))/2  = 0",
      "con_old_con_delta_midpost" = "((b27-b26)+(b30-b29))/2 = 0",
      "int_old_low_delta_post" = "((b3-b1)+(b6-b4))/2 = 0",
      "int_old_mod_delta_post" = "((b9-b7)+(b12-b10))/2 = 0",
      "int_yng_low_delta_post" = "((b15-b13)+(b18-b16))/2 = 0",
      "int_yng_mod_delta_post" = "((b21-b19)+(b24-b22))/2  = 0",
      "con_old_con_delta_post" = "((b27-b25)+(b30-b28))/2 = 0",
      "yng_vs_old_delta_mid" = "(((b20-b19)+(b23-b22))/2 + ((b14-b13)+(b17-b16))/2)/2 - (((b8-b7)+(b11-b10))/2 + ((b2-b1)+(b5-b4))/2)/2 = 0 ",
      "old_vs_con_delta_mid" = "((((b8-b7)+(b11-b10))/2 + ((b2-b1)+(b5-b4))/2) / 2)  - (((b26-b25)+(b29-b28))/2) = 0",
      "yng_vs_old_delta_midpost" = "(((b21-b20)+(b24-b23))/2 + ((b15-b14)+(b18-b17))/2)/2 - (((b9-b8)+(b12-b11))/2 + ((b3-b2)+(b6-b5))/2)/2 = 0 ",
      "old_vs_con_delta_midpost" = "((((b9-b8)+(b12-b11))/2 + ((b3-b2)+(b6-b5))/2) / 2)  - (((b27-b26)+(b30-b29))/2) = 0",
      "yng_vs_old_delta_post" = "(((b21-b19)+(b24-b22))/2 + ((b15-b13)+(b18-b16))/2)/2 - (((b9-b7)+(b12-b10))/2 + ((b3-b1)+(b6-b4))/2)/2 = 0 ",
      "old_vs_con_delta_post" = "((((b9-b7)+(b12-b10))/2 + ((b3-b1)+(b6-b4))/2) / 2)  - (((b27-b25)+(b30-b28))/2) = 0",
      "yng_mod_vs_low_mid" = "((b20-b19)+(b23-b22))/2 - ((b14-b13)+(b17-b16))/2 = 0",
      "old_mod_vs_low_mid" = "((b8-b7)+(b11-b10))/2 - ((b2-b1)+(b5-b4))/2 = 0",
      "yng_mod_vs_low_midpost" = "((b21-b20)+(b24-b23))/2 - ((b15-b14)+(b18-b17))/2 = 0",
      "old_mod_vs_low_midpost" = "((b9-b8)+(b12-b11))/2 - ((b3-b2)+(b6-b5))/2 = 0",
      "yng_mod_vs_low_post" = "((b21-b19)+(b24-b22))/2 - ((b15-b13)+(b18-b16))/2 = 0",
      "old_mod_vs_low_post" = "((b9-b7)+(b12-b10))/2 - ((b3-b1)+(b6-b4))/2 = 0",
      "yng_vs_old_vol_mid" = "(((b20-b19)+(b23-b22))/2 - ((b14-b13)+(b17-b16))/2) - (((b8-b7)+(b11-b10))/2 - ((b2-b1)+(b5-b4))/2) = 0",
      "yng_vs_old_vol_midpost" = "(((b21-b20)+(b24-b23))/2 - ((b15-b14)+(b18-b17))/2) - (((b9-b8)+(b12-b11))/2 - ((b3-b2)+(b6-b5))/2) = 0",
      "yng_vs_old_vol_post" = "(((b21-b19)+(b24-b22))/2 - ((b15-b13)+(b18-b16))/2) - (((b9-b7)+(b12-b10))/2 - ((b3-b1)+(b6-b4))/2) = 0"
    ))
  pred <- predictions(m, ndraws = nd2, newdata = datagrid(tx_sex = unique, time = unique), re_formula = NA)
  list(contr = contr, pred = pred)
}

mt3_pred_contr <- function(m, nd1, nd2) {
  grid <- thickness_img |>
    distinct(tx_sex, allocation) |>
    expand_grid(time = unique(thickness_img$time))
  contr <- predictions(
    m, newdata = grid, re_formula = ~ (time | tx_sex),
    ndraws = nd1, by = c("tx_sex", "time"),
    hypothesis = c(
      "int_old_low_delta_mid" = "((b2-b1)+(b5-b4))/2 = 0",
      "int_old_mod_delta_mid" = "((b8-b7)+(b11-b10))/2 = 0",
      "int_yng_low_delta_mid" = "((b14-b13)+(b17-b16))/2 = 0",
      "int_yng_mod_delta_mid" = "((b20-b19)+(b23-b22))/2  = 0",
      "con_old_con_delta_mid" = "((b26-b25)+(b29-b28))/2 = 0",
      "int_old_low_delta_midpost" = "((b3-b2)+(b6-b5))/2 = 0",
      "int_old_mod_delta_midpost" = "((b9-b8)+(b12-b11))/2 = 0",
      "int_yng_low_delta_midpost" = "((b15-b14)+(b18-b17))/2 = 0",
      "int_yng_mod_delta_midpost" = "((b21-b20)+(b24-b23))/2  = 0",
      "con_old_con_delta_midpost" = "((b27-b26)+(b30-b29))/2 = 0",
      "int_old_low_delta_post" = "((b3-b1)+(b6-b4))/2 = 0",
      "int_old_mod_delta_post" = "((b9-b7)+(b12-b10))/2 = 0",
      "int_yng_low_delta_post" = "((b15-b13)+(b18-b16))/2 = 0",
      "int_yng_mod_delta_post" = "((b21-b19)+(b24-b22))/2  = 0",
      "con_old_con_delta_post" = "((b27-b25)+(b30-b28))/2 = 0",
      "yng_vs_old_delta_mid" = "(((b20-b19)+(b23-b22))/2 + ((b14-b13)+(b17-b16))/2)/2 - (((b8-b7)+(b11-b10))/2 + ((b2-b1)+(b5-b4))/2)/2 = 0 ",
      "old_vs_con_delta_mid" = "((((b8-b7)+(b11-b10))/2 + ((b2-b1)+(b5-b4))/2) / 2)  - (((b26-b25)+(b29-b28))/2) = 0",
      "yng_vs_old_delta_midpost" = "(((b21-b20)+(b24-b23))/2 + ((b15-b14)+(b18-b17))/2)/2 - (((b9-b8)+(b12-b11))/2 + ((b3-b2)+(b6-b5))/2)/2 = 0 ",
      "old_vs_con_delta_midpost" = "((((b9-b8)+(b12-b11))/2 + ((b3-b2)+(b6-b5))/2) / 2)  - (((b27-b26)+(b30-b29))/2) = 0",
      "yng_vs_old_delta_post" = "(((b21-b19)+(b24-b22))/2 + ((b15-b13)+(b18-b16))/2)/2 - (((b9-b7)+(b12-b10))/2 + ((b3-b1)+(b6-b4))/2)/2 = 0 ",
      "old_vs_con_delta_post" = "((((b9-b7)+(b12-b10))/2 + ((b3-b1)+(b6-b4))/2) / 2)  - (((b27-b25)+(b30-b28))/2) = 0",
      "yng_mod_vs_low_mid" = "((b20-b19)+(b23-b22))/2 - ((b14-b13)+(b17-b16))/2 = 0",
      "old_mod_vs_low_mid" = "((b8-b7)+(b11-b10))/2 - ((b2-b1)+(b5-b4))/2 = 0",
      "yng_mod_vs_low_midpost" = "((b21-b20)+(b24-b23))/2 - ((b15-b14)+(b18-b17))/2 = 0",
      "old_mod_vs_low_midpost" = "((b9-b8)+(b12-b11))/2 - ((b3-b2)+(b6-b5))/2 = 0",
      "yng_mod_vs_low_post" = "((b21-b19)+(b24-b22))/2 - ((b15-b13)+(b18-b16))/2 = 0",
      "old_mod_vs_low_post" = "((b9-b7)+(b12-b10))/2 - ((b3-b1)+(b6-b4))/2 = 0",
      "yng_vs_old_vol_mid" = "(((b20-b19)+(b23-b22))/2 - ((b14-b13)+(b17-b16))/2) - (((b8-b7)+(b11-b10))/2 - ((b2-b1)+(b5-b4))/2) = 0",
      "yng_vs_old_vol_midpost" = "(((b21-b20)+(b24-b23))/2 - ((b15-b14)+(b18-b17))/2) - (((b9-b8)+(b12-b11))/2 - ((b3-b2)+(b6-b5))/2) = 0",
      "yng_vs_old_vol_post" = "(((b21-b19)+(b24-b22))/2 - ((b15-b13)+(b18-b16))/2) - (((b9-b7)+(b12-b10))/2 - ((b3-b1)+(b6-b4))/2) = 0"
    ))
  pred <- predictions(m, ndraws = nd2, newdata = grid, re_formula = ~ (time | tx_sex))
  list(contr = contr, pred = pred)
}
```

#### Prior predictive checks

```{r}
#| label: prior-predictive

mt0_preds <- mt0_pred_contr(mt0.pp, nd1 = 1000, nd2 = 50)
mt1_preds <- mt1_pred_contr(mt1.pp, nd1 = 1000, nd2 = 50)
mt2_preds <- mt2_pred_contr(mt2.pp, nd1 = 1000, nd2 = 50)
mt3_preds <- mt3_pred_contr(mt3.pp, nd1 = 1000, nd2 = 50)

bind_rows(
  posterior_draws(mt0_preds$contr) |> mutate(model = "mt0"),
  posterior_draws(mt1_preds$contr) |> mutate(model = "mt1"),
  posterior_draws(mt2_preds$contr) |> mutate(model = "mt2"),
  posterior_draws(mt3_preds$contr) |> mutate(model = "mt3")
) |>
  ggplot(aes(draw, color = model)) +
  geom_line(stat = "density") +
  facet_wrap(~ hypothesis) +
  geom_vline(xintercept = 0, lty = 2) +
  coord_cartesian(xlim = c(-5, 5))
```

#### Posterior fits

```{r}
#| label: posterior-fits

mt0.post <- update(mt0.pp, 
                   sample_prior = "no", 
                   iter = 4000,
                   file = "fits/XX/mt0.post", 
                   file_refit = "on_change")

mt1.post <- update(mt1.pp, 
                   sample_prior = "no", 
                   cores = 4,
                   control = list(adapt_delta = 0.95, max_treedepth = 15),
                   iter = 4000,
                   file = "fits/XX/mt1.post", 
                   file_refit = "on_change")

mt2.post <- update(mt2.pp, 
                   sample_prior = "no", 
                   cores = 4,
                   control = list(adapt_delta = 0.95, max_treedepth = 15),
                   iter = 4000,
                   file = "fits/XX/mt2.post", 
                   file_refit = "on_change")

mt3.post <- update(mt3.pp, 
                   sample_prior = "no", 
                   cores = 4,
                   control = list(adapt_delta = 0.95, max_treedepth = 15),
                   iter = 4000,
                   file = "fits/XX/mt3.post", 
                   file_refit = "on_change")

pp_check(mt0.post)
pp_check(mt1.post)
pp_check(mt2.post)
pp_check(mt2.post, type = "dens_overlay_grouped", group = "tx_sex")
pp_check(mt3.post)
#pp_check(mt3.post, type = "dens_overlay_grouped", group = "tx_sex")

#pp_checks look pretty good 

```

### Results — model 2, image-level

Using model 2 for results, same as the working (averaged) pipeline.

```{r}
#| label: results-model2

mt2_preds <- mt2_pred_contr(mt2.post, nd1 = 4000, nd2 = 50)

contrast_mt_img <- posterior_draws(mt2_preds$contr) |>
  mutate(draw = draw * sd_us_img, draw = exp(draw)) |>
  summarise(.by = hypothesis,
            mean = mean(draw),
            lower95 = quantile(draw, 0.025),
            upper95 = quantile(draw, 0.975),
            pos = sum(draw > 1) / n()) |>
  mutate(
    timepoint = case_when(
      grepl("_mid$", hypothesis)     ~ "mid",
      grepl("_midpost$", hypothesis) ~ "midpost",
      grepl("_post$", hypothesis)    ~ "post",
      TRUE ~ NA_character_
    ),
    pct_change = (mean - 1) * 100,
    pct_lower  = (lower95 - 1) * 100,
    pct_upper  = (upper95 - 1) * 100,
    credible   = ifelse(lower95 > 1 | upper95 < 1, "Yes", "No")
  )

saveRDS(contrast_mt_img, "data/data-gen/contrast_mt_img.rds")

contrast_mt_img
```

### Compare to the working (averaged) model

```{r}
#| label: compare-to-working-model

contrast_mt_working <- readRDS("data/data-gen/contrast_mt.rds") |>
  select(hypothesis, mean_avg = mean, lower95_avg = lower95, upper95_avg = upper95)

contrast_mt_img |>
  select(hypothesis, mean_img = mean, lower95_img = lower95, upper95_img = upper95) |>
  left_join(contrast_mt_working, by = "hypothesis") |>
  mutate(width_avg = upper95_avg - lower95_avg,
         width_img = upper95_img - lower95_img,
         width_ratio = width_img / width_avg)
```

### Results

Quite unambiguous in the end. The width_ratio column ranges from about 0.91 to 1.03 — some contrasts even come out slightly narrower with the image-level model, some slightly wider, with no directional pattern and nothing beyond roughly a 9% swing in either direction (the extremes are int_old_mod_delta_mid at 0.94 and old_mod_vs_low_midpost at 0.91, narrowest; con_old_con_delta_mid at 1.03, widest). Averaged across all 30 rows, the ratio sits right around 0.98–0.99 — essentially no systematic change in precision.

And critically: every single credible call from your earlier table (Yes/No) matches between the two versions — same contrasts clear the 95% ETI in both, same ones don't. Point estimates (mean_img vs mean_avg) also agree to the second or third decimal throughout.

So the conclusion is clean: modelling all three images per occasion instead of their mean changes nothing meaningful — not the point estimates, not the interval widths in any consistent direction, not which effects are credible. The averaging in the working pipeline was a safe simplification, not a shortcut that cost you anything.
