---
title: "Untitled"
format: html
editor: source
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

```
 


```{r}
library(here)
library(tidyverse)   
library(ggdist)      
library(patchwork)   
library(cowplot)     


```

## Results, in a short format

### Muscle Volume

```{r}
#| label: fig-muscle-volume
#| fig-cap: "Posterior distribution of five between-group contrasts for the ratio (post/pre) of thigh muscle volume. Each row shows the full posterior density, with the point marking the mean ratio and the black bars the 50% and 95% CI. The dashed line marks no change (ratio = 1); the density is shaded blue where the ratio exceeds 1. The percentage label gives the posterior probability that the ratio exceeds 1 (pos), as a measure of directional confidence."
#| fig-width: 10
#| fig-height: 6

readRDS(here("figures", "fig-muscle-volume-combined.RDS"))

#use fig-muscle-volume.RDS for the manuscript version

```

- Old: Intervention vs. Control — weak-moderate evidence (3.8%, pos 97.6%)
- Young vs. Old (intervention) — weak evidence (2.2%, pos 94.0%)
- Young: Mod vs. Low — strong evidence (1.9%, pos >99.9%)
- Old: Mod vs. Low — strong evidence (1.6%, pos 99.9%)
- Volume effect: Young vs. Old — no evidence (0.3%, pos 68.5%)

### Muscle Thickness


```{r}
#| label: fig-muscle-thickness
#| fig-cap: ""
#| fig-width: 10
#| fig-height: 6

readRDS(here("figures", "fig-muscle-thickness.RDS"))

#this is only for the supplementary version

```


- Old: Intervention vs. Control — no/weak evidence (post: 4.2%, pos 84.8%; mid: 1.8%, pos 66.0%)
- Young vs. Old (intervention) — weak evidence (post: 2.8%, pos 83.8%; mid: 3.5%, pos 88.7%)
- Young: Mod vs. Low — weak evidence (post: 2.9%, pos 83.9%; mid: 3.3%, pos 88.4%)
- Old: Mod vs. Low — strong evidence (post: 8.8%, pos 99.8%; mid: 7.3%, pos 99.7%)
- Volume effect: Young vs. Old — weak evidence, favoring old (post: pos 91.9% for old>young; mid: pos 84.0%)

### Upper-Body Lean Mass

```{r}
#| label: fig-ub-lean-mass
#| fig-cap: ""
#| fig-width: 10
#| fig-height: 6

readRDS(here("figures", "fig-ub-lean-mass.RDS"))

#this is only for the supplementary version

```


- Old: Intervention vs. Control — no/weak evidence (post: 0.3%, pos 57.8%; mid: 1.6%, pos 87.8%)
- Young vs. Old (intervention) — weak evidence (post: 1.9%, pos 95.5%; mid: no evidence, pos 74.1%)
- Young: Mod vs. Low — no evidence (pos 63.1%)
- Old: Mod vs. Low — no evidence (pos 52.0%)
- Volume effect: Young vs. Old — no evidence at either timepoint (pos 57–60%)

### Peak Torque (force–velocity, speed-dependent)

```{r}
#| label: fig-force-velocity
#| fig-cap: "Model-predicted peak isokinetic and isometric torque as a function of angular velocity (force–velocity relationship) in young and older adults across the resistance-training intervention. Lines show posterior median predictions from the Bayesian multilevel model, colored by training volume (low/moderate) and styled by timepoint (pre/mid/post). Top left: young intervention participants. Top right: older intervention participants. Bottom left: older passive control. Bottom right: pooled contrasts (intervention vs. control, and young vs. old, both pooled across volume condition) at the four tested angular velocities, shown as posterior mean change from pre-intervention (Δ, Nm) with 95% credible intervals, at the mid- and post-intervention timepoints."
#| fig-width: 11
#| fig-height: 12

readRDS(here("figures/fig-force-velocity.rds"))

```


- Old: Intervention vs. Control — weak evidence, all speeds (post: pos 79–86%; mid: pos 75–91%)
- Young vs. Old (intervention) — weak evidence overall; stronger and credible specifically in the low-volume condition (mid: pos 97.7–98.3%; post: pos 91.5–97.1%), not credible under moderate volume
- Young: Mod vs. Low — strong evidence, smaller increase under moderate, all speeds (pos 98.8–99.9%)
- Old: Mod vs. Low — no evidence of difference, any speed (pos ~50–60%)
- Volume effect: Young vs. Old — moderate-strong evidence (less favorable to moderate volume in young) at isometric/60°/120°/s (pos 98.2–99.6%), weak at 240°/s (pos ~94%)

### 1RM Leg Press

```{r}
#| label: fig-leg-press
#| fig-cap: ""
#| fig-width: 10
#| fig-height: 6

readRDS(here("figures", "fig-leg-press.RDS"))

#this is only for the supplementary version

```


- Old: Intervention vs. Control — no evidence (post: pos 60.6%; mid: pos 73.7%)
- Young vs. Old (intervention) — no evidence (pos ~52%, both timepoints)
- Young: Mod vs. Low — no evidence (pos 70.4%)
- Old: Mod vs. Low — no evidence (pos 53.0%)
- Volume effect: Young vs. Old — no evidence at either timepoint (pos 64–65%)

### Bicep Curl MVC

```{r}
#| label: fig-bicep-curl
#| fig-cap: ""
#| fig-width: 10
#| fig-height: 6

readRDS(here("figures", "fig-bicep-curl.RDS"))

#this is only for the supplementary version

```


- Old: Intervention vs. Control — strong evidence (post: 14.6%, pos >99.9%; mid: 14.1%, pos 99.9%) — driven mainly by control-group decline
- Young vs. Old (intervention) — no/weak evidence (post: pos 71.1%; mid: pos 80.3%)
- Young: Mod vs. Low — weak evidence, likely smaller under moderate (pos 92.4%)
- Old: Mod vs. Low — no evidence (pos 74.7%)
- Volume effect: Young vs. Old — weak-moderate evidence (post: pos 93.2% for old>young; mid: pos 85.0%)

### OGTT

```{r}
#| label: fig-ogtt
#| fig-cap: "Oral glucose tolerance test (OGTT) responses before and after the training intervention, by age group. Top panels: blood glucose concentrations (mmol L⁻¹) across the 2-hour sampling period, shown for pre-intervention (left) and post-intervention (right) timepoints, in young (20–30 years) and old (70+ years) participants. Shaded ribbons represent the 10th, 25th, 75th, and 90th percentiles of the posterior distribution around the mean (solid line). Bottom panel: the estimated difference in training-induced change in blood glucose (post minus pre) between young and old participants, across the sampling period. Shaded ribbons represent the 2.5th, 10th, 25th, 75th, 90th, and 97.5th percentiles of the posterior distribution around the mean difference (solid line); the dashed horizontal line indicates no difference between age groups."
#| fig-width: 10
#| fig-height: 8

readRDS(here("figures/fig-ogtt.RDS"))

```

- Old: Intervention vs. Control — credible reduction from ~38 min onward through end of window (pos ~97.5–97.7%)
- Young vs. Old (training response) — credible difference 30–78 min post-ingestion (peak pos 100% at 54 min); not credible outside this window

### UBF Protein

```{r}
#| label: fig-protein-rna
#| fig-cap: "UBF and c-Myc (high peak, ~60 kDa) protein content, and total RNA yield, before and 3 weeks into (and after, where applicable) the training intervention. Panels A, C, and E: predicted values at each measured timepoint, corrected area (arbitrary units) for UBF and c-Myc, and total RNA (ng, adjusted for tissue weight). Error bars represent the 95% credible interval around the posterior mean (point and line). Panels B, D, and F: posterior distributions for five key between-group and between-condition contrasts — Old: Intervention vs. Control, Young vs. Old (intervention), Young: Mod vs. Low, Old: Mod vs. Low, and Volume effect: Young vs. Old, expressed as the ratio of change between the two groups or conditions being compared. For UBF and c-Myc, only the pre-to-week 3 change is shown, reflecting the timepoints measured. For total RNA, both the pre-to-week 3 change (grey) and the pre-to-post change (blue) are shown, per the legend. Points show the posterior mean ratio with 95% credible intervals; the dashed vertical line indicates no difference between groups (ratio = 1). Percentages indicate the posterior probability that the ratio exceeds 1 — i.e., the probability that the first-named group or condition in each comparison showed the larger increase."
#| fig-width: 16
#| fig-height: 10

readRDS(here("figures/fig-protein-rna.rds"))

```


- Old: Intervention vs. Control — no evidence (pos 58.6%, wrong direction/not credible)
- Young vs. Old — no evidence (pos 56.8%)
- Young: Mod vs. Low — no evidence (pos 53.9%)
- Old: Mod vs. Low — no evidence (pos 67.8%)
- Volume effect: Young vs. Old — no evidence (pos 67.2%)

### c-Myc Protein (high peak)

- Old: Intervention vs. Control — strong evidence (99.8%, pos 99.3%) — driven mainly by control-group decrease
- Young vs. Old — no evidence (pos 61.5%)
- Young: Mod vs. Low — weak evidence (pos 83.5%)
- Old: Mod vs. Low — weak evidence (pos 84.5%)
- Volume effect: Young vs. Old — no evidence (pos 46.7%)

### Total RNA

- Old: Intervention vs. Control — no evidence at post (pos 52.4%); weak evidence at w3 (pos 87.6%)
- Young vs. Old — no evidence at post (pos 65.3%); weak evidence at w3 (pos 88.9%, young > old); reverses direction between w3 and post (pos 91.9% for young falling relative to old)
- Young: Mod vs. Low — no evidence at either timepoint (pos 56–66%)
- Old: Mod vs. Low — no evidence at either timepoint (pos 60–67%)
- Volume effect: Young vs. Old — no evidence at either timepoint (pos 57–68%)

### RNA-sequencing

```{r}
#| label: fig-rnaseq-combined
#| fig-cap: "Baseline and training-related transcriptional differences between age groups. (A) Baseline age differences (Young − Old): volcano plot of gene-level estimates (log₂ fold-change) against FDR (log₂ scale), with the dashed line marking FDR = 5%. Genes with the strongest evidence of differential expression are labeled. (B) Rank-based gene-set enrichment analysis (GSEA) of the baseline age contrast, showing the running enrichment score (top) and per-gene contrast effect size (z-score, bottom) across gene rank position, for three representative gene sets (oxidative phosphorylation, heterochromatin formation, complement activation, classical pathway). (C) Age × training-response difference (Young − Old): volcano plot of gene-level estimates for the interaction contrast, in the same format as (A). (D) GSEA of the age × training-response interaction contrast, in the same format as (B), for three representative gene sets (RNA splicing, angiogenesis, aerobic respiration)."
#| fig-width: 10
#| fig-height: 10


library(seqwrap)
m2r <- readRDS(here("data-out/m2-rnaseq.RDS"))
m2s <- seqwrap_summarise(m2r, verbose = FALSE)

n_diff_expressed <- m2s$summaries |> 
  dplyr::filter(term == "ageyoung") |> 
  dplyr::mutate(p.adj = p.adjust(p.value, method = "BH")) |> 
  dplyr::filter(p.adj < 0.05) |> 
  nrow()



n_sig_young_t3 <- m2s$summaries |> 
  dplyr::filter(term == "young_t3") |> 
  dplyr::mutate(p.adj = p.adjust(p.value, method = "BH")) |> 
  dplyr::filter(p.adj < 0.05) |> 
  nrow()

n_sig_old_t3 <- m2s$summaries |> 
  dplyr::filter(term == "old_t3") |> 
  dplyr::mutate(p.adj = p.adjust(p.value, method = "BH")) |> 
  dplyr::filter(p.adj < 0.05) |> 
  nrow()

n_sig_old_volume_t3 <- m2s$summaries |> 
  dplyr::filter(term == "old_volume_t3") |> 
  dplyr::mutate(p.adj = p.adjust(p.value, method = "BH")) |> 
  dplyr::filter(p.adj < 0.05) |> 
  nrow()



readRDS(here("figures/fig-rnaseq.RDS"))


```

- Baseline age difference — large number of genes differentially expressed (FDR<5%); young higher in oxidative phosphorylation, lower in complement activation and heterochromatin formation
- Training effect (baseline→post) — large, broad response in both age groups (thousands of genes each); dominant biology (ECM/collagen up, RNA-processing/ribosome biogenesis down) shared between groups
- Age × training-response interaction — essentially null at the gene level (1 gene at FDR 5%), but GSEA shows subtle magnitude differences (RNA splicing down-regulation, angiogenesis/aerobic respiration up-regulation more pronounced in young)
- Volume effect (Old only) — substantial number of genes differ by volume in old (mitochondrial/oxidative up, contractile/translation down with higher volume); essentially no volume effect detected in young
