---
title: "Beta regression simulations"
format: typst
bibliography: bibliography.bib
bibliographystyle: "the-lancet"
editor_options: 
  chunk_output_type: console
---

## The Beta distribution

The Beta distribution is parameterized by $\alpha$ and $\beta$, called shape parameters. These are positive numbers and related to the central tendency $\mu$ and inverse dispersion (or precision [@douma2019]), $\phi$ of the distribution:

$$
\begin{align}
\alpha &= \mu \times \phi \\
\beta &= (1 - \mu) \times \phi
\end{align}
$$


$\mu$ is modelled on the logit scale where $\eta$ is the linear predictor. To get $\mu$ from $\eta$ we apply the inverse logit transformation:

$$\mu = \frac{\operatorname{exp}(\eta)}{1+\operatorname{exp}(\eta)}$$


The inverse dispersion parameter $\phi$ is modeled on the log scale:

$$\phi = \operatorname{exp}(1/\sigma)$$
To simulate values from the beta distribution we can use `rbeta()`, e.g.:

```{r}
#| message: false
#| warning: false


library(tidyverse)
library(gt)

y <- rbeta(100, 2, 5)


data.frame(y) |>
  ggplot(aes(y)) + geom_histogram()

```


The mean of the Beta distribution ($\mu$) is related to the linear predictor, $\eta$, as shown below.


```{r}
eta <- seq(from = -6, to = 6, by = 0.1)

data.frame(eta, mu = plogis(eta)) |>
  ggplot(aes(eta, mu)) + geom_line()
```


## Simulating beta-distributed data

In our study we want to explore the effect of age (~~a continuous variable~~ a factor variable) and training status (a factor variable) on the proportion of retained introns (IR; $\text{IR} \in [0,1]$). We are doing this in a repeated design, meaning that the same participants are evaluated before and after training (time). We add a varying intercept ($b_0 \sim \operatorname{Normal}(0, \sigma_b)$) to the linear predictor to induce correlations in the data set.

 
$$
\begin{align}
\text{IR} &\sim \text{Beta}(\alpha, \beta) \\
\alpha &= \mu \times \phi \\
\beta &= (1 - \mu) \times \phi \\
\mu &= \frac{\operatorname{exp}(\eta)}{1+\operatorname{exp}(\eta)} \\
\phi &= \operatorname{exp}(1/\sigma) \\
\eta &= \beta_0 + \beta_1 \times \text{age} + \beta_2 \times \text{training} + \beta_3 \times (\text{age} \times \text{time}) + b_0 \\
\sigma &= 2 \\
b_0 &\sim \operatorname{Normal}(0, \sigma_b)
\end{align}
$$

We can convert this model to a simulation function that will give us a data set of size N, balanced between groups. 

```{r}


ir_sim1 <- function(n = 10, 
                    beta0 = 0, 
                    beta1 = -2, 
                    beta2 = 2, 
                    beta3 = -2, 
                    sigma = 2, 
                    sigma_b0 = 1) {
  
  if(!n %%2 == 0) stop("n must be an even number")
  
  age <- c(rep(0, n/2), rep(1, n/2))
  age <- rep(age, 2)
  training <- c(rep(0, n), rep(1, n))
  b0 <- rnorm(n, 0, sigma_b0)
  b0 <- rep(b0, 2)
  
  eta <- beta0 + beta1 * age + beta2 * training + beta3 * age * training + b0
  
  phi <- exp(sigma)
  mu <- exp(eta) / (1 + exp(eta))
  
  alpha <- mu * phi
  beta <- (1 - mu) * phi
  
  y <- rbeta(n*2, alpha, beta)
  
  df <- data.frame(age, training, id = rep(1:n, 2), y = y)
  return(df)
}
```

The raw data from the simulation function produces outcome on the scale $y\in[0,1]$ based on the parameters for the linear predictor of $\eta$.

```{r}
## Test the function 




dat <- ir_sim1(n = 100, beta0 = 0, beta1 = 2, beta2 = 2, beta3 = -2, sigma = 0.1, sigma_b0 = 0.05)



dat |>
  ggplot(aes(as.factor(training), y)) + 
  geom_point(position = position_jitter(width = 0.2)) + 
  facet_wrap(~ age)


```

The `rbeta` function returns values in the interval $[0,1]$ including boudaries (0, and 1). However, modelling with the `beta_family` distribution does not allow for boundary values (0,1). We have to compress the data prior to fitting.


```{r}


compress <- function(x, e = 10^-4, additative = TRUE) {
  
  n <- length(x)
  
  if(additative) {
    xout <- x * (1 - 2*e) + e
  } else {
    xout <- (x * (n - 1) + 0.5) / n
  }
  
  xout
  
}


dat$y2 <- compress(dat$y)



library(glmmTMB)


m <- glmmTMB(y2 ~ age + training + age:training + (1|id), data = dat, 
             family = "beta_family")


summary(m)



```

The simulation function together with compression seems to return a data sets that we are able to model. In the construction of the parameters of the Beta distribution ($\alpha$ and $\beta$), the dispersion parameter $\phi$ is of importance. We will likely not get the coefficients from $\eta$ back from the model due to $\phi$. To explore the effect of $\phi$:


```{r}

sigmas <- seq(from = 0.1, to = 10, by = 0.2)

results <- list()


for(i in seq_along(sigmas)) {
  
  # Simulate data
  dat <- ir_sim1(n = 100, beta0 = 3, beta1 = -3, beta2 = -3, beta3 = 3, sigma = sigmas[i], sigma_b0 = 0.25)

  
  dat$y2 <- compress(dat$y)
  
  # Fit model
  m <- glmmTMB(y2 ~ age + training + age : training, data = dat, 
             family = "beta_family")

  results[[i]] <- data.frame(param = c("b0", "b1", "b2", "b3"), 
             coef = fixef(m)$cond,
             sigma = sigmas[i],
             row.names = NULL)
  
}

results <- bind_rows(results) 


results |>
  mutate(eta_beta = rep(c('beta[0]*" = 3"',
                          'beta[1]*" = -3"', 
                          'beta[2]*" = -3"', 
                          'beta[3]*" = 3"'), length.out = length(sigmas) * 4)) |>
  
  ggplot(aes(sigma, coef)) + geom_point() + 
  facet_wrap(~ eta_beta, labeller = label_parsed)
  



```

As $\phi$ increases, the precision increases. This means that we should think about $\sigma$ in the simulation as being specified as $1/\sigma$. A large $\sigma$ will result in small precision (and larger variation in the coefficients).


### Specifying predictions and contrasts

Using the `marginaleffects` package we can do predictions and compute contrasts. This can be useful for visualizations and inference. Below we first specify a contrast for the design we've simulated above and the apply it as afunction in a seqwrap call. 

In `marginaleffects::predictions` we can specify a data grid from where we want predictions. Using the `datagrid` function we can specify the available variables in the data set, age and training, using the unique function. The table shows the output, we have specified "response" as the scale on which we want the predictions.




```{r}
#| label: tbl-predictions
#| tbl-cap: "Output from a marginaleffects prediction."
# Load marginaleffects
library(marginaleffects)



# Simulate data and compress the observations
dat <- ir_sim1(n = 100, beta0 = 3, beta1 = -3, beta2 = -3, beta3 = 3, sigma = 3, sigma_b0 = 0.5)

dat$y2 <- compress(dat$y)

# Fit model
m <- glmmTMB(y2 ~ age + training + age : training, data = dat, 
             family = "beta_family")



# Do predictions
preds <-  marginaleffects::predictions(
  m,  # The model object
  re.form = NA, # The form of the random effects - This gives us the average per group
  type = "response", # Predict on the response scale (observed scale)
  # Use the data from the model data grid
  newdata = marginaleffects::datagrid(
    training = unique, 
    age = unique))


preds |> 
  gt() |> 
  fmt_auto() |> 
  opt_table_font(size = 10)
  
```

We can use the prediction data table to specify contrasts as hypotheses. In the `hypothesis` slot we list a key (e.g., `"train_old"`) and a contrast as differences between rows in the prediction table (e.g., the first row corresponding to `training == 0` and `age == 0` is `"b1"`). The hypothesis is `= 0` meaning that the null-hypothesis is that there are no differences. The `predictions` function will handle the SE calculation and create corresponding confidence intervals.  

```{r}
# Disable a warning message regarding specifying contrasts
options(marginaleffects_safe = FALSE)

contr <-  marginaleffects::predictions(
  m,  # The model object
  re.form = NA, # The form of the random effects - This gives us the average per group
  type = "response", # Predict on the response scale (observed scale)
  # Use the data from the model data grid
  newdata = marginaleffects::datagrid(
    training = unique, 
    age = unique), 
  hypothesis = c(
    # Training effect within old
    "train_old" = "b3 - b1 = 0",
    # Training effect within young
    "train_young" = "b4 - b2 = 0", 
    # Training effect young vs old
    "train_old_young" = "(b3 - b1) - (b4 - b2) = 0" 
  ))



```

The plot below shows the simulated data, predictions and contrasts.

```{r}
#| fig-cap: "Simulated data points (black), predictions (orange) and contrasts (blue) with 95% confidence intervals."
#| label: fig-prediction-example

# Plot 1: Predictions
p1 <- dat |>
  ggplot(aes(as.factor(training), y)) + 
  geom_point(position = position_jitter(width = 0.1)) + 
  
  
  geom_errorbar(data = data.frame(preds), 
             aes(as.factor(training), estimate, 
                 ymin = conf.low, ymax = conf.high), 
             width = 0.2) +
  
  geom_point(data = data.frame(preds), 
             aes(as.factor(training), estimate), 
             shape = 21, size = 4, fill = "orange") +
  
  facet_wrap(~ age)

# Plot 2: Contrasts
p2 <- contr |> 
  data.frame() |> 
  ggplot(aes(hypothesis, estimate)) + 
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), 
                width = 0.1) + 
  geom_point(shape = 21, fill = "steelblue", 
             size = 4)


cowplot::plot_grid(p1, p2, ncol = 1)



```

Next we can define a function to get the marginal predictions out from a seqwrap model object. To showcase this, first we need a simulated data set and metadat.


```{r}
# A single data set with random combinations of parameter values can be simulated as
dat <- ir_sim1(n = 80, beta0 = rnorm(1,0,2), 
               beta1 = rnorm(1,0,3), beta2 = rnorm(1,0,3), 
               beta3 = rnorm(1,0,1), sigma = rnorm(1,0,3), 
               sigma_b0 = abs(rnorm(1,0.4, 0.2)))

# We can wrap it in a function ...
sim_wrap <- function() {
  dat <- ir_sim1(n = 80, beta0 = rnorm(1,0,2), beta1 = rnorm(1,0,3), beta2 = rnorm(1,0,3), beta3 = rnorm(1,0,1), sigma = rnorm(1,0,3), sigma_b0 = abs(rnorm(1,0.4, 0.2)))
  
  dat$y2 <- compress(dat$y)
  
  return(dat)
}

# And simulate musltiple data sets
# Return N simulations and collect them in a data frame
sims <- replicate(100, sim_wrap(), simplify = FALSE)
sims_df <- do.call(rbind, lapply(sims, function(df) t(df[, 5, drop = FALSE]))) |> 
  data.frame(row.names = NULL)

# Set seq_sample_id (reuse dat from above)
colnames(sims_df) <- paste0("s", dat$id, "_", dat$training) 

sims_df$target <- paste0("target", 1:100)

sims_df <- sims_df |> 
  dplyr::select(target, starts_with("s")) 

metadata <- dat |> 
  mutate(seq_sample_id  = paste0("s", id, "_", training) ) |> 
  dplyr::select(id, age, training, seq_sample_id)


```


We now specify the marginaleffects prediction-contrast-function as a summary function. Notice that I'm adding predictions to the hypothesis slot. **Importantly** when using the marginaleffects package in seqwrap we need to specify in the function that warnings should be disabled. 

```{r}

sum_fun <- function(m) {
  
  # Disable a warning message regarding specifying contrasts
  options(marginaleffects_safe = FALSE)
  
  
  contr <-  marginaleffects::predictions(
  m,  # The model object
  re.form = NA, # The form of the random effects - This gives us the average per group
  type = "response", # Predict on the response scale (observed scale)
  # Use the data from the model data grid
  newdata = marginaleffects::datagrid(
    training = unique, 
    age = unique), 
  hypothesis = c(
    "old_baseline" = "b1 = 0",
    "young_baseline" = "b2 = 0",
    "old_post" = "b3 = 0",
    "young_post" = "b4 = 0",
    
    # Training effect within old
    "train_old" = "b3 - b1 = 0",
    # Training effect within young
    "train_young" = "b4 - b2 = 0", 
    # Training effect young vs old
    "train_old_young" = "(b3 - b1) - (b4 - b2) = 0" 
  )) |> 
    data.frame()
  
  return(contr)
  
}


```

The seqwrap setup is constructed as in the vignette


```{r}
library(seqwrap)


m1 <- seqwrap_compose(
  data = sims_df, 
  metadata = metadata, 
  modelfun = glmmTMB::glmmTMB, 
  arguments = alist(
    formula = y ~ age * training + (1|id), 
    family = glmmTMB::beta_family
  ), 
  summary_fun = sum_fun)

m1r <- seqwrap(m1, 
               cores = 10, 
               return_models = FALSE)

m1s <- seqwrap_summarise(m1r)

```


Now we have a seqwrap summary with contrasts and predictions specified using marginaleffects. These can of course be plotted with the outcome on the response scale.


```{r}

# Plotting targets with significant interaction effect
targets <- m1s$summaries |> 
  filter(hypothesis == "train_old_young" & p.value < 0.01) |> 
  pull(target)
  
m1s$summaries |> 
  filter(target %in% targets) |> 
  filter(grepl("baseline|post", hypothesis)) |> 
  separate(hypothesis, into = c("age", "time")) |> 
  ggplot(aes(estimate, target, fill = age, shape = time)) + geom_point() + 
  scale_shape_manual(values = c(21, 24))
  
```




## Multilevel Beta in a meta-analysis context

How can we interpret coefficients in the multilevel version of the Beta model?

Interpretation of our data will benefit from interpretation using a multilevel model. Data comes from a number of different studies, with multiple samples from participants in each study. 


```{r}

# beta0, global intercept
# beta1, global slope for age
# beta2, global training effect
# sigma0/sigma1/sigma2, SD of global effects between studies
# sigma_participant, SD of between participant intercepts
# sigma, "residual" variation should be entered as precision 1/sigma
# age/training/study/participants, design variables
#
#
# returns a vector of simulated IR values based on the design variables.
ir_sim2 <- function(beta0 = 3, 
                    beta1 = -3, 
                    beta2 = 3, 
                    sigma0 = 1, 
                    sigma1 = 1, 
                    sigma2 = 1, 
                    sigma = 1/0.5, 
                    sigma_participant = 1,
                    age = c(1, 2, 1, 2), 
                    training = c(0,0,0,0), 
                    study = c(1, 2, 3, 4), 
                    participant = c(1, 2, 4, 5)) {
  
  # Save n observations
  n <- length(participant)
  # Check that all design variables are of equal length
  if(length(unique(lengths(list(participant, study, training, age)))) != 1) stop("All design variables (age, training, study, participant) must be of the same length.") 
   
  
  # List all studies
  studies <- unique(study)
  
  ## Draw intercepts per study
  beta0s <- rnorm(length(studies), beta0, sigma0)
  
  ## Draw age-slopes per study
  beta1s <- rnorm(length(studies), beta1, sigma1)
  
  ## Draw training slopes per study
  beta2s <- rnorm(length(studies), beta2, sigma2)
  
  ## List all participants
  participants <- unique(paste0(participant, "_", study))
  
  ## Draw intercept per participant
  b0s <- rnorm(length(participants), 0, sigma_participant) 
  
  # Combine all drawn values
  
  partdat <- data.frame(participant_study = participants, 
                     b0 = b0s) |>
    tidyr::separate(participant_study, 
                    into = c("participant", "study"), 
                    sep = "_")
  
  studydat <- data.frame(study = as.character(studies), 
                      beta0 = beta0s, 
                      beta1 = beta1s, 
                      beta2 = beta2s)
  
  eta <- partdat |>
    dplyr::inner_join(studydat, by = "study") |>
    dplyr::inner_join(data.frame(age, 
                                 training, 
                                 study = as.character(study), 
                                 participant = as.character(participant))) |>
    dplyr::mutate(eta = beta0 + beta1 * age + beta2 * training + b0) |>
    dplyr::pull(eta)
  
  phi <- exp(sigma)
  mu <- exp(eta) / (1 + exp(eta))
  
  alpha <- mu * phi
  beta <- (1 - mu) * phi
  
  y <- rbeta(n, alpha, beta)
  
  return(y)
  
}


```

### Multilevel Beta regression with simulated data


```{r}

design <- rbind(
  # Study 1
  data.frame(study = 1, 
             participant = rep(1:20, each = 2), 
             age = rep(runif(20, 0.5, 1), each = 2), 
             training = c(0,1)), 
  # Study 2
   data.frame(study = 2, 
             participant = rep(1:20, each = 2), 
             age = rep(runif(20, 0.1, 0.6), each = 2), 
             training = c(0,1)), 
  # Study 3
  data.frame(study = 3, 
             participant = rep(1:20, each = 2), 
             age = rep(runif(20, 0.1, 0.9), each = 2), 
             training = c(0,1)), 
  # Study 4
    data.frame(study = 4, 
             participant = rep(1:20, each = 2), 
             age = rep(runif(20, 0.01, 0.3), each = 2), 
             training = c(0,1)), 
  # Study 5
     data.frame(study = 5, 
             participant = rep(1:20, each = 2), 
             age = rep(runif(20, 0.01, 0.99), each = 2), 
             training = c(0,1))
  
)


```


Next we simulate (and compress) the data.

```{r}

y <- ir_sim2(beta0 = 3, beta1 = -1.5, beta2 = 2, 
             sigma0 = 2, sigma1 = 1, sigma2 = 2,
        sigma = 1/0.5, 
        sigma_participant = 0.66, 
        age = design$age, 
        training = design$training, 
        participant = design$participant, 
        study = design$study)


dat1 <- design |>
  mutate(participant = paste0(study, "_", participant))

dat1$y <- compress(y)


```


```{r}

dat1 |>
  ggplot(aes(age, y, 
             color = as.factor(study))) + 
  geom_smooth(se = FALSE) +
  geom_point()


```


```{r}

m <- glmmTMB(y ~ age * training + 
               (1|study) + (1|participant), 
             data = dat1, 
             family = beta_family() )


summary(m)
VarCorr(m)

```





















