---
title: "Muscle strength data visualization"
format: html
editor_options: 
  chunk_output_type: console
---


```{r, setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

```

## Purpose

The purpose is to inspect and process the muscle strength data from the "ReLiEf" project. The data consists of peak torque and power measurements from the humac norm dynamometer (60, 120, 240 d/s + isometric), 1RM legpress (possibly power) from keiser air press and maximal bicep isometric force from the MuscleLab.


```{r, data import}

library(tidyverse)
library(dplyr)
library(cowplot)

# Importing the data from the data package
humac.dat <- reliefdata::relief_humac
keiser.dat <- reliefdata::relief_1rm
bicep.dat <- reliefdata::relief_isobc
condition <- reliefdata::relief_volume %>%
  mutate(participant = as.character(participant))
participants <- reliefdata::relief_participants %>%
  mutate(age_group = if_else(age < 40, "yng", "old"))

################ HUMAC DATA #########################
# Joining condition and participant info
humac.dat <- humac.dat %>%
  full_join(condition) %>%
  select(-arm) %>% 
  full_join(participants) %>%
  mutate(age_group = factor(age_group, levels = c("yng", "old")),
         condition = factor(condition, levels = c("low", "mod")),
         time = factor(time, levels = c("pre", "mid",  "post")),
         speed = factor(speed, levels = c("0", "60", "120", "240")),
         allocation = factor(allocation, levels = c("int", "con")),
         condition = case_when(participant %in% c("3071", "3075", "3076",
                                                  "3083", "3084", "3085",
                                                  "3086", "3090", "3091",
                                                  "3094") ~ "none", TRUE ~ condition))


# Create a safe max function (due to NA's that na.rm = TRUE wont fix). 
# This creates a function the returns proper NA values where all values are NA for a group
safe_max <- function(x) {
  if (all(is.na(x))) return(NA_real_)
  max(x, na.rm = TRUE)
}

safe_min <- function(x) {
  if (all(is.na(x))) return(NA_real_)
  min(x, na.rm = TRUE)
}


#Step-by-step explanation:
#1. function(x) - This creates a function that takes one input called x (which will be your vector of values)
#2. if (all(is.na(x))) - This checks the condition:

#is.na(x) checks each element and returns TRUE/FALSE for each
#all() checks if ALL elements are TRUE
#So this asks: "Are ALL values in x missing (NA)?"

#3. return(NA_real_) - If all values are NA:

#Return NA_real_ (which is specifically a numeric NA, as opposed to NA_character_ or NA_integer_)
#return() immediately exits the function

#4. max(x, na.rm = TRUE) - If NOT all values are NA:

#Calculate the maximum, ignoring any NAs that exist
#This line only runs if the if condition is FALSE

# Get max values for all outcomes

max.dat <- humac.dat %>%
  filter(!participant == "3012") %>% # Missing data
  group_by(participant, time, leg, speed, condition, age_group, allocation, sex) %>%
  summarise(pt_max = safe_max(pt),
            power_max = safe_max(rep_power),
            work_max = safe_max(rep_work),
            tt_min = safe_min(pt_tt),
            angle = mean(pt_angle, na.rn = TRUE),
            .groups = "drop")


# Reshape the data to long format for faceting

long.dat <- max.dat %>%
  pivot_longer(names_to = "outcome",
               values_to = "value",
               cols = c(pt_max, power_max, work_max, tt_min, angle))

## Filter by speed and parameter

# Isometric

ptisom <- long.dat %>%
  filter(speed == "0", outcome == "pt_max")

#pwisom <- long.dat %>%
#  filter(speed == "0", outcome == "power_max") No data here


# 60 d/s

pt60 <- long.dat %>%
  filter(speed == "60", outcome == "pt_max")

pw60 <- long.dat %>%
  filter(speed == "60", outcome == "power_max")

  
# 120 d/s

pt120 <- long.dat %>%
  filter(speed == "120", outcome == "pt_max")

pw120 <- long.dat %>%
  filter(speed == "120", outcome == "power_max")


# Isometric

pt240 <- long.dat %>%
  filter(speed == "240", outcome == "pt_max")

pw240 <- long.dat %>%
  filter(speed == "240", outcome == "power_max")


################# KEISER DATA ###############

# Merging keiser and condition + cleaning up

keiser.clean <- keiser.dat %>%
  select(participant, timep, leg, maxrm, allocation, age_group = age.grp, sex) %>%
  filter(!timep == "famil",
         !is.na(timep)) %>%
  full_join(condition) %>%
  select(-arm) %>%
  filter(!participant %in% c("3004", "3005", "3010", "3051", "3045", "3050")) %>% # Missing data, error in sampling for 3045 and 3050
  # Create a clean timepoint variable
  mutate(time_clean = case_when(
    timep == "pre" ~ "pre",
    timep %in% c("mid1", "mid2") ~ "mid",
    timep %in% c("post1", "post2") ~ "post",
    TRUE ~ timep
  )) %>%
  # Take the max for each participant, leg, and clean timepoint
  group_by(participant, leg, allocation, age_group, sex, condition, time_clean) %>%
  summarise(maxrm = max(maxrm, na.rm = TRUE), .groups = "drop") %>%
  # Rename back to timep
  rename(timep = time_clean) %>%

  mutate(age_group = factor(age_group, levels = c("yng", "old")),
         condition = factor(condition, levels = c("low", "mod")),
         timep = factor(timep, levels = c("pre", "mid",  "post")),
         allocation = factor(allocation, levels = c("int", "con")),
         condition = case_when(participant %in% c("3071", "3075", "3076",
                                                  "3083", "3084", "3085",
                                                  "3086", "3090", "3091",
                                                  "3094") ~ "none", TRUE ~ condition))


################## ISO Bicep DATA ####################

bicep.clean <- bicep.dat %>%
  select(participant, timep, arm, mf1, mf2, mf3, allocation, age_grp = age.grp, sex) %>%
  filter(!timep == "famil",
         !is.na(timep)) %>% 
  full_join(condition) %>%
  filter(!participant %in% c("3004", "3005", "3017", "3089")) %>% # 3017 too few RT sessions or dropout
  select(-leg) %>%
  mutate(
    mf = pmax(mf1, mf2, mf3, na.rm = TRUE)) %>% # taking highest observation per test
  select(-mf1, mf2, mf3) %>%
  # Create a clean timepoint variable
  mutate(time_clean = case_when(
    timep == "pre" ~ "pre",
    timep %in% c("mid1", "mid2") ~ "mid",
    timep %in% c("post1", "post2") ~ "post",
    TRUE ~ timep
  )) %>%
  # Take the max for each participant, leg, and clean timepoint
  group_by(participant, arm, allocation, age_grp, sex, condition, time_clean) %>%
  summarise(maxmf = max(mf, na.rm = TRUE), .groups = "drop") %>%
  # Rename back to timep
  rename(timep = time_clean) %>%
  mutate(age_grp = factor(age_grp, levels = c("yng", "old")),
         condition = factor(condition, levels = c("low", "mod")),
         timep = factor(timep, levels = c("pre", "mid",  "post")),
         allocation = factor(allocation, levels = c("int", "con")),
         condition = case_when(participant %in% c("3071", "3075", "3076",
                                                  "3083", "3084", "3085",
                                                  "3086", "3090", "3091",
                                                  "3094") ~ "none", TRUE ~ condition))



```


<!-- ### Individual Changes in Peak Torque at 0, 60, 120 and 240 d/s -->

```{r, individual change peak torque plot}

library(grid)
library(gridExtra)

## Peak torque Isometric

ptisom.plot <- ptisom %>%
  ggplot(aes(x = time, y = value, group = interaction(participant, leg))) +
  geom_line(aes(color = condition), alpha = 0.3) +
  geom_point(aes(color = condition), alpha = 0.3, size = 1) +
  facet_wrap(vars(age_group, allocation), scales = "fixed") +
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +
  labs(title = "",
       subtitle = "",
       x = "",
       y = "Peak Torque (Nm)") +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"),
        axis.line.x = element_blank(),
        axis.text.x = element_blank(),
        strip.text = element_blank())

## Peak torque 60

pt60.plot <- pt60 %>%
  ggplot(aes(x = time, y = value, group = interaction(participant, leg))) +
  geom_line(aes(color = condition), alpha = 0.3) +
  geom_point(aes(color = condition), alpha = 0.3, size = 1) +
  facet_wrap(vars(age_group, allocation), scales = "fixed") +
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +
  labs(title = "",
       subtitle = "",
       x = "",
       y = "") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"),
        axis.line.x = element_blank(),
        axis.text.x = element_blank())

## Peak torque 120

pt120.plot <- pt120 %>%
  ggplot(aes(x = time, y = value, group = interaction(participant, leg))) +
  geom_line(aes(color = condition), alpha = 0.3) +
  geom_point(aes(color = condition), alpha = 0.3, size = 1) +
  facet_wrap(vars(age_group, allocation), scales = "fixed") +
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +
  labs(title = "",
       subtitle = "",
       x = "Time",
       y = "Peak Torque (Nm)") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"))

## Peak torque 240

pt240.plot <- pt240 %>%
  ggplot(aes(x = time, y = value, group = interaction(participant, leg))) +
  geom_line(aes(color = condition), alpha = 0.3) +
  geom_point(aes(color = condition), alpha = 0.3, size = 1) +
  facet_wrap(vars(age_group, allocation), scales = "fixed") +
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +
  labs(title = "",
       subtitle = "", #Each line represents one leg from one participant
       x = "Time",
       y = "") +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.box.margin = margin(0, 0, 0, 0)) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"))




# Combinding the plots

ptfig <- plot_grid(
  ptisom.plot + theme(legend.position = "none"),
  pt60.plot + theme(legend.position = "none"),
  pt120.plot,
  pt240.plot,
  ncol = 2,
  nrow = 2,
  labels = c("A) Isometric", "B) 60 d/s", "C) 120 d/s", "D) 240 d/s"),
  rel_heights = c(0.8, 1) 
)


#ptfig



```

<!-- ### Mean Change in Peak Torque at 0, 60, 120 and 240 d/s -->

```{r, mean change peak torque plot}


# Isometric 

ptisommean <- ptisom %>%
  #summarising means
  group_by(age_group, allocation, condition, time) %>%
  summarise(mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            .groups = "drop" ) %>%
  
  #creating the plot
  ggplot(aes(x = time, y = mean, color = condition, group = condition)) +
  
  geom_line(linewidth = 1, position = position_dodge(.2)) +
  geom_point(size = 3, position = position_dodge(.2)) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                width = 0.1, linewidth = 0.8, position = position_dodge(.2)) +
  
  facet_wrap(vars(age_group, allocation), scales = "fixed") +
  
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +
  
  labs(title = "",
       subtitle = "",
       x = "",
       y = "Peak Torque (Nm)") +
  
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"),
        axis.text.x = element_blank(),
        axis.line.x = element_blank())


# 60 d/s
pt60mean <- pt60 %>%
  #summarising means
  group_by(age_group, allocation, condition, time) %>%
  summarise(mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            .groups = "drop" ) %>%
  
  #creating the plot
  ggplot(aes(x = time, y = mean, color = condition, group = condition)) +
  
  geom_line(linewidth = 1, position = position_dodge(.2)) +
  geom_point(size = 3, position = position_dodge(.2)) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                width = 0.1, linewidth = 0.8, position = position_dodge(.2)) +
  
  facet_wrap(vars(age_group, allocation), scales = "fixed") +
  
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +
  
  labs(title = "",
       subtitle = "",
       x = "",
       y = "") +
  
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"),
        axis.text.x = element_blank(),
        axis.line.x = element_blank())


# 120 d/s
pt120mean <- pt120 %>%
  #summarising means
  group_by(age_group, allocation, condition, time) %>%
  summarise(mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            .groups = "drop" ) %>%
  
  #creating the plot
  ggplot(aes(x = time, y = mean, color = condition, group = condition)) +
  
  geom_line(linewidth = 1, position = position_dodge(.2)) +
  geom_point(size = 3, position = position_dodge(.2)) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                width = 0.1, linewidth = 0.8, position = position_dodge(.2)) +
  
  facet_wrap(vars(age_group, allocation), scales = "fixed") +
  
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +
  
  labs(title = "",
       subtitle = "",
       x = "Time",
       y = "Peak Torque (Nm)") +
  
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"))


# 120 d/s
pt240mean <- pt240 %>%
  #summarising means
  group_by(age_group, allocation, condition, time) %>%
  summarise(mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            .groups = "drop" ) %>%
  
  #creating the plot
  ggplot(aes(x = time, y = mean, color = condition, group = condition)) +
  
  geom_line(linewidth = 1, position = position_dodge(.2)) +
  geom_point(size = 3, position = position_dodge(.2)) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                width = 0.1, linewidth = 0.8, position = position_dodge(.2)) +
  
  facet_wrap(vars(age_group, allocation), scales = "fixed") +
  
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +
  
  labs(title = "",
       subtitle = "",
       x = "Time",
       y = "") +
  
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"))


# Combinding the plots

ptmeanfig <- plot_grid(
  ptisommean + theme(legend.position = "none"),
  pt60mean + theme(legend.position = "none"),
  pt120mean,
  pt240mean,
  ncol = 2,
  nrow = 2,
  labels = c("A) Isometric", "B) 60 d/s", "C) 120 d/s", "D) 240 d/s"),
  rel_heights = c(0.8, 1) 
)


#ptmeanfig



```


<!-- ### Mean Peak Torque Change with Individual Raw Data -->

```{r, Combined change peak torque plot}

# Set seed for jitter

set.seed(1)

# Isometric 

ptisommean <- ptisom %>%
  #summarising means
  group_by(age_group, allocation, condition, time) %>%
  summarise(mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            .groups = "drop" ) %>%
  
  #creating the plot
  ggplot(aes(x = time, y = mean, color = condition, group = condition)) +
  
  geom_line(linewidth = 1, position = position_dodge(.5)) +
  geom_point(shape = 21, aes(fill = condition), color = "black",
             size = 3, stroke = 1, position = position_dodge(.5)) +
#  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
#                width = 0.1, linewidth = 0.8, position = position_dodge(.2)) +

  # adding in the raw data
  geom_point(data = ptisom,
             aes(x = time, y = value,
                 color = condition, group = condition),
                 size = 1.5,
                 alpha = 0.3,
             position = position_jitter(width = 0.2)) +
  
  facet_wrap(vars(age_group, allocation), scales = "fixed") +
  
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +  # <-- hide the color legend
  
  scale_fill_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                    guide = "none") +  # <-- add fill scale
  
  labs(title = "",
       subtitle = "",
       x = "",
       y = "Peak Torque (Nm)") +
  
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"),
        axis.text.x = element_blank(),
        axis.line.x = element_blank())


# 60 d/s
pt60mean <- pt60 %>%
  #summarising means
  group_by(age_group, allocation, condition, time) %>%
  summarise(mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            .groups = "drop" ) %>%
  
  #creating the plot
  ggplot(aes(x = time, y = mean, color = condition, group = condition)) +
  
  geom_line(linewidth = 1, position = position_dodge(.5)) +
  geom_point(shape = 21, aes(fill = condition), color = "black",
             size = 3, stroke = 1, position = position_dodge(.5)) +
 # geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
#                width = 0.1, linewidth = 0.8, position = position_dodge(.2)) +
  
  # adding in the raw data
  geom_point(data = pt60,
             aes(x = time, y = value,
                 color = condition, group = condition),
                 size = 1.5,
                 alpha = 0.3,
             position = position_jitter(width = 0.2)) +
  
  facet_wrap(vars(age_group, allocation), scales = "fixed") +
  
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +  # <-- hide the color legend
  
  scale_fill_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                    guide = "none") +  # <-- add fill scale
  
  labs(title = "",
       subtitle = "",
       x = "",
       y = "") +
  
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"),
        axis.text.x = element_blank(),
        axis.line.x = element_blank())


# 120 d/s
pt120mean <- pt120 %>%
  #summarising means
  group_by(age_group, allocation, condition, time) %>%
  summarise(mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            .groups = "drop" ) %>%
  
  #creating the plot
  ggplot(aes(x = time, y = mean, color = condition, group = condition)) +
  
  geom_line(linewidth = 1, position = position_dodge(.5)) +
  geom_point(shape = 21, aes(fill = condition), color = "black",
             size = 3, stroke = 1, position = position_dodge(.5)) +
 # geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
  #              width = 0.1, linewidth = 0.8, position = position_dodge(.2)) +
  
  # adding in the raw data
  geom_point(data = pt120,
             aes(x = time, y = value,
                 color = condition, group = condition),
                 size = 1.5,
                 alpha = 0.3,
             position = position_jitter(width = 0.2)) +
  
  facet_wrap(vars(age_group, allocation), scales = "fixed") +
  
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +  # <-- hide the color legend
  
  scale_fill_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                    guide = "none") +  # <-- add fill scale
  
  labs(title = "",
       subtitle = "",
       x = "Time",
       y = "Peak Torque (Nm)") +
  
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"))


# 240 d/s
pt240mean <- pt240 %>%
  #summarising means
  group_by(age_group, allocation, condition, time) %>%
  summarise(mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            .groups = "drop" ) %>%
  
  #creating the plot
  ggplot(aes(x = time, y = mean, color = condition, group = condition)) +
  
  geom_line(linewidth = 1, position = position_dodge(.5)) +
  geom_point(shape = 21, aes(fill = condition), color = "black",
             size = 3, stroke = 1, position = position_dodge(.5)) +
  #geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
   #             width = 0.1, linewidth = 0.8, position = position_dodge(.2)) +
  
  # adding in the raw data
  geom_point(data = pt240,
             aes(x = time, y = value,
                 color = condition, group = condition),
                 size = 1.5,
                 alpha = 0.3,
             position = position_jitter(width = 0.2)) +
  
  facet_wrap(vars(age_group, allocation), scales = "fixed") +
  
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +  # <-- hide the color legend
  
  scale_fill_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                    guide = "none") +  # <-- add fill scale
  
  labs(title = "",
       subtitle = "",
       x = "Time",
       y = "") +
  
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"))



# Combinding the plots

ptmeanfig <- plot_grid(
  ptisommean + theme(legend.position = "none"),
  pt60mean + theme(legend.position = "none"),
  pt120mean + theme(legend.position = "none"),
  pt240mean + theme(legend.position = "none"),
  ncol = 2,
  nrow = 2,
  labels = c("A) Isometric", "B) 60 d/s", "C) 120 d/s", "D) 240 d/s"),
  label_size = 8
)





```


Raw data (small, less visible dots) included to show the spread.

### Changes in Peak Torque at Different Angular Velocities

```{r}

# Calculating group means
curve.sum <- long.dat %>%
  filter(outcome == "pt_max") %>%
  group_by(age_group, allocation, time, speed) %>%
  summarise(mean_torque = mean(value, na.rm = TRUE),
            se_torque = sd(value, na.rm = TRUE) / sqrt(n()),
            .groups = "drop")

# Creating a combined group variable for easier plotting
curve.sum <- curve.sum %>%
  mutate(group = paste(age_group, allocation, sep = "_"))


# Plot 1: Comparing changes over time (baseline + halfway + post)
# Separate panels for each group

plot_time <- curve.sum %>%
  ggplot(aes(x = speed, y = mean_torque, color = time, group = time)) +
  geom_line(linewidth = 1) +
  #geom_point(size = 3) +
  #geom_errorbar(aes(ymin = mean_torque - se_torque, 
  #                  ymax = mean_torque + se_torque), 
  #              width = 10) +
  facet_wrap(~ age_group + allocation, ncol = 3) +
  labs(x = "Angular Velocity (d/s)",
       y = "Peak Torque (Nm)",
       color = "Time Point",
       title = "Force-Velocity Curves Across Training") +
  theme_classic() +
  theme(legend.position = "bottom")


# Plot 2: Single plot showing everything (might be busy)
plot_all <- curve.sum %>%
  ggplot(aes(x = speed, y = mean_torque, color = group, linetype = time, 
             group = interaction(group, time))) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(x = "Angular Velocity (d/s)",
       y = "Peak Torque (Nm)",
       color = "Group",
       linetype = "Time Point",
       title = "") +
  scale_color_manual(values = c("yng_int" = "#E69F00",
                                  "old_int" = "#56B4E9",
                                  "old_con" = "gray50"),
                     labels = c("yng_int" = "Young Intervention", 
                                "old_int" = "Old Intervention", 
                                "old_con" = "Old Control")) +
  theme_classic() +
  theme(legend.position = "bottom")


plot_all




```



<!-- ### Mean Change in Power -->

```{r, Combined change in power plot}

# Set seed for jitter

set.seed(1)

# 60 d/s
pw60mean <- pw60 %>%
  #summarising means
  group_by(age_group, allocation, condition, time) %>%
  summarise(mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            .groups = "drop" ) %>%

  #creating the plot
  ggplot(aes(x = time, y = mean, color = condition, group = condition)) +
  
  geom_line(linewidth = 1, position = position_dodge(.5)) +
  geom_point(shape = 21, aes(fill = condition), color = "black",
             size = 3, stroke = 1, position = position_dodge(.5)) +
 # geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
#                width = 0.1, linewidth = 0.8, position = position_dodge(.2)) +
  
  # adding in the raw data
  geom_point(data = pw60,
             aes(x = time, y = value,
                 color = condition, group = condition),
                 size = 1.5,
                 alpha = 0.3,
             position = position_jitter(width = 0.2)) +
  
  facet_wrap(vars(age_group, allocation), scales = "fixed") +
  
  
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +  # <-- hide the color legend
  
  scale_fill_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                    guide = "none") +  # <-- add fill scale
  
  labs(title = "",
       subtitle = "",
       x = "",
       y = "") +
  
  theme_minimal() +
  theme(legend.position = "right") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"),
        axis.text.x = element_blank(),
        axis.line.x = element_blank())


# 120 d/s
pw120mean <- pw120 %>%
  #summarising means
  group_by(age_group, allocation, condition, time) %>%
  summarise(mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            .groups = "drop" ) %>%
  
  #creating the plot
  ggplot(aes(x = time, y = mean, color = condition, group = condition)) +
  
  geom_line(linewidth = 1, position = position_dodge(.5)) +
  geom_point(shape = 21, aes(fill = condition), color = "black",
             size = 3, stroke = 1, position = position_dodge(.5)) +
 # geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
  #              width = 0.1, linewidth = 0.8, position = position_dodge(.2)) +
  
  # adding in the raw data
  geom_point(data = pw120,
             aes(x = time, y = value,
                 color = condition, group = condition),
                 size = 1.5,
                 alpha = 0.3,
             position = position_jitter(width = 0.2)) +
  
  facet_wrap(vars(age_group, allocation), scales = "fixed") +
  
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +  # <-- hide the color legend
  
  scale_fill_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                    guide = "none") +  # <-- add fill scale
  
  labs(title = "",
       subtitle = "",
       x = "Time",
       y = "Peak Torque (Nm)") +
  
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"))


# 240 d/s
pw240mean <- pw240 %>%
  #summarising means
  group_by(age_group, allocation, condition, time) %>%
  summarise(mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            .groups = "drop" ) %>%
  
  #creating the plot
  ggplot(aes(x = time, y = mean, color = condition, group = condition)) +
  
  geom_line(linewidth = 1, position = position_dodge(.5)) +
  geom_point(shape = 21, aes(fill = condition), color = "black",
             size = 3, stroke = 1, position = position_dodge(.5)) +
  #geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
   #             width = 0.1, linewidth = 0.8, position = position_dodge(.2)) +
  
  # adding in the raw data
  geom_point(data = pw240,
             aes(x = time, y = value,
                 color = condition, group = condition),
                 size = 1.5,
                 alpha = 0.3,
             position = position_jitter(width = 0.2)) +
  
  facet_wrap(vars(age_group, allocation), scales = "fixed") +
  
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +  # <-- hide the color legend
  
  scale_fill_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                    guide = "none") +  # <-- add fill scale
  
  labs(title = "",
       subtitle = "",
       x = "Time",
       y = "") +
  
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"))



# Combinding the plots

upperfig <- plot_grid(pw60mean,
                      NULL,
                      ncol = 2,
                      rel_widths = c(1.4, 0.6),
                      labels = c("A) 60 d/s", ""),
                      label_size = 8)


lowerfig <- plot_grid(
  pw120mean + theme(legend.position = "none"),
  pw240mean + theme(legend.position = "none"),
  ncol = 2,
  labels = c("B) 120 d/s", "C) 240 d/s"),
  label_size = 8
  )

pwmeanfig <- plot_grid(
  upperfig,
  lowerfig,
  nrow = 2
)





```

<!-- Raw data (small, less visible dots) included to show the spread. -->

<!--
### Individual Change 1RM (Keiser) -->

```{r, Individual change 1RM}


rm.plot <- keiser.clean %>%
  ggplot(aes(x = timep, y = maxrm, group = interaction(participant, leg))) +
  geom_line(aes(color = condition), alpha = 0.3) +
  geom_point(aes(color = condition), alpha = 0.3, size = 1) +
  #geom_text(aes(label = paste(participant, leg)), size = 2, nudge_y = 5) +
  facet_wrap(vars(age_group, allocation), scales = "fixed") +
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +
  labs(title = "",
       subtitle = "",
       x = "Time",
       y = "1RM Keiser Airpress") +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"))

```


### Mean Change 1RM Keiser

```{r, Mean change 1RM}

meanrm.plot <- keiser.clean %>%
  #summarising means
  group_by(age_group, allocation, condition, timep) %>%
  summarise(mean = mean(maxrm, na.rm = TRUE),
            .groups = "drop" ) %>%

  #creating the plot
  ggplot(aes(x = timep, y = mean, color = condition, group = condition)) +
  
  geom_line(linewidth = 1, position = position_dodge(.5)) +
  geom_point(shape = 21, aes(fill = condition), color = "black",
             size = 3, stroke = 1, position = position_dodge(.5)) +
 # geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
#                width = 0.1, linewidth = 0.8, position = position_dodge(.2)) +
  
  # adding in the raw data
  geom_point(data = keiser.clean,
             aes(x = timep, y = maxrm,
                 color = condition, group = condition),
                 size = 1.5,
                 alpha = 0.3,
             position = position_jitter(width = 0.2)) +
  
  facet_wrap(vars(age_group, allocation), scales = "fixed") +
  
  
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +  # <-- hide the color legend
  
  scale_fill_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                    guide = "none") +  # <-- add fill scale
  
  labs(title = "",
       subtitle = "",
       x = "Time",
       y = "1RM Keiser Airpress") +
  
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"))

meanrm.plot

```

Raw data (small, less visible dots) included to show the spread.

### Mean Change Isometric Muscle Force Bicep Curl

```{r, Mean change bicep}

bicep.plot <- bicep.clean %>%
  ggplot(aes(x = timep, y = maxmf, group = interaction(participant, arm))) +
  geom_line(aes(color = condition), alpha = 0.3) +
  geom_point(aes(color = condition), alpha = 0.3, size = 1) +
  #geom_text(aes(label = paste(participant, arm)), size = 2, nudge_y = 5) +
  facet_wrap(vars(age_grp, allocation), scales = "fixed") +
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +
  labs(title = "",
       subtitle = "",
       x = "",
       y = "Peak Torque (Nm)") +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"),
        axis.line.x = element_blank(),
        axis.text.x = element_blank(),
        strip.text = element_blank())



meanbicep.plot <- bicep.clean %>%
  #summarising means
  group_by(age_grp, allocation, condition, timep) %>%
  summarise(mean = mean(maxmf, na.rm = TRUE),
            .groups = "drop" ) %>%

  #creating the plot
  ggplot(aes(x = timep, y = mean, color = condition, group = condition)) +
  
  geom_line(linewidth = 1, position = position_dodge(.5)) +
  geom_point(shape = 21, aes(fill = condition), color = "black",
             size = 3, stroke = 1, position = position_dodge(.5)) +
 # geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
#                width = 0.1, linewidth = 0.8, position = position_dodge(.2)) +
  
  # adding in the raw data
  geom_point(data = bicep.clean,
             aes(x = timep, y = maxmf,
                 color = condition, group = condition),
                 size = 1.5,
                 alpha = 0.3,
             position = position_jitter(width = 0.2)) +
  
  facet_wrap(vars(age_grp, allocation), scales = "fixed") +
  
  
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +  # <-- hide the color legend
  
  scale_fill_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                    guide = "none") +  # <-- add fill scale
  
  labs(title = "",
       subtitle = "",
       x = "Time",
       y = "Isometric Force (N)") +
  
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"))

meanbicep.plot

```

Raw data (small, less visible dots) included to show the spread.


## Questions
1. Same as muscle mass, make an index for lower body, separate comparison for upper body
2. Possibly compare lower and upper body

<!-- Klar til statistikk, metode beskrevet -->

