---
title: "Muscle mass data visualization"
format: html
editor_options: 
  chunk_output_type: console
---

```{r, setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

```

## Purpose

The purpose is to inspect and process the muscle mass data from the "ReLiEf" project. The data (muscle volume from MRI scans, muscle thickness from ultrasound) will be imported from the "reliefdata" package. Trying some different plots to look at the data, both muscle volume from MRI and muscle thickness from ultrasound.

```{r, data import and clean up}

library(tidyverse)
library(ggrepel)

mr.dat <- reliefdata::relief_mr %>% 
  mutate(participant = as.character(participant))
us.dat <- reliefdata::relief_thickness
condition <- reliefdata::relief_volume %>%
  mutate(participant = as.character(participant))
participants <- reliefdata::relief_participants %>%
  mutate(age_group = if_else(age < 40, "yng", "old"))

mr.dat <- mr.dat %>%
  full_join(condition) %>%
  select(-arm) %>% 
  full_join(participants)

us.dat <- us.dat %>%
  full_join(condition) %>%
  select(-arm) %>%
  full_join(participants)

# Check data structure
#str(mr.dat)
#head(mr.dat)
#summary(mr.dat)

#str(us.dat)
#head(us.dat)
#summary(us.dat)

## Check for missing values
# need to filter out the measurements marked as "both" condition, seems like 
# the masters students didnt make those measurements


mr.clean <- mr.dat %>%
  filter(!leg == "both",
         !participant %in% c("3070", "3092", "3093", "3069", #These are all missing 
                             "3068", "3056", "3053", "3046", #measurements
                             "3045", "3035", "3034", "3033", 
                             "3028", "3026", "3025", "3022", 
                             "3021", "3020", "3019", "3017", 
                             "3005", "3004")) %>%
  mutate(age_group = factor(age_group, levels = c("yng", "old")),
         condition = factor(condition, levels = c("low", "mod")),
         time = factor(time, levels = c("pre", "post")),
         allocation = factor(allocation, levels = c("int", "con")),
         condition = case_when(participant %in% c("3071", "3075", "3076",
                                                  "3083", "3084", "3085",
                                                  "3086", "3090", "3091",
                                                  "3094") ~ "none", TRUE ~ condition))



#colSums(is.na(mr.clean)) 

# No NAs after filtering, but some of these are only missing
# e.g. post measurements - worth keeping baseline somehow?
# 3004, 3005, 3019, 3026, 3033, 3034, 3046, 3053, 3069,
# and 3093 dropped out due to different reasons (health/time
# issues primarily). 3017 and 3092 had too few RT sessions (<20) 
# and is thus excluded. 3020, 3021, 3022, 3025, 3028, 
# and 3056 had incomplete analyses (the data was either 
# not on the CD or only arms). 3035, 3045, 3068, and 3070 
# missed their appointment for measurement and rescheduling 
# was not possible.


us.dat2 <- us.dat %>% 
  filter(!participant %in% c("3046", "3033")) # These two dropped out

# 3.1 Create composite muscle thickness (VL + VI average)
comp.dat <- us.dat2 %>%
  mutate(
    composite_thickness = (VLthick + VIthick) / 2
  )

# 3.2 Create limb-specific baseline values
# These will be constant for each leg (Volume condition) within each person
composite.dat <- comp.dat %>%
  group_by(participant, condition) %>%
  mutate(
    # Baseline values for muscle thickness
    baseline_VL = mean(VLthick[time == "pre"]),
    baseline_VI = mean(VIthick[time == "pre"]),
    baseline_composite = mean(composite_thickness[time == "pre"])) %>%
  ungroup()

# 3.3 Ensure proper factor coding with correct reference levels
us.clean <- composite.dat %>%
  mutate(
    age_grp = factor(age.grp, levels = c("yng", "old")),
    condition = factor(condition, levels = c("low", "mod")),
    time = factor(time, levels = c("pre", "mid", "post")),
    allocation = factor(allocation, levels = c("int", "con")),
         condition = case_when(participant %in% c("3071", "3075", "3076",
                                                  "3083", "3084", "3085",
                                                  "3086", "3090", "3091",
                                                  "3094") ~ "none", TRUE ~ condition)) %>%
  group_by(participant, leg, time, age_grp, height, weight, allocation, condition) %>%
  summarise(thickness = mean(composite_thickness, na.rm = TRUE))


#colSums(is.na(us.clean))


```


<!--## Plot 1: Individual Response Plot/Spaghetti Plot

This plot shows every participant's trajectory from baseline to post.-->


```{r, MRI plot1}

## Young vs. Old

mrplot1 <- mr.clean %>%
  filter(!compartment %in% c("fat", "intrafat", "bone")) %>%
  ggplot(aes(x = time, y = volume, group = interaction(participant, leg))) +
  geom_line(aes(color = condition), alpha = 0.3) +
  geom_point(aes(color = condition), alpha = 0.3, size = 1) +
  facet_wrap(vars(age_group, allocation), scales = "fixed") +
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +
  labs(title = "Individual Thigh Muscle Volume Changes",
       subtitle = "Each line represents one leg from one participant",
       x = "Time",
       y = "Muscle Volume (cm3)") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"))
  

#mrplot1


```


```{r, US plot1}

usplot1 <- us.clean %>%
  ggplot(aes(x = time, y = thickness, group = interaction(participant, leg))) +
  geom_line(aes(color = condition), alpha = 0.3) +
  geom_point(aes(color = condition), alpha = 0.3, size = 1) +
  facet_wrap(vars(age_grp, allocation), scales = "fixed") +
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +
 # geom_text_repel(aes(label = participant), size = 2) +
  labs(title = "Individual Muscle Thickness (v. laterlais + intermedius) Changes",
       subtitle = "Each line represents one leg from one participant",
       x = "Time",
       y = "Muscle Thickness") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"))

#usplot1



```


## Plot 2: Mean change with error bars

This plot displays group averages with standard deviation.


```{r, MRI plot2}

mrplot2 <- mr.clean %>%
  filter(!compartment %in% c("fat", "intrafat", "bone")) %>%
  group_by(age_group, allocation, condition, time) %>%
  summarise(mean = mean(volume, na.rm = TRUE),
            sd = sd(volume, na.rm = TRUE),
            .groups = "drop" ) %>%
  ggplot(aes(x = time, y = mean, color = condition, group = condition)) +
  geom_line(linewidth = 1, position = position_dodge(.2)) +
  geom_point(size = 3, position = position_dodge(.2)) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                width = 0.1, linewidth = 0.8, position = position_dodge(.2)) +
  facet_wrap(vars(age_group, allocation), scales = "fixed") +
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +
  labs(title = "Mean Thigh Muscle Volume Changes",
       subtitle = "Error bars show standard deviation",
       x = "Time",
       y = "Mean Muscle Volume (cm3)") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"))
  
mrplot2

```


```{r, US plot2}

usplot2 <- us.clean %>%
  group_by(age_grp, allocation, condition, time) %>%
  summarise(mean = mean(thickness, na.rm = TRUE),
            sd = sd(thickness, na.rm = TRUE),
            .groups = "drop" ) %>%
  ggplot(aes(x = time, y = mean, color = condition, group = condition)) +
  geom_line(linewidth = 1, position = position_dodge(.2)) +
  geom_point(size = 3, position = position_dodge(.2)) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),
                width = 0.1, linewidth = 0.8, position = position_dodge(.2)) +
  facet_wrap(vars(age_grp, allocation), scales = "fixed") +
  scale_color_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                     name = "Training Volume") +
  labs(title = "Mean Muscle Thickness (v. lateralis + intermedius) Changes",
       subtitle = "Error bars show standard deviation",
       x = "Time",
       y = "Mean Muscle Thickness") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"))
  
usplot2

```

## Questions

1. Is there a difference between age and/or training volume?
2. Does age influence the effekt of training volume?
- Should also add LBM data, at least upper body
<!-- Status: Data klar til statistikk, metode beskrevet -->








<!--## Plot 3: Change Score Plot (Work in progress)

This plot focuses on the magnitude of change (post minus pre) using boxplots with individual points overlaid.  These need some work -->

```{r, MRI plot3}

mr.change <- mr.clean %>%
  select(participant, age_group, leg, allocation, condition, time, volume, compartment) %>%
  filter(!compartment %in% c("fat", "intrafat", "bone")) %>%
  pivot_wider(names_from = time,
              values_from = volume) %>%
  mutate(change = post - pre,
         percent_change = (post - pre) / pre * 100)

mrplot3 <- mr.change %>%
  ggplot(aes(x = condition, y = change, fill = condition)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 2) +
  facet_wrap(vars(age_group, allocation), scales = "fixed") +
  scale_fill_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9", "none" = "gray50"),
                    name = "Training Volume") +
  labs(title = "Absolute Change in Thigh Muscle Volume",
       subtitle = "Each point represents one leg",
       x = "Training Volume",
       y = "Change in Muscle Volume (cm3)") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"))

#mrplot3


```


```{r, US plot3}

us.change <- us.clean %>%
  select(participant, age_grp, leg, allocation, condition, time, thickness) %>%
  pivot_wider(names_from = time,
              values_from = thickness) %>%
  ungroup() %>%
  mutate(change.mid = mid - pre,
         change.post = post - pre,
         pchange.mid = (mid - pre) / pre * 100,
         pchange.post = (post - pre) / pre * 100) %>%
  pivot_longer(names_to = "time",
               values_to = "change",
               cols = c(change.mid, change.post)) %>%
  select(participant, leg, allocation, age_grp, condition, time, change)

usplot3 <- us.change %>%
  ggplot(aes(x = time, y = change, fill = condition)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 2) +
  facet_wrap(vars(age_grp, allocation), scales = "fixed") +
  scale_fill_manual(values = c("low" = "#E69F00", "mod" = "#56B4E9"),
                    name = "Training Volume") +
  labs(title = "Absolute Change in Muscle Thickness (v. laterlis + intermedius)",
       subtitle = "Each point represents one leg",
       x = "Training Volume",
       y = "Change in Muscle Thickness") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"))

#usplot3


```














<!-- #########################THESE NEED MORE WORK####################### -->
<!-- ### Plot 4: Paired Comparison

This plot shows pre and post side-by-side to make it easier to visually compare the starting- and endpoints.

*Muscle Volume plot*

```{r, MRI plot4}

mrplot4 <- mr.clean %>%
  filter(!compartment %in% c("fat", "intrafat", "bone")) %>%
  ggplot(aes(x = condition, y = volume, fill = time)) +
  geom_boxplot(alpha = 0.6, position = position_dodge(0.8)) +
  geom_point(aes(group = interaction(participant, leg)),
             position = position_dodge(0.8),
             alpha = 0.3, size = 1.5) +
  facet_wrap(~age_group, scales = "free_y") +
  scale_fill_manual(values = c("pre" = "#999999", "post" = "#009E73"),
                    name = "Time Point") +
  labs(title = "Muscle Volume: Pre vs Post Training",
       subtitle = "Boxplots with individual data points",
       x = "Training Volume",
       y = "Muscle Volume (cm3)") +
  theme_classic() +
  theme(legend.position = "bottom")


mrplot4


```


*Muscle Thickness plot*

```{r, US plot4}

usplot4 <- us.clean %>%
  ggplot(aes(x = condition, y = thickness, fill = time)) +
  geom_boxplot(alpha = 0.6, position = position_dodge(0.8)) +
  geom_point(aes(group = interaction(participant, leg)),
             position = position_dodge(0.8),
             alpha = 0.3, size = 1.5) +
  facet_wrap(~age_grp, scales = "free_y") +
  scale_fill_manual(values = c("pre" = "#999999", "mid" = "#56B4E9", "post" = "#009E73"),
                    name = "Time Point") +
  labs(title = "Muscle Thickness: Pre, Mid and Post Training",
       subtitle = "Boxplots with individual data points",
       x = "Training Volume",
       y = "Muscle Thickness") +
  theme_classic() +
  theme(legend.position = "bottom")


usplot4


```



### Plot 5: Interaction Plot

This plot is specifically designed to show whether young and old participants responded differently to the two training volumes. Parallel lines = no interaction, crossing/diverging lines = interaction effect.

*Muscle Volume plot*

```{r, MRI plot5}

mr.interaction <- mr.clean %>%
  filter(!compartment %in% c("fat", "intrafat", "bone")) %>%
  group_by(age_group, condition, time) %>%
  summarise(mean = mean(volume, na.rm = TRUE),
            .groups = "drop")

mrplot5 <- mr.interaction %>%
  ggplot(aes(x = condition, y = mean, 
             color = age_group, group = age_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 4) +
  facet_wrap(~time) +
  scale_color_manual(values = c("yng" = "#D55E00", "old" = "#0072B2"),
                     name = "Age Group") +
  labs(title = "Age and Training Volume Interaction",
       subtitle = "Mean muscle volume across conditions",
       x = "Training Volume",
       y = "Mean Muscle Volume (cm3)") +
  theme_classic() +
  theme(legend.position = "bottom")


mrplot5

```


*Muscle Volume plot*

```{r, US plot5}

us.interaction <- us.clean %>%
  group_by(age_grp, condition, time) %>%
  summarise(mean = mean(thickness, na.rm = TRUE),
            .groups = "drop")

usplot5 <- us.interaction %>%
  ggplot(aes(x = condition, y = mean, 
             color = age_grp, group = age_grp)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 4) +
  facet_wrap(~time) +
  scale_color_manual(values = c("yng" = "#D55E00", "old" = "#0072B2"),
                     name = "Age Group") +
  labs(title = "Age and Training Volume Interaction",
       subtitle = "Mean muscle thickness across conditions",
       x = "Training Volume",
       y = "Mean Muscle Thickness") +
  theme_classic() +
  theme(legend.position = "bottom")


usplot5

```

-->

